### 2018-08-20 paper/paper.tex
Template for paper that includes genotypes plot
Initial _eLife_ template for paper. At this point
only includes Title, Authors, Abstract, and a
figure showing the genotypes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.
% Additionally, the use of the \newcommand function should be limited.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Single-cell virus sequencing of influenza variants that trigger innate immunity}

\author[1]{Alistair B. Russell}
\author[1,2,3*]{Jesse D. Bloom}
\affil[1]{Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States}
\affil[2]{Department of Genome Sciences, University of Washington, Seattle, United States}
\affil[3]{Howard Hughes Medical Institute, Fred Hutchinson Cancer Research Center, Seattle, United States}

\corr{jbloom@fredhutch.org}{JDB}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
The outcome of viral infection is extremely heterogeneous at the cellular level, and many viruses trigger innate immunity in only a fraction of infected cells. 
Here we develop an approach to determine how the genetic variation inherent in viral populations contributes to this heterogeneity.
We simultaneously determine the cellular transcriptome and full sequences of all viral genes in hundreds of influenza-infected cells.
Infections that trigger immunity are enriched for several features: absence of the gene encoding the virus's primary immune antagonist, internal deletions in viral polymerase genes, and mutations in viral proteins involved in replication and nuclear export. 
However, immune activation remains stochastic in cells infected by viruses with these genetic lesions, and sometimes occurs even in cells infected with fully wildtype virions.
Overall, our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation.
\end{abstract}

\section{Introduction}

\citep{russell2018extreme}

\section{Results}

\subsection{Subsection}
\FIG{genotypes} and \FIGSUPP[genotypes]{genotypes_by_ifn} and \FIGDATA[genotypes]{genotypes}

\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[width=0.9\linewidth]{figures/p_genotypes.pdf}
}
\caption{Viral genotypes and infection outcomes in single cells.}
\label{fig:genotypes}

\figsupp[Genotypes and infection outcomes plotted separately for IFN+ and IFN- cells.]
{This plot shows the same data as \FIG{genotypes}, but with cells separated into columns based on whether they are IFN- or IFN+.}
{\includegraphics[height=0.95\textheight]{figures/p_genotypes_by_ifn.pdf}}
\label{figsupp:genotypes_by_ifn}

\figdata{A CSV file giving the genotypes is in \texttt{figures/genotypes.csv}.}
\label{figdata:genotypes}

\end{fullwidth}
\end{figure}

\section{Discussion}

Add text

\section{Methods and Materials}

Add text

\section{Acknowledgments}

Add this

\nolinenumbers

\bibliography{references}

\end{document}

### 2018-08-22 paper/paper.tex
outline of first few parts of paper

@@
 \usepackage{siunitx}
+\usepackage[export]{adjustbox}
 \DeclareSIUnit\Molar{M}
+
+\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
 
@@
 We simultaneously determine the cellular transcriptome and full sequences of all viral genes in hundreds of influenza-infected cells.
-Infections that trigger immunity are enriched for several features: absence of the gene encoding the virus's primary immune antagonist, internal deletions in viral polymerase genes, and mutations in viral proteins involved in replication and nuclear export.
+Infections that trigger immunity are associated with several features: absence of the gene encoding the virus's primary immune antagonist, internal deletions in viral polymerase genes, and point mutations in viral proteins involved in replication and nuclear export.
 However, immune activation remains stochastic in cells infected by viruses with these genetic lesions, and sometimes occurs even in cells infected with fully wildtype virions.
@@
 
-\citep{russell2018extreme}
+\citep{russell2018extreme} and \citep{steuerman2018dissection}.
 
+\clearpage
 \section{Results}
 
-\subsection{Subsection}
+\subsection{A method to obtain the complete genotypes of the virions infecting single cells that do and do not express IFN}
+
+Two challenges:
+
+IFN expression is rare in infected cells (\FIG{IFNrare}).
+Apparent in single-cell \textit{in vitro}~\citep{russell2018extreme}, single-cell \textit{in vivo}~\citep{steuerman2018dissection}, and reporter assays by others~\citep{killip2017single} and ourselves (\FIG{IFNrare}).
+
+%%% start IFNrare figure %%%
+\begin{figure}
+\centerline{
+{\bf \Large A}\includegraphics[width=0.25\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
+{\bf \Large B}
+{\bf \Large C}
+}
+\caption{
+Triggering of IFN production is rare in influenza-infected cells.
+{\bf (A)} Number of influenza-infected, NS-deficient, and IFN+ cells in our prior single-cell study of influenza infection of A549 cells.
+The Venn diagram shows data aggregated over all samples from \citet{russell2018extreme}.
+{\bf (B)} Number of IFN+ cells in the single-cell analysis of influenza-infected mice by \citep{steuerman2018dissection}.
+\jdbcomment{We could possibly add a \emph{small simple} schematic of reporter line here as panel {\bf (C)} and move the current {\bf (C)} to {\bf (D)}, and then have more details in \FIGSUPP[IFNrare]{reporter_cells}.}
+Alternatively, we could just show everything in \FIGSUPP[IFNrare]{reporter_cells}.
+{\bf (C)} An A549 fluorescent-reporter cell-line confirms that IFN expression is very rare in cells infected with our relatively pure viral stock.
+\jdbcomment{This should be a simple figure, such as a plot that just shows the fraction IFN+ among flu infected cells for a few replicates.
+If you have suitable data, you could show for low, medium, and high-inducing stock and indicate which we use.
+Then put detailed flow data similar to what is in your writeup in \FIGSUPP[IFNrare]{flow}.
+For that flow data, I think contour plots are better than overplotting.
+Also, don't subsample but show all points, as that will give better statistics.}
+}
+\label{fig:IFNrare}
+
+\figsupp[Creation and validation of A549 fluorescent-reporter cell lines for type I and type III IFN expression.]
+{CAPTION}
+{IMAGE}
+\label{figsupp:reporter_cells}
+
+\figsupp[Expression of type I and type III IFN are highly correlated in influenza-infected A549 cells.]
+{CAPTION}
+{IMAGE}
+\label{figsupp:type_I_vs_III}
+
+\figsupp[Full flow cytometry data for \FIG{IFNrare}C.]
+{CAPTION}
+{IMAGE}
+\label{figsupp:flow}
+
+\end{figure}
+%%% end IFNrare figure %%%
+
+All past single-cell transcriptomics of viral infected cells~\citep{russell2018extreme,zanini2018single,steuerman2018dissection,zanini2018virus} have focused on counting the number of viral transcripts rather than determining their full sequences.
+We came up with a scheme to determine full genotypes of infecting viruses in IFN+ cells (\FIG{approach}).
+
+%%% start approach figure
+\begin{figure}
+
+\centerline{\jdbcomment{FIGURE}}
+
+\caption{
+Schematic of experimental approach.
+}
+
+\end{figure}
+%%% end approach figure
+
+\subsection{Viral genotypes in IFN+ and IFN- influenza-infected cells}
 \FIG{genotypes} and \FIGSUPP[genotypes]{genotypes_by_ifn} and \FIGDATA[genotypes]{genotypes}
 
+%%% start genotypes figure %%%
 \begin{figure}
@@
 }
-\caption{Viral genotypes and infection outcomes in single cells.}
+\caption{
+Viral genotypes and infection outcomes in single cells.
+}
 \label{fig:genotypes}
@@
 \end{figure}
+%%% end genotypes figure %%%
 

### 2018-08-23 paper/paper.tex
clean up paper repo
Remove `*.blg` file, tiny tweaks to `paper.tex`
added re-analysis of Steuerman et al
Added subdirectory (`figures/IFN_stochastic/SteuermanReanalysis/`)
that re-analyses data of Steuerman et al on single-cell
RNA-seq of flu-infected mice.

Then added Figure supplement (currently Figure 1 - Figure supplement 4)
that includes a summary of this re-analysis.

@@
 The Venn diagram shows data aggregated over all samples from \citet{russell2018extreme}.
-{\bf (B)} Number of IFN+ cells in the single-cell analysis of influenza-infected mice by \citep{steuerman2018dissection}.
-\jdbcomment{We could possibly add a \emph{small simple} schematic of reporter line here as panel {\bf (C)} and move the current {\bf (C)} to {\bf (D)}, and then have more details in \FIGSUPP[IFNrare]{reporter_cells}.}
-Alternatively, we could just show everything in \FIGSUPP[IFNrare]{reporter_cells}.
-{\bf (C)} An A549 fluorescent-reporter cell-line confirms that IFN expression is very rare in cells infected with our relatively pure viral stock.
+{\bf (B)} An A549 fluorescent-reporter cell-line confirms that IFN expression is very rare in cells infected with our relatively pure viral stock.
 \jdbcomment{This should be a simple figure, such as a plot that just shows the fraction IFN+ among flu infected cells for a few replicates.
@@
 For that flow data, I think contour plots are better than overplotting.
-Also, don't subsample but show all points, as that will give better statistics.}
+Also, don't subsample but show all points, as that will give better statistics.
+We could possibly add a \emph{small simple} schematic of reporter line here as panel {\bf (B)} and move the current {\bf (B)} to {\bf (C)}, and then have more details in \FIGSUPP[IFNrare]{reporter_cells}.
+Alternatively, we could just show everything in \FIGSUPP[IFNrare]{reporter_cells}.}
 }
@@
 {CAPTION}
-{IMAGE}
+{\jdbcomment{ADD FIGURE}}
 \label{figsupp:reporter_cells}
@@
 {CAPTION}
-{IMAGE}
+{\jdbcomment{ADD FIGURE}}
 \label{figsupp:type_I_vs_III}
 
-\figsupp[Full flow cytometry data for \FIG{IFNrare}C.]
+\figsupp[Full flow cytometry data for \FIG{IFNrare}B.]
 {CAPTION}
-{IMAGE}
+{\jdbcomment{ADD FIGURE}}
 \label{figsupp:flow}
+
+\figsupp[Few cells express detectable IFN in the single-cell analysis influenza-infected mice by \citet{steuerman2018dissection}.]
+{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell mRNA-sequencing of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
+Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
+Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (1 replicate)---we do not show cells from the control mice.
+As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
+An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
+Nonetheless, there is detectable expression of key interferon-stimulated gene (ISG) mRNAs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN$\alpha$ and IFN$\beta$) or type III (IFN$\lambda$) mRNAs, and only at low levels.
+The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
+}
+{\includegraphics[width=0.8\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
+\label{figsupp:mice}
 
@@
 
-\centerline{\jdbcomment{FIGURE}}
+\centerline{\jdbcomment{FIGURE 2 HERE}}
 
 \caption{
-Schematic of experimental approach.
+{\bf (A)} Schematic of experimental approach.
+{\bf (B)} Some summary stats on numbers of cells, etc.
 }

### 2018-08-25 paper/paper.tex
cleaned IFN calling and figure creation
- Clean up and simplify IFN calling.

- Place figures from `monocle_analysis.ipynb`
  in separate directory

@@
 {\centering
-\includegraphics[width=0.9\linewidth]{figures/p_genotypes.pdf}
+\includegraphics[width=0.9\linewidth]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 {This plot shows the same data as \FIG{genotypes}, but with cells separated into columns based on whether they are IFN- or IFN+.}
-{\includegraphics[height=0.95\textheight]{figures/p_genotypes_by_ifn.pdf}}
+{\includegraphics[height=0.95\textheight]{figures/single_cell_figures/p_genotypes_by_ifn.pdf}}
 \label{figsupp:genotypes_by_ifn}
 
-\figdata{A CSV file giving the genotypes is in \texttt{figures/genotypes.csv}.}
+\figdata{A CSV file giving the genotypes is in \texttt{figures/single_cell_figures/genotypes.csv}.}
 \label{figdata:genotypes}

### 2018-08-26 paper/paper.tex
figures summarizing cells / infection / IFN
Cleaned up `monocle_analysis.ipynb` to use a common
color scheme and make a merged figure summarizing
calling of the cells, the infected cells, analysis
of IFN expression, coinfection rates, ISG expression,
and the tSNE.

Also updated other Venn diagram figure to be consistent
with this coloring scheme.

Also made LaTex paper include the new figures and add
appropriate legends.

@@
 The Venn diagram shows data aggregated over all samples from \citet{russell2018extreme}.
-{\bf (B)} An A549 fluorescent-reporter cell-line confirms that IFN expression is very rare in cells infected with our relatively pure viral stock.
+{\bf (B)} An A549 fluorescent-reporter cell-line confirms that IFN expression is rare in cells infected with our viral stock.
 \jdbcomment{This should be a simple figure, such as a plot that just shows the fraction IFN+ among flu infected cells for a few replicates.
@@
 
-\figsupp[Expression of type I and type III IFN are highly correlated in influenza-infected A549 cells.]
+\figsupp[Expression of type I and type III IFN are highly correlated in influenza-infected A549 cells.
+\jdbcomment{Alternatively, we could leave this out if you don't have clean data for this point and just refer to \FIGSUPP[experiment]{type_I_III_correlation}}]
 {CAPTION}
@@
 Nonetheless, there is detectable expression of key interferon-stimulated gene (ISG) mRNAs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN$\alpha$ and IFN$\beta$) or type III (IFN$\lambda$) mRNAs, and only at low levels.
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels.
 The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
 }
-{\includegraphics[width=0.8\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
+{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
 \label{figsupp:mice}
@@
 All past single-cell transcriptomics of viral infected cells~\citep{russell2018extreme,zanini2018single,steuerman2018dissection,zanini2018virus} have focused on counting the number of viral transcripts rather than determining their full sequences.
-We came up with a scheme to determine full genotypes of infecting viruses in IFN+ cells (\FIG{approach}).
+We came up with a scheme to determine full genotypes of infecting viruses in IFN+ cells (\FIG{experiment}).
 
@@
 \begin{figure}
+\begin{fullwidth}
 
-\centerline{\jdbcomment{FIGURE 2 HERE}}
+{\centering
+{\bf A} \jdbcomment{PANEL WITH EXPERIMENTAL OUTLINE SCHEMATIC} \vspace{0.2in}\\
+
+\includegraphics[width=\linewidth, clip=false]{figures/single_cell_figures/p_cell_summary.pdf}
+}
 
 \caption{
-{\bf (A)} Schematic of experimental approach.
-{\bf (B)} Some summary stats on numbers of cells, etc.
+Experimental overview.
+{\bf (A)} Schematic of approach.
+{\bf (B)} Number of cells for which transcriptomes were obtained.
+Most transcriptomes are from human cells, but a few are from the uninfected canine cells spiked into the experiment as a control.
+From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was captured in an emulsion.
+These cross-celltype multiplets are excluded from subsequent analyses.
+{\bf (C)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
+Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
+Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (possibly low-quality emulsions or multiplets), and are excluded from subsequent analyses.
+{\bf (D)} Distribution across cells of the fraction of all mRNA derived from influenza.
+Cells called as infected are in blue, while other cells are in green.
+The inset shows the amount of viral mRNA among the human cells called as infected.
+{\bf (E)} The number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
+The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
+\FIGSUPP[experiment]{frac_has_gene} shows the frequency that each viral gene is detected in infected cells.
+{\bf (F)} Relative expression of viral genes among infected cells.
+{\bf (G)} The number of cells infected with wildtype virus, synonymously barcoded virus, or both.
+From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
+{\bf (H)} Fraction of cellular mRNA derived from type I or type III IFN across all cells, faceted by whether the cells are infected.
+Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
+We grouped type I and type III IFN together because their expression is highly correlated (\FIGSUPP[experiment]{type_I_III_correlation}).
+Many cells that do not express IFN itself still express interferon-stimulated genes (\FIGSUPP[experiment]{ISG}).
 }
+\label{fig:experiment}
 
+\figsupp[Fraction of infected cells that detectably express each influenza gene.]
+{Bars show the fraction of infected cells that are called as expressing each viral gene.
+The gray dotted line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
+Each viral gene is detected in $\approx$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+The exception is NP, which is detected in virtually all infected cells.
+The much higher frequency of detecting NP could reflect a biological phenomenon, or it could be because cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments and analysis.}
+{\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}}
+\label{figsupp:frac_has_gene}
+
+\figsupp[Expression of type I and type III IFN genes is highly correlated in single cells in our experiments.]
+{The correlation between the fraction of cellular mRNA derived from type I (IFN-$\alpha$ and IFN-$\beta$) and type III (IFN-$\lambda$) IFN in the A549 cells in our experiments.
+Each point represents a single cell.
+The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
+Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
+}
+{\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}}
+\label{figsupp:type_I_III_correlation}
+
+\figsupp[Expression of interferon-stimulated genes (ISGs).]
+{
+For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs that are highly expressed A549 cells (IFIT1, ISG15, CCL5, and Mx1). 
+{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
+We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
+Comparison to \FIG{experiment}H shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
+This is probably because paracrine signaling leads to ISG expression in some cells that are not themselves expressing IFN.
+{\bf (B)} The correlation between the fraction of cellular mRNA derived from IFN and from ISGs.
+Each point represents a single cell, and the Pearson correlation coefficient is shown.
+IFN and ISG expression is more correlated for infected than uninfected cells, probably because in the latter the ISG expression is often due to paracrine signaling that does not induce expression of IFN itself.
+But among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but very few cells that express high levels of IFN without also substantially expressing ISGs.
+}
+{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}}
+\label{figsupp:ISG}
+
+\figsupp[Unsupervised t-SNE clustering shows that cell-to-cell variation in influenza genes, IFN genes, and ISGs explains a substantial part of the structure of the data.]
+{To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
+In the t-SNE plot, each point is a different cell and cells with similar transcriptomes are closer together.
+Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA shown on a log (top) or linear (bottom) scale.
+As is clear from this plot, expression of influenza genes, IFN, and ISGs help explain the structure of the data, since cells with high expression of these genes clearly group together.
+}
+{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_tsne.pdf}}
+\label{figsupp:tSNE}
+
+\end{fullwidth}
 \end{figure}
@@
 \figsupp[Genotypes and infection outcomes plotted separately for IFN+ and IFN- cells.]
-{This plot shows the same data as \FIG{genotypes}, but with cells separated into columns based on whether they are IFN- or IFN+.}
-{\includegraphics[height=0.95\textheight]{figures/single_cell_figures/p_genotypes_by_ifn.pdf}}
+{This plot shows the same data as \FIG{genotypes}, but with cells separated into columns based on whether they are IFN- (left column) or IFN+ (right column).}
+{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_genotypes_by_ifn.pdf}}
 \label{figsupp:genotypes_by_ifn}
 
-\figdata{A CSV file giving the genotypes is in \texttt{figures/single_cell_figures/genotypes.csv}.}
+\figdata{A CSV file giving the genotypes is in \texttt{figures/single\_cell\_figures/genotypes.csv}.}
 \label{figdata:genotypes}

### 2018-08-26 paper/paper.tex
color scheme for Venn diagram and Steuerman plot
Normalized the color scheme for these two plots to
match the rest of the figures.

Also, adjusted the Venn diagram to only show infected
and IFN+, as we'll discuss NS- later.

@@
 {\bf \Large B}
-{\bf \Large C}
 }
@@
 Triggering of IFN production is rare in influenza-infected cells.
-{\bf (A)} Number of influenza-infected, NS-deficient, and IFN+ cells in our prior single-cell study of influenza infection of A549 cells.
+{\bf (A)} Number of influenza-infected and IFN+ cells in our prior single-cell study of influenza infection of A549 cells.
 The Venn diagram shows data aggregated over all samples from \citet{russell2018extreme}.
@@
 Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
-Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (1 replicate)---we do not show cells from the control mice.
+Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
 As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
@@
 {Bars show the fraction of infected cells that are called as expressing each viral gene.
-The gray dotted line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
+The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
 Each viral gene is detected in $\approx$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.

### 2018-08-29 paper/paper.tex
genotypes plot indicates low-freq mutations
Refactored the genotypes plots made by
`monocle_analysis.ipynb` so the size of indels /
mutations is proportional to their frequency.
updated genotypes figure with supplements
Slightly changed cutoffs for calling mutations in
`pacbio_analysis.ipynb`.

Added a schematic of how strand exchange is identified.

Added a figure generated by this notebook that shows
the number of CCSs and strand exchange.

Updated the paper with additional figures related to
PacBio and genotype calling.

@@
 
+\figdata{Genbank files giving the sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
+\label{seqs}
+
 \end{fullwidth}
@@
 {\centering
-\includegraphics[width=0.9\linewidth]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 Viral genotypes and infection outcomes in single cells.
+Each row shows a single infected cell.
+Arrows indicate the presence of a viral gene from a single viral barcode variant (light blue) or both barcode variants (dark blue).
+Circles and boxes indicate mutations or indels as described in the legend.
+Circle areas and box heights are proportional to the fraction of PacBio CCSs that report the mutation (see \FIGDATA[genotypes]{genotypes} for numerical data).
+For dual-barcode infections, mutations / indels for the wildtype viral variant are on the top half of the arrow and those for the synonymously barcoded variant are on the bottom half. 
+Green boxes to the left show the percent of all mRNA in that cell derived from virus.
+Orange boxes show the percent of cellular mRNA derived from IFN, with boxes framed in orange indicating cells classified as IFN+ in \FIG{experiment}H.
 }
 \label{fig:genotypes}
+
+\figsupp[Strategy for detecting strand exchange during sequencing of full-length viral genes.]
+{The library preparation for PacBio sequencing of the cDNA for the full-length viral genes requires many cycles of PCR in order to produce a sufficient amount of DNA for sequencing.
+A major concern is whether strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
+We can detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wildtype virus and virus carrying synonymous barcodes near both termini of each gene.
+If there is no strand exchange, all molecules should either be wildtype or have the synonymous mutations at the viral barcode sites.
+But strand exchange will create some molecules that have a mix of wildtype nucleotides in the viral barcode at one termini and synonymous mutations in the viral barcode at the other termini.
+\FIGSUPP[genotypes]{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
+Note that since the rate of homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
+}
+{\includegraphics[width=0.6\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}}
+\label{figsupp:StrandExchange}
+
+\figsupp[Number of PacBio circular consensus sequences and PCR strand exchange rate.]
+{
+The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
+Note that these sequences were obtained using several PacBio runs, some of which were loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
+Therefore, unlike the 10X transcript count data in \FIG{experiment}H, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
+Especially for the polymerase genes (PB2, PB1, and PA), many of the CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were presumably preferentially amplified during PCR.
+Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
+Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences for the same gene, and cell barcodes to collapse duplicate sequences from the same cell.
+The bars are colored by whether the sequence can be identified as being derived from the wildtype viral variant, the synonymously barcoded variant, or represents a mixed barcode molecule (see \FIGSUPP[genotypes]{StrandExchange}).
+From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
+Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
+A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcode.
+}
+{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
+\label{figsupp:CCSs}
 
@@
 
-\figdata{A CSV file giving the genotypes is in \texttt{figures/single\_cell\_figures/genotypes.csv}.}
+\figdata{A CSV file giving the genotypes is in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
 \label{figdata:genotypes}

### 2018-08-29 paper/paper.tex
Finished updating genotype figure with supplements
Added a supplement that shows for how many cells
we obtained complete viral genotypes, and shows
how we tended to get more viral genotypes for
cells that were **not** co-infected, and for
cells with higher viral mRNA expression.

@@
 
+\figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
+{Cells for which we could determine the full sequences of all genes expressed by the infecting virus(es).
+{\bf (A)} We could call the complete genotypes of the infecting virus(es) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
+{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNAs than cells for which we could not call complete genotypes.
+This makes sense, as cells with more viral mRNA are more likely to have their viral genes captured in the PacBio sequencing, which is only able to capture a small fraction of the total transcripts identified by the 3'-sequencing of the 10X platform.
+The lower calling rate for dual-barcode co-infections relative to single barcode infectionis is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing.
+An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
+Note also that this plot is limited to the cells that were called as infected (\FIG{experiment}D) and could clearly be classified as IFN- or IFN+ (\FIG{experiment}H).
+}
+{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}
+
 \figsupp[Genotypes and infection outcomes plotted separately for IFN+ and IFN- cells.]

### 2018-09-03 paper/paper.tex
viral genetics contributes to flu heterogeneity
Added plot showing that cells infected with fully
wildtype virus have less heterogeneity in fraction
flu than cells infected with viruses containing
amino-acid mutations or indels.
added mutations figure to paper
This figure (currently Figure 4) shows
the association of specific viral mutation
types with IFN induction.

@@
 {\centering
-\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.8\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 
-\figdata{A CSV file giving the genotypes is in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
+\figdata{A CSV file giving the genotypes is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
 \label{figdata:genotypes}
@@
 %%% end genotypes figure %%%
+
+%%% begin mutations figure %%%
+\begin{figure}
+\begin{fullwidth}
+{\centering
+\includegraphics[width=\linewidth]{figures/single_cell_figures/p_mutations.pdf}
+}
+\caption{
+Viral genetic variation partially explains the heterogeneity in viral burden and IFN induction among the infected cells for which we sequenced all expressed viral genes (\FIG{genotypes}).
+{\bf (A)} 
+The percent of all mRNA derived from virus, faceted by whether the cells express wildtype copies of all eight viral genes, or fail to express a wildtype copy of one or more genes (due to gene absence, amino-acid substitution, insertion, or deletion).
+Cells infected by full wildtype virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
+{\bf (B)}
+There is a trend for cells infected by full wildtype virus to express less IFN.
+{\bf (C)}
+Some specific forms of viral genetic variation are associated with IFN induction.
+The top panel show the percent of IFN+ and IFN- cells that fail to express each viral gene.
+The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each viral gene, conditioned on the cell expressing that gene.
+Numbers above the bars give P-values (Fisher's exact test) for rejecting the hypothesis that percents are equal among IFN- and IFN+ cells. 
+Absence of NS is strongly associated with IFN induction; amino-acid substitutions in PB1 or NS and deletions in HA are weakly associated with IFN induction.
+\FIGSUPP[mutations]{complete_annotated} gives exact numbers and multiple-hypothesis corrections on the P-values; only the absence of NS remains significant at a false-discovery rate of 10\%.
+\FIGSUPP[mutations]{all_annotated} shows that the trends are largely the same if we include infected cells with incomplete sequence information for some expressed viral genes, with the addition of a trend towards IFN induction associated with \emph{presence} of PB2 and PA.
+Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
+{\bf (D)}
+There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
+Note that throughout this figure, we only consider as substitution mutations that are non-synonymous.
+}
+\label{fig:mutations}
+
+\figsupp[A version of \FIG{mutations}C with cell counts and multiple-hypothesis corrections.]
+{The numbers at the top show the Q-values as well as the P-values.
+Only the association of absence of NS with IFN induction remains significant at a false-discovery rate of 0.1.
+The fractions written in small numbers on the bars indicate the number of cells that have the indicated mutation over the total number of cells for that IFN state.
+}
+{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_muts_ifn_complete_annot.pdf}}
+\label{figsupp:complete_annotated}
+
+\figsupp[Analysis as in \FIG{mutations}C but including infected cells with incomplete viral sequence information.]
+{This plot differs from \FIG{mutations}C and \FIGSUPP[mutations]{all_annotated} in that it also includes data from cells for which some viral genes were not fully sequenced.
+For incompletely sequenced cells, deletions and substitutions are included in the counts when that particular viral gene is sequenced.
+The major trends in \FIG{mutations}C are also true for the larger dataset in this figure.
+In addition, there is now a modest trend for infected cells that fail to express PB2 and PA \emph{not} to express IFN.
+}
+{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_muts_ifn_all_annot.pdf}}
+\label{figsupp:all_annotated}
+
+\figdata{A CSV file giving the viral mutations and related information in \FIG{mutations} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.}
+\label{figdata:mutations}
+
+\end{fullwidth}
+\end{figure}
+%%% end mutations figure %%%
 

### 2018-09-03 paper/paper.tex
cleaned up plots displaying mutations vs IFN
The plots with IFN versus mutations were very
cluttered as they had P values, Q values, and
raw counts. Got rid of the raw counts and showed
both P and Q values on the same line in small text.

@@
 The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each viral gene, conditioned on the cell expressing that gene.
-Numbers above the bars give P-values (Fisher's exact test) for rejecting the hypothesis that percents are equal among IFN- and IFN+ cells. 
-Absence of NS is strongly associated with IFN induction; amino-acid substitutions in PB1 or NS and deletions in HA are weakly associated with IFN induction.
-\FIGSUPP[mutations]{complete_annotated} gives exact numbers and multiple-hypothesis corrections on the P-values; only the absence of NS remains significant at a false-discovery rate of 10\%.
-\FIGSUPP[mutations]{all_annotated} shows that the trends are largely the same if we include infected cells with incomplete sequence information for some expressed viral genes, with the addition of a trend towards IFN induction associated with \emph{presence} of PB2 and PA.
+Numbers above the bars give P-values (Fisher's exact test) and associated Q-values~\citep{storey2003statistical} for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
+Absence of NS is strongly associated with IFN induction, and amino-acid substitutions in PB1 or NS and deletions in HA are weakly associated with IFN induction.
+Only the absence of NS remains significant at a false-discovery rate of 10\%.
+\FIGSUPP[mutations]{allcells} shows that the trends remain largely the same if we also analyze infected cells with incomplete viral sequence information, with the addition of a trend towards IFN induction associated with \emph{presence} of PB2 and PA.
 Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
@@
 
-\figsupp[A version of \FIG{mutations}C with cell counts and multiple-hypothesis corrections.]
-{The numbers at the top show the Q-values as well as the P-values.
-Only the association of absence of NS with IFN induction remains significant at a false-discovery rate of 0.1.
-The fractions written in small numbers on the bars indicate the number of cells that have the indicated mutation over the total number of cells for that IFN state.
-}
-{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_muts_ifn_complete_annot.pdf}}
-\label{figsupp:complete_annotated}
-
 \figsupp[Analysis as in \FIG{mutations}C but including infected cells with incomplete viral sequence information.]
-{This plot differs from \FIG{mutations}C and \FIGSUPP[mutations]{all_annotated} in that it also includes data from cells for which some viral genes were not fully sequenced.
+{This plot differs from \FIG{mutations}C in that it also includes data from cells for which some viral genes were not fully sequenced.
 For incompletely sequenced cells, deletions and substitutions are included in the counts when that particular viral gene is sequenced.
@@
 {\includegraphics[width=\textwidth]{figures/single_cell_figures/p_muts_ifn_all_annot.pdf}}
-\label{figsupp:all_annotated}
+\label{figsupp:allcells}
 

### 2018-09-03 paper/paper.tex
added plot on ISG / virus associations to paper
Added figure supplement to Figure 4 that shows
the association of viral gene absence / mutations
with ISG expression. Trends are mostly similar
as for IFN expression.

@@
 \FIGSUPP[mutations]{allcells} shows that the trends remain largely the same if we also analyze infected cells with incomplete viral sequence information, with the addition of a trend towards IFN induction associated with \emph{presence} of PB2 and PA.
+\FIGSUPP[mutations]{isg} shows that the trends also remain largely similar if we analyze the association of viral genetic variation with ISG expression rather than IFN expression.
 Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
@@
 }
-{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_muts_ifn_all_annot.pdf}}
+{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_muts_ifn_all.pdf}}
 \label{figsupp:allcells}
+
+\figsupp[Association of viral genetic variation and ISG expression.]
+{This plot shows the same information as panels \FIG{mutations}B-D except that it shows the association of viral gene absence / mutations with ISG expression rather than IFN expression.
+Cells are classified as ISG+ or ISG- as in \FIGSUPP[experiment]{ISG}.
+The major viral features that associate with IFN induction also associate with ISG expression.
+Specifically, cells that express wildtype copies of all eight viral genes tend to express ISGs less frequently and at lower levels
+The absence of NS is strongly associated with higher ISG expression.
+Viral burden is not associated with ISG expression levels in NS-sufficient infections, but is in NS-deficient infections.
+}
+{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_mutations_isg.pdf}}
+\label{figsupp:isg}
 

### 2018-09-04 paper/paper.tex
Figure 1B (IFN reporter) and supplements
Added images showing creation and validation
of the IFN reporter cell lines. These
are Figure 1B and associated supplements.

@@
 {\bf \Large A}\includegraphics[width=0.25\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
-{\bf \Large B}
+\hspace{0.05\textwidth}
+{\bf \Large B}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
+\hspace{0.05\textwidth}
+{\bf \Large C} \jdbcomment{ADD FIGURE PANEL}
 }
@@
 The Venn diagram shows data aggregated over all samples from \citet{russell2018extreme}.
-{\bf (B)} An A549 fluorescent-reporter cell-line confirms that IFN expression is rare in cells infected with our viral stock.
-\jdbcomment{This should be a simple figure, such as a plot that just shows the fraction IFN+ among flu infected cells for a few replicates.
-If you have suitable data, you could show for low, medium, and high-inducing stock and indicate which we use.
-Then put detailed flow data similar to what is in your writeup in \FIGSUPP[IFNrare]{flow}.
-For that flow data, I think contour plots are better than overplotting.
-Also, don't subsample but show all points, as that will give better statistics.
-We could possibly add a \emph{small simple} schematic of reporter line here as panel {\bf (B)} and move the current {\bf (B)} to {\bf (C)}, and then have more details in \FIGSUPP[IFNrare]{reporter_cells}.
-Alternatively, we could just show everything in \FIGSUPP[IFNrare]{reporter_cells}.}
+{\bf (B)} Creation of an A549 reporter cell line to efficiently identify single cells expressing IFN.
+The cells are engineered to contain an IFN promoter that drives expression of a fluorescent protein and cell-surface protein amenable to MACS (magnetic-activated cell sorting).
+We created reporter cell lines with both type I (IFN-$\beta$) and type III (IFN-$\lambda$) promoters (\FIGDATA[IFNrare]{reporter_sequences}).
+The reporters were efficiently activated when infected with a virus known to potently induce IFN (\FIGSUPP[IFNrare]{reporter_validation}).
+Expression of type I and type III IFN is highly correlated in the reporter cell lines upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
+{\bf (C)}
+Frequency of IFN induction upon infection with the influenza virus stocks used in this paper as quantified using the reporter cell lines.
+\jdbcomment{This should be a simple figure, such as a plot that just shows the fraction IFN+ among flu infected cells for the virus stocks we use, plus positive and negative controls.
+Then put detailed flow data in \FIGSUPP[IFNrare]{flow}. }
 }
@@
 
-\figsupp[Creation and validation of A549 fluorescent-reporter cell lines for type I and type III IFN expression.]
-{CAPTION}
-{\jdbcomment{ADD FIGURE}}
-\label{figsupp:reporter_cells}
+\figsupp[Validation of IFN reporter cell lines.]
+{To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
+Activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR).
+Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal for the IFN-$\lambda$ reporter driving ZsGreen.
+}
+{\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}}
+\label{figsupp:reporter_validation}
 
-\figsupp[Expression of type I and type III IFN are highly correlated in influenza-infected A549 cells.
-\jdbcomment{Alternatively, we could leave this out if you don't have clean data for this point and just refer to \FIGSUPP[experiment]{type_I_III_correlation}}]
-{CAPTION}
-{\jdbcomment{ADD FIGURE}}
+\figsupp[Expression of type I and type III IFN are highly correlated in influenza-infected A549 cells.]
+{A549 cells were dually transduced with the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
+The cells were then infected with two different stocks of ``wildtype'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
+After 13 hours, cells were analyzed by flow cytometry.
+As shown in the FACS plots, expression of the IFN-$\beta$ and IFN-$\lambda$ reporters is highly correlated in all cases.}
+{\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}}
 \label{figsupp:type_I_vs_III}
 
-\figsupp[Full flow cytometry data for \FIG{IFNrare}B.]
-{CAPTION}
+\figsupp[Full flow cytometry data for \FIG{IFNrare}C.]
+{\jdbcomment{ADD CAPTION}}
 {\jdbcomment{ADD FIGURE}}
@@
 \label{figsupp:mice}
+
+\figdata{Sequences of the reporters schematized in \FIG{IFNrare}B.
+The MACS sortable surface marker is a truncated LNGFR.
+For the IFN-$\beta$ reporter, the fluorescent protein is mNeonGreen.
+For the IFN-$\lambda$ (IL29) reporter, the fluorescent protein is ZsGreen.
+The sequences of plasmids carrying these reporters are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1517_pHAGE2_IFNbeta_prom_LNGFR_P2A_mNeongreen.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1767_pHAGE2_IL29_promnkozak_LNGFR_P2A_zsGreen.gb}.
+}
+\label{figdata:reporter_sequences}
 

### 2018-09-04 paper/paper.tex
added outline of initial Methods

@@
 \figdata{Genbank files giving the sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
-\label{seqs}
+\label{figdata:virus_seqs}
 
@@
 
-Add text
+\subsection{IFN reporter cell lines}
+\jdbcomment{Alistair writes this}
+Reporter sequences \FIGDATA[IFNrare]{reporter_sequences}
+
+\subsection{Viral stocks}
+\jdbcomment{Alistair writes this}
+Wildtype and synonymously barcoded viruses in \FIGDATA[experiment]{virus_seqs}
+
+\subsection{Single-cell infections and transcriptomics using 10X Chromium on IFN-enriched cells}
+\jdbcomment{Alistair writes this}
+
+\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
+\jdbcomment{Alistair writes this}
+
+\subsection{Computational analysis of single-cell transcriptomic and PacBio data}
+\jdbcomment{Jesse writes this}
 

### 2018-09-04 paper/paper.tex
merged Alistair schematic into main paper

@@
 
-{\centering
-{\bf A} \jdbcomment{PANEL WITH EXPERIMENTAL OUTLINE SCHEMATIC} \vspace{0.2in}\\
+{\bf A} \includegraphics[width=0.9\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
+\vspace{0.1in}
 
 \includegraphics[width=\linewidth, clip=false]{figures/single_cell_figures/p_cell_summary.pdf}
-}
 
 \caption{
-Experimental overview.
-{\bf (A)} Schematic of approach.
+Single-cell virus sequencing of influenza infections that induce IFN.
+{\bf (A)} Experimental workflow.
+(1) IFN reporter A549 cells are infected with a mix of wildtype virus and virus carrying synonymous mutations at both ends of each gene segment.
+IFN+ cells are enriched by MACS, and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
+(2) The mRNAs from individual cells are tagged with cell-specific barcodes using a 10X Chromium device.
+(3) The entire cellular transcriptomes are quantified using the standard single-cell mRNA-seq protocol, and viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
+(4) The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
 {\bf (B)} Number of cells for which transcriptomes were obtained.
@@
 
-\subsection{Computational analysis of single-cell transcriptomic and PacBio data}
-\jdbcomment{Jesse writes this}
+\subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
+A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}
 

### 2018-09-05 paper/paper.tex
added computational methods
Added the computational methods section
as well as Supplementary files 1 and 2.

@@
 \usepackage[export]{adjustbox}
+\usepackage{float}
 \DeclareSIUnit\Molar{M}
@@
 \newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
+
+\newfloat{suppfile}{thp}{lofsupfile}
+%\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
+\floatname{suppfile}{Supplementary file}
 
@@
 
-{\bf A} \includegraphics[width=0.9\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
+\hspace{0.04in} {\bf A} \includegraphics[width=0.9\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
 \vspace{0.1in}
@@
 {\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}
+\label{figsupp:ncells}
 
@@
 \subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
-A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}
+A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}~\citep{russell2018github}. 
+This pipeline is orchestrated by \texttt{Snakemake}~\citep{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
+The sequencing data and annotated cell-gene matrix are available on the GEO repository under \jdbcomment{add accession}.
+
+Briefly, the raw deep sequencing data from the Illumina 3'-end sequencing were processed using the 10X Genomics software package \texttt{cellranger} (version 2.2.0). 
+We built a multi-species alignment reference consisting of a concatenation of the human and influenza virus transcriptomes (the first ``species'') and the canine transcriptome (the second ``species'').
+The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
+The influenza virus transcriptome consisted of the mRNAs for the wildtype A/WSN/1933 virus strain in \FIGDATA[experiment]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wildtype and synonymously barcoded viral variants to this transcriptome).
+The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
+The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{experiment}B), and generate a matrix giving the expression of each gene in each single cell.
+We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wildtype or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.
+
+The PacBio sequences of the full-length viral genes were analyzed as follows.
+First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
+We further processed these CCSs using custom Python code and the \texttt{minimap2}~\citep{li2018minimap2} long-read aligner (version 2.11-r797).
+The Python code has been implemented in the API of \texttt{dms\_tools2}~\citep[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
+A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:pacbio_analysis}.
+We refer the reader to this notebook for a detailed description and extensive plots showing the results at each step.
+Here is a brief summary: we filtered for CCSs that had the expected 5' termini (from the influenza-specific primers) and 3' termini (corresponding to the 10X adaptor), and for which we could identify the cell barcode, UMI, and polyA tail.
+We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
+At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\FIGSUPP[genotypes]{StrandExchange}), and reassuringly found that strand exchange was rare (\FIGSUPP[genotypes]{CCSs}).
+The small number of CCSs with identifiable strand exchange were filtered from further analysis.
+We then further filtered for CCSs that contained valid cell barcodes as identified by the \texttt{cellranger} pipeline, and kept just one CCS per UMI (preferentially retaining high-quality CCSs that aligned to full-length cDNAs).
+Finally, we used the CCSs to call the sequence of the viral gene in each cell, calling mutations separately for each viral barcode variant.
+We called mutations (insertions, deletions, and substitutions) in the viral gene sequences as follows:
+\begin{enumerate}
+\item Mutations with accuracies less than 0.999 (which constitute $<$0.5\% of all mutations) were ignored.
+\item If all CCSs for a particular viral-barcode variant of a gene in a cell were wildtype, it was called as wildtype.
+\item If any CCSs for a particular viral-barcode variant of gene in a cell had a mutation, then require at least two CCSs to call the sequence.
+\item If at least two and $>$30\% of the CCSs had a specific mutation, then call that mutation as present and note its frequency among the CCSs.
+The exception was single-nucleotide indels in homopolymers, for which we required three CCSs to call a mutation (the reason is that the main mode of PacBio sequencing errors is short indels in homopolymers).
+\end{enumerate}
+The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
+We could call the sequences of all expressed viral genes in about half of the infected cells (\FIGSUPP[genotypes]{ncells}).
+The mutations called using this pipeline are shown in \FIG{genotypes}, and \FIGDATA[genotypes]{genotypes} gives the number of CCSs supporting each mutation call.
+The called sequences of the viral genes were added to the annotated cell-gene matrix.
+
+Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
+This analysis utilized a variety of R and Bioconductor~\citep{huber2015orchestrating} packages, including \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
+A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:monocle_analysis}.
+We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
+Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{experiment}C.
+We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{experiment}D).
+We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{experiment}E and \FIGSUPP[experiment]{frac_has_gene}).
+To identify cells that were co-infected with both viral barcodes (\FIG{experiment}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived form the more common viral barcode.
+We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{experiment}H and \FIGSUPP[experiment]{ISG}, respectively.
+We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
+We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
+The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).
 
@@
 
-Add this
+\jdbcomment{Add this}
 
@@
 
+\clearpage
+
+\begin{suppfile}
+\caption{\label{suppfile:pacbio_analysis}
+An HTML rendering of the Jupyter notebook that analyzes the PacBio data to call the viral sequences in infected cells is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
+This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
+The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb.}}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{\label{suppfile:monocle_analysis}
+An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
+This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
+The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb.}}
+\end{suppfile}
+
 \end{document}

### 2018-09-06 paper/paper.tex
Working on methods section, still rough

@@
 \jdbcomment{Alistair writes this}
+For our type I interferon reporters, A 1kb region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector, with a NotI site immediately downstream serving as an artificial kozak sequence (REF).
+Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeongreen, and mNeongreen translationally linked to LNGFR$\delta$C by a p2A linker (REFS).
+For our type III interferon reporters, a 1.2kb region upstream of the human IL29 (IFNL1) gene was cloned into the pHAGE2 vector, with the native kozak sequence left at the 3' end. 
+Downstream of this promoter zsGreen translationally linked to LNGFR$\delta$C via a p2A linker was cloned. 
+Replication-incompetent lentivirus was generated from these vectors as previously described (REF).
+A monolayer of A549 cells was transduced with each lentivirus, and sorted as single cells. 
+These single cell clones were grown and tested for reporter activity by poly I:C transfection ? a potent agonist of the RIG-I pathway.
+Critically, cells propagated and used in experiments were not exposed to poly I:C,  thus avoiding selection on survival post interferon-induction.
+For our dual- type I and type III reporter, a single-cell clone of our type III reporter strain was transduced with our type I reporter bearing the mCherry fluorescent marker and subsequently propagated as a single cell clone and validated as above.
+
 Reporter sequences \FIGDATA[IFNrare]{reporter_sequences}
@@
 \jdbcomment{Alistair writes this}
-Wildtype and synonymously barcoded viruses in \FIGDATA[experiment]{virus_seqs}
+All viruses were generated using reverse-genetics as previously described.
+Briefly, bidirectional plasmids producing viral mRNA and vRNA are transfected into a co-culture of MDCK-SIAT1 and 293t cells grown in D10.
+After 24h, media is changed from D10 to influenza growth media.
+Between 48h and 72h post-infection, depending on viral growth kinetics, supernatant is harvested and titered by TCID50.
+Virus is subsequently expanded on MDCK-SIAT1 cells as described below.
+For all viral variants, the described mutations were introduced into reverse-genetics plasmids using PCR mutagenesis. 
+For the HA-flanked mCherry pseudovirus, a plasmid was generated similarly to a previously-described HA-flanked eGFP vector.
+Synonymously-barcoded virus utilized previously described 3' synonymous mutations, and two newly-introduced single-nucleotide synonymous mutations at the 5' end of each viral gene.
+Wild-type, synonymously-barcoded, NS1-inactivated, and single-nucleotide variant viruses were all grown in unmodified MDCK-SIAT1 and 293t cells.
+Viral variants bearing large internal deleltions in the PB1 segment were grown in MDCK-SIAT1 and 293t cells expressing PB1 as previously described.
+Pseudovirus expressing mCherry in place of eGFP was grown in MDCK-SIAT1 and 293t cells expressing HA as previously described.
+
+Wildtype and synonymously barcoded viruses in \FIGDATA[experiment]{virus_seqs} obtained from reverse-genetic transfections was subsequently propagated at an MOI of 0.001 for 72h twice in MDCK-SIAT1 cells (wild-type) or once at an MOI of 0.01 (barcoded) for 60h in order to generate potential immunostimulatory diversity.
+For analyses of interferon induction, an additional wild-type stock of virus was grown at an MOI of 0.01 for 36h in order to generate a population relatively free of defective particles.
+(WILL REPEAT WITH 48H STOCK AS DESCRIBED BELOW)
+Viruses bearing SNPs, deletions in PB1, NS1-inactivation mutations, and HA-mCherry pseudovirus were all propagated at an MOI of 0.01 for 48h owing to slower growth kinetics. 
+All stocks were subsequently titered by TCID50, using complementing cells where necessary.
+Additionally, for qPCR experiments, stocks were titered by qPCR against the viral HA segment using a plasmid standard to determine relative genome copy numbers between viral stocks.
 

### 2018-09-06 paper/paper.tex
A few minor changes

@@
 \jdbcomment{Alistair writes this}
-For our type I interferon reporters, A 1kb region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector, with a NotI site immediately downstream serving as an artificial kozak sequence (REF).
+For our type I interferon reporters, A 1kb region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector, with a NotI site immediately downstream serving as an artificial Kozak sequence (REF).
 Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeongreen, and mNeongreen translationally linked to LNGFR$\delta$C by a p2A linker (REFS).
@@
 Replication-incompetent lentivirus was generated from these vectors as previously described (REF).
-A monolayer of A549 cells was transduced with each lentivirus, and sorted as single cells. 
-These single cell clones were grown and tested for reporter activity by poly I:C transfection ? a potent agonist of the RIG-I pathway.
+A monolayer of A549 cells in D10 was transduced with each lentivirus in the presence of 5 $\mu$g polybrene, and sorted as single cells. 
+These single cell clones were grown and tested for reporter activity by poly I:C transfection---a potent agonist of the RIG-I pathway.
 Critically, cells propagated and used in experiments were not exposed to poly I:C,  thus avoiding selection on survival post interferon-induction.
@@
 After 24h, media is changed from D10 to influenza growth media.
-Between 48h and 72h post-infection, depending on viral growth kinetics, supernatant is harvested and titered by TCID50.
-Virus is subsequently expanded on MDCK-SIAT1 cells as described below.
+Between 48h and 72h post-infection, depending on viral growth kinetics, supernatant was harvested and titered by TCID50.
+Virus was subsequently expanded on MDCK-SIAT1 cells as described below.
 For all viral variants, the described mutations were introduced into reverse-genetics plasmids using PCR mutagenesis. 

### 2018-09-07 paper/paper.tex
Small changes to methods made before recent comments, will continue to update!

@@
 For our type I interferon reporters, A 1kb region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector, with a NotI site immediately downstream serving as an artificial Kozak sequence (REF).
-Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeongreen, and mNeongreen translationally linked to LNGFR$\delta$C by a p2A linker (REFS).
+Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeongreen, and mNeongreen translationally linked to low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C) by a p2A linker (REFS).
 For our type III interferon reporters, a 1.2kb region upstream of the human IL29 (IFNL1) gene was cloned into the pHAGE2 vector, with the native kozak sequence left at the 3' end. 
-Downstream of this promoter zsGreen translationally linked to LNGFR$\delta$C via a p2A linker was cloned. 
+Downstream of this promoter zsGreen translationally linked to LNGFR$\Delta$C via a p2A linker was cloned. 
 Replication-incompetent lentivirus was generated from these vectors as previously described (REF).
@@
 All viruses were generated using reverse-genetics as previously described.
+All infections were performed in influenza growth media as previously described.
 Briefly, bidirectional plasmids producing viral mRNA and vRNA are transfected into a co-culture of MDCK-SIAT1 and 293t cells grown in D10.
@@
 \subsection{Single-cell infections and transcriptomics using 10X Chromium on IFN-enriched cells}
+Type I interferon reporter A549 cells (mNeongreen-p2A-LNGFR$/Delta$C) were infected with a mixture of barcoded (MOI 0.1) and wild-type (MOI 0.02) virus. 
+These differing MOI's were chosen to roughly match the fraction of HA-positive cells in flow cytometry experiments (and thus biologically active particles) titering these two stocks. 
+Infections were allowed to proceed for 12h.
+Cells were trypsonized, trypsin was quenched with D10 media, and cells were resuspended in de-gassed PBS supplemented with 0.5\% bovine serum albumin and 5 mM EDTA. 
+Cells were then incubated with anti-LNGFR MACSelect Microbeads and passed over an MS magnetic column twice, selecting the positive population each time.
+The original, unsorted, population was then added back in to approximately 10 percent of the final cell fraction in order to ensure the presence of interferon negative cells.
+At this point, uninfected MDCK-SIAT1 cells were also added to approximately 10 percent of the final cell fraction for doublet and lysis metrics. 
+The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics instrument at a targeted capture of ~1500 cells.
+
+This sample was then processed according to the manufacturer's protocol using the Chromium Single Cell 3?
+Library and Gel Bead Kit v2 with one important modification; rather than process all full-length cDNA through enzymatic fragmentation, several nanograms were retained for targeted sequencing as described below.
+
 \jdbcomment{Alistair writes this}
@@
 \subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
+Approximately 1ng of full-length 10x-barcoded cDNA was used as template in an emulsion PCR using the
+
 \jdbcomment{Alistair writes this}

### 2018-09-10 paper/paper.tex
cleaned up methods for IFN reporter cells

@@
 {To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
-Activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR).
+At \jdbcomment{?} hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR).
 Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal for the IFN-$\lambda$ reporter driving ZsGreen.
@@
 \subsection{IFN reporter cell lines}
-\jdbcomment{Alistair writes this}
-For our type I interferon reporters, A 1kb region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector, with a NotI site immediately downstream serving as an artificial Kozak sequence (REF).
-Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeongreen, and mNeongreen translationally linked to low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C) by a p2A linker (REFS).
-For our type III interferon reporters, a 1.2kb region upstream of the human IL29 (IFNL1) gene was cloned into the pHAGE2 vector, with the native kozak sequence left at the 3' end. 
-Downstream of this promoter zsGreen translationally linked to LNGFR$\Delta$C via a p2A linker was cloned. 
-Replication-incompetent lentivirus was generated from these vectors as previously described (REF).
-A monolayer of A549 cells in D10 was transduced with each lentivirus in the presence of 5 $\mu$g polybrene, and sorted as single cells. 
-These single cell clones were grown and tested for reporter activity by poly I:C transfection---a potent agonist of the RIG-I pathway.
-Critically, cells propagated and used in experiments were not exposed to poly I:C,  thus avoiding selection on survival post interferon-induction.
-For our dual- type I and type III reporter, a single-cell clone of our type III reporter strain was transduced with our type I reporter bearing the mCherry fluorescent marker and subsequently propagated as a single cell clone and validated as above.
+We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}B).
+The parental A549 cell line used to create these reporters was obtained from ATCC (CCL-185), and was tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core and authenticated using the ATCC STR profiling service.
+The cells were maintained in D10 media (DMEM supplemented with 10\% heat-inactivated fetal bovine serum, 2 mM L-glutamine, 100 U of penicillin / ml, and 100 $\mu$g of streptomycin / ml) at 37$^{\circ}$C and 5\% carbon dioxide.
 
-Reporter sequences \FIGDATA[IFNrare]{reporter_sequences}
+To create the type I interferon reporters, a 1kb promoeter region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector~\citep{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
+Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and mNeonGreen translationally linked to low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} by a P2A linker~\citep{kim2011high}.
+The sequence of the last of these constructs is provided in \FIGDATA[IFNrare]{reporter_sequences}.
 
-\subsection{Viral stocks}
+To create the type III interferon reporters, a 1.2kb region upstream of the human IL29 (IFNL1) gene was cloned into the pHAGE2 vector, with the native Kozak sequence retained at the 3' end. 
+Downstream of this promoter we cloned ZsGreen translationally linked to LNGFR$\Delta$C via a P2A linker.
+The sequence of this construct is provided in \FIGDATA[IFNrare]{reporter_sequences}.
+
+We used these constructs to generate lentiviral vectors~\citep{oconnell2010lentiviral} which were used to transduce of A549 cells in the presence of 5 $\mu$g polybrene.
+We then sorted single transduced cells and expanded them.
+A portion of the expanded cells were tested for reporter activity by transfecting poly(I:C) (a potent agonist of the RIG-I pathway), and we retained clones with strong activation.
+Importantly, the cells that we retained for further use were not the same portion that were tested by poly(I:C) treatment, but rather a separate split of the same population---this avoids any selection on the cells from transient activation of IFN.
+For the dual type I / type III reporter used in \FIGSUPP[IFNrare]{type_I_vs_III}, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
+
+\FIGSUPP[IFNrare]{reporter_validation} shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
+For detection of the cell-surface bound LNGFR$\Delta$C, \jdbcomment{cells were stained with?}.
+
+\subsection{Viruses for single-cell experiments}
 \jdbcomment{Alistair writes this}
@@
 \jdbcomment{Alistair writes this}
+
+\subsection{Viruses for validation experiments}
+\jdbcomment{Alistair writes this}
+All viruses were generated using reverse-genetics as previously described.
+All infections were performed in influenza growth media as previously described.
+Briefly, bidirectional plasmids producing viral mRNA and vRNA are transfected into a co-culture of MDCK-SIAT1 and 293t cells grown in D10.
+After 24h, media is changed from D10 to influenza growth media.
+Between 48h and 72h post-infection, depending on viral growth kinetics, supernatant was harvested and titered by TCID50.
+Virus was subsequently expanded on MDCK-SIAT1 cells as described below.
+For all viral variants, the described mutations were introduced into reverse-genetics plasmids using PCR mutagenesis. 
+For the HA-flanked mCherry pseudovirus, a plasmid was generated similarly to a previously-described HA-flanked eGFP vector.
+Synonymously-barcoded virus utilized previously described 3' synonymous mutations, and two newly-introduced single-nucleotide synonymous mutations at the 5' end of each viral gene.
+Wild-type, synonymously-barcoded, NS1-inactivated, and single-nucleotide variant viruses were all grown in unmodified MDCK-SIAT1 and 293t cells.
+Viral variants bearing large internal deleltions in the PB1 segment were grown in MDCK-SIAT1 and 293t cells expressing PB1 as previously described.
+Pseudovirus expressing mCherry in place of eGFP was grown in MDCK-SIAT1 and 293t cells expressing HA as previously described.
+
+Wildtype and synonymously barcoded viruses in \FIGDATA[experiment]{virus_seqs} obtained from reverse-genetic transfections was subsequently propagated at an MOI of 0.001 for 72h twice in MDCK-SIAT1 cells (wild-type) or once at an MOI of 0.01 (barcoded) for 60h in order to generate potential immunostimulatory diversity.
+For analyses of interferon induction, an additional wild-type stock of virus was grown at an MOI of 0.01 for 36h in order to generate a population relatively free of defective particles.
+(WILL REPEAT WITH 48H STOCK AS DESCRIBED BELOW)
+Viruses bearing SNPs, deletions in PB1, NS1-inactivation mutations, and HA-mCherry pseudovirus were all propagated at an MOI of 0.01 for 48h owing to slower growth kinetics. 
+All stocks were subsequently titered by TCID50, using complementing cells where necessary.
+Additionally, for qPCR experiments, stocks were titered by qPCR against the viral HA segment using a plasmid standard to determine relative genome copy numbers between viral stocks.
 

### 2018-09-10 paper/paper.tex
revised methods for viruses for single-cell

@@
 \subsection{Viruses for single-cell experiments}
-\jdbcomment{Alistair writes this}
-All viruses were generated using reverse-genetics as previously described.
-All infections were performed in influenza growth media as previously described.
-Briefly, bidirectional plasmids producing viral mRNA and vRNA are transfected into a co-culture of MDCK-SIAT1 and 293t cells grown in D10.
-After 24h, media is changed from D10 to influenza growth media.
-Between 48h and 72h post-infection, depending on viral growth kinetics, supernatant was harvested and titered by TCID50.
-Virus was subsequently expanded on MDCK-SIAT1 cells as described below.
-For all viral variants, the described mutations were introduced into reverse-genetics plasmids using PCR mutagenesis. 
-For the HA-flanked mCherry pseudovirus, a plasmid was generated similarly to a previously-described HA-flanked eGFP vector.
-Synonymously-barcoded virus utilized previously described 3' synonymous mutations, and two newly-introduced single-nucleotide synonymous mutations at the 5' end of each viral gene.
-Wild-type, synonymously-barcoded, NS1-inactivated, and single-nucleotide variant viruses were all grown in unmodified MDCK-SIAT1 and 293t cells.
-Viral variants bearing large internal deleltions in the PB1 segment were grown in MDCK-SIAT1 and 293t cells expressing PB1 as previously described.
-Pseudovirus expressing mCherry in place of eGFP was grown in MDCK-SIAT1 and 293t cells expressing HA as previously described.
+We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
+The cells were infected with a roughly equal mix of the wildtype virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
+We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to allow us to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
+The sequences of all gene segments from the wildtype and synonymously barcoded viral strains are in \FIGDATA[experiment]{virus_seqs}.
 
-Wildtype and synonymously barcoded viruses in \FIGDATA[experiment]{virus_seqs} obtained from reverse-genetic transfections was subsequently propagated at an MOI of 0.001 for 72h twice in MDCK-SIAT1 cells (wild-type) or once at an MOI of 0.01 (barcoded) for 60h in order to generate potential immunostimulatory diversity.
-For analyses of interferon induction, an additional wild-type stock of virus was grown at an MOI of 0.01 for 36h in order to generate a population relatively free of defective particles.
-(WILL REPEAT WITH 48H STOCK AS DESCRIBED BELOW)
-Viruses bearing SNPs, deletions in PB1, NS1-inactivation mutations, and HA-mCherry pseudovirus were all propagated at an MOI of 0.01 for 48h owing to slower growth kinetics. 
-All stocks were subsequently titered by TCID50, using complementing cells where necessary.
-Additionally, for qPCR experiments, stocks were titered by qPCR against the viral HA segment using a plasmid standard to determine relative genome copy numbers between viral stocks.
+Both viral strains were generated by reverse genetics using the pHW18* series of bi-directional plasmids~\citep{hoffmann2000dna}.
+We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\citep{xue2016propagation}.
+The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 in influenza growth media and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
+Specifically, after generation by reverse genetics, the wildtype variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
+The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of \citet{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
+
+For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
+We estimated titers of transcriptionally active virions by staining for HA expression in virus-infected A549 cells.
+Specifically, we infected A549 cells in influenza growth medium, and at \jdbcomment{?} hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\citep{doud2017complete}.
+After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC, washed, fixed in 1\% formaldehyde in PBS, washed again, and then analyzed by flow cytometry to determine the fraction expressing detectable HA protein.
 
@@
 \subsection{Viruses for validation experiments}
-\jdbcomment{Alistair writes this}
+\jdbcomment{Alistair writes this. I've copied text from prior consolidated section, but it needs to be clarified by more clearly breaking things down by virus.}
 All viruses were generated using reverse-genetics as previously described.

### 2018-09-10 paper/paper.tex
revised methods for single-cell infection, MACS, and 10X processing

@@
 (1) IFN reporter A549 cells are infected with a mix of wildtype virus and virus carrying synonymous mutations at both ends of each gene segment.
-IFN+ cells are enriched by MACS, and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
+IFN+ cells are enriched by MACS (\FIGSUPP[experiment]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
 (2) The mRNAs from individual cells are tagged with cell-specific barcodes using a 10X Chromium device.
@@
 \label{fig:experiment}
+
+\figsupp[MACS enrichment of IFN+ cells]
+{\jdbcomment{Add caption}}
+{\jdbcomment{Add figure showing example MACS sorting}}
+\label{figsupp:MACS}
 
@@
 To create the type I interferon reporters, a 1kb promoeter region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector~\citep{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
-Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and mNeonGreen translationally linked to low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} by a P2A linker~\citep{kim2011high}.
+Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\citep{kim2011high}.
 The sequence of the last of these constructs is provided in \FIGDATA[IFNrare]{reporter_sequences}.
@@
 
-\subsection{Single-cell infections and transcriptomics using 10X Chromium on IFN-enriched cells}
-Type I interferon reporter A549 cells (mNeongreen-p2A-LNGFR$/Delta$C) were infected with a mixture of barcoded (MOI 0.1) and wild-type (MOI 0.02) virus. 
-These differing MOI's were chosen to roughly match the fraction of HA-positive cells in flow cytometry experiments (and thus biologically active particles) titering these two stocks. 
-Infections were allowed to proceed for 12h.
-Cells were trypsonized, trypsin was quenched with D10 media, and cells were resuspended in de-gassed PBS supplemented with 0.5\% bovine serum albumin and 5 mM EDTA. 
-Cells were then incubated with anti-LNGFR MACSelect Microbeads and passed over an MS magnetic column twice, selecting the positive population each time.
-The original, unsorted, population was then added back in to approximately 10 percent of the final cell fraction in order to ensure the presence of interferon negative cells.
-At this point, uninfected MDCK-SIAT1 cells were also added to approximately 10 percent of the final cell fraction for doublet and lysis metrics. 
-The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics instrument at a targeted capture of ~1500 cells.
+\subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
+The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the LNGFR$\Delta$C-P2A-mNeonGreen reporter.
+A schematic of the experiment is shown in \FIG{experiment}A.
 
-This sample was then processed according to the manufacturer's protocol using the Chromium Single Cell 3?
-Library and Gel Bead Kit v2 with one important modification; rather than process all full-length cDNA through enzymatic fragmentation, several nanograms were retained for targeted sequencing as described below.
+The wildtype and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
+The cells were then infected with this mixture at a dose designed to infect about 30\% of cells \jdbcomment{Is this correct that you targeted 30\%?} (this is about what ended up being the rate of detectable infection in the final sorted and sequenced cells, see \FIG{experiment}D).
+\jdbcomment{Were the cells washed after infection?}
+Infections were allowed to proceed for 12 hours.
+The cells were then trypsinized, the trypsin was quenched with D10 media, and cells were resuspended in de-gassed PBS supplemented with 0.5\% bovine serum albumin and 5 mM EDTA. 
+To enrich IFN+ cells, the cells were then incubated with anti-LNGFR MACSelect Microbeads (Miltenyi Biotec) and twice passed over an MS magnetic column (Miltenyi Biotec), retaining the bound (and presumably IFN-enriched) population each time.
+This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \FIGSUPP[experiment]{MACS}.
+The original, unsorted, population was then added back in to approximately 10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
+At this point, uninfected MDCK-SIAT1 cells were also added to approximately 10\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{experiment}B) and background viral mRNA in uninfected cells (\FIG{experiment}D).
+The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.
 
-\jdbcomment{Alistair writes this}
+This sample was then processed to create libraries for Illumina 3'-end sequencing according to the 10X Genomics protocol using the Chromium Single Cell 3' Library and Gel Bead Kit v2 with one important modification: rather than process all full-length cDNA through enzymatic fragmentation, several nanograms were retained for targeted full-length viral cDNA sequencing as described below.
+The single-cell mRNA-seq library was sequenced on an Illumina HiSeq 2500, and the data analyzed as described below.
 
@@
 Approximately 1ng of full-length 10x-barcoded cDNA was used as template in an emulsion PCR using the 
-
 \jdbcomment{Alistair writes this}

### 2018-09-10 paper/paper.tex
added MACS enrichment figure

@@
 \figsupp[MACS enrichment of IFN+ cells]
-{\jdbcomment{Add caption}}
-{\jdbcomment{Add figure showing example MACS sorting}}
+{
+Example MACS enrichments of IFN+ influenza-infected cells.
+The A549 cells with the IFN-$\beta$ LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype A/WSN/1933 influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
+After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
+{\bf (A)}
+After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
+The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
+As indicated by the percentages shown for the original and MACS-sorted population, this process led to between 50- and 350-fold enrichment for IFN+ cells.
+We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
+{\bf (B)}
+Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
+The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
+The qPCR was performed in quadruplicate (hence the four points for each sample).
+}
+{\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}}
 \label{figsupp:MACS}
@@
 \jdbcomment{Were the cells washed after infection?}
-Infections were allowed to proceed for 12 hours.
+Infections were allowed to proceed for 12 hours \jdbcomment{Clarify: this says 12 hours, but other places it says 13 hours.
+Were they harvested at 12 but the idea is that the whole process added another hour, or were they harvested at 13 hours?}.
 The cells were then trypsinized, the trypsin was quenched with D10 media, and cells were resuspended in de-gassed PBS supplemented with 0.5\% bovine serum albumin and 5 mM EDTA. 
@@
 Approximately 1ng of full-length 10x-barcoded cDNA was used as template in an emulsion PCR using the 
+\jdbcomment{Alistair writes this}
+
+\subsection{qPCR for viral genes and IFN}
 \jdbcomment{Alistair writes this}

### 2018-09-10 paper/paper.tex
Addressed minor comments

@@
 {To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
-At \jdbcomment{?} hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR).
+At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
 Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal for the IFN-$\lambda$ reporter driving ZsGreen.
@@
 \FIGSUPP[IFNrare]{reporter_validation} shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
-For detection of the cell-surface bound LNGFR$\Delta$C, \jdbcomment{cells were stained with?}.
+For detection of the cell-surface bound LNGFR$\Delta$C, cells were stained with PE-conjugated anti-LNGFR (CD271) antibody from Miltenyi Biotec.
 
@@
 We estimated titers of transcriptionally active virions by staining for HA expression in virus-infected A549 cells.
-Specifically, we infected A549 cells in influenza growth medium, and at \jdbcomment{?} hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\citep{doud2017complete}.
+Specifically, we infected A549 cells in influenza growth medium, and at 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\citep{doud2017complete}.
 After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC, washed, fixed in 1\% formaldehyde in PBS, washed again, and then analyzed by flow cytometry to determine the fraction expressing detectable HA protein.
@@
 The wildtype and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
-The cells were then infected with this mixture at a dose designed to infect about 30\% of cells \jdbcomment{Is this correct that you targeted 30\%?} (this is about what ended up being the rate of detectable infection in the final sorted and sequenced cells, see \FIG{experiment}D).
-\jdbcomment{Were the cells washed after infection?}
-Infections were allowed to proceed for 12 hours \jdbcomment{Clarify: this says 12 hours, but other places it says 13 hours.
-Were they harvested at 12 but the idea is that the whole process added another hour, or were they harvested at 13 hours?}.
+The cells were then infected with this mixture at a dose designed to infect about 50\% of cells.
+Infections were allowed to proceed for 12 hours.
 The cells were then trypsinized, the trypsin was quenched with D10 media, and cells were resuspended in de-gassed PBS supplemented with 0.5\% bovine serum albumin and 5 mM EDTA. 
@@
 At this point, uninfected MDCK-SIAT1 cells were also added to approximately 10\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{experiment}B) and background viral mRNA in uninfected cells (\FIG{experiment}D).
+Preparation of samples for single cell analysis added an extra hour to the process, and thus the cells were analyzed at a final time point of 13 hours post infection.
 The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.

### 2018-09-10 paper/paper.tex
very minor tweaks to Methods wording

@@
 The wildtype and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
-The cells were then infected with this mixture at a dose designed to infect about 50\% of cells.
+The cells were then infected with this mixture at a dose designed to infect about half the cells (\FIG{experiment}D suggests that the actual rate of detectable infection was slightly lower).
 Infections were allowed to proceed for 12 hours.
@@
 At this point, uninfected MDCK-SIAT1 cells were also added to approximately 10\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{experiment}B) and background viral mRNA in uninfected cells (\FIG{experiment}D).
-Preparation of samples for single cell analysis added an extra hour to the process, and thus the cells were analyzed at a final time point of 13 hours post infection.
+Preparation of samples for single cell analysis (which was performed at room temperature) added an extra hour to the process after cell collection, and thus we consider the cells to have been analyzed at 13 hours post infection.
 The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.

### 2018-09-10 paper/paper.tex
Added section on PacBio PCR enrichment. Will keep working through on things. I will probably need to add some kind of schematic as well as a primer table?

--- 2018-09-10 paper/paper.tex
@@
 \subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
-Approximately 1ng of full-length 10x-barcoded cDNA was used as template in an emulsion PCR using the 
-\jdbcomment{Alistair writes this}
+All PCR amplifications were performed using Kapa HiFi Hotstart Readymix.
+1ng of full-length 10x-barcoded cDNA was used as template in a multiplex PCR reaction using  eight segment-specific 5' primers and a shared, 10x-specific 3' primer.
+To reduce strand-exchange and generate more even amplification across segments, this reaction was performed in an emulsion PCR using the Micellula DNA Emulsion Kit from Roboklon, supplementing our PCR reaction with additional BSA to a final concentration of 0.1mg/ml.
+For all emulsion PCRs, 30 rounds of PCR was performed in order to saturate emulsion materials---unlike open PCR reactions, the physical occlusion of amplified material reduces artifacts at high cycle numbers.
+This amplified material was then subjected to eight, individual, emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 10x-specific 3' primer, using 1ng of material in each reaction.
+This material was pooled to try and obtain an equimolar ratio of segments and sequenced in one SMRTcell in the PacBio RSII, and one SMRTcell of the PacBio Sequel. 
+All remaining SMRT-sequencing described in this section was performed on the PacBio Sequel.
+To reduce biases in segment representation, we repooled this material and ran another SMRTcell consisting of only HA, NP, NA, M, and NS  segments, and two SMRTcells consisting of PB2, PB1, and PA segments.
+As our depth on the three longest segments remained low, we further amplified our already-amplified material using a non-emulsion PCR using standard conditions for PB2, PB1, and PA, and subjecting the resulting material to a rigorous SPRI-bead selection using SPRIselect beads at a bead-PCR volume ratio of 0.4.
+All non-emulsion PCRs were recovered at sub $/{\mu}$g yields to prevent the occurrence of artifacts at PCR saturation.
+We sequenced this bead-selected and repooled material on an additional SMRTcell.
+Finally, to improve recovery of full-length PB1, PB2, and PA due to over-representation of internally deleted sequences deriving from defective viruses, we first used our already-amplified segment-specific material in a low-cycle PCR to append circularization adapters, circularized 10ng of this material using NEBuilder in a 20µl reaction volume using an extended reaction time of 50 minutes, incubated these reactions a further 1 hour at 37$^{\circ}$C with exonuclease V to remove linear concatemers and uncircularized material, and bead-cleaned using SPRIselect beads at a volume ratio of 0.4.
+This circularized material was then used in three non-emulsion PCRs specific to PB2, PB1, and PA, using two segment-specific primers that align to the central portion of each gene but in apposition to one-another, and bead-cleaned using 0.4 volumes.
+This results in successful amplification only if the central portion of the gene is deleted, and if the cDNA is circularized, while still amplifying across cell-specific barcode regions. 
+This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRTcell in the PacBio Sequel.
 
 \subsection{qPCR for viral genes and IFN}
+
 \jdbcomment{Alistair writes this}

### 2018-09-10 paper/paper.tex
fixed two small symbol typos

@@
 To reduce biases in segment representation, we repooled this material and ran another SMRTcell consisting of only HA, NP, NA, M, and NS  segments, and two SMRTcells consisting of PB2, PB1, and PA segments.
+
 As our depth on the three longest segments remained low, we further amplified our already-amplified material using a non-emulsion PCR using standard conditions for PB2, PB1, and PA, and subjecting the resulting material to a rigorous SPRI-bead selection using SPRIselect beads at a bead-PCR volume ratio of 0.4.
-All non-emulsion PCRs were recovered at sub $/{\mu}$g yields to prevent the occurrence of artifacts at PCR saturation.
+All non-emulsion PCRs were recovered at sub ${\mu}$g yields to prevent the occurrence of artifacts at PCR saturation.
 We sequenced this bead-selected and repooled material on an additional SMRTcell.
-Finally, to improve recovery of full-length PB1, PB2, and PA due to over-representation of internally deleted sequences deriving from defective viruses, we first used our already-amplified segment-specific material in a low-cycle PCR to append circularization adapters, circularized 10ng of this material using NEBuilder in a 20µl reaction volume using an extended reaction time of 50 minutes, incubated these reactions a further 1 hour at 37$^{\circ}$C with exonuclease V to remove linear concatemers and uncircularized material, and bead-cleaned using SPRIselect beads at a volume ratio of 0.4.
+Finally, to improve recovery of full-length PB1, PB2, and PA due to over-representation of internally deleted sequences deriving from defective viruses, we first used our already-amplified segment-specific material in a low-cycle PCR to append circularization adapters, circularized 10ng of this material using NEBuilder in a 20${\mu}$l reaction volume using an extended reaction time of 50 minutes, incubated these reactions a further 1 hour at 37$^{\circ}$C with exonuclease V to remove linear concatemers and uncircularized material, and bead-cleaned using SPRIselect beads at a volume ratio of 0.4.
 This circularized material was then used in three non-emulsion PCRs specific to PB2, PB1, and PA, using two segment-specific primers that align to the central portion of each gene but in apposition to one-another, and bead-cleaned using 0.4 volumes.

### 2018-09-11 paper/paper.tex
Draft of qPCR, requires a few additions, references, primer sequences

@@
 \subsection{qPCR for viral genes and IFN}
+\abrcomment{Assuming we use the cycloheximide and ribavirin data as-is, can be modified if not}
+For figures \abrcomment{whichever figures} A549 cells were seeded at a density of 10,000 cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+Immediately prior to infection D10 media was removed and replaced with influenza growth media and virus was added.
+Viral infections were normalized to an MOI of a relatively pure stock of virus (0.4 for (cycloheximide figure) and 0.2 for (ribavirin figure), and identical particle numbers (calculated by qPCR, described below) of other viruses were added.
+For experimental treatments of ribavirin and cycloheximde, these compounds were added to cells at the time of infection to a final concentration of 100$\mu$M and 50$\mu$g/ml, respectively.
+At 8 hours post infection media was removed and cell lysates generated using the CellAmp RT-PCR for Direct Prep Kit (Real Time) \& Protein Analysis, following manufacturer's instructions. 
+cDNA was synthesized using an oligoDT primer and the SuperScript III first-strand synthesis supermix from ThermoFisher using the manufacturer’s protocol. 
+Transcript abundance was measured using SYBR green PCR master mix, using a combined anneal/extension step of 60 for one minute with the following primers: HA: 5’-GGCCCAACCACACATTCAAC-3’, 5’-GCTCATCACTGCTAGACGGG-3’, IFNB1: 5’-AAACTCATGAGCAGTCTGCA-3’, 5’-AGGAGATCTTCAGTTTCGGAGG-3’, L32: 5’-AGCTCCCAAAAATAGACGCAC-3’, 5’-TTCATAGCAGTAGGCACAAAGG-3’.
+
+For titering virus by qPCR, vRNA was harvested from 80 μl of viral supernatant by the addition of 600 μl of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol.
+The cDNA was generated using SuperScript III first-strand synthesis supermix using the manufacturer’s protocol, and using the universal vRNA primers of Hoffmann et al.
+(add Hoffman ref) with the modifications described in Xue et al.
+(add Katherine ref).
+The qPCR was then performed as for mRNA measurements
 
@@
 \end{document}
+ent}

### 2018-09-11 paper/paper.tex
minor tweaks to LaTex so it compiles

@@
 \newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
+\newcommand{\abrcomment}[1]{\emph{\color{blue} [#1]}}
 
@@
 To create the type III interferon reporters, a 1.2kb region upstream of the human IL29 (IFNL1) gene was cloned into the pHAGE2 vector, with the native Kozak sequence retained at the 3' end. 
-Downstream of this promoter we cloned ZsGreen translationally linked to LNGFR$\Delta$C via a P2A linker.
+Downstream of this promoter we cloned LNGFR$\Delta$C linked to ZsGreen via a P2A linker.
 The sequence of this construct is provided in \FIGDATA[IFNrare]{reporter_sequences}.
@@
 
-For titering virus by qPCR, vRNA was harvested from 80 μl of viral supernatant by the addition of 600 μl of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol.
+For titering virus by qPCR, vRNA was harvested from 80 $\mu$l of viral supernatant by the addition of 600 $\mu$l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol.
 The cDNA was generated using SuperScript III first-strand synthesis supermix using the manufacturer’s protocol, and using the universal vRNA primers of Hoffmann et al.

### 2018-09-11 paper/paper.tex
added Fig 1C showing % IFN+
Analyzed flow data for uninfected, flu-infected,
and Sendai-infected cells and added figure
and figure supplement.

@@
 {\bf \Large A}\includegraphics[width=0.25\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
-\hspace{0.05\textwidth}
-{\bf \Large B}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
-\hspace{0.05\textwidth}
-{\bf \Large C} \jdbcomment{ADD FIGURE PANEL}
+\hspace{0.02\textwidth}
+{\bf \Large B}\includegraphics[width=0.32\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
+\hspace{0.02\textwidth}
+{\bf \Large C} \includegraphics[width=0.18\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
 }
@@
 {\bf (C)}
-Frequency of IFN induction upon infection with the influenza virus stocks used in this paper as quantified using the reporter cell lines.
-\jdbcomment{This should be a simple figure, such as a plot that just shows the fraction IFN+ among flu infected cells for the virus stocks we use, plus positive and negative controls.
-Then put detailed flow data in \FIGSUPP[IFNrare]{flow}. }
+Frequency of IFN induction upon infection with the influenza virus stocks used in the single-cell studies in this paper as quantified using the IFN reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
+The plot also shows uninfected cells as a negative control, and cells infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai} as a positive control.
+The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
 }
@@
 \figsupp[Full flow cytometry data for \FIG{IFNrare}C.]
-{\jdbcomment{ADD CAPTION}}
-{\jdbcomment{ADD FIGURE}}
+{Flow cytometry data for \FIG{IFNrare}C.
+\jdbcomment{add details}
+}
+{\includegraphics[width=0.75\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}}
 \label{figsupp:flow}

### 2018-09-12 paper/paper.tex
finished adding legends to Figure 1 and supplements

@@
 The Venn diagram shows data aggregated over all samples from \citet{russell2018extreme}.
-{\bf (B)} Creation of an A549 reporter cell line to efficiently identify single cells expressing IFN.
-The cells are engineered to contain an IFN promoter that drives expression of a fluorescent protein and cell-surface protein amenable to MACS (magnetic-activated cell sorting).
-We created reporter cell lines with both type I (IFN-$\beta$) and type III (IFN-$\lambda$) promoters (\FIGDATA[IFNrare]{reporter_sequences}).
-The reporters were efficiently activated when infected with a virus known to potently induce IFN (\FIGSUPP[IFNrare]{reporter_validation}).
+\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell mRNA-sequencing study of influenza-infected mice by \citet{steuerman2018dissection}.
+{\bf (B)} Creation of an A549 reporter cell line to identify single cells expressing IFN.
+The cells contain an IFN promoter that drives expression of a cell-surface protein amenable to MACS (magnetic-activated cell sorting) and a fluorescent protein.
+We created reporter cell lines with type I (IFN-$\beta$) and type III (IFN-$\lambda$) promoters (\FIGDATA[IFNrare]{reporter_sequences}).
+The reporters were efficiently activated when infected with a strain of Sendai virus~\citep{strahle2006sendai} known to potently induce IFN (\FIGSUPP[IFNrare]{reporter_validation}).
 Expression of type I and type III IFN is highly correlated in the reporter cell lines upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
 {\bf (C)}
-Frequency of IFN induction upon infection with the influenza virus stocks used in the single-cell studies in this paper as quantified using the IFN reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
-The plot also shows uninfected cells as a negative control, and cells infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai} as a positive control.
+Frequency of IFN induction upon infection with the influenza virus stocks used in the single-cell studies in this paper quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
+The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus as a positive control.
 The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
@@
 \label{fig:IFNrare}
+
+\figsupp[Few cells express detectable IFN in the single-cell analysis of influenza-infected mice by \citet{steuerman2018dissection}.]
+{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell mRNA-sequencing of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
+Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
+Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
+As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
+An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
+Nonetheless, there is detectable expression of key interferon-stimulated gene (ISG) mRNAs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one mRNA transcript detected).
+The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
+}
+{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
+\label{figsupp:mice}
 
@@
 {Flow cytometry data for \FIG{IFNrare}C.
-\jdbcomment{add details}
+The A549 cells with the IFN-$\lambda$ reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
+After 13 hours, the cells were stained for expression of HA protein and analyzed for the IFN-$\lambda$ reporter driven ZsGreen expression by FACS.
+Each condition was done in triplicate.
+The contour plots show the density of all cells, and all IFN+ cells are also indicated by orange dots.
+Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
+For the influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
+For the uninfected and Sendai-virus infected, the percentage IFN+ was calculated among all cells, since these cells do not express HA.
 }
@@
 \label{figsupp:flow}
-
-\figsupp[Few cells express detectable IFN in the single-cell analysis influenza-infected mice by \citet{steuerman2018dissection}.]
-{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell mRNA-sequencing of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
-Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
-Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
-As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
-An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key interferon-stimulated gene (ISG) mRNAs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels.
-The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
-}
-{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
-\label{figsupp:mice}
 

### 2018-09-12 paper/paper.tex
Rewrote pacbio and qPCR sections. Will need to add references and supplemental information to these sections still

--- 2018-09-11 paper/paper.tex
@@
 All PCR amplifications were performed using Kapa HiFi Hotstart Readymix.
-1ng of full-length 10x-barcoded cDNA was used as template in a multiplex PCR reaction using  eight segment-specific 5' primers and a shared, 10x-specific 3' primer.
+In order to enrich full-length cDNA derived from influenza transcripts, 1ng of full-length 10x-barcoded cDNA was used as template in a multiplex PCR reaction using eight segment-specific 5' primers and a shared, 10x-specific 3' primer \abrcomment{supplemental text}.
 To reduce strand-exchange and generate more even amplification across segments, this reaction was performed in an emulsion PCR using the Micellula DNA Emulsion Kit from Roboklon, supplementing our PCR reaction with additional BSA to a final concentration of 0.1mg/ml.
-For all emulsion PCRs, 30 rounds of PCR was performed in order to saturate emulsion materials---unlike open PCR reactions, the physical occlusion of amplified material reduces artifacts at high cycle numbers.
-This amplified material was then subjected to eight, individual, emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 10x-specific 3' primer, using 1ng of material in each reaction.
+Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
+This process physically separates disparate template molecules, preventing strand exchange, and, moreover, allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules.
+To reach this level of reagent exhaustion, for all emulsion PCRs described here, 30 rounds of PCR was performed in order to saturate emulsion materials---unlike open PCR reactions, the physical occlusion of amplified material reduces artifacts at high cycle numbers.
+
+As multiplex enrichment would still result in highly uneven reads across influenza transcripts due to differences in gene segment expression, this amplified material was then subjected to eight, individual, emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 10x-specific 3' primer, using 1ng of material in each reaction.
+This permitted us to selectively enrich transcripts derived from each influenza segment and then recombine for more even sequencing across the influenza genome.
 This material was pooled to try and obtain an equimolar ratio of segments and sequenced in one SMRTcell in the PacBio RSII, and one SMRTcell of the PacBio Sequel. 
 All remaining SMRT-sequencing described in this section was performed on the PacBio Sequel.
-To reduce biases in segment representation, we repooled this material and ran another SMRTcell consisting of only HA, NP, NA, M, and NS  segments, and two SMRTcells consisting of PB2, PB1, and PA segments.
+Our theoretically equimolar pool of PCR products still produced highly biased ratios of segment coverage, and so to reduce the variability in length of our input pool to try and reduce bias in SMRT library preparation we produced two new sequencing pools, one consisting of the segments HA, NP, NA, M, and NS, and the other consisting of the longer PB2, PB1, and PA transcripts.
+The latter was sequenced on two additional SMRTcells, and the former on a single SMRTcell.
 
 As our depth on the three longest segments remained low, we further amplified our already-amplified material using a non-emulsion PCR using standard conditions for PB2, PB1, and PA, and subjecting the resulting material to a rigorous SPRI-bead selection using SPRIselect beads at a bead-PCR volume ratio of 0.4.
+This bead-selection was an attempt to correct for the high level of internally-deleted PB2, PB1, and PA segments dominating our sequencing, and enrich for the longer, full-length segments.
 All non-emulsion PCRs were recovered at sub ${\mu}$g yields to prevent the occurrence of artifacts at PCR saturation.
 We sequenced this bead-selected and repooled material on an additional SMRTcell.
-Finally, to improve recovery of full-length PB1, PB2, and PA due to over-representation of internally deleted sequences deriving from defective viruses, we first used our already-amplified segment-specific material in a low-cycle PCR to append circularization adapters, circularized 10ng of this material using NEBuilder in a 20${\mu}$l reaction volume using an extended reaction time of 50 minutes, incubated these reactions a further 1 hour at 37$^{\circ}$C with exonuclease V to remove linear concatemers and uncircularized material, and bead-cleaned using SPRIselect beads at a volume ratio of 0.4.
-This circularized material was then used in three non-emulsion PCRs specific to PB2, PB1, and PA, using two segment-specific primers that align to the central portion of each gene but in apposition to one-another, and bead-cleaned using 0.4 volumes.
-This results in successful amplification only if the central portion of the gene is deleted, and if the cDNA is circularized, while still amplifying across cell-specific barcode regions. 
+
+Enrichment of full-length PB2, PB1, and PA was aided by bead-selection but these segments still exhibited relatively low PCR efficiency and comprised a small component of the remaining material.
+Finally, to improve recovery of full-length PB1, PB2, and PA and overcome poor PCR efficiency and over-representation of internally deleted sequences deriving from defective viruses, we undertook a technique allowing us to use two segment-specific primers rather than one, in a PCR that would exclude internally-deleted sequences.
+Namely, we sought to circularize our template molecules, allowing us to still amplify across the whole transcript, including the cell-specific barcode, while using two segment-specific primers in apposition that lie within a region absent in most internally-deleted segments.
+First, we first used our already-amplified segment-specific material in a low-cycle PCR to append circularization adapters, and circularized 10ng of this material using NEBuilder in a 20${\mu}$l reaction volume using an extended reaction time of 50 minutes.
+We then further incubated these reactions a  1 hour at 37$^{\circ}$C with exonuclease V to remove linear concatemers and uncircularized material, and bead-cleaned using SPRIselect beads at a volume ratio of 0.4.
+This circularized material was then used in three non-emulsion PCRs specific to PB2, PB1, and PA, using two segment-specific primers that align to the central portion of each gene but in apposition to one-another, and bead-cleaned using a ratio of 0.4 volumes.
+This results in successful amplification only if the central portion of the gene is present, while still amplifying across the entire transcript as well as cell-specific barcode regions. 
 This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRTcell in the PacBio Sequel.
@@
 For figures \abrcomment{whichever figures} A549 cells were seeded at a density of 10,000 cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
-Immediately prior to infection D10 media was removed and replaced with influenza growth media and virus was added.
-Viral infections were normalized to an MOI of a relatively pure stock of virus (0.4 for (cycloheximide figure) and 0.2 for (ribavirin figure), and identical particle numbers (calculated by qPCR, described below) of other viruses were added.
-For experimental treatments of ribavirin and cycloheximde, these compounds were added to cells at the time of infection to a final concentration of 100$\mu$M and 50$\mu$g/ml, respectively.
+Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at the indicated MOI.
+Viral infections were normalized to an MOI of a relatively-deletion free stock of virus that we previously characterized (ref), and identical particle numbers (calculated by qPCR, described below) of other viruses were added.
+Crucially, by controlling for particle number rather than infectious units we avoid the confounding problem of differing immune stimulation by simply infecting at a higher rate of biologically active but non-infectious particles.
+For experimental treatments with ribavirin and cycloheximde, these compounds were added to cells at the time of infection to a final concentration of 100$\mu$M and 50$\mu$g/ml, respectively, both of which are sufficient to block downstream viral replication (references).
 At 8 hours post infection media was removed and cell lysates generated using the CellAmp RT-PCR for Direct Prep Kit (Real Time) \& Protein Analysis, following manufacturer's instructions. 
@@
 (add Katherine ref).
-The qPCR was then performed as for mRNA measurements
+The qPCR was then performed as for mRNA measurements.
 

### 2018-09-12 paper/paper.tex
minor edits to methods

@@
 
-To create the type I interferon reporters, a 1kb promoeter region upstream of the human IFNB1 gene was cloned into the pHAGE2 lentiviral vector~\citep{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
+To create the type I interferon reporters, a 1kb promoter region upstream of the human IFNB1 gene were cloned into the pHAGE2 lentiviral vector~\citep{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
 Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\citep{kim2011high}.
@@
 We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
-The cells were infected with a roughly equal mix of the wildtype virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
+We used both the wildtype virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
 We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
-In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to allow us to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
 The sequences of all gene segments from the wildtype and synonymously barcoded viral strains are in \FIGDATA[experiment]{virus_seqs}.
+These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
 
@@
 
+\subsection{Flow cytometry analyses for HA expression}
 For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
 We estimated titers of transcriptionally active virions by staining for HA expression in virus-infected A549 cells.
-Specifically, we infected A549 cells in influenza growth medium, and at 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\citep{doud2017complete}.
+Specifically, we infected A549 cells (or one of the A549 reporter cell line variants as indicated) in influenza growth medium, and at 13 to 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\citep{doud2017complete}.
 After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC, washed, fixed in 1\% formaldehyde in PBS, washed again, and then analyzed by flow cytometry to determine the fraction expressing detectable HA protein.
@@
 The original, unsorted, population was then added back in to approximately 10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
-At this point, uninfected MDCK-SIAT1 cells were also added to approximately 10\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{experiment}B) and background viral mRNA in uninfected cells (\FIG{experiment}D).
-Preparation of samples for single cell analysis (which was performed at room temperature) added an extra hour to the process after cell collection, and thus we consider the cells to have been analyzed at 13 hours post infection.
+At this point, uninfected canine (MDCK-SIAT1) cells were also added to approximately 10\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{experiment}B) and background viral mRNA in uninfected cells (\FIG{experiment}D).
+We begin this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
 The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.

### 2018-09-13 paper/paper.tex
Jesse revisions to PacBio methods

@@
 {\centering
-\includegraphics[height=0.8\textheight]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.76\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 \label{figsupp:genotypes_by_ifn}
+
+\figdata{A text file giving the primers used to amplify the influenza transcripts for PacBio sequencing are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
+}
+\label{figdata:primer_sequences}
 
@@
 \subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
-All PCR amplifications were performed using Kapa HiFi Hotstart Readymix.
-In order to enrich full-length cDNA derived from influenza transcripts, 1ng of full-length 10x-barcoded cDNA was used as template in a multiplex PCR reaction using eight segment-specific 5' primers and a shared, 10x-specific 3' primer \abrcomment{supplemental text}.
-To reduce strand-exchange and generate more even amplification across segments, this reaction was performed in an emulsion PCR using the Micellula DNA Emulsion Kit from Roboklon, supplementing our PCR reaction with additional BSA to a final concentration of 0.1mg/ml.
+We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
+All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{experiment}A for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} for at detailed schematic).
+However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
+We therefore needed to enrich for the viral molecules, a process made more challenging by the fact that the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end means that an enrichment PCR could only be semi-specific (the primer at the 5' end can be specific, but that the 3' end must bind the common 10X adaptor shared among all molecules).
+
+We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (although some viral segments produce multiple splice forms, all mRNAs from a given segment share the same 5' end).
+The sequences of these primers are in \FIGDATA[genotypes]{primer_sequences}.
+A major concern during these PCRs is strand exchange (see \FIGSUPP[genotypes]{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
+To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
 Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
-This process physically separates disparate template molecules, preventing strand exchange, and, moreover, allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules.
-To reach this level of reagent exhaustion, for all emulsion PCRs described here, 30 rounds of PCR was performed in order to saturate emulsion materials---unlike open PCR reactions, the physical occlusion of amplified material reduces artifacts at high cycle numbers.
+This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\citep{Boers:2015emulsion}.
+We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .
+Both the common 3' primer and the multiplex mix of eight 5' primers were added to a final concentration of 0.5$\mu$M.
+We performed 30 cycles of PCR, using an extension time of 2 minutes 15 seconds at 67$^{\circ}$C, and a melting temperature of 95$^{\circ}C$.
+This melting temperature is lower than the standard 98$^{\circ}$C melting step suggested by the manufacturer for Kapa HiFi because we wanted to avoid collapse of emulsion integrity at high temperature.
+We performed 30 cycles in order to saturate the reagents in each emulsion---unlike for open PCR reactions, the physical occlusion of amplified material in emulsin PCR reduces artifacts at high cycle numbers.
 
-As multiplex enrichment would still result in highly uneven reads across influenza transcripts due to differences in gene segment expression, this amplified material was then subjected to eight, individual, emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 10x-specific 3' primer, using 1ng of material in each reaction.
-This permitted us to selectively enrich transcripts derived from each influenza segment and then recombine for more even sequencing across the influenza genome.
-This material was pooled to try and obtain an equimolar ratio of segments and sequenced in one SMRTcell in the PacBio RSII, and one SMRTcell of the PacBio Sequel. 
-All remaining SMRT-sequencing described in this section was performed on the PacBio Sequel.
-Our theoretically equimolar pool of PCR products still produced highly biased ratios of segment coverage, and so to reduce the variability in length of our input pool to try and reduce bias in SMRT library preparation we produced two new sequencing pools, one consisting of the segments HA, NP, NA, M, and NS, and the other consisting of the longer PB2, PB1, and PA transcripts.
-The latter was sequenced on two additional SMRTcells, and the former on a single SMRTcell.
+Because were were concerned that the multiplex PCR would still result in highly uneven amplification of different influenza cDNAs due to differences in their expression levels and lengths, the product of this multiplex PCR was subjected to eight additional individual emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 3' primer, using 1 ng of material in each reaction.
+The material from these eight segment-specific PCRs was then pooled with the goal of obtaining a equimolar ratio of segments, and sequenced on one SMRT Cell in a PacBio RS II and one SMRT Cell of a PacBio Sequel.
+Detailed results from the analysis of these first two sequencing runs is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and Supplementary~file~\ref{suppfile:pacbio_analysis}.
+These results showed that although the PCRs substantially enriched for influenza molecules, the relative coverage of the different viral genes was still very uneven, with the longer genes (especially the polymerase genes) severely under-sampled.
 
-As our depth on the three longest segments remained low, we further amplified our already-amplified material using a non-emulsion PCR using standard conditions for PB2, PB1, and PA, and subjecting the resulting material to a rigorous SPRI-bead selection using SPRIselect beads at a bead-PCR volume ratio of 0.4.
-This bead-selection was an attempt to correct for the high level of internally-deleted PB2, PB1, and PA segments dominating our sequencing, and enrich for the longer, full-length segments.
-All non-emulsion PCRs were recovered at sub ${\mu}$g yields to prevent the occurrence of artifacts at PCR saturation.
-We sequenced this bead-selected and repooled material on an additional SMRTcell.
+To try to improve coverage of the polymerase genes, we produced two new sequencing pools: one consisting of the five shortest viral segments (HA, NP, NA, M, and NS) from the aforementioned segment-specific emulsion PCRs, and the other consisting of the three longer polymerase segments (PB2, PB1, and PA).
+The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\citep{xue2016propagation} and are presumably preferentially amplified during PCR.
 
-Enrichment of full-length PB2, PB1, and PA was aided by bead-selection but these segments still exhibited relatively low PCR efficiency and comprised a small component of the remaining material.
-Finally, to improve recovery of full-length PB1, PB2, and PA and overcome poor PCR efficiency and over-representation of internally deleted sequences deriving from defective viruses, we undertook a technique allowing us to use two segment-specific primers rather than one, in a PCR that would exclude internally-deleted sequences.
-Namely, we sought to circularize our template molecules, allowing us to still amplify across the whole transcript, including the cell-specific barcode, while using two segment-specific primers in apposition that lie within a region absent in most internally-deleted segments.
-First, we first used our already-amplified segment-specific material in a low-cycle PCR to append circularization adapters, and circularized 10ng of this material using NEBuilder in a 20${\mu}$l reaction volume using an extended reaction time of 50 minutes.
-We then further incubated these reactions a  1 hour at 37$^{\circ}$C with exonuclease V to remove linear concatemers and uncircularized material, and bead-cleaned using SPRIselect beads at a volume ratio of 0.4.
-This circularized material was then used in three non-emulsion PCRs specific to PB2, PB1, and PA, using two segment-specific primers that align to the central portion of each gene but in apposition to one-another, and bead-cleaned using a ratio of 0.4 volumes.
-This results in successful amplification only if the central portion of the gene is present, while still amplifying across the entire transcript as well as cell-specific barcode regions. 
-This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRTcell in the PacBio Sequel.
+To obtain more reads for longer full-length polymerase variants, we therefore subjected 10 ng of our amplified material for each polymerase segment to a bead selection using SPRIselect beads at a volume ratio of 0.4 to select for larger molecules.
+This selection removes most low-molecular weight DNA species including internally-deleted defective segments.
+Material from this selection was then amplified using 16 (PB1) or 14 (PB2 and PA) cycles of a non-emulsion PCR using the standard conditions recommended by the Kapa HiFi Hotstart ReadyMix (extension at 67$^{\circ}$C for 2 minutes 15 seconds, and melting at 98$^{\circ}C$).
+The use of relatively few PCR cycles was designed to prevent the occurrence of the artifacts (including strand exchange) that occur in non-emulsion PCRs as the reactions approach saturation.
+We pooled the products of these reactions from this size-selection and sequenced on a SMRT Cell of a PacBio Sequel.
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), this sequencing yielded modestly more full-length polymerase variants, but they were still heavily undersampled compared to other viral genes.
+
+To further to improve recovery of full-length PB1, PB2, and PA, we therefore took an alternative approach that allowed us to perform a specific PCR for full-length polymerase variants.
+Specifically, we circularized the template molecules (including the full length of genes plus the cell barcode, UMI, and 10X adaptor), and then used two segment-specific primers that annealed in apposition near the center of each polymerase gene to linearize these circular molecules.
+Only molecules that contain the middle of the polymerase genes (which are typically full-length) are linearized by this process.
+In the downstream computational analysis, we can then determine the full sequence of the gene as well as the cell barcode of the initial molecule from which the linearized molecule is derived.
+Specifically, we first used 2.5 ng of our already-amplified segment-specific material in a 10-cycle PCR to append circularization adapters (see \FIGDATA[genotypes]{primer_sequences} for sequences), and cleaned the resultant mixture using SPRIselect beads at a volume ratio of 0.4.
+We then used 10 ng of this amplified material in a 20$\mu$l NEBuilder reaction using an extended reaction time of 50 minutes in order to circularize the molecules.
+We next incubated these reactions for 1 hour at 37$^{\circ}$C with exonuclease V and additional ATP to a final increase in concentration of 1 mM to digest all non-circularized molecules.
+The circularized and digested material was then cleaned using SPRIselect beads at a volume ratio of 0.4.
+This material was then used as template for three non-emulsion PCRs specific to PB2, PB1, or PA, using two segment-specific primers that align to the central portion of each gene but in apposition to each another (see \FIGDATA[genotypes]{primer_sequences} for sequences).
+These linearization reactions used 20 (PB2) or 26 (PB1 and PA) PCR cycles, and the resulting products were cleaned using SPRIselect beads at a volume ratio of 1.0.
+This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRT Cell of PacBio Sequel. 
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), this process efficiently yielded many full-length polymerase variants.
+
+The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[experiment]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[experiment]{StrandExchange} for an illustration of how this is determined).
+A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio run is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}).
+Importantly, the various PCR biases and enrichment schemes mean that the coverage of molecules by the PacBio sequencing is not proportional to their original abundance in the starting mRNA.
+However, as described in the computational analysis section below, the final analyses use the cell barcodes and UMIs in conjunction with the standard 10X Illumina sequencing to ensure that none of the conclusions are affected by the disproportionate amplification of some molecules during the PacBio library preparation (for instance, duplicate UMIs are removed from the PacBio data, and all conclusions about gene abundance or absence are based on the Illumina data).
 
@@
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at the indicated MOI.
-Viral infections were normalized to an MOI of a relatively-deletion free stock of virus that we previously characterized (ref), and identical particle numbers (calculated by qPCR, described below) of other viruses were added.
-Crucially, by controlling for particle number rather than infectious units we avoid the confounding problem of differing immune stimulation by simply infecting at a higher rate of biologically active but non-infectious particles.
-For experimental treatments with ribavirin and cycloheximde, these compounds were added to cells at the time of infection to a final concentration of 100$\mu$M and 50$\mu$g/ml, respectively, both of which are sufficient to block downstream viral replication (references).
+Viral infections were normalized to an MOI of a previously-characterized relatively-deletion free stock of virus~\citep{russell2018extreme}, with all other strains matched for viral genome copy number as calculated by qPCR, described below.
+Crucially, by controlling for genome copy number rather than infectious units we avoid the confounding problem of differing immune stimulation by simply infecting at a higher rate of biologically active but non-infectious particles.
+For experimental treatments with ribavirin and cycloheximde, these compounds were added to cells at the time of infection to a final concentration of 100$\mu$M and 50$\mu$g/ml, respectively, both of which are sufficient to block downstream viral replication~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg,killip2014activation}.
 At 8 hours post infection media was removed and cell lysates generated using the CellAmp RT-PCR for Direct Prep Kit (Real Time) \& Protein Analysis, following manufacturer's instructions. 
-cDNA was synthesized using an oligoDT primer and the SuperScript III first-strand synthesis supermix from ThermoFisher using the manufacturer’s protocol. 
-Transcript abundance was measured using SYBR green PCR master mix, using a combined anneal/extension step of 60 for one minute with the following primers: HA: 5’-GGCCCAACCACACATTCAAC-3’, 5’-GCTCATCACTGCTAGACGGG-3’, IFNB1: 5’-AAACTCATGAGCAGTCTGCA-3’, 5’-AGGAGATCTTCAGTTTCGGAGG-3’, L32: 5’-AGCTCCCAAAAATAGACGCAC-3’, 5’-TTCATAGCAGTAGGCACAAAGG-3’.
+cDNA was synthesized using an oligoDT primer and the SuperScript III first-strand synthesis supermix from ThermoFisher using the manufacturer's protocol. 
+Transcript abundance was measured using SYBR green PCR master mix, using a combined anneal/extension step of 60 for one minute with the following primers: HA: 5'-GGCCCAACCACACATTCAAC-3', 5'-GCTCATCACTGCTAGACGGG-3', IFNB1: 5'-AAACTCATGAGCAGTCTGCA-3', 5'-AGGAGATCTTCAGTTTCGGAGG-3', L32: 5'-AGCTCCCAAAAATAGACGCAC-3', 5'-TTCATAGCAGTAGGCACAAAGG-3'.
 
 For titering virus by qPCR, vRNA was harvested from 80 $\mu$l of viral supernatant by the addition of 600 $\mu$l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol. 
-The cDNA was generated using SuperScript III first-strand synthesis supermix using the manufacturer’s protocol, and using the universal vRNA primers of Hoffmann et al.
-(add Hoffman ref) with the modifications described in Xue et al.
-(add Katherine ref).
-The qPCR was then performed as for mRNA measurements.
+The cDNA was generated using SuperScript III first-strand synthesis supermix using the manufacturer's protocol, and using the universal vRNA primers of Hoffmann et al.
+(add Hoffman ref) with the modifications described in Xue et al.\citep{Xue:2017dla}.
+The qPCR was then performed as for mRNA measurements using the HA qPCR primers to measure genome dose.
 

### 2018-09-13 paper/paper.tex
Jesse revisions to qPCR methods

@@
 \subsection{qPCR for viral genes and IFN}
+We performed qPCR for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
+For the qPCR, we used the SYBR Green PCR Master Mix (Thermo Fisher) according to the manufacturer's protocol.
+The qPCR primers were: HA primer 1, 5'-\texttt{GGCCCAACCACACATTCAAC}-3'; HA primer 2, 5'-\texttt{GCTCATCACTGCTAGACGGG}-3'; IFNB1 primer 1, 5'-\texttt{AAACTCATGAGCAGTCTGCA}-3'; IFNB1 primer 2, 5'-\texttt{AGGAGATCTTCAGTTTCGGAGG}-3'; L32 primer 1, 5'-\texttt{AGCTCCCAAAAATAGACGCAC}-3'; L32 primer 2, 5'-\texttt{TTCATAGCAGTAGGCACAAAGG}-3'.
+
+For titering virus by qPCR, vRNA was harvested from 80 $\mu$l of viral supernatant by the addition of 600 $\mu$l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini Kit protocol. 
+The cDNA was generated using SuperScript III First-Strand Synthesis SuperMix (Thermo Fisher) using the manufacturer's protocol, and using the universal vRNA primers of \citet{hoffmann2001universal} with the modifications described in \citet{xue2017parallel}.
+The number of HA molecules were then quantified by qPCR to measure genome dose.
+
 \abrcomment{Assuming we use the cycloheximide and ribavirin data as-is, can be modified if not}
@@
 cDNA was synthesized using an oligoDT primer and the SuperScript III first-strand synthesis supermix from ThermoFisher using the manufacturer's protocol.
-Transcript abundance was measured using SYBR green PCR master mix, using a combined anneal/extension step of 60 for one minute with the following primers: HA: 5'-GGCCCAACCACACATTCAAC-3', 5'-GCTCATCACTGCTAGACGGG-3', IFNB1: 5'-AAACTCATGAGCAGTCTGCA-3', 5'-AGGAGATCTTCAGTTTCGGAGG-3', L32: 5'-AGCTCCCAAAAATAGACGCAC-3', 5'-TTCATAGCAGTAGGCACAAAGG-3'.
-
-For titering virus by qPCR, vRNA was harvested from 80 $\mu$l of viral supernatant by the addition of 600 $\mu$l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol. 
-The cDNA was generated using SuperScript III first-strand synthesis supermix using the manufacturer's protocol, and using the universal vRNA primers of Hoffmann et al.
-(add Hoffman ref) with the modifications described in Xue et al.\citep{Xue:2017dla}.
-The qPCR was then performed as for mRNA measurements using the HA qPCR primers to measure genome dose.
-
-\jdbcomment{Alistair writes this}
 

### 2018-09-13 paper/paper.tex
minor tweaks to abstract

@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous at the cellular level, and many viruses trigger innate immunity in only a fraction of infected cells. 
-Here we develop an approach to determine how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We simultaneously determine the cellular transcriptome and full sequences of all viral genes in hundreds of influenza-infected cells.
-Infections that trigger immunity are associated with several features: absence of the gene encoding the virus's primary immune antagonist, internal deletions in viral polymerase genes, and point mutations in viral proteins involved in replication and nuclear export.
-However, immune activation remains stochastic in cells infected by viruses with these genetic lesions, and sometimes occurs even in cells infected with fully wildtype virions.
-Overall, our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation.
+The outcome of viral infection is extremely heterogeneous at the cellular level: infected cells vary widely in the amount of viral material they produce, and only sometimes activate innate immunity.  
+Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
+We determine the cellular transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Infections that trigger immunity at the single-cell level are associated with defects in the infecting viruses that include failure to express the primary immune antagonist, internal deletions in polymerase genes, and point mutations in proteins involved in replication and nuclear export.
+However, immune activation remains stochastic in cells infected by viruses with these genetic defects, and sometimes occurs even in cells infected with wildtype virus.
+Our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation.
 \end{abstract}

### 2018-09-14 paper/paper.tex
initial draft of section 1 of Results (_A system to identify and enrich for IFN+ positive cells_)

@@
 We determine the cellular transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that trigger immunity at the single-cell level are associated with defects in the infecting viruses that include failure to express the primary immune antagonist, internal deletions in polymerase genes, and point mutations in proteins involved in replication and nuclear export.
+Infections that trigger immunity at the single-cell level are associated with defects in the infecting viruses that include failure to express the primary immune antagonist, internal deletions in polymerase genes, and mutations in proteins involved in replication and nuclear export.
 However, immune activation remains stochastic in cells infected by viruses with these genetic defects, and sometimes occurs even in cells infected with wildtype virus.
-Our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation.
+Our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation in single cells.
 \end{abstract}
@@
 
+The antiviral IFNs~\citep[type I and type III IFNs;][]{kotenko2017contribution}
+Early in infection
+
 \clearpage
@@
 
-\subsection{A method to obtain the complete genotypes of the virions infecting single cells that do and do not express IFN}
+\subsection{A system to identify and enrich for IFN+ positive cells}
+A challenge in studying the induction of IFN by influenza is that infection with this virus only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
+For instance, we previously performed single-cell mRNA sequencing of hundreds of influenza-infected A549 cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
+Only one of these cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
+Importantly, the paucity of IFN+ cells identified in our prior single-cell mRNA sequencing was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
 
-Two challenges:
-
-IFN expression is rare in infected cells (\FIG{IFNrare}).
-Apparent in single-cell \textit{in vitro}~\citep{russell2018extreme}, single-cell \textit{in vivo}~\citep{steuerman2018dissection}, and reporter assays by others~\citep{killip2017single} and ourselves (\FIG{IFNrare}).
+The rareness of IFN induction by influenza infection is not merely a feature of A549 cells in culture.
+Prior studies using fluorescent reporters have found that IFN expression is very rare at the single-cell level in influenza-infected mice at 24 hours post-infection~\citep{kallfass2013visualizing}.
+In further support of these findings, we re-analyzed the single-cell mRNA sequencing of influenza-infected mice by \citet{steuerman2018dissection}.
+As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed interferon-stimulated genes, which are induced autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
+But only a handful of infected cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
 
@@
 \caption{
-Triggering of IFN production is rare in influenza-infected cells.
-{\bf (A)} Number of influenza-infected and IFN+ cells in our prior single-cell study of influenza infection of A549 cells.
-The Venn diagram shows data aggregated over all samples from \citet{russell2018extreme}.
-\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell mRNA-sequencing study of influenza-infected mice by \citet{steuerman2018dissection}.
-{\bf (B)} Creation of an A549 reporter cell line to identify single cells expressing IFN.
-The cells contain an IFN promoter that drives expression of a cell-surface protein amenable to MACS (magnetic-activated cell sorting) and a fluorescent protein.
+IFN expression is rare in influenza-infected cells.
+{\bf (A)} Number of influenza-infected and IFN+ cells in our prior single-cell study of influenza infection of A549 cells~\citep{russell2018extreme}.
+\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell mRNA-sequencing of influenza-infected mice by \citet{steuerman2018dissection}.
+{\bf (B)} An A549 reporter cell line to identify single cells expressing IFN.
+The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
 We created reporter cell lines with type I (IFN-$\beta$) and type III (IFN-$\lambda$) promoters (\FIGDATA[IFNrare]{reporter_sequences}).
-The reporters were efficiently activated when infected with a strain of Sendai virus~\citep{strahle2006sendai} known to potently induce IFN (\FIGSUPP[IFNrare]{reporter_validation}).
-Expression of type I and type III IFN is highly correlated in the reporter cell lines upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
+The reporters were efficiently activated when infected with an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
+Activation of the type I and type III IFN reporters is highly correlated in single cells upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
 {\bf (C)}
-Frequency of IFN induction upon infection with the influenza virus stocks used in the single-cell studies in this paper quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
-The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus as a positive control.
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
+The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
 The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
@@
 
-\figsupp[Few cells express detectable IFN in the single-cell analysis of influenza-infected mice by \citet{steuerman2018dissection}.]
+\figsupp[Few cells express detectable IFN in the single-cell mRNA-sequencing of influenza-infected mice by \citet{steuerman2018dissection}.]
 {A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell mRNA-sequencing of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
@@
 
-\figsupp[Expression of type I and type III IFN are highly correlated in influenza-infected A549 cells.]
+\figsupp[Expression of type I and type III IFN is highly correlated in A549 cells.]
 {A549 cells were dually transduced with the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
@@
 
-\figdata{Sequences of the reporters schematized in \FIG{IFNrare}B.
-The MACS sortable surface marker is a truncated LNGFR.
-For the IFN-$\beta$ reporter, the fluorescent protein is mNeonGreen.
-For the IFN-$\lambda$ (IL29) reporter, the fluorescent protein is ZsGreen.
-The sequences of plasmids carrying these reporters are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1517_pHAGE2_IFNbeta_prom_LNGFR_P2A_mNeongreen.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1767_pHAGE2_IL29_promnkozak_LNGFR_P2A_zsGreen.gb}.
+\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1517_pHAGE2_IFNbeta_prom_LNGFR_P2A_mNeongreen.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1767_pHAGE2_IL29_promnkozak_LNGFR_P2A_zsGreen.gb}.
 }
@@
 
+In order to efficiently identify and enrich cells that express IFN, we integrated IFN reporters into A549 cells (\FIG{IFNrare}B).
+These reporters consisted of a type I (IFNB1) or type III (IFNL1) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
+Cells that activate IFN will express the cell-surface protein, enabling them to be enriched by magnetic-activated cell sorting (MACS).
+The frequency of IFN+ cells can also be determined by flow cytometry for the fluorescent protein.
+We validated that these A549 reporter cell lines were efficiently activated by infection with a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
+We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result that is further validated by the single-cell mRNA sequencing described below.
+Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.
+
+We then used one of the reporter cell lines to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza.
+As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
+This frequency of IFN induction was roughly comparable to what we had observed in our prior single-cell mRNA sequencing of influenza-infected cells (\FIG{IFNrare}B), with the slightly higher rate in \FIG{IFNrare}C probably due to using a different viral stock and slightly later timepoint.
+
+\subsection{new section}
 All past single-cell transcriptomics of viral infected cells~\citep{russell2018extreme,zanini2018single,steuerman2018dissection,zanini2018virus} have focused on counting the number of viral transcripts rather than determining their full sequences.

### 2018-09-14 paper/paper.tex
slight tweaks to workflow schematic

@@
 
-\hspace{0.04in} {\bf A} \includegraphics[width=0.9\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
+\hspace{0.04in} {\bf A} \includegraphics[width=0.95\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
 \vspace{0.1in}
@@
 \caption{
-Single-cell virus sequencing of influenza infections that induce IFN.
+Single-cell sequencing of influenza infections that induce IFN.
 {\bf (A)} Experimental workflow.
-(1) IFN reporter A549 cells are infected with a mix of wildtype virus and virus carrying synonymous mutations at both ends of each gene segment.
+(1) IFN reporter A549 cells are infected with a mix of wildtype and synonymously barcoded viruses.
 IFN+ cells are enriched by MACS (\FIGSUPP[experiment]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
@@
 {\bf (B)} Number of cells for which transcriptomes were obtained.
-Most transcriptomes are from human cells, but a few are from the uninfected canine cells spiked into the experiment as a control.
-From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was captured in an emulsion.
-These cross-celltype multiplets are excluded from subsequent analyses.
+From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion.
 {\bf (C)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
-{\bf (H)} Fraction of cellular mRNA derived from type I or type III IFN across all cells, faceted by whether the cells are infected.
+{\bf (H)} Fraction of cellular mRNA derived from IFN across all cells, faceted by whether the cells are infected.
 Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
-We grouped type I and type III IFN together because their expression is highly correlated (\FIGSUPP[experiment]{type_I_III_correlation}).
-Many cells that do not express IFN itself still express interferon-stimulated genes (\FIGSUPP[experiment]{ISG}).
+Many cells that do not express IFN itself still express ISGs (\FIGSUPP[experiment]{ISG}).
 }

### 2018-09-15 paper/paper.tex
Split experiment figure into workflow and transcriptomics

@@
 
-\subsection{A system to identify and enrich for IFN+ positive cells}
+\subsection{A system to identify and enrich for rare IFN+ positive cells}
 A challenge in studying the induction of IFN by influenza is that infection with this virus only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
-For instance, we previously performed single-cell mRNA sequencing of hundreds of influenza-infected A549 cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
-Only one of these cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
-Importantly, the paucity of IFN+ cells identified in our prior single-cell mRNA sequencing was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
+For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
+Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
+Importantly, the paucity of IFN+ cells identified in our prior single-cell transcriptomics was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
 
@@
 Prior studies using fluorescent reporters have found that IFN expression is very rare at the single-cell level in influenza-infected mice at 24 hours post-infection~\citep{kallfass2013visualizing}.
-In further support of these findings, we re-analyzed the single-cell mRNA sequencing of influenza-infected mice by \citet{steuerman2018dissection}.
-As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed interferon-stimulated genes, which are induced autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
+In further support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
+As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes, which are induced autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
 But only a handful of infected cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
@@
 IFN expression is rare in influenza-infected cells.
-{\bf (A)} Number of influenza-infected and IFN+ cells in our prior single-cell study of influenza infection of A549 cells~\citep{russell2018extreme}.
-\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell mRNA-sequencing of influenza-infected mice by \citet{steuerman2018dissection}.
+{\bf (A)} Number of infected and IFN+ cells in our prior single-cell transcriptomics of influenza-infected A549 cells~\citep{russell2018extreme}.
+\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
 {\bf (B)} An A549 reporter cell line to identify single cells expressing IFN.
@@
 
-\figsupp[Few cells express detectable IFN in the single-cell mRNA-sequencing of influenza-infected mice by \citet{steuerman2018dissection}.]
-{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell mRNA-sequencing of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
+\figsupp[Few cells express detectable IFN in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.]
+{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
 Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
@@
 
-In order to efficiently identify and enrich cells that express IFN, we integrated IFN reporters into A549 cells (\FIG{IFNrare}B).
-These reporters consisted of a type I (IFNB1) or type III (IFNL1) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
+In order to efficiently identify and enrich cells that express IFN, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B).
+These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
 Cells that activate IFN will express the cell-surface protein, enabling them to be enriched by magnetic-activated cell sorting (MACS).
 The frequency of IFN+ cells can also be determined by flow cytometry for the fluorescent protein.
-We validated that these A549 reporter cell lines were efficiently activated by infection with a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
-We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result that is further validated by the single-cell mRNA sequencing described below.
+We validated that the IFN reporters were efficiently activated by infection with a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
+We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result that is further validated by our single-cell transcriptomics described below.
 Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.
 
-We then used one of the reporter cell lines to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza.
+We then used the reporter cells to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
 As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
-This frequency of IFN induction was roughly comparable to what we had observed in our prior single-cell mRNA sequencing of influenza-infected cells (\FIG{IFNrare}B), with the slightly higher rate in \FIG{IFNrare}C probably due to using a different viral stock and slightly later timepoint.
+This frequency of IFN induction was roughly comparable to what we had observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B), with the slightly higher rate in \FIG{IFNrare}C probably due to using a different viral stock and slightly later timepoint.
 
-\subsection{new section}
-All past single-cell transcriptomics of viral infected cells~\citep{russell2018extreme,zanini2018single,steuerman2018dissection,zanini2018virus} have focused on counting the number of viral transcripts rather than determining their full sequences.
-We came up with a scheme to determine full genotypes of infecting viruses in IFN+ cells (\FIG{experiment}).
+\subsection{Simultaneous transcriptomics and virus-sequencing of single infected cells}
+We next set out to determine whether features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
+Standard single-cell transcriptomic approaches are inadequate to address this question, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
+But simply measuring the abundance of viral transcripts is insufficient for our purposes: we want to know the full sequences of all viral genes in each infected cell, so that we can determine if features of these genes (e.g., mutations) contribute to heterogeneity in infection outcome.
 
-%%% start approach figure
+%%% start workflow figure
 \begin{figure}
@@
 
-\hspace{0.04in} {\bf A} \includegraphics[width=0.95\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
-\vspace{0.1in}
-
-\includegraphics[width=\linewidth, clip=false]{figures/single_cell_figures/p_cell_summary.pdf}
+\includegraphics[width=\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
 
 \caption{
-Single-cell sequencing of influenza infections that induce IFN.
-{\bf (A)} Experimental workflow.
-(1) IFN reporter A549 cells are infected with a mix of wildtype and synonymously barcoded viruses.
-IFN+ cells are enriched by MACS (\FIGSUPP[experiment]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
-(2) The mRNAs from individual cells are tagged with cell-specific barcodes using a 10X Chromium device.
-(3) The entire cellular transcriptomes are quantified using the standard single-cell mRNA-seq protocol, and viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
-(4) The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
-{\bf (B)} Number of cells for which transcriptomes were obtained.
-From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion.
-{\bf (C)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
-Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
-Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (possibly low-quality emulsions or multiplets), and are excluded from subsequent analyses.
-{\bf (D)} Distribution across cells of the fraction of all mRNA derived from influenza.
-Cells called as infected are in blue, while other cells are in green.
-The inset shows the amount of viral mRNA among the human cells called as infected.
-{\bf (E)} The number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
-The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
-\FIGSUPP[experiment]{frac_has_gene} shows the frequency that each viral gene is detected in infected cells.
-{\bf (F)} Relative expression of viral genes among infected cells.
-{\bf (G)} The number of cells infected with wildtype virus, synonymously barcoded virus, or both.
-From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
-{\bf (H)} Fraction of cellular mRNA derived from IFN across all cells, faceted by whether the cells are infected.
-Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
-Many cells that do not express IFN itself still express ISGs (\FIGSUPP[experiment]{ISG}).
+Approach for simultaneous transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
+{\bf (A)}
+IFN reporter A549 cells are infected with a mix of wildtype and synonymously barcoded viruses.
+IFN+ cells are enriched by MACS (\FIGSUPP[workflow]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
+{\bf (B)}
+The mRNAs from individual cells are tagged with cell-specific barcodes using a 10X Chromium device.
+{\bf (C)}
+The entire cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
+{\bf (D)}
+viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
+{\bf (E)}
+The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
 }
-\label{fig:experiment}
+\label{fig:workflow}
 
-\figsupp[MACS enrichment of IFN+ cells]
+\figsupp[MACS enrichment of IFN+ cells.]
 {
 Example MACS enrichments of IFN+ influenza-infected cells.
-The A549 cells with the IFN-$\beta$ LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype A/WSN/1933 influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
+The A549 cells with the IFN-$\beta$ LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
 After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
@@
 \label{figsupp:MACS}
+
+\figdata{Genbank files giving the sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
+\label{figdata:virus_seqs}
+
+\end{fullwidth}
+\end{figure}
+%%% end workflow figure
+
+We developed the experimental approach in \FIG{workflow} to simultaneously determine the entire transcriptome and the full sequences of the infecting virions in single cells.
+First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain: the wildtype virus and a ``synonymously barcoded'' variant that contains two engineered synonymous mutations near each end of each gene (\FIGDATA[workflow]{virus_seqs}).
+We have previously shown that synonymously barcoding viral genes near the 3' end of the mRNA enables quantification of the co-infection rate in single-cell transcriptomic data, since about half the cells infected by multiple virions express a mix of wildtype and synonymously barcoded viral transcripts~\citep{russell2018extreme}.
+New to this study, we also added synonymous barcodes near the 5' end of the mRNA; as explained below, having barcodes at both ends provides an important control for PCR artifacts during full-length sequencing of viral transcripts.
+
+We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{transcriptomics}A), using a viral dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
+From 12 to 13 hours post-infection, we used magnetic columns to enrich for cells that expressed the cell-surface protein driven by the IFN reporter.
+Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to about 5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}; the MACS enrichment efficiency is imperfect so many sorted cells are still IFN-).
+In order to ensure the presence of IFN- cell, we also added back non-sorted cells to $\sim$10\% of the total.
+We also added uninfected canine cells to $\sim$5\% of the total to serve as an internal control for multiplets and to estimate the background amount of viral mRNA observed in the transcriptomes of truly uninfected cells.
+
+%%% start transcriptomics figure
+\begin{figure}
+\begin{fullwidth}
+
+\includegraphics[width=\linewidth, clip=false]{figures/single_cell_figures/p_cell_summary.pdf}
+
+\caption{
+Single-cell transcriptomics of influenza-infected cells enriched for IFN positivity.
+{\bf (A)} 
+Number of cells for which transcriptomes were obtained.
+From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion.
+{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
+Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
+Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (possibly low-quality emulsions or multiplets), and are excluded from subsequent analyses.
+{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
+Cells called as infected are in blue, while other cells are in green.
+The inset shows the amount of viral mRNA among the human cells called as infected.
+{\bf (D)} The number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
+The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
+\FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected in infected cells.
+{\bf (E)} Relative expression of viral genes among infected cells.
+{\bf (F)} The number of cells infected with wildtype virus, synonymously barcoded virus, or both.
+From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
+{\bf (G)} Fraction of cellular mRNA derived from IFN across all cells, faceted by whether the cells are infected.
+Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
+Many cells that do not express IFN itself still express ISGs (\FIGSUPP[transcriptomics]{ISG}).
+}
+\label{fig:transcriptomics}
 
@@
 We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
-Comparison to \FIG{experiment}H shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
+Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
 This is probably because paracrine signaling leads to ISG expression in some cells that are not themselves expressing IFN.
@@
 
-\figdata{Genbank files giving the sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
-\label{figdata:virus_seqs}
-
 \end{fullwidth}
 \end{figure}
-%%% end approach figure
+%%% end transcriptomics figure
 
@@
 Green boxes to the left show the percent of all mRNA in that cell derived from virus.
-Orange boxes show the percent of cellular mRNA derived from IFN, with boxes framed in orange indicating cells classified as IFN+ in \FIG{experiment}H.
+Orange boxes show the percent of cellular mRNA derived from IFN, with boxes framed in orange indicating cells classified as IFN+ in \FIG{transcriptomics}G.
 }
@@
 Note that these sequences were obtained using several PacBio runs, some of which were loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
-Therefore, unlike the 10X transcript count data in \FIG{experiment}H, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
+Therefore, unlike the 10X transcript count data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
 Especially for the polymerase genes (PB2, PB1, and PA), many of the CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were presumably preferentially amplified during PCR.
@@
 An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
-Note also that this plot is limited to the cells that were called as infected (\FIG{experiment}D) and could clearly be classified as IFN- or IFN+ (\FIG{experiment}H).
+Note also that this plot is limited to the cells that were called as infected (\FIG{transcriptomics}C) and could clearly be classified as IFN- or IFN+ (\FIG{transcriptomics}G).
 }
@@
 {This plot shows the same information as panels \FIG{mutations}B-D except that it shows the association of viral gene absence / mutations with ISG expression rather than IFN expression.
-Cells are classified as ISG+ or ISG- as in \FIGSUPP[experiment]{ISG}.
+Cells are classified as ISG+ or ISG- as in \FIGSUPP[transcriptomics]{ISG}.
 The major viral features that associate with IFN induction also associate with ISG expression.
@@
 In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
-The sequences of all gene segments from the wildtype and synonymously barcoded viral strains are in \FIGDATA[experiment]{virus_seqs}.
+The sequences of all gene segments from the wildtype and synonymously barcoded viral strains are in \FIGDATA[workflow]{virus_seqs}.
 These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
@@
 The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the LNGFR$\Delta$C-P2A-mNeonGreen reporter.
-A schematic of the experiment is shown in \FIG{experiment}A.
+A schematic of the experiment is shown in \FIG{workflow}.
 
 The wildtype and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
-The cells were then infected with this mixture at a dose designed to infect about half the cells (\FIG{experiment}D suggests that the actual rate of detectable infection was slightly lower).
+The cells were then infected with this mixture at a dose designed to infect about half the cells (\FIG{transcriptomics}C suggests that the actual rate of detectable infection was slightly lower).
 Infections were allowed to proceed for 12 hours.
@@
 To enrich IFN+ cells, the cells were then incubated with anti-LNGFR MACSelect Microbeads (Miltenyi Biotec) and twice passed over an MS magnetic column (Miltenyi Biotec), retaining the bound (and presumably IFN-enriched) population each time.
-This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \FIGSUPP[experiment]{MACS}.
-The original, unsorted, population was then added back in to approximately 10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
-At this point, uninfected canine (MDCK-SIAT1) cells were also added to approximately 10\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{experiment}B) and background viral mRNA in uninfected cells (\FIG{experiment}D).
+This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \FIGSUPP[workflow]{MACS}.
+The original, unsorted, population was then added back in to $\sim$10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
+At this point, uninfected canine (MDCK-SIAT1) cells were also added to $\sim$5\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{transcriptomics}A) and background viral mRNA in uninfected cells (\FIG{transcriptomics}C).
 We begin this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
@@
 This sample was then processed to create libraries for Illumina 3'-end sequencing according to the 10X Genomics protocol using the Chromium Single Cell 3' Library and Gel Bead Kit v2 with one important modification: rather than process all full-length cDNA through enzymatic fragmentation, several nanograms were retained for targeted full-length viral cDNA sequencing as described below.
-The single-cell mRNA-seq library was sequenced on an Illumina HiSeq 2500, and the data analyzed as described below.
+The single-cell transcriptomics library was sequenced on an Illumina HiSeq 2500, and the data analyzed as described below.
 
@@
 We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
-All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{experiment}A for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} for at detailed schematic).
+All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} for at detailed schematic).
 However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
@@
 The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
-The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[experiment]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[experiment]{StrandExchange} for an illustration of how this is determined).
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[genotypes]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[genotypes]{StrandExchange} for an illustration of how this is determined).
 A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio run is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}).
@@
 
-Wildtype and synonymously barcoded viruses in \FIGDATA[experiment]{virus_seqs} obtained from reverse-genetic transfections was subsequently propagated at an MOI of 0.001 for 72h twice in MDCK-SIAT1 cells (wild-type) or once at an MOI of 0.01 (barcoded) for 60h in order to generate potential immunostimulatory diversity.
+Wildtype and synonymously barcoded viruses in \FIGDATA[workflow]{virus_seqs} obtained from reverse-genetic transfections was subsequently propagated at an MOI of 0.001 for 72h twice in MDCK-SIAT1 cells (wild-type) or once at an MOI of 0.01 (barcoded) for 60h in order to generate potential immunostimulatory diversity.
 For analyses of interferon induction, an additional wild-type stock of virus was grown at an MOI of 0.01 for 36h in order to generate a population relatively free of defective particles.
@@
 The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
-The influenza virus transcriptome consisted of the mRNAs for the wildtype A/WSN/1933 virus strain in \FIGDATA[experiment]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wildtype and synonymously barcoded viral variants to this transcriptome).
+The influenza virus transcriptome consisted of the mRNAs for the wildtype A/WSN/1933 virus strain in \FIGDATA[workflow]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wildtype and synonymously barcoded viral variants to this transcriptome).
 The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
-The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{experiment}B), and generate a matrix giving the expression of each gene in each single cell.
+The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{transcriptomics}A), and generate a matrix giving the expression of each gene in each single cell.
 We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wildtype or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.
@@
 We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
-Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{experiment}C.
-We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{experiment}D).
-We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{experiment}E and \FIGSUPP[experiment]{frac_has_gene}).
-To identify cells that were co-infected with both viral barcodes (\FIG{experiment}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived form the more common viral barcode.
-We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{experiment}H and \FIGSUPP[experiment]{ISG}, respectively.
+Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{transcriptomics}B.
+We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
+We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \FIGSUPP[transcriptomics]{frac_has_gene}).
+To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived form the more common viral barcode.
+We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}, respectively.
 We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.

### 2018-09-15 paper/paper.tex
partial text for section 2 of results (workflow)

@@
 \subsection{Simultaneous transcriptomics and virus-sequencing of single infected cells}
-We next set out to determine whether features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
-Standard single-cell transcriptomic approaches are inadequate to address this question, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
-But simply measuring the abundance of viral transcripts is insufficient for our purposes: we want to know the full sequences of all viral genes in each infected cell, so that we can determine if features of these genes (e.g., mutations) contribute to heterogeneity in infection outcome.
+We next set out to develop a method to determine whether features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
+Standard single-cell transcriptomic approaches are inadequate to for this purpose, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
+These transcriptomic measurements of abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
 
@@
 The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
-As indicated by the percentages shown for the original and MACS-sorted population, this process led to between 50- and 350-fold enrichment for IFN+ cells.
+As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment for IFN+ cells.
 We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
@@
 
-We developed the experimental approach in \FIG{workflow} to simultaneously determine the entire transcriptome and the full sequences of the infecting virions in single cells.
-First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain: the wildtype virus and a ``synonymously barcoded'' variant that contains two engineered synonymous mutations near each end of each gene (\FIGDATA[workflow]{virus_seqs}).
-We have previously shown that synonymously barcoding viral genes near the 3' end of the mRNA enables quantification of the co-infection rate in single-cell transcriptomic data, since about half the cells infected by multiple virions express a mix of wildtype and synonymously barcoded viral transcripts~\citep{russell2018extreme}.
-New to this study, we also added synonymous barcodes near the 5' end of the mRNA; as explained below, having barcodes at both ends provides an important control for PCR artifacts during full-length sequencing of viral transcripts.
+We therefore developed the experimental approach in \FIG{workflow} to simultaneously determine the entire transcriptome and the full sequences of the infecting virions in single cells.
+First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain of influenza: the wildtype virus and a ``synonymously barcoded'' variant that contains two engineered synonymous mutations near each end of each gene (\FIGDATA[workflow]{virus_seqs}).
+We have previously used synonymous barcoding near the 3' end of viral mRNAs to quantify the co-infection rate in single-cell transcriptomic data~\citep[about half of cells infected by multiple virions express a mix of wildtype and synonymously barcoded viral transcripts;][]{russell2018extreme}.
+New to this study, we also added synonymous barcodes near the 5' end of the mRNAs; as explained below, having viral barcodes at both ends provides an important control for PCR artifacts during full-length sequencing of viral transcripts.
 
-We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{transcriptomics}A), using a viral dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
+We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{transcriptomics}A) at a dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
 From 12 to 13 hours post-infection, we used magnetic columns to enrich for cells that expressed the cell-surface protein driven by the IFN reporter.
-Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to about 5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}; the MACS enrichment efficiency is imperfect so many sorted cells are still IFN-).
-In order to ensure the presence of IFN- cell, we also added back non-sorted cells to $\sim$10\% of the total.
-We also added uninfected canine cells to $\sim$5\% of the total to serve as an internal control for multiplets and to estimate the background amount of viral mRNA observed in the transcriptomes of truly uninfected cells.
+Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to $\sim$5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}).
+In order to ensure the presence of IFN- cells, after the MACS enrichment we added back non-sorted cells to $\sim$10\% of the total.
+We also added uninfected canine cells to $\sim$5\% of the total to serve as an internal control for multiplets and enable estimation of the background amount of viral mRNA detected in the transcriptomes of truly uninfected cells.
+
+We then processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and performs reverse-transcription of poly-adenylated mRNAs that appends a unique cell barcode to all cDNAs originating from a particular droplet, and a unique molecular identifier (UMI) to each cDNA molecules (\FIG{workflow}B).
+Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
+Furthermore, because virtually all the influenza genome is transcribed into mRNA, the cell-barcoded transcripts cover nearly the entire viral genome: the only portions that are not captured in the cell-barcoded transcripts are one highly conserved nucleotide upstream of the transcription start site at nucleotide 2~\citep{koppstein2015sequencing}
 

### 2018-09-15 paper/paper.tex
complete draft of second Results section: _Simultaneous transcriptomics and virus-sequencing of single infected cells_

@@
 We next set out to develop a method to determine whether features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
-Standard single-cell transcriptomic approaches are inadequate to for this purpose, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
+Standard single-cell transcriptomic approaches are inadequate for this purpose, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
 These transcriptomic measurements of abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
@@
 
-We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{transcriptomics}A) at a dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
+We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{workflow}A) at a dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
 From 12 to 13 hours post-infection, we used magnetic columns to enrich for cells that expressed the cell-surface protein driven by the IFN reporter.
@@
 
-We then processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and performs reverse-transcription of poly-adenylated mRNAs that appends a unique cell barcode to all cDNAs originating from a particular droplet, and a unique molecular identifier (UMI) to each cDNA molecules (\FIG{workflow}B).
+We then processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and then reverse transcribes polyadenylated mRNAs in a way that appends a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
 Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
-Furthermore, because virtually all the influenza genome is transcribed into mRNA, the cell-barcoded transcripts cover nearly the entire viral genome: the only portions that are not captured in the cell-barcoded transcripts are one highly conserved nucleotide upstream of the transcription start site at nucleotide 2~\citep{koppstein2015sequencing}
+Furthermore, because virtually all the influenza genome is transcribed into mRNA, the cell-barcoded cDNA spans nearly the entire 13,581-nucleotide segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.
+
+We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
+But we also took a portion of the cell-barcoded cDNA and enriched for full-length viral molecules (\FIG{workflow}D) using PCR as detailed below.
+We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
+These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique copies of each viral gene in each cell.
+Because most cells are infected by just one or two unique virions, we can therefore build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of the infecting virions.
+Combining this information with the 3'-end sequencing simultaneously determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).
 
@@
 We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \FIGSUPP[transcriptomics]{frac_has_gene}).
-To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived form the more common viral barcode.
+To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
 We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}, respectively.

### 2018-09-15 paper/paper.tex
minor tweaks to first two sections of Results

@@
 \subsection{A system to identify and enrich for rare IFN+ positive cells}
-A challenge in studying the induction of IFN by influenza is that infection with this virus only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
+A challenge in studying the induction of IFN by influenza is that infection only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
 For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
 Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
-Importantly, the paucity of IFN+ cells identified in our prior single-cell transcriptomics was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
+Importantly, the paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
 
 The rareness of IFN induction by influenza infection is not merely a feature of A549 cells in culture.
-Prior studies using fluorescent reporters have found that IFN expression is very rare at the single-cell level in influenza-infected mice at 24 hours post-infection~\citep{kallfass2013visualizing}.
+Prior studies using fluorescent reporters have found that IFN expression is rare at the single-cell level in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
 In further support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
-As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes, which are induced autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
+As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes (ISGs), which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
 But only a handful of infected cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
@@
 IFN expression is rare in influenza-infected cells.
-{\bf (A)} Number of infected and IFN+ cells in our prior single-cell transcriptomics of influenza-infected A549 cells~\citep{russell2018extreme}.
+{\bf (A)} Number of infected and IFN+ A549 cells in our prior single-cell transcriptomics~\citep{russell2018extreme}.
 \FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
-{\bf (B)} An A549 reporter cell line to identify single cells expressing IFN.
+{\bf (B)} An A549 reporter cell line to identify cells expressing IFN.
 The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
-We created reporter cell lines with type I (IFN-$\beta$) and type III (IFN-$\lambda$) promoters (\FIGDATA[IFNrare]{reporter_sequences}).
-The reporters were efficiently activated when infected with an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
+We created reporters with type I and type III promoters (\FIGDATA[IFNrare]{reporter_sequences}).
+The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
 Activation of the type I and type III IFN reporters is highly correlated in single cells upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
@@
 
-\figsupp[Few cells express detectable IFN in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.]
+\figsupp[Few cells express detectable IFN in influenza-infected mice.]
 {A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
-Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
+Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
 Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
@@
 An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key interferon-stimulated gene (ISG) mRNAs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one mRNA transcript detected).
+Nonetheless, there is detectable expression of key ISGs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one transcript detected).
 The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
@@
 {To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
-At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
-Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal for the IFN-$\lambda$ reporter driving ZsGreen.
+At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
+Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
 }
@@
 {Flow cytometry data for \FIG{IFNrare}C.
-The A549 cells with the IFN-$\lambda$ reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
+The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
 After 13 hours, the cells were stained for expression of HA protein and analyzed for the IFN-$\lambda$ reporter driven ZsGreen expression by FACS.
@@
 
-In order to efficiently identify and enrich cells that express IFN, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B).
+To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B).
 These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
-Cells that activate IFN will express the cell-surface protein, enabling them to be enriched by magnetic-activated cell sorting (MACS).
+Cells that activate IFN express the cell-surface protein, enabling them to be enriched by magnetic-activated cell sorting (MACS).
 The frequency of IFN+ cells can also be determined by flow cytometry for the fluorescent protein.
-We validated that the IFN reporters were efficiently activated by infection with a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
-We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result that is further validated by our single-cell transcriptomics described below.
+We validated that the reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
+We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result further validated by our single-cell transcriptomics described below.
 Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.
 
-We then used the reporter cells to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
+We used the reporter cells to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
 As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
-This frequency of IFN induction was roughly comparable to what we had observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B), with the slightly higher rate in \FIG{IFNrare}C probably due to using a different viral stock and slightly later timepoint.
+This frequency of IFN induction is roughly comparable to what we had observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B).
 
 \subsection{Simultaneous transcriptomics and virus-sequencing of single infected cells}
-We next set out to develop a method to determine whether features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
+We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
 Standard single-cell transcriptomic approaches are inadequate for this purpose, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
-These transcriptomic measurements of abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
+These measurements of transcriptomic abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
 
@@
 {\bf (B)}
-The mRNAs from individual cells are tagged with cell-specific barcodes using a 10X Chromium device.
+The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes using a 10X Chromium.
 {\bf (C)}
@@
 Example MACS enrichments of IFN+ influenza-infected cells.
-The A549 cells with the IFN-$\beta$ LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
+A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
 After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
@@
 The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
-As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment for IFN+ cells.
+As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
 We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
@@
 
-\figdata{Genbank files giving the sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
+\figdata{Genbank files giving sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
 \label{figdata:virus_seqs}
@@
 
-We therefore developed the experimental approach in \FIG{workflow} to simultaneously determine the entire transcriptome and the full sequences of the infecting virions in single cells.
+We therefore used the approach in \FIG{workflow} to simultaneously determine the entire transcriptome and the full sequences of the infecting virions in single cells.
 First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain of influenza: the wildtype virus and a ``synonymously barcoded'' variant that contains two engineered synonymous mutations near each end of each gene (\FIGDATA[workflow]{virus_seqs}).
-We have previously used synonymous barcoding near the 3' end of viral mRNAs to quantify the co-infection rate in single-cell transcriptomic data~\citep[about half of cells infected by multiple virions express a mix of wildtype and synonymously barcoded viral transcripts;][]{russell2018extreme}.
+We have previously used synonymous barcoding near the 3' end of viral mRNAs to identify co-infections in single-cell transcriptomic data~\citep[about half of cells infected by multiple virions express a mix of wildtype and synonymously barcoded viral transcripts;][]{russell2018extreme}.
 New to this study, we also added synonymous barcodes near the 5' end of the mRNAs; as explained below, having viral barcodes at both ends provides an important control for PCR artifacts during full-length sequencing of viral transcripts.
@@
 We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{workflow}A) at a dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
-From 12 to 13 hours post-infection, we used magnetic columns to enrich for cells that expressed the cell-surface protein driven by the IFN reporter.
+From 12 to 13 hours post-infection, we used magnetic columns to enrich cells that expressed the cell-surface protein driven by the IFN reporter.
 Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to $\sim$5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}).
-In order to ensure the presence of IFN- cells, after the MACS enrichment we added back non-sorted cells to $\sim$10\% of the total.
+To ensure the presence of IFN- cells, after the MACS enrichment we added back non-sorted cells to $\sim$10\% of the total.
 We also added uninfected canine cells to $\sim$5\% of the total to serve as an internal control for multiplets and enable estimation of the background amount of viral mRNA detected in the transcriptomes of truly uninfected cells.
@@
 We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
-These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique copies of each viral gene in each cell.
-Because most cells are infected by just one or two unique virions, we can therefore build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of the infecting virions.
+These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
+Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
 Combining this information with the 3'-end sequencing simultaneously determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).
+
+\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}
 
@@
 \section{Acknowledgments}
-
-\jdbcomment{Add this}
+Andy Marty and Genomics Core.
+Cole Trapnell.
+Jason Underwood.
+AJ Velthuis. 
+\jdbcomment{Finish this}
 

### 2018-09-15 paper/paper.tex
draft Results section 3: _Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells_

@@
 After 13 hours, cells were analyzed by flow cytometry.
-As shown in the FACS plots, expression of the IFN-$\beta$ and IFN-$\lambda$ reporters is highly correlated in all cases.}
+As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.}
 {\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}}
@@
 The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
-After 13 hours, the cells were stained for expression of HA protein and analyzed for the IFN-$\lambda$ reporter driven ZsGreen expression by FACS.
+After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for expression of the ZsGreen driven by the \textit{IFNL1} reporter.
 Each condition was done in triplicate.
@@
 \caption{
-Single-cell transcriptomics of influenza-infected cells enriched for IFN positivity.
+Single-cell transcriptomics of IFN-enriched influenza-infected cells.
 {\bf (A)} 
 Number of cells for which transcriptomes were obtained.
-From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion.
+From the number of human and canine transcriptomes, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion.
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
-{\bf (G)} Fraction of cellular mRNA derived from IFN across all cells, faceted by whether the cells are infected.
+{\bf (G)} Fraction of cellular mRNA from IFN across all cells, faceted by whether the cells are infected.
 Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
@@
 \figsupp[Fraction of infected cells that detectably express each influenza gene.]
-{Bars show the fraction of infected cells that are called as expressing each viral gene.
+{The fraction of infected cells that are called as expressing each viral gene.
 The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
-Each viral gene is detected in $\approx$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
 The exception is NP, which is detected in virtually all infected cells.
-The much higher frequency of detecting NP could reflect a biological phenomenon, or it could be because cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments and analysis.}
+The much higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments because the number of viral mRNAs is below the detection limit.}
 {\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}}
@@
 
-\figsupp[Expression of type I and type III IFN genes is highly correlated in single cells in our experiments.]
-{The correlation between the fraction of cellular mRNA derived from type I (IFN-$\alpha$ and IFN-$\beta$) and type III (IFN-$\lambda$) IFN in the A549 cells in our experiments.
-Each point represents a single cell.
+\figsupp[Expression of type I and type III IFN is highly correlated in single cells in our experiment.]
+{The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
+Each point represents one cell.
 The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
@@
 
-\figsupp[Expression of interferon-stimulated genes (ISGs).]
+\figsupp[Expression of ISGs in single infected and un-infected cells.]
 {
-For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs that are highly expressed A549 cells (IFIT1, ISG15, CCL5, and Mx1). 
+For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
 {\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
@@
 Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
-This is probably because paracrine signaling leads to ISG expression in some cells that are not themselves expressing IFN.
-{\bf (B)} The correlation between the fraction of cellular mRNA derived from IFN and from ISGs.
-Each point represents a single cell, and the Pearson correlation coefficient is shown.
-IFN and ISG expression is more correlated for infected than uninfected cells, probably because in the latter the ISG expression is often due to paracrine signaling that does not induce expression of IFN itself.
-But among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but very few cells that express high levels of IFN without also substantially expressing ISGs.
+This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
+{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
+Each point represents one cell, and the Pearson correlation coefficient is shown.
+IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression more often due to paracrine signaling that does not induce expression of IFN itself.
+Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
 }
@@
 
-\figsupp[Unsupervised t-SNE clustering shows that cell-to-cell variation in influenza genes, IFN genes, and ISGs explains a substantial part of the structure of the data.]
-{To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
-In the t-SNE plot, each point is a different cell and cells with similar transcriptomes are closer together.
-Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA shown on a log (top) or linear (bottom) scale.
-As is clear from this plot, expression of influenza genes, IFN, and ISGs help explain the structure of the data, since cells with high expression of these genes clearly group together.
+\figsupp[Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.]
+{To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
+In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
+Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA, shown on a log (top) or linear (bottom) scale.
+As is clear from this plot, expression of influenza, IFN, and ISG genes contributes substantially to the structure of the data, since cells with high expression of these genes clearly group together.
 }

### 2018-09-16 paper/paper.tex
draft of results section on transcriptomics

@@
 \newcommand{\abrcomment}[1]{\emph{\color{blue} [#1]}}
+
+\renewcommand{\floatpagefraction}{.9}
 
@@
 \subsection{A system to identify and enrich for rare IFN+ positive cells}
-A challenge in studying the induction of IFN by influenza is that infection only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
-For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
+Influenza infection only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
+For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
 Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
@@
 The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
-Activation of the type I and type III IFN reporters is highly correlated in single cells upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
 {\bf (C)}
@@
 We validated that the reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
-We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result further validated by our single-cell transcriptomics described below.
+We also found that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result further validated by our single-cell transcriptomics described below.
 Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.
@@
 We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
-Standard single-cell transcriptomic approaches are inadequate for this purpose, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
+Standard single-cell transcriptomic approaches are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
 These measurements of transcriptomic abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
@@
 Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to $\sim$5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}).
-To ensure the presence of IFN- cells, after the MACS enrichment we added back non-sorted cells to $\sim$10\% of the total.
-We also added uninfected canine cells to $\sim$5\% of the total to serve as an internal control for multiplets and enable estimation of the background amount of viral mRNA detected in the transcriptomes of truly uninfected cells.
+To ensure the presence of IFN- cells after the MACS enrichment, we added back non-sorted cells to $\sim$10\% of the total.
+We also added uninfected canine cells to $\sim$5\% of the total as an internal control for multiplets and to estimate the background amount of viral mRNA detected in the transcriptomes of truly uninfected cells.
 
-We then processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and then reverse transcribes polyadenylated mRNAs in a way that appends a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
+We processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and then reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
 Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
-Furthermore, because virtually all the influenza genome is transcribed into mRNA, the cell-barcoded cDNA spans nearly the entire 13,581-nucleotide segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.
+Furthermore, because virtually all the influenza genome is transcribed, the cell-barcoded cDNA spans nearly the entire 13,581-nucleotide segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.
 
 We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
-But we also took a portion of the cell-barcoded cDNA and enriched for full-length viral molecules (\FIG{workflow}D) using PCR as detailed below.
+But we also took a portion and enriched for full-length viral molecules (\FIG{workflow}D) using PCR as detailed below.
 We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
@@
 Number of cells for which transcriptomes were obtained.
-From the number of human and canine transcriptomes, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion.
+From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
 Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
-Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (possibly low-quality emulsions or multiplets), and are excluded from subsequent analyses.
+Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
 {\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
@@
 The inset shows the amount of viral mRNA among the human cells called as infected.
-{\bf (D)} The number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
+{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
 The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
-\FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected in infected cells.
+\FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected.
 {\bf (E)} Relative expression of viral genes among infected cells.
-{\bf (F)} The number of cells infected with wildtype virus, synonymously barcoded virus, or both.
+{\bf (F)} Number of cells infected with wildtype virus, synonymously barcoded virus, or both.
 From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
-{\bf (G)} Fraction of cellular mRNA from IFN across all cells, faceted by whether the cells are infected.
+{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
 Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
-Many cells that do not express IFN itself still express ISGs (\FIGSUPP[transcriptomics]{ISG}).
+A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
+Many cells that do not express IFN still express ISGs (\FIGSUPP[transcriptomics]{ISG}).
 }
@@
 
-\figsupp[Expression of ISGs in single infected and un-infected cells.]
+\figsupp[Expression of ISGs in single infected and uninfected cells.]
 {
@@
 %%% end transcriptomics figure
+
+We first analyzed the transcriptomic data to characterize the infection and IFN status of the cells in our experiment.
+Overall, we obtained transcriptomes for 1,614 human (A549) cells (some of which were infected with influenza), and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
+We also obtained 12 transcriptomes with a mix of human and canine transcripts; these mixes arise when multiple cells are captured in a single droplet.
+From the number of mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the human transcriptomes are derived from multiple cells.
+To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B).
+For the remainder of this paper, we analyze the 1,490 human cells that passed this filtering.
+
+To identify infected cells, we examined the fraction of each transcriptome that was derived from virus (\FIG{transcriptomics}B).
+As expected, only a small fraction ($\sim$0.7\%) of the transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction.
+Using this test, we classified 290 of the 1,490 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
+We classified the other cells as uninfected, although we cannot rule out that some of them were infected with virions that produced too little mRNA to be detected in our experiment.
+The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
+As we observed previously~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but for a few cells viral mRNA comprises over half the transcriptome.
+
+We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
+Although ten transcripts are encoded by the eight viral genes, we call presence / absence at the level of genes rather than transcripts.
+The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable in the 3'-end sequencing transcriptomic data.
+\FIG{transcriptomics}D (top panel) shows that the majority (162 of 290) of infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
+This measured frequency of infected cells expressing all eight genes is higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
+The difference is sufficiently modest that it could simply be due to methodological differences in the thresholds used to call gene presence.
+But it is important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could conceivably alter the distribution of viral gene expression from a fully random sample of infected cells.
+
+The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel), consistent with our prior work~\citep{russell2018extreme}.
+However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived over 50\% of their mRNA from influenza, while other cells that expressed all viral genes had only a few percent of their mRNA from influenza (\FIG{transcriptomics}D, bottom panel).
+Also consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent across cells (\FIG{transcriptomics}E) and generally matched values obtained by older bulk studies~\citep{hatada1989control}.
+
+Using the synonymous barcodes near the 3' termini of transcripts from one of the two infecting viral variants, we determined that 38\% of cells were co-infected with wildtype and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
+From the numbers in \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode and so cannot be directly identified as co-infected from the transcriptomic data.
+Interestingly, this co-infection rate is substantially higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
+We speculate that the discrepancy arises because the MACS for IFN+ cells might also enrich for co-infected cells, because infection itself may not actually be Poisson, or because co-infection might increase the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
+
+We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
+Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}C, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
+Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected but produced too few viral transcripts to be detected above background.
+Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\FIGSUPP[transcriptomics]{ISG}A).
+The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express significant IFN (\FIGSUPP[transcriptomics]{ISG}B).
+Overall, these results are consistent with the established knowledge that IFN is expressed only in cells that directly detect that they have been infected, but that ISGs are also expressed via paracrine signaling in cells that do not detect viral infection~\citep{stetson2006type,honda2006type}.
+
+Finally, we qualitatively examined whether expression of viral genes, IFN, and ISGs play a major role in explaining the structure of the high-dimensional transcriptomic data.
+\FIGSUPP[transcriptomics]{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
+Although the clustering is entirely unsupervised, cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with expression of these genes involves uninfected and IFN- cells.
+This fact provides some qualitative justification for the focus of this paper on examining viral factors associated with IFN expression.
 

### 2018-09-16 paper/paper.tex
Draft of result section 4: _The full genotypes of viruses in single IFN+ and IFN- cells_

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Single-cell virus sequencing of influenza variants that trigger innate immunity}
+%\title{Single-cell virus sequencing of influenza variants that trigger innate immunity}
+\title{Combined single-cell transcriptomics and virus sequencing of influenza infections that trigger innate immunity}
 
@@
 We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
-But we also took a portion and enriched for full-length viral molecules (\FIG{workflow}D) using PCR as detailed below.
+But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
 We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
@@
 
-We first analyzed the transcriptomic data to characterize the infection and IFN status of the cells in our experiment.
+We first analyzed the transcriptomic data.
 Overall, we obtained transcriptomes for 1,614 human (A549) cells (some of which were infected with influenza), and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
 We also obtained 12 transcriptomes with a mix of human and canine transcripts; these mixes arise when multiple cells are captured in a single droplet.
-From the number of mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the human transcriptomes are derived from multiple cells.
+From the number of mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
 To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B).
@@
 
-To identify infected cells, we examined the fraction of each transcriptome that was derived from virus (\FIG{transcriptomics}B).
+To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}B).
 As expected, only a small fraction ($\sim$0.7\%) of the transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
@@
 The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
-As we observed previously~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but for a few cells viral mRNA comprises over half the transcriptome.
+As described previously~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but for a few cells viral mRNA comprises over half the transcriptome.
 
@@
 
-Using the synonymous barcodes near the 3' termini of transcripts from one of the two infecting viral variants, we determined that 38\% of cells were co-infected with wildtype and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From the numbers in \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode and so cannot be directly identified as co-infected from the transcriptomic data.
+Using the synonymous barcodes near the 3' termini of transcripts from one of the two infecting viral variants, we determined that 38\% of cells were co-infected with wildtype and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected the null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
+From the numbers in \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be directly identified from the transcriptomic data).
 Interestingly, this co-infection rate is substantially higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
-We speculate that the discrepancy arises because the MACS for IFN+ cells might also enrich for co-infected cells, because infection itself may not actually be Poisson, or because co-infection might increase the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
+This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
 
@@
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express significant IFN (\FIGSUPP[transcriptomics]{ISG}B).
-Overall, these results are consistent with the established knowledge that IFN is expressed only in cells that directly detect that they have been infected, but that ISGs are also expressed via paracrine signaling in cells that do not detect viral infection~\citep{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect that they have been infected, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect viral infection~\citep{stetson2006type,honda2006type}.
 
@@
 
-\subsection{Viral genotypes in IFN+ and IFN- influenza-infected cells}
-\FIG{genotypes} and \FIGSUPP[genotypes]{genotypes_by_ifn} and \FIGDATA[genotypes]{genotypes}
+\subsection{The full genotypes of viruses in single IFN+ and IFN- cells}
 
@@
 {\centering
-\includegraphics[height=0.76\textheight]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.78\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 Viral genotypes and infection outcomes in single cells.
-Each row shows a single infected cell.
-Arrows indicate the presence of a viral gene from a single viral barcode variant (light blue) or both barcode variants (dark blue).
-Circles and boxes indicate mutations or indels as described in the legend.
-Circle areas and box heights are proportional to the fraction of PacBio CCSs that report the mutation (see \FIGDATA[genotypes]{genotypes} for numerical data).
-For dual-barcode infections, mutations / indels for the wildtype viral variant are on the top half of the arrow and those for the synonymously barcoded variant are on the bottom half. 
-Green boxes to the left show the percent of all mRNA in that cell derived from virus.
-Orange boxes show the percent of cellular mRNA derived from IFN, with boxes framed in orange indicating cells classified as IFN+ in \FIG{transcriptomics}G.
+Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
+Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
+For dual-barcode infections, mutations / indels for the wildtype and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively. 
+Green boxes at left show the percent of all mRNA in that cell derived from virus.
+Orange boxes show the percent of cellular mRNA derived from IFN, with orange box frames indicating cells classified as IFN+ in \FIG{transcriptomics}G.
 }
@@
 
+\figsupp[Number of PacBio CCSs and PCR strand exchange rate.]
+{
+The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
+Note that these sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
+(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in Supplementary~file~\ref{suppfile:pacbio_analysis}.)
+Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
+Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
+Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
+Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences derived from the same cDNA, and cell barcodes were used to collapse duplicate sequences from the same cell.
+The bars in the plot are colored by whether the sequence is derived from the wildtype viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \FIGSUPP[genotypes]{StrandExchange} for more explanation).
+From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
+Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
+A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
+}
+{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
+\label{figsupp:CCSs}
+
 \figsupp[Strategy for detecting strand exchange during sequencing of full-length viral genes.]
-{The library preparation for PacBio sequencing of the cDNA for the full-length viral genes requires many cycles of PCR in order to produce a sufficient amount of DNA for sequencing.
-A major concern is whether strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
+{The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
+A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
 We can detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wildtype virus and virus carrying synonymous barcodes near both termini of each gene.
-If there is no strand exchange, all molecules should either be wildtype or have the synonymous mutations at the viral barcode sites.
-But strand exchange will create some molecules that have a mix of wildtype nucleotides in the viral barcode at one termini and synonymous mutations in the viral barcode at the other termini.
+If there is no strand exchange, all molecules should either be wildtype or have the synonymous barcoding mutations at \emph{both} termini.
+But strand exchange will create some molecules that have wildtype nucleotides at one termini and synonymous barcoding mutations at the other termini.
 \FIGSUPP[genotypes]{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
@@
 
-\figsupp[Number of PacBio circular consensus sequences and PCR strand exchange rate.]
-{
-The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
-Note that these sequences were obtained using several PacBio runs, some of which were loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
-Therefore, unlike the 10X transcript count data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
-Especially for the polymerase genes (PB2, PB1, and PA), many of the CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were presumably preferentially amplified during PCR.
-Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
-Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences for the same gene, and cell barcodes to collapse duplicate sequences from the same cell.
-The bars are colored by whether the sequence can be identified as being derived from the wildtype viral variant, the synonymously barcoded variant, or represents a mixed barcode molecule (see \FIGSUPP[genotypes]{StrandExchange}).
-From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
-Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
-Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
-A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcode.
-}
-{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
-\label{figsupp:CCSs}
-
 \figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
@@
 {\bf (A)} We could call the complete genotypes of the infecting virus(es) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
-{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNAs than cells for which we could not call complete genotypes.
-This makes sense, as cells with more viral mRNA are more likely to have their viral genes captured in the PacBio sequencing, which is only able to capture a small fraction of the total transcripts identified by the 3'-sequencing of the 10X platform.
-The lower calling rate for dual-barcode co-infections relative to single barcode infectionis is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing.
+{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
+Both facts makes sense.
+Cells with more viral mRNA are more likely to have their viral cDNA captured in the PacBio sequencing, which is only captures a small fraction of the total transcripts identified by the 3'-end sequencing transcriptomic sequencing.
+The lower calling rate for dual-barcode co-infections is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing.
 An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
@@
 
-\figdata{A text file giving the primers used to amplify the influenza transcripts for PacBio sequencing are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
+\figdata{A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
 }
@@
 %%% end genotypes figure %%%
+
+The foregoing transcriptomic analyses identify which viral genes are expressed in each infected cell, but they do not provide information on the sequences of these genes.
+This is a major shortcoming, since influenza's high mutation rate means that although the \emph{consensus} sequence of the viral stock used in our experiment corresponds to WSN influenza, the specific virions infecting individual cells will often carry mutations~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+We therefore used the approach schematized in \FIG{workflow}D to determine the full-length sequences of all viral genes expressed in the majority of infected cells.
+
+By employing several PCR enrichment strategies (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to influenza genes (\FIGSUPP[genotypes]{CCSs}).
+Crucially, the fact that about half the infecting virions encoded genes that were tagged with synonymous viral barcodes at both termini enabled us to estimate the rate of PCR strand exchange among cDNAs that we amplified for PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
+Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link mutations in viral transcripts to cell barcodes and UMIs that identify the cell and molecule of origin.
+
+We used these CCSs to call the sequences of viral genes in single cells.
+We called a mutation as present in a viral gene in a particular cell if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
+For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
+This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant, as well as identifying high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+The strategy will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency among viral transcripts in that cell.
+Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is an inherent limit to our detection threshold: we can identify viral mutations that are at high frequency within single infected cells, but not mutations that are present on just a small fraction of the viral transcripts in a cell.
+
+We were able to call the sequences of all expressed viral genes in the majority of infected cells (\FIGSUPP[genotypes]{ncells}).
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were not co-infected by both viral barcode variants (\FIGSUPP[genotypes]{ncells}).
+But despite this bias towards calling full viral genotypes from cells with these characteristics, we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
+
+The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
+Simple visual inspection of this figure reveals a wealth of information.
+For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed all eight genes and did not induce detectable IFN.
+But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that either had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
+The two cells that expressed the most IFN in our experiment (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS segment, which encodes influenza virus's primary innate immune antagonist NS1.
+Many other IFN+ cells had other defects, including large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+However, \FIG{genotypes} also shows that expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} still only managed to express viral mRNA to only 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN expression.
+
+\subsection{Systematic examination of viral features associated with infection outcome}
 
@@
 %%% end mutations figure %%%
+
+\subsection{Validation that multiple types of viral mutations increase IFN induction}
 
@@
 We then further filtered for CCSs that contained valid cell barcodes as identified by the \texttt{cellranger} pipeline, and kept just one CCS per UMI (preferentially retaining high-quality CCSs that aligned to full-length cDNAs).
+We then removed from the CCSs the barcoding synonymous mutations that we had engineered into one of the two viral variants.
 Finally, we used the CCSs to call the sequence of the viral gene in each cell, calling mutations separately for each viral barcode variant.
@@
 This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
-The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb.}}
+The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.}
 \end{suppfile}
@@
 This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
-The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb.}}
+The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.}
 \end{suppfile}

### 2018-09-17 paper/paper.tex
draft of _Systematic examination of viral features associated with infection outcome_ (Results section 5)

@@
 \caption{
-Viral genetic variation partially explains the heterogeneity in viral burden and IFN induction among the infected cells for which we sequenced all expressed viral genes (\FIG{genotypes}).
+Viral genetic variation partially explains heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.
 {\bf (A)} 
-The percent of all mRNA derived from virus, faceted by whether the cells express wildtype copies of all eight viral genes, or fail to express a wildtype copy of one or more genes (due to gene absence, amino-acid substitution, insertion, or deletion).
+Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight viral genes, or fail to express an unmutated copy of one or more genes.
 Cells infected by full wildtype virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
@@
 Some specific forms of viral genetic variation are associated with IFN induction.
-The top panel show the percent of IFN+ and IFN- cells that fail to express each viral gene.
+The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
 The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each viral gene, conditioned on the cell expressing that gene.
 Numbers above the bars give P-values (Fisher's exact test) and associated Q-values~\citep{storey2003statistical} for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
-Absence of NS is strongly associated with IFN induction, and amino-acid substitutions in PB1 or NS and deletions in HA are weakly associated with IFN induction.
-Only the absence of NS remains significant at a false-discovery rate of 10\%.
-\FIGSUPP[mutations]{allcells} shows that the trends remain largely the same if we also analyze infected cells with incomplete viral sequence information, with the addition of a trend towards IFN induction associated with \emph{presence} of PB2 and PA.
-\FIGSUPP[mutations]{isg} shows that the trends also remain largely similar if we analyze the association of viral genetic variation with ISG expression rather than IFN expression.
-Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
+\FIGSUPP[mutations]{allcells} shows that the trends remain largely the same if we include infected cells with incomplete viral sequence information. 
+\FIGSUPP[mutations]{isg} shows that the trends remain largely the same if we categorize cells by ISG expression rather than IFN expression.
 {\bf (D)}
 There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
-Note that throughout this figure, we only consider as substitution mutations that are non-synonymous.
+Throughout this figure, we only consider substitutions that are non-synonymous.
+Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
 }
@@
 %%% end mutations figure %%%
+
+To more systematically assess how viral genetic variation affected infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
+\FIG{mutations}A shows that the 49 cells infected by full wildtype viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
+Therefore, viral genetic variation is a major contributor to the heterogeneity in viral transcriptional burden among cells.
+
+Viral genetic variation also contributed to the heterogeneity in IFN induction, with cells infected by full wildtype virions less likely to express IFN than cells infected by incomplete or mutated virions (\FIG{mutations}B).
+Furthermore, specific viral genetic defects were associated with IFN induction (\FIG{mutations}C).
+For instance, absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were more weakly enriched.
+Only the absence of NS remained associated with IFN induction at a false discovery rate of 10\% (\FIG{mutations}C), although a Bayesian-minded virologist might argue that controlling for all the hypotheses in \FIG{mutations}C is too stringent since the association of defects in certain genes (particularly NS and the polymerase genes) with IFN induction is more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
+In any case, the modest number of cells with full viral genotype information precludes drawing strong conclusions from statistical analyses alone---but \FIG{genotypes} and \FIG{mutations}C do motivate compelling hypotheses about specific viral mutations that contribute to IFN induction, and for most of the rest of this paper we pursue these hypotheses using experimental follow-up.
+
+One other interesting trend emerges from analyzing the combined single-cell transcriptomic and viral genetic information.
+There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
+The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of the IFN induction, but the experimental follow-up below will also shed more light on this question.
 

### 2018-09-17 paper/paper.tex
Made edits to viral rescue for variants of interest (non SNP). Will keep working on any outstanding methods and next set of figures

--- 2018-09-16 paper/paper.tex
@@
 For the dual type I / type III reporter used in \FIGSUPP[IFNrare]{type_I_vs_III}, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
+All reporter lines tested negative for mycoplasma contamination by the Fred Hutch Genomics Core.
 
@@
 \jdbcomment{Alistair writes this. I've copied text from prior consolidated section, but it needs to be clarified by more clearly breaking things down by virus.}
-All viruses were generated using reverse-genetics as previously described.
-All infections were performed in influenza growth media as previously described.
-Briefly, bidirectional plasmids producing viral mRNA and vRNA are transfected into a co-culture of MDCK-SIAT1 and 293t cells grown in D10.
-After 24h, media is changed from D10 to influenza growth media.
-Between 48h and 72h post-infection, depending on viral growth kinetics, supernatant was harvested and titered by TCID50.
-Virus was subsequently expanded on MDCK-SIAT1 cells as described below.
-For all viral variants, the described mutations were introduced into reverse-genetics plasmids using PCR mutagenesis. 
-For the HA-flanked mCherry pseudovirus, a plasmid was generated similarly to a previously-described HA-flanked eGFP vector.
-Synonymously-barcoded virus utilized previously described 3' synonymous mutations, and two newly-introduced single-nucleotide synonymous mutations at the 5' end of each viral gene.
-Wild-type, synonymously-barcoded, NS1-inactivated, and single-nucleotide variant viruses were all grown in unmodified MDCK-SIAT1 and 293t cells.
-Viral variants bearing large internal deleltions in the PB1 segment were grown in MDCK-SIAT1 and 293t cells expressing PB1 as previously described.
-Pseudovirus expressing mCherry in place of eGFP was grown in MDCK-SIAT1 and 293t cells expressing HA as previously described.
 
-Wildtype and synonymously barcoded viruses in \FIGDATA[workflow]{virus_seqs} obtained from reverse-genetic transfections was subsequently propagated at an MOI of 0.001 for 72h twice in MDCK-SIAT1 cells (wild-type) or once at an MOI of 0.01 (barcoded) for 60h in order to generate potential immunostimulatory diversity.
-For analyses of interferon induction, an additional wild-type stock of virus was grown at an MOI of 0.01 for 36h in order to generate a population relatively free of defective particles.
-(WILL REPEAT WITH 48H STOCK AS DESCRIBED BELOW)
-Viruses bearing SNPs, deletions in PB1, NS1-inactivation mutations, and HA-mCherry pseudovirus were all propagated at an MOI of 0.01 for 48h owing to slower growth kinetics. 
-All stocks were subsequently titered by TCID50, using complementing cells where necessary.
-Additionally, for qPCR experiments, stocks were titered by qPCR against the viral HA segment using a plasmid standard to determine relative genome copy numbers between viral stocks.
+All viruses were generated using reverse-genetics as previously described using plasmids derived from pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
+To generate viruses from these plasmids, or variants described below, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
+At 24 hr post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml).
+At 48-72h post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 min, and stored aliquots of the clarified viral supernatant at 80$^{\circ}$C.
+We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of Reed and Muench as described above.
+
+For comparisons of interferon stimulation a relatively pure stock of WSN influenza was derived by harvesting reverse-genetics supernatants at 48h post infection and passaging virus on MDCK-SIAT1 cells for 36h at an MOI of 0.05---conditions which we previously validated to lead to stocks relatively depleted in defective particles~\citep{russell2018extreme}.
+
+To generate an NS1-inactivated strain of influenza, six nucleotide changes resulting in the addition of five in-frame stop codons introduced immediately downstream of the 5' splice donor site, interrupting translation of NS1 while leaving NS2 (NEP) intact.
+The viral RNA produced by this vector is described in (sequence)\abrcomment{I have uploaded the genbank files, just waiting for appropriate figure to append}.
+This modified plasmid was used in a reverse-genetics approach to recapitulate mutant virus, harvesting transfection supernatant after 72h.
+This virus was then passaged in MDCK-SIAT1 cells for 48h at an MOI of 0.05---due to slower growth kinetics than the wild-type virus this longer incubation was necessary.
+Genome copy number of this virus relative to infectious titer was determined as described below by qPCR.
+
+To generate viruses bearing internal deletions in the PB1 segment, indicated regions of the PB1 segment were excised by PCR from our reverse genetics plasmids.
+Sequences are described in (sequence) \abrcomment{I have uploaded the genbank files and modified coordinates in map names to match mRNA coordinates, just waiting for appropriate figure to append}. 
+To rescue these viruses, we used 293t cells and MDCK-SIAT1 cells transduced to constituitively express PB1 as previously described!\citep{Bloom2010oseltamivir}. 
+Virus was harvested from transfections at 72h, and passaged twice in MDCK-SIAT1 cells expressing PB1 constituitively at an MOI of 0.001 for 72h and 0.01 for 48h. 
+This passaging was necessary as viral titers from transfections were too low to generate sufficient virus from a single passage.
+This virus was titered on complementing MDCK-SIAT1 cells.
+Genome copy number of this virus relative to infectious titer was determined as described below by qPCR.
+
+To generate pseudovirus that expresses fluorescent mCherry instead of HA, a plasmid was generated similarly to a previously-described eGFP-expressing pseudovirus~\citep{russell2018extreme}. 
+To rescue this virus, we used MDCK-SIAT1 cells transduced to constituitively express HA and included an HA protein expression plasmid in our reverse-genetics transfection mixture.
+Virus was harvested from transfections at 72h and passaged for 48 at an MOI of 0.01 on MDCK-SIAT1 cells that express HA. 
+This virus was titered on complementing MDCK-SIAT1 cells.
 
@@
 ent}
+_figures/pacbio_analysis.html}.
+This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
+The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb.}}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{\label{suppfile:monocle_analysis}
+An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
+This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
+The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb.}}
+\end{suppfile}
+
+\end{document}
+ent}
+mlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb.}}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{\label{suppfile:monocle_analysis}
+An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
+This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
+The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb.}}
+\end{suppfile}
+
+\end{document}
+ent}

### 2018-09-19 paper/paper.tex
remove duplicated text at end of paper (somehow very last text with suppfiles had gotten repeated a few times)

@@
 \end{document}
-ent}
-_figures/pacbio_analysis.html}.
-This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
-The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb.}}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{\label{suppfile:monocle_analysis}
-An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
-This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
-The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb.}}
-\end{suppfile}
-
-\end{document}
-ent}
-mlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb.}}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{\label{suppfile:monocle_analysis}
-An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
-This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
-The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb.}}
-\end{suppfile}
-
-\end{document}
-ent}

### 2018-09-20 paper/paper.tex
Added descriptions of SNP-virus reconstitutitions to methods as well as flow cytometry analyses.

@@
 This melting temperature is lower than the standard 98$^{\circ}$C melting step suggested by the manufacturer for Kapa HiFi because we wanted to avoid collapse of emulsion integrity at high temperature.
-We performed 30 cycles in order to saturate the reagents in each emulsion---unlike for open PCR reactions, the physical occlusion of amplified material in emulsin PCR reduces artifacts at high cycle numbers.
+We performed 30 cycles in order to saturate the reagents in each emulsion---unlike for open PCR reactions, the physical occlusion of amplified material in emulsion PCR reduces artifacts at high cycle numbers.
 
@@
 \subsection{Viruses for validation experiments}
-\jdbcomment{Alistair writes this. I've copied text from prior consolidated section, but it needs to be clarified by more clearly breaking things down by virus.}
 
@@
 At 24 hr post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml).
-At 48-72h post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 min, and stored aliquots of the clarified viral supernatant at 80$^{\circ}$C.
+At 48-72h post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
 We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of Reed and Muench as described above.
@@
 Sequences are described in (sequence) \abrcomment{I have uploaded the genbank files and modified coordinates in map names to match mRNA coordinates, just waiting for appropriate figure to append}. 
-To rescue these viruses, we used 293t cells and MDCK-SIAT1 cells transduced to constituitively express PB1 as previously described!\citep{Bloom2010oseltamivir}. 
+To rescue these viruses, we used 293t cells and MDCK-SIAT1 cells transduced to constituitively express PB1 as previously described~\citep{Bloom2010oseltamivir}. 
 Virus was harvested from transfections at 72h, and passaged twice in MDCK-SIAT1 cells expressing PB1 constituitively at an MOI of 0.001 for 72h and 0.01 for 48h. 
@@
 This virus was titered on complementing MDCK-SIAT1 cells.
+
+Viruses bearing potentially innate-immune stimulatory SNPs as identified in our single-cell analyses were recapitulated by first introducing these mutations to plasmids encoding the viral segments PB1 or NS by PCR.
+Sequences of these plasmids are described in (sequence).
+\abrcomment{I have uploaded the genbank files and modified coordinates in map names to match mRNA coordinates, just waiting for appropriate figure to append}.
+These viruses were rescued via reverse-transfection in a coculture of MDCK-SIAT1 and 293t cells alongside a matched wild-type control.
+Each virus was rescued in two seperate transfection reactions. 
+Transfection supernatants were harvested at either 72 hours post transfection (biological replicate \#2 of all SNPs tested) or 48-50h post infection (biological replicate \#1 of all SNPs tested and both replicates of wild-type influenza).
+Virus recovered from transfections was titered by hemagglutination, and MOI set to infectious dose as measured in the wild-type virus as measured by TCID50 on MDCK-SIAT1 cells. 
+Particle-normalized viruses recovered from transfections were used directly in experiments in \abrcomment{Figure}.
+
+\subsection{flow cytometry analysis of interferon induction}
+Our A549 lambda reporter cell line was seeded in D10 media 24h prior to infection.
+At the time of infection, D10 media was replaced with influenza growth media and virus added to the indicated titers.
+Cells were harvested 13hpi and stained for HA as described above.
+For NS1 staining, HA-stained cells were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 $\mu$g/ml, washed, stained with a goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405, washed, and analyzed.
+After channels were compensated and cells gated to exclude multiplets and debris in FlowJo, data were analyzed using custom python pipelines described in \abrcomment{wherever we put those notebooks}.
+Briefly, gating thresholds for interferon, mCherry, and HA staining were set to 99.9\% of an uninfected control.
+For NS1, owing to sample-to-sample staining variation, thresholds were dynamically set on an individual sample to encompass an equal proportion of HA-expressing and NS1-expressing cells.
 

### 2018-09-20 paper/paper.tex
organized mutant plasmid maps and fixed numbering of PB1del names

@@
 
-To generate an NS1-inactivated strain of influenza, six nucleotide changes resulting in the addition of five in-frame stop codons introduced immediately downstream of the 5' splice donor site, interrupting translation of NS1 while leaving NS2 (NEP) intact.
+To generate an NS1-inactivated strain of influenza, six nucleotide changes resulting in the addition of five in-frame stop codons starting 10 nucleotides downstream of the 5' splice donor site, interrupting translation of NS1 while leaving NS2 (NEP) intact.
 The viral RNA produced by this vector is described in (sequence)\abrcomment{I have uploaded the genbank files, just waiting for appropriate figure to append}.

### 2018-09-20 paper/paper.tex
slight edits to Methods, then realized better to wait until figures made

@@
 
+%%% begin validation figure %%%
+\begin{figure}
+
+\jdbcomment{add figure}
+\caption{\jdbcomment{add caption}}
+\label{fig:validation}
+
+\figdata{Genbank plasmid maps for the mutant genes in this figure cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
+\label{figdata:sequences}
+
+\end{figure}
+%%% end validation figure %%%
+
 \section{Discussion}
@@
 We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\citep{xue2016propagation}.
-The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 in influenza growth media and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
+The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 in influenza growth media ((Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
 Specifically, after generation by reverse genetics, the wildtype variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
@@
 \subsection{Viruses for validation experiments}
-
-All viruses were generated using reverse-genetics as previously described using plasmids derived from pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
-To generate viruses from these plasmids, or variants described below, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
-At 24 hr post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml).
-At 48-72h post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
-We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of Reed and Muench as described above.
+We generated the mutant viruses for the validation experiments in \FIG{validation} by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wildtype plasmid for that gene.
+To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
+At 24 hours post-transfection, we changed media from D10 to influenza growth media. 
+At 48 to 72 hours post-transfection, we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
+We then thawed an aliquot of each virus and titered by TCID50 as described above.
 

### 2018-09-21 paper/paper.tex
some modest tweaks to Results text

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\title{Single-cell virus sequencing of influenza variants that trigger innate immunity}
-\title{Combined single-cell transcriptomics and virus sequencing of influenza infections that trigger innate immunity}
+\title{Single-cell virus sequencing of influenza infections that trigger innate immunity}
 
@@
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We determine the cellular transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that trigger immunity at the single-cell level are associated with defects in the infecting viruses that include failure to express the primary immune antagonist, internal deletions in polymerase genes, and mutations in proteins involved in replication and nuclear export.
-However, immune activation remains stochastic in cells infected by viruses with these genetic defects, and sometimes occurs even in cells infected with wildtype virus.
+We do this by determining the cellular transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Infections that trigger immunity in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
+However, immune activation remains stochastic in cells infected by viruses with these defects, and sometimes occurs even in cells infected with a virus that expresses unmutated copies of all genes.
 Our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation in single cells.
@@
 We validated that the reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
-We also found that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result further validated by our single-cell transcriptomics described below.
+We also found that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by our single-cell transcriptomics described below).
 Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.
@@
 
-\subsection{Simultaneous transcriptomics and virus-sequencing of single infected cells}
+\subsection{Combined transcriptomics and virus-sequencing of single infected cells}
 We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
@@
 First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain of influenza: the wildtype virus and a ``synonymously barcoded'' variant that contains two engineered synonymous mutations near each end of each gene (\FIGDATA[workflow]{virus_seqs}).
-We have previously used synonymous barcoding near the 3' end of viral mRNAs to identify co-infections in single-cell transcriptomic data~\citep[about half of cells infected by multiple virions express a mix of wildtype and synonymously barcoded viral transcripts;][]{russell2018extreme}.
-New to this study, we also added synonymous barcodes near the 5' end of the mRNAs; as explained below, having viral barcodes at both ends provides an important control for PCR artifacts during full-length sequencing of viral transcripts.
+These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts.
 
@@
 Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
-Furthermore, because virtually all the influenza genome is transcribed, the cell-barcoded cDNA spans nearly the entire 13,581-nucleotide segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.
+Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.
 
@@
 
-We first analyzed the transcriptomic data.
-Overall, we obtained transcriptomes for 1,614 human (A549) cells (some of which were infected with influenza), and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
+We obtained transcriptomes for 1,614 human (A549) cells (some of which were infected with influenza), and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
 We also obtained 12 transcriptomes with a mix of human and canine transcripts; these mixes arise when multiple cells are captured in a single droplet.
@@
 The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
-As described previously~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but for a few cells viral mRNA comprises over half the transcriptome.
+As in our prior work~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but for a few cells viral mRNA comprises over half the transcriptome.
 
@@
 
-Using the synonymous barcodes near the 3' termini of transcripts from one of the two infecting viral variants, we determined that 38\% of cells were co-infected with wildtype and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected the null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From the numbers in \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be directly identified from the transcriptomic data).
-Interestingly, this co-infection rate is substantially higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
+By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wildtype and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
+From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
+Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
 This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
@@
 
-Finally, we qualitatively examined whether expression of viral genes, IFN, and ISGs play a major role in explaining the structure of the high-dimensional transcriptomic data.
+Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
 \FIGSUPP[transcriptomics]{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
 Although the clustering is entirely unsupervised, cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with expression of these genes involves uninfected and IFN- cells.
-This fact provides some qualitative justification for the focus of this paper on examining viral factors associated with IFN expression.
 
@@
 
-The foregoing transcriptomic analyses identify which viral genes are expressed in each infected cell, but they do not provide information on the sequences of these genes.
+The foregoing transcriptomic analyses identify which viral genes are expressed in each cell, but they do not provide information on the sequences of these genes.
 This is a major shortcoming, since influenza's high mutation rate means that although the \emph{consensus} sequence of the viral stock used in our experiment corresponds to WSN influenza, the specific virions infecting individual cells will often carry mutations~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
-We therefore used the approach schematized in \FIG{workflow}D to determine the full-length sequences of all viral genes expressed in the majority of infected cells.
+We therefore used the approach schematized in \FIG{workflow}D to determine the full-length sequences of the viral genes expressed in the majority of infected cells.
 
@@
 We used these CCSs to call the sequences of viral genes in single cells.
-We called a mutation as present in a viral gene in a particular cell if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
+We called a mutation as present in a particular cell if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
 For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
-This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant, as well as identifying high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant (estimated to be $\sim$75\% of infected cells), as well as identifying high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
 The strategy will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency among viral transcripts in that cell.
@@
 But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that either had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
-The two cells that expressed the most IFN in our experiment (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS segment, which encodes influenza virus's primary innate immune antagonist NS1.
-Many other IFN+ cells had other defects, including large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
-However, \FIG{genotypes} also shows that expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} still only managed to express viral mRNA to only 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN expression.
+The two cells that expressed the most IFN in our experiment (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS segment, which encodes influenza virus's primary innate immune antagonist.
+Many other IFN+ cells had different defects, including large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+However, \FIG{genotypes} also shows that expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN.
 
@@
 \caption{
-Viral genetic variation partially explains heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.
+Viral genetic variation associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.
 {\bf (A)} 
@@
 
-To more systematically assess how viral genetic variation affected infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
+To more systematically assess how viral genetic variation associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
 \FIG{mutations}A shows that the 49 cells infected by full wildtype viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
@@
 
-Viral genetic variation also contributed to the heterogeneity in IFN induction, with cells infected by full wildtype virions less likely to express IFN than cells infected by incomplete or mutated virions (\FIG{mutations}B).
+Viral genetic variation also associated with IFN induction, since cells infected by full wildtype virions less IFN than cells infected by incomplete or mutated virions (\FIG{mutations}B).
 Furthermore, specific viral genetic defects were associated with IFN induction (\FIG{mutations}C).
 For instance, absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were more weakly enriched.
-Only the absence of NS remained associated with IFN induction at a false discovery rate of 10\% (\FIG{mutations}C), although a Bayesian-minded virologist might argue that controlling for all the hypotheses in \FIG{mutations}C is too stringent since the association of defects in certain genes (particularly NS and the polymerase genes) with IFN induction is more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
-In any case, the modest number of cells with full viral genotype information precludes drawing strong conclusions from statistical analyses alone---but \FIG{genotypes} and \FIG{mutations}C do motivate compelling hypotheses about specific viral mutations that contribute to IFN induction, and for most of the rest of this paper we pursue these hypotheses using experimental follow-up.
+Only the absence of NS remained associated with IFN induction at a false discovery rate of 10\% (\FIG{mutations}C), although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent since the association of defects in certain genes with IFN induction is more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
+In any case, the modest number of cells with full viral genotype information precludes drawing strong conclusions from statistical analyses alone---but \FIG{genotypes} and \FIG{mutations}C do motivate compelling hypotheses about specific viral mutations that contribute to IFN induction, and in the next few sections, we pursue these hypotheses using experimental follow-up.
 
-One other interesting trend emerges from analyzing the combined single-cell transcriptomic and viral genetic information.
+One other interesting trend emerges from the combined single-cell data.
 There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of the IFN induction, but the experimental follow-up below will also shed more light on this question.
+The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of the IFN induction, but the experimental follow-up below sheds more light on this question.
 

### 2018-09-22 paper/paper.tex
fixed flow for Figure 1 to not trim values <= 0
incorporated Figure 6 and supplements into paper

@@
 \hspace{0.02\textwidth}
-{\bf \Large C} \includegraphics[width=0.18\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
+{\bf \Large C} \includegraphics[width=0.2\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
 }
@@
 The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
-After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for expression of the ZsGreen driven by the \textit{IFNL1} reporter.
+After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
 Each condition was done in triplicate.
@@
 
-\jdbcomment{add figure}
-\caption{\jdbcomment{add caption}}
+\centerline{\includegraphics[width=0.85\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
+\caption{Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.
+{\bf (A)}
+Percent of cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line (see \FIGSUPP[validation]{SNPflow} for details).
+The numbers indicate the median of four measurements for each viral mutant.
+The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for the wildtype is indicated with a dashed blue line.
+The points are colored orange if the mutant virus stock induces IFN more frequently than the wildtype viral stock (t-test, $P < 0.01$), and blue if the frequency of IFN induction is not significantly different than wildtype at this cutoff.
+{\bf (B)}
+Similar to the first panel, but validates the increased rate of IFN induction for a large internal deletion in the PB1 gene.
+See \FIGSUPP[validation]{delflow} for detailed flow cytometry data, and \FIGSUPP[validation]{delqPCR} for details of how we validated that the infections were performed with a similar doses of each viral mutant.
+The experiments in the two panels were performed on different days, and so exact numerical values can be reliably compared within each panel but not between panels.
+}
 \label{fig:validation}
+
+\figsupp[Flow cytometry data for \FIG{validation}A.]
+{A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected (three replicates) or infected with stocks of the indicated viral mutant in quadruplicate.
+After 13 hours, cells were stained for expression of HA protein and analyzed by FACS for HA and the ZsGreen driven by the \textit{IFNL1} reporter.
+The contour plots show the density of all cells, and IFN+ cells are also indicated by orange dots.
+Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
+For all influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
+For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
+For each viral mutant, two independently generated stocks were each assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other viral stock).
+The infections with replicate \#1 of the wildtype virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
+The HA staining indicates that the number of transcriptionally active viral units per cell was similar across most mutants, and calculating percentage IFN+ only among HA+ cells largely corrects for modest variation in titer across mutants.
+}
+{\includegraphics[height=0.75\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}}
+\label{figsupp:SNPflow}
+
+\figsupp[Flow cytometry data for \FIG{validation}B.]
+{This figure is similar to \FIGSUPP[validation]{SNPflow} except that it shows the flow cytometry data for \FIG{validation}B.
+Each infection was performed in triplicate.
+A complicating factor for these data is that the virus with the deletion in PB1 (PB1del385to2163) cannot perform secondary transcription because it does not express PB1 protein, and so cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
+Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wildtype and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
+Then qPCR was used to show that the ratio of viral RNA to TCID50 was similar for all viral variants (data not shown).
+Finally, the approach in \FIGSUPP[validation]{delqPCR} was used to show that all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
+All this normalization suggests that all infections were performed at a similar dose of particles active for primary transcription.
+Therefore, the percent IFN+ was calculated from the flow data in this figure for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for the PB1del385to2163 virus.
+}
+{\includegraphics[width=0.7\textwidth]{figures/Validation_Figure/del_flow_plot.pdf}}
+\label{figsupp:delflow}
+
+\figsupp[Validation of viral titer normalization for \FIG{validation}B.]
+{Validation that the infections in \FIG{validation}B and \FIGSUPP[validation]{delflow} were performed at similar doses of virions capable of initiating primary transcription.
+In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \FIGSUPP[validation]{delflow}), and then mRNA was harvested for qPCR on oligo-dT primed reverse transcription.
+The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
+These infections were performed in the presence of absence of cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins.
+In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
+But in the presence of cycloheximide, all viruses produced similar amounts of viral mRNA, indicating that the dose of particles active for primary transcription is roughly equivalent across variants.
+Each measurement was performed in quadruplicate.
+}
+{\includegraphics[width=0.6\textwidth]{figures/Validation_Figure/qpcr_plot.pdf}}
+\label{figsupp:delqPCR}
 
@@
 Sequences are described in (sequence) \abrcomment{I have uploaded the genbank files and modified coordinates in map names to match mRNA coordinates, just waiting for appropriate figure to append}. 
-To rescue these viruses, we used 293t cells and MDCK-SIAT1 cells transduced to constituitively express PB1 as previously described~\citep{Bloom2010oseltamivir}. 
+To rescue these viruses, we used 293t cells and MDCK-SIAT1 cells transduced to constituitively express PB1 as previously described~\citep{bloom2010permissive}. 
 Virus was harvested from transfections at 72h, and passaged twice in MDCK-SIAT1 cells expressing PB1 constituitively at an MOI of 0.001 for 72h and 0.01 for 48h. 

### 2018-09-22 paper/paper.tex
updated methods after adding Figure 6

@@
 \subsection{qPCR for viral genes and IFN}
-We performed qPCR for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
-For the qPCR, we used the SYBR Green PCR Master Mix (Thermo Fisher) according to the manufacturer's protocol.
+We performed qPCR on reverse-transcribed mRNA for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
+For the qPCR, we used the SYBR Green PCR Master Mix (Thermo Fisher) according to the manufacturer's protocol using oligo-dT primers.
 The qPCR primers were: HA primer 1, 5'-\texttt{GGCCCAACCACACATTCAAC}-3'; HA primer 2, 5'-\texttt{GCTCATCACTGCTAGACGGG}-3'; IFNB1 primer 1, 5'-\texttt{AAACTCATGAGCAGTCTGCA}-3'; IFNB1 primer 2, 5'-\texttt{AGGAGATCTTCAGTTTCGGAGG}-3'; L32 primer 1, 5'-\texttt{AGCTCCCAAAAATAGACGCAC}-3'; L32 primer 2, 5'-\texttt{TTCATAGCAGTAGGCACAAAGG}-3'.
@@
 
-\abrcomment{Assuming we use the cycloheximide and ribavirin data as-is, can be modified if not}
-For figures \abrcomment{whichever figures} A549 cells were seeded at a density of 10,000 cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+For the qPCR in \FIGSUPP[validation]{delqPCR}, A549 cells were seeded at a density of 10,000 cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at the indicated MOI.
-Viral infections were normalized to an MOI of a previously-characterized relatively-deletion free stock of virus~\citep{russell2018extreme}, with all other strains matched for viral genome copy number as calculated by qPCR, described below.
-Crucially, by controlling for genome copy number rather than infectious units we avoid the confounding problem of differing immune stimulation by simply infecting at a higher rate of biologically active but non-infectious particles.
-For experimental treatments with ribavirin and cycloheximde, these compounds were added to cells at the time of infection to a final concentration of 100$\mu$M and 50$\mu$g/ml, respectively, both of which are sufficient to block downstream viral replication~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg,killip2014activation}.
-At 8 hours post infection media was removed and cell lysates generated using the CellAmp RT-PCR for Direct Prep Kit (Real Time) \& Protein Analysis, following manufacturer's instructions. 
-cDNA was synthesized using an oligoDT primer and the SuperScript III first-strand synthesis supermix from ThermoFisher using the manufacturer's protocol.
+For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml~\citep[a concentration sufficient to block secondary transcription;][]{killip2014activation} at the time of infection.
+After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.
 
-\subsection{Viruses for validation experiments}
-We generated the mutant viruses for the validation experiments in \FIG{validation} by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wildtype plasmid for that gene.
+\jdbcomment{These details can be added once the appropriate figure is in place: ribavirin was added to cells at the time of infection to a final concentration of 100$\mu$M, respectively, which is sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.}
+
+\subsection{Viruses and experiments for validation experiments}
+In \FIG{validation}, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
+Four of these viruses contained point mutations in PB1, two contained point mutations in NS, and one contained an internal deletion in PB1.
+In addition, we created a virus with an inactivated NS1 to mimic the infections that failed to express NS (it was not possible to generate a viral stock that completely lacked the NS segment).
+This NS1stop virus contained six nucleotide changes resulting in the addition of five in-frame stop codons in NS1 starting 10 nucleotides downstream of the 5' splice donor site, thereby disrupting NS1 while leaving NS2 (NEP) intact.
+All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\citep{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
+ \FIGDATA[validation]{sequences} provides the full sequences for all of these plasmids.
+
+We generated the wildtype and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wildtype plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
 At 24 hours post-transfection, we changed media from D10 to influenza growth media. 
-At 48 to 72 hours post-transfection, we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
-We then thawed an aliquot of each virus and titered by TCID50 as described above.
+At 50 hours post-transfection (for the replicate 1 viruses in \FIGSUPP[validation]{SNPflow}) or 72 hours (for the replicate 2 viruses in \FIGSUPP[validation]{SNPflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
+We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
+For the infections in \FIGSUPP[validation]{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent HI titer on \jdbcomment{red blood cells from what species where, and is there citation to protocol?}.
+We then performed infections of the A549 reporter cell line at equivalent HI titer and analyzed the data as described in \FIGSUPP[validation]{SNPflow}.
 
-For comparisons of interferon stimulation a relatively pure stock of WSN influenza was derived by harvesting reverse-genetics supernatants at 48h post infection and passaging virus on MDCK-SIAT1 cells for 36h at an MOI of 0.05---conditions which we previously validated to lead to stocks relatively depleted in defective particles~\citep{russell2018extreme}.
+To generate the NS1stop mutant virus and the wildtype and PB1del385to2163 mutant viruses in \FIGSUPP[validation]{delflow}, we used slightly different procedures.
+The wildtype virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
+The NS1stop virus was similarly generated, but was passaged for 48 rather than 36 hours, since it had slower growth kinetics and so needed a longer period of time to reach high titers.
+The viruses with deletions in the PB1 segment could not be generated in normal 293T and MDCK-SIAT1 cells, since they required the exogenous expression of the PB1 protein.
+Therefore, these viruses were generated in previously described 293T and MDCK-SIAT1 cells that had been engineered to constitutively express PB1~\citep{bloom2010permissive}.
+These viruses were harvested from transfections at 72 hours, and passaged twice in the MDCK-SIAT1 cells constitutively expressing PB1 at a MOI of 0.001 for 72 hours and 0.01 for 48 hours. 
+This passaging was necessary as viral titers from transfections were too low to generate sufficient virus from a single passage.
+The wildtype and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
+The infections in \FIGSUPP[validation]{delflow} were performed at equivalent TCID50s as described in the legend to that figure.
+That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \FIGSUPP[validation]{delqPCR}.
 
-To generate an NS1-inactivated strain of influenza, six nucleotide changes resulting in the addition of five in-frame stop codons starting 10 nucleotides downstream of the 5' splice donor site, interrupting translation of NS1 while leaving NS2 (NEP) intact.
-The viral RNA produced by this vector is described in (sequence)\abrcomment{I have uploaded the genbank files, just waiting for appropriate figure to append}.
-This modified plasmid was used in a reverse-genetics approach to recapitulate mutant virus, harvesting transfection supernatant after 72h.
-This virus was then passaged in MDCK-SIAT1 cells for 48h at an MOI of 0.05---due to slower growth kinetics than the wild-type virus this longer incubation was necessary.
-Genome copy number of this virus relative to infectious titer was determined as described below by qPCR.
-
-To generate viruses bearing internal deletions in the PB1 segment, indicated regions of the PB1 segment were excised by PCR from our reverse genetics plasmids.
-Sequences are described in (sequence) \abrcomment{I have uploaded the genbank files and modified coordinates in map names to match mRNA coordinates, just waiting for appropriate figure to append}. 
-To rescue these viruses, we used 293t cells and MDCK-SIAT1 cells transduced to constituitively express PB1 as previously described~\citep{bloom2010permissive}. 
-Virus was harvested from transfections at 72h, and passaged twice in MDCK-SIAT1 cells expressing PB1 constituitively at an MOI of 0.001 for 72h and 0.01 for 48h. 
-This passaging was necessary as viral titers from transfections were too low to generate sufficient virus from a single passage.
-This virus was titered on complementing MDCK-SIAT1 cells.
-Genome copy number of this virus relative to infectious titer was determined as described below by qPCR.
-
+\subsection{\jdbcomment{Not yet used, can add when section with this virus is added}}
 To generate pseudovirus that expresses fluorescent mCherry instead of HA, a plasmid was generated similarly to a previously-described eGFP-expressing pseudovirus~\citep{russell2018extreme}. 
@@
 
-Viruses bearing potentially innate-immune stimulatory SNPs as identified in our single-cell analyses were recapitulated by first introducing these mutations to plasmids encoding the viral segments PB1 or NS by PCR.
-Sequences of these plasmids are described in (sequence).
-\abrcomment{I have uploaded the genbank files and modified coordinates in map names to match mRNA coordinates, just waiting for appropriate figure to append}.
-These viruses were rescued via reverse-transfection in a coculture of MDCK-SIAT1 and 293t cells alongside a matched wild-type control.
-Each virus was rescued in two seperate transfection reactions. 
-Transfection supernatants were harvested at either 72 hours post transfection (biological replicate \#2 of all SNPs tested) or 48-50h post infection (biological replicate \#1 of all SNPs tested and both replicates of wild-type influenza).
-Virus recovered from transfections was titered by hemagglutination, and MOI set to infectious dose as measured in the wild-type virus as measured by TCID50 on MDCK-SIAT1 cells. 
-Particle-normalized viruses recovered from transfections were used directly in experiments in \abrcomment{Figure}.
-
-\subsection{flow cytometry analysis of interferon induction}
-Our A549 lambda reporter cell line was seeded in D10 media 24h prior to infection.
-At the time of infection, D10 media was replaced with influenza growth media and virus added to the indicated titers.
-Cells were harvested 13hpi and stained for HA as described above.
+\subsection{\jdbcomment{NS1 staining}}
+\jdbcomment{Most details about flow are now in figure supplements, but we do need to add details on NS1 staining after we see how that goes into paper}
 For NS1 staining, HA-stained cells were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 $\mu$g/ml, washed, stained with a goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405, washed, and analyzed.
-After channels were compensated and cells gated to exclude multiplets and debris in FlowJo, data were analyzed using custom python pipelines described in \abrcomment{wherever we put those notebooks}.
-Briefly, gating thresholds for interferon, mCherry, and HA staining were set to 99.9\% of an uninfected control.
+After channels were compensated and cells gated to exclude multiplets and debris in FlowJo, data were analyzed using custom python pipelines described in 
 For NS1, owing to sample-to-sample staining variation, thresholds were dynamically set on an individual sample to encompass an equal proportion of HA-expressing and NS1-expressing cells.

### 2018-09-24 paper/paper.tex
added hemagglutination details

@@
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \FIGSUPP[validation]{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent HI titer on \jdbcomment{red blood cells from what species where, and is there citation to protocol?}.
-We then performed infections of the A549 reporter cell line at equivalent HI titer and analyzed the data as described in \FIGSUPP[validation]{SNPflow}.
+For the infections in \FIGSUPP[validation]{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells ~\citep{hirs1941tHI}.
+ Briefly, a solution of 10\% v/v red blood cells was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
+Two-fold serial dilutions of virus were added to an equal volume of diluted red blood cells, and titer was measured as the highest dilution of viral stock at which complete hemagglutination of red blood cells was observed.
+We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \FIGSUPP[validation]{SNPflow}.
 

### 2018-09-24 paper/paper.tex
Added source of RBCs

@@
 For the infections in \FIGSUPP[validation]{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells ~\citep{hirs1941tHI}.
- Briefly, a solution of 10\% v/v red blood cells was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
+ Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
 Two-fold serial dilutions of virus were added to an equal volume of diluted red blood cells, and titer was measured as the highest dilution of viral stock at which complete hemagglutination of red blood cells was observed.

### 2018-09-25 paper/paper.tex
minor tweaks to Results text, start of validation text

@@
 
-The foregoing transcriptomic analyses identify which viral genes are expressed in each cell, but they do not provide information on the sequences of these genes.
-This is a major shortcoming, since influenza's high mutation rate means that although the \emph{consensus} sequence of the viral stock used in our experiment corresponds to WSN influenza, the specific virions infecting individual cells will often carry mutations~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
-We therefore used the approach schematized in \FIG{workflow}D to determine the full-length sequences of the viral genes expressed in the majority of infected cells.
+The foregoing transcriptomic analyses identify which viral genes are expressed in each cell, but not the sequences of these genes.
+This is a major shortcoming, since influenza's high mutation rate means that although the \emph{consensus} sequence of the viral stock corresponds to WSN influenza, the virions infecting individual cells will often carry mutations~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+We therefore used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in the majority of infected cells.
 
 By employing several PCR enrichment strategies (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to influenza genes (\FIGSUPP[genotypes]{CCSs}).
-Crucially, the fact that about half the infecting virions encoded genes that were tagged with synonymous viral barcodes at both termini enabled us to estimate the rate of PCR strand exchange among cDNAs that we amplified for PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange among PacBio-sequenced cDNAs (\FIGSUPP[genotypes]{StrandExchange}).
 Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link mutations in viral transcripts to cell barcodes and UMIs that identify the cell and molecule of origin.
@@
 We used these CCSs to call the sequences of viral genes in single cells.
-We called a mutation as present in a particular cell if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
+We called a mutation if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
 For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
-This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant (estimated to be $\sim$75\% of infected cells), as well as identifying high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
-The strategy will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency among viral transcripts in that cell.
-Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is an inherent limit to our detection threshold: we can identify viral mutations that are at high frequency within single infected cells, but not mutations that are present on just a small fraction of the viral transcripts in a cell.
+This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant (estimated to be $\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+The strategy will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
+Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is an inherent limit to our detection threshold: we can identify viral mutations that are at high frequency within single infected cells, but not mutations that occur on just a small fraction of the viral transcripts in a cell.
 
-We were able to call the sequences of all expressed viral genes in the majority of infected cells (\FIGSUPP[genotypes]{ncells}).
-We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were not co-infected by both viral barcode variants (\FIGSUPP[genotypes]{ncells}).
-But despite this bias towards calling full viral genotypes from cells with these characteristics, we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
+We could call the sequences of all expressed viral genes in the majority of infected cells (\FIGSUPP[genotypes]{ncells}).
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by by only one viral barcode variant (\FIGSUPP[genotypes]{ncells}).
+But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
 
@@
 \caption{
-Viral genetic variation associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.
+Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.
 {\bf (A)} 
-Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight viral genes, or fail to express an unmutated copy of one or more genes.
+Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
 Cells infected by full wildtype virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
 {\bf (B)}
-There is a trend for cells infected by full wildtype virus to express less IFN.
+Cells infected by full wildtype virus to express less IFN.
 {\bf (C)}
-Some specific forms of viral genetic variation are associated with IFN induction.
+Specific viral defects associated with IFN induction.
 The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
-The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each viral gene, conditioned on the cell expressing that gene.
-Numbers above the bars give P-values (Fisher's exact test) and associated Q-values~\citep{storey2003statistical} for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
-\FIGSUPP[mutations]{allcells} shows that the trends remain largely the same if we include infected cells with incomplete viral sequence information. 
-\FIGSUPP[mutations]{isg} shows that the trends remain largely the same if we categorize cells by ISG expression rather than IFN expression.
+The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
+Numbers give P-values (Fisher's exact test) and associated Q-values~\citep{storey2003statistical} for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
+\FIGSUPP[mutations]{allcells} shows the trends are similar if we include infected cells with incomplete viral sequence information. 
+\FIGSUPP[mutations]{isg} shows the trends are similar if we categorize cells by ISG expression rather than IFN expression.
 {\bf (D)}
@@
 
-To more systematically assess how viral genetic variation associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
+To more systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
 \FIG{mutations}A shows that the 49 cells infected by full wildtype viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
@@
 
-Viral genetic variation also associated with IFN induction, since cells infected by full wildtype virions less IFN than cells infected by incomplete or mutated virions (\FIG{mutations}B).
-Furthermore, specific viral genetic defects were associated with IFN induction (\FIG{mutations}C).
+Viral genetic variation also associated with IFN induction, since cells infected by full wildtype virions expressed less IFN than cells infected by incomplete or mutated virions (\FIG{mutations}B).
+Specific viral genetic defects were especially associated with IFN induction (\FIG{mutations}C).
 For instance, absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were more weakly enriched.
-Only the absence of NS remained associated with IFN induction at a false discovery rate of 10\% (\FIG{mutations}C), although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent since the association of defects in certain genes with IFN induction is more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
-In any case, the modest number of cells with full viral genotype information precludes drawing strong conclusions from statistical analyses alone---but \FIG{genotypes} and \FIG{mutations}C do motivate compelling hypotheses about specific viral mutations that contribute to IFN induction, and in the next few sections, we pursue these hypotheses using experimental follow-up.
+Only the absence of NS remained associated with IFN induction at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are much more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
+In any case, \FIG{genotypes} and \FIG{mutations}C motivate compelling hypotheses about specific viral mutations that contribute to IFN induction, and below we pursue these hypotheses using experimental follow-up.
 
@@
 Similar to the first panel, but validates the increased rate of IFN induction for a large internal deletion in the PB1 gene.
-See \FIGSUPP[validation]{delflow} for detailed flow cytometry data, and \FIGSUPP[validation]{delqPCR} for details of how we validated that the infections were performed with a similar doses of each viral mutant.
-The experiments in the two panels were performed on different days, and so exact numerical values can be reliably compared within each panel but not between panels.
+See \FIGSUPP[validation]{delflow} and \FIGSUPP[validation]{delqPCR} for more detailed experimental data.
+The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
 }
@@
 
-\figdata{Genbank plasmid maps for the mutant genes in this figure cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
+\figdata{Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
 \label{figdata:sequences}
@@
 %%% end validation figure %%%
+
+To test if the viral defects identified in single IFN+ cells played a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these mutations.
+The greatest enrichment of amino-acid substitutions in IFN+ cells was in the PB1 and NS genes (\FIG{mutations}C), so we...
 

### 2018-09-26 paper/paper.tex
tiny bit more text in Results validation section

@@
 
-\subsection{The full genotypes of viruses in single IFN+ and IFN- cells}
+\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells}
 
@@
 
-\subsection{Systematic examination of viral features associated with infection outcome}
+\subsection{Viral features associated with infection outcome in single cells}
 
@@
 
-\subsection{Validation that multiple types of viral mutations increase IFN induction}
+\subsection{Defects in viruses infecting single IFN+ cells often increase IFN induction}
 
@@
 
-To test if the viral defects identified in single IFN+ cells played a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these mutations.
-The greatest enrichment of amino-acid substitutions in IFN+ cells was in the PB1 and NS genes (\FIG{mutations}C), so we...
+To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
+
+The viral feature most strongly associated with IFN induction in single cells was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
+Although it is sometimes possible to use complementing cell lines to generate influenza viruses lacking a specific viral gene~\citep{fujii2003selective,marsh2007specific}, we were unsuccessful in using this approach to generate viruses that lacked NS.
+The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza virus's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
+We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term \emph{NS1stop}) that had multiple stop codons early in the NS1 coding sequence.
+
+The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were associated with IFN induction (\FIG{genotypes}, \FIG{mutations}C), so we also created mutant viruses for each amino-acid substitution found in these genes among IFN+ cells that did not also lack the NS gene.
+Overall, there were six \jdbcomment{actually, seven with S704P} such point mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, \jdbcomment{what about PB1-S704P?} NS1-A122V, and NS2-E47G (\FIGSUPP[genotypes]{genotypes_by_ifn}).
 

### 2018-09-27 paper/paper.tex
text draft for validation section of Results

--- 2018-09-25 paper/paper.tex
@@
 
-\subsection{Validation that multiple types of viral mutations increase IFN induction}
+\subsection{Validation that viral defects in single IFN+ cells often cause increased IFN induction}
 
@@
 The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for the wildtype is indicated with a dashed blue line.
-The points are colored orange if the mutant virus stock induces IFN more frequently than the wildtype viral stock (t-test, $P < 0.01$), and blue if the frequency of IFN induction is not significantly different than wildtype at this cutoff.
+The points are colored orange if the mutant virus stock induces IFN more frequently than the wildtype viral stock (one-sided t-test, $P < 0.01$), and blue if the frequency of IFN induction is not significantly different than wildtype at this cutoff.
 {\bf (B)}
@@
 
-To test if the viral defects identified in single IFN+ cells played a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these mutations.
-The greatest enrichment of amino-acid substitutions in IFN+ cells was in the PB1 and NS genes (\FIG{mutations}C), so we...
+To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
+
+The viral defect most strongly associated with IFN induction in single cells was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
+Although it is sometimes possible to use complementing cell lines to generate influenza viruses lacking a specific viral gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
+The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza virus's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
+We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
+
+The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of amino-acid substitutions found in these genes among IFN+ cells.
+Overall, we created six such point-mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
+
+Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a large deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+We therefore created a virus carrying this deletion that we were able to propagate in cells constitutively expressing the PB1 protein~\citep{bloom2010permissive}.
+
+We next tested the rate of IFN induction by each viral stock using the reporter cell lines.
+As shown in \FIG{validation}, five of the eight mutant viruses induced IFN more frequently than a wildtype viral stock.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the six point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than the wildtype viral stock.
+These results show that the viral defects in single IFN+ cells often cause increased IFN production.
+
+It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of particular viral defects among the IFN+ cells in the single-cell data.
+However, these comparisons are confounded by the fact that the measurements in \FIG{validation} are made using \emph{bulk} stocks of each viral variant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
+For instance, our single-cell experiment used a ``wildtype'' viral stock, but the vast majority of IFN+ cells (31 of 40) still failed to express wildtype copies of all viral genes (\FIGSUPP[genotypes]{genotypes_by_ifn}).
+Therefore, although the increased IFN induction by most viral variant stocks in \FIG{validation} does indicate that these variants are causally associated with IFN, without single-cell virus-sequencing it remains impossible to know how many of the IFN+ cells for each viral stock are infected by virions that also have other defects.
 
@@
 In \FIG{validation}, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
-Four of these viruses contained point mutations in PB1, two contained point mutations in NS, and one contained an internal deletion in PB1.
-In addition, we created a virus with an inactivated NS1 to mimic the infections that failed to express NS (it was not possible to generate a viral stock that completely lacked the NS segment).
+For point-mutant viruses, we created variants for all amino-acid substitutions found in PB1 and NS among IFN+ cells that did not also lack NS.
+One of these mutants (amino-acid substitution S704P in PB1) did not reach sufficient titers in a single attempt to generate it by reverse genetics, and so was dropped from the experiment (note that we did not attempt replicates of the reverse genetics for this mutant, and so are \emph{not} confident in drawing strong conclusions about its actual attenuation).
+This left six point-mutant viruses: four with point mutations in PB1, and two with point mutations in NS.
+We also created a mutant virus that contained the internal deletion in PB1 found in an IFN+ cell.
+In addition, we created a virus with an inactivated NS1 to mimic the infections that failed to express NS (we were unable to use complementing cells to generate a viral stock that completely lacked the NS segment).
 This NS1stop virus contained six nucleotide changes resulting in the addition of five in-frame stop codons in NS1 starting 10 nucleotides downstream of the 5' splice donor site, thereby disrupting NS1 while leaving NS2 (NEP) intact.

### 2018-09-28 paper/paper.tex
some modest tweaks to Results text

@@
 \author[1]{Alistair B. Russell}
+\author[1]{Jacob R. Kowalsky}
 \author[1,2,3*]{Jesse D. Bloom}
@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous at the cellular level: infected cells vary widely in the amount of viral material they produce, and only sometimes activate innate immunity.  
+The outcome of viral infection is extremely heterogeneous at the cellular level, and infected cells only sometimes activate innate immunity.  
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We do this by determining the cellular transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that trigger immunity in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
-However, immune activation remains stochastic in cells infected by viruses with these defects, and sometimes occurs even in cells infected with a virus that expresses unmutated copies of all genes.
-Our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation in single cells.
+We do this by developing a new approach to determine both the cellular transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Infections that activate an interferon response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
+However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
+Our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation in single influenza-infected cells.
 \end{abstract}
@@
 The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
-We created reporters with type I and type III promoters (\FIGDATA[IFNrare]{reporter_sequences}).
+We created reporters with type I and type III IFN promoters (\FIGDATA[IFNrare]{reporter_sequences}).
 The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
@@
 These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
-Cells that activate IFN express the cell-surface protein, enabling them to be enriched by magnetic-activated cell sorting (MACS).
-The frequency of IFN+ cells can also be determined by flow cytometry for the fluorescent protein.
-We validated that the reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
-We also found that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by our single-cell transcriptomics described below).
-Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.
+Cells that activate the IFN reporter can enriched by magnetic-activated cell sorting (MACS) for the cell-surface protein, or identified by flow cytometry for the fluorescent protein.
+The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics described below).
+Therefore, for the rest of this paper, we will use ``IFN expression'' to refer to the combined expression of type I and III IFNs.
 
@@
 Standard single-cell transcriptomic approaches are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
-These measurements of transcriptomic abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
+Such measurements of transcriptomic abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
 
@@
 \caption{
-Approach for simultaneous transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
+Approach for combined transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
 {\bf (A)}
@@
 
-We therefore used the approach in \FIG{workflow} to simultaneously determine the entire transcriptome and the full sequences of the infecting virions in single cells.
-First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain of influenza: the wildtype virus and a ``synonymously barcoded'' variant that contains two engineered synonymous mutations near each end of each gene (\FIGDATA[workflow]{virus_seqs}).
-These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts.
+We therefore developed the approach in \FIG{workflow} to determine both the entire transcriptome \emph{and} the full sequences of the infecting virions in single cells.
+First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain of influenza: the wildtype virus and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
+These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 
-We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{workflow}A) at a dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
-From 12 to 13 hours post-infection, we used magnetic columns to enrich cells that expressed the cell-surface protein driven by the IFN reporter.
+We next used this viral stock to infect A549 cells carrying an IFN reporter (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of the cells.
+From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter.
 Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to $\sim$5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}).
-To ensure the presence of IFN- cells after the MACS enrichment, we added back non-sorted cells to $\sim$10\% of the total.
-We also added uninfected canine cells to $\sim$5\% of the total as an internal control for multiplets and to estimate the background amount of viral mRNA detected in the transcriptomes of truly uninfected cells.
+To ensure the presence of IFN- cells after the MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
+We also added uninfected canine cells to $\sim$5\% of the total as an internal control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
 
@@
 
-We obtained transcriptomes for 1,614 human (A549) cells (some of which were infected with influenza), and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
+We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
 We also obtained 12 transcriptomes with a mix of human and canine transcripts; these mixes arise when multiple cells are captured in a single droplet.
@@
 Using this test, we classified 290 of the 1,490 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
-We classified the other cells as uninfected, although we cannot rule out that some of them were infected with virions that produced too little mRNA to be detected in our experiment.
+We classified the other cells as uninfected, although it is possible that some of them were infected with virions that produced too little mRNA to be detected in our experiment.
 The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
-As in our prior work~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but for a few cells viral mRNA comprises over half the transcriptome.
+As in our prior work~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
 
@@
 The difference is sufficiently modest that it could simply be due to methodological differences in the thresholds used to call gene presence.
-But it is important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could conceivably alter the distribution of viral gene expression from a fully random sample of infected cells.
+But it is also important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could alter the distribution of viral gene expression from a fully random sample of infected cells.
 
@@
 
-By employing several PCR enrichment strategies (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to influenza genes (\FIGSUPP[genotypes]{CCSs}).
+By employing several PCR enrichment strategies (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
 Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange among PacBio-sequenced cDNAs (\FIGSUPP[genotypes]{StrandExchange}).
-Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link mutations in viral transcripts to cell barcodes and UMIs that identify the cell and molecule of origin.
+Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
 
@@
 We could call the sequences of all expressed viral genes in the majority of infected cells (\FIGSUPP[genotypes]{ncells}).
-We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by by only one viral barcode variant (\FIGSUPP[genotypes]{ncells}).
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\FIGSUPP[genotypes]{ncells}).
 But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
@@
 Simple visual inspection of this figure reveals a wealth of information.
-For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed all eight genes and did not induce detectable IFN.
+For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wildtype copies all eight genes and did not induce detectable IFN.
 But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that either had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
-The two cells that expressed the most IFN in our experiment (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS segment, which encodes influenza virus's primary innate immune antagonist.
-Many other IFN+ cells had different defects, including large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
-However, \FIG{genotypes} also shows that expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN.
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN.
 
@@
 Viral genetic variation also associated with IFN induction, since cells infected by full wildtype virions expressed less IFN than cells infected by incomplete or mutated virions (\FIG{mutations}B).
-Specific viral genetic defects were especially associated with IFN induction (\FIG{mutations}C).
+Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
 For instance, absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were more weakly enriched.
-Only the absence of NS remained associated with IFN induction at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are much more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
+Only the absence of NS remained significant at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are much more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
 In any case, \FIG{genotypes} and \FIG{mutations}C motivate compelling hypotheses about specific viral mutations that contribute to IFN induction, and below we pursue these hypotheses using experimental follow-up.
 
-One other interesting trend emerges from the combined single-cell data.
+One other interesting trend emerges from the single-cell data.
 There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
@@
 The numbers indicate the median of four measurements for each viral mutant.
-The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for the wildtype is indicated with a dashed blue line.
-The points are colored orange if the mutant virus stock induces IFN more frequently than the wildtype viral stock (one-sided t-test, $P < 0.01$), and blue if the frequency of IFN induction is not significantly different than wildtype at this cutoff.
+The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wildtype is indicated with a dashed blue line.
+Points are colored orange if the mutant virus stock induces IFN more frequently than the wildtype viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
 {\bf (B)}
-Similar to the first panel, but validates the increased rate of IFN induction for a large internal deletion in the PB1 gene.
+Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
 See \FIGSUPP[validation]{delflow} and \FIGSUPP[validation]{delqPCR} for more detailed experimental data.
@@
 
-The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of amino-acid substitutions found in these genes among IFN+ cells.
+The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of the amino-acid substitutions found in these genes among IFN+ cells.
 Overall, we created six such point-mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
@@
 Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a large deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
-We therefore created a virus carrying this deletion that we were able to propagate in cells constitutively expressing the PB1 protein~\citep{bloom2010permissive}.
+We therefore created a virus carrying this deletion that we were able to propagate in cells constitutively expressing the PB1 protein.
 
 We next tested the rate of IFN induction by each viral stock using the reporter cell lines.
-As shown in \FIG{validation}, five of the eight mutant viruses induced IFN more frequently than a wildtype viral stock.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the six point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than the wildtype viral stock.
+As shown in \FIG{validation}, five of the eight mutant viruses induced IFN more frequently than wildtype.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the six point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wildtype.
 These results show that the viral defects in single IFN+ cells often cause increased IFN production.
@@
 For instance, our single-cell experiment used a ``wildtype'' viral stock, but the vast majority of IFN+ cells (31 of 40) still failed to express wildtype copies of all viral genes (\FIGSUPP[genotypes]{genotypes_by_ifn}).
-Therefore, although the increased IFN induction by most viral variant stocks in \FIG{validation} does indicate that these variants are causally associated with IFN, without single-cell virus-sequencing it remains impossible to know how many of the IFN+ cells for each viral stock are infected by virions that also have other defects.
+Therefore, although the increased IFN induction by most viral variant stocks in \FIG{validation} does indicate that these variants are causally associated with IFN, without single-cell virus-sequencing it is impossible to know how many of the IFN+ cells induced by each viral stock are infected by virions that also have other defects.
+For instance, most of the 0.67\% of cells infected with the ``wildtype'' viral stock that turned IFN+ in \FIG{validation}A are probably infected by virions that have some type of genetic defect.
 
@@
 \subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
-The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the LNGFR$\Delta$C-P2A-mNeonGreen reporter.
+The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-P2A-mNeonGreen reporter.
 A schematic of the experiment is shown in \FIG{workflow}.

### 2018-09-30 paper/paper.tex
added viral burden vs IFN for NS-deficient viruses (first part Fig 7)

@@
 Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wildtype and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
-Then qPCR was used to show that the ratio of viral RNA to TCID50 was similar for all viral variants (data not shown).
-Finally, the approach in \FIGSUPP[validation]{delqPCR} was used to show that all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
+Finally, the approach in \FIGSUPP[validation]{delqPCR} was used to show that equivalent TCID50s, all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
 All this normalization suggests that all infections were performed at a similar dose of particles active for primary transcription.
@@
 {Validation that the infections in \FIG{validation}B and \FIGSUPP[validation]{delflow} were performed at similar doses of virions capable of initiating primary transcription.
-In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \FIGSUPP[validation]{delflow}), and then mRNA was harvested for qPCR on oligo-dT primed reverse transcription.
+In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \FIGSUPP[validation]{delflow}), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins.
+These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins.
 In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
@@
 For instance, most of the 0.67\% of cells infected with the ``wildtype'' viral stock that turned IFN+ in \FIG{validation}A are probably infected by virions that have some type of genetic defect.
+
+\subsection{Distinct IFN-inducing characteristics of different viral variants}
+
+%%% begin mutantcomparison figure %%%
+\begin{figure}
+
+\centerline{\includegraphics[width=0.85\textwidth]{figures/MutantComparison/p_merged.pdf}}
+\caption{
+For NS1-deficient infections, higher viral gene expression causes more IFN induction.
+{\bf (A)}
+Flow cytometry shows that for viruses with defects in NS1, there is increased IFN induction among infected cells that express high levels of HA protein relative to those that express low levels of HA.
+However, for wildtype virus and viruses with mutations in PB1, there is only a modest increase in IFN induction with increasing HA.
+The y-axis shows the ratio of the percent of cells that are IFN+ in the highest HA-expression quartile relative to the lowest HA-expression quartile.
+Points indicate four replicates, and lines indicate the mean.
+See \FIGSUPP[mutantcomparison]{ha_vs_ifn} for details.
+{\bf (B)}
+Higher viral gene expression \emph{causes} more IFN induction for NS1-deficient infections, since blocking viral secondary transcription strongly decreases IFN levels in cells infected with NS1stop virus, but not cells infected with the virus carrying a deletion in PB1.
+A549 cells were infected with both viral stocks at doses that induced equivalent IFN, and ribavirin was added to half the samples at a concentration (100 $\mu$M) sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.
+After eight hours, IFNB1 transcript was quantified relative to the housekeeping gene L32 by qPCR. 
+Points indicate four replicates, and lines indicate the mean.
+The dashed gray line indicates the limit of detection.
+\FIGSUPP[mutantcomparison]{cycloheximide} shows that similar results are obtained if secondary transcription is blocked with cycloheximide rather than ribavirin.
+}
+\label{fig:mutantcomparison}
+
+\figsupp[Flow cytometry analysis showing that viral burden strongly correlates with IFN induction in NS1-deficient infections.]
+{Re-analysis of the flow cytometry data in \FIGSUPP[validation]{SNPflow} indicates that a higher viral burden is associated with increased IFN-induction in NS1-deficient infections.
+We re-analyzed all the infected cells (defining ``infected'' using the same HA-gating scheme in \FIGSUPP[validation]{SNPflow}) for the wildtype virus, the NS1stop virus, and the three amino-acid mutants with significantly increased IFN induction.
+For each virus and replicate, we binned the infected cells into HA expression quartiles.
+We then calculated the percent of cells that were IFN+ in each quartile (defining ``IFN+'' using the same gating scheme in \FIGSUPP[validation]{SNPflow}).
+The plots show the mean HA expression in each quartile versus the percent of cells that are IFN+.
+The results clearly indicate that for the NS1stop virus, increased viral burden (more HA) correlates with IFN induction.
+A similar trend also holds for the virus with an amino-acid substitution in NS1 (NS1-A122V). 
+But for wildtype and the two PB1 mutants, there is no clear trend for higher viral burden to correlate with IFN induction.
+\FIG{mutantcomparison}A summarizes the data in this figure by simply showing the ratio of the percent IFN+ of the highest HA-expression quartile to the lowest HA-expression quartile for each sample and replicate.}
+{\includegraphics[width=0.9\textwidth]{figures/MutantComparison/p_ifn_vs_ha.pdf}}
+\label{figsupp:ha_vs_ifn}
+
+\figsupp[Similar to \FIG{mutantcomparison}B, but uses cycloheximide rather than ribavirin.]
+{Preventing viral secondary transcription by addition of the cycloheximide also decreases the rate of IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
+However, the results are more complex here, since cycloheximide alone increases IFN production~\citep{raj1981analysis, maroteaux1983cycloheximide, killip2014activation}.
+Therefore, we see an increase in IFN levels upon cycloheximide treatment for the uninfected cells and cells infected with the PB1 deletion virus, but a decrease in cells infected with the NS1stop virus.
+This plot shows qPCR results for the same samples as in \FIGSUPP[validation]{cycloheximide}, but with the qPCR for expression of IFNB1 transcripts relative to the housekeeping gene L32.
+As described in \FIGSUPP[validation]{cycloheximide}, A549 cells were infected at an MOI of 0.4, such that HA expression was equivalent between viral variants when secondary transcription was blocked, and some of the samples treated with 50 $\mu$g/ml of cycloheximide.
+After 8 hours, transcript levels were quantified by qPCR.
+Points indicate four replicates, and lines indicate the mean.
+The dashed line shows the limit of detection, and points below this limit where set to that value.
+}
+{\includegraphics[width=0.6\textwidth]{figures/MutantComparison/cyclo_qPCR_plot.pdf}}
+\label{figsupp:cycloheximide}
+
+\end{figure}
+%%% end mutantcomparison figure%%%
 
@@
 We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\citep{xue2016propagation}.
-The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 in influenza growth media ((Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
+The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
 Specifically, after generation by reverse genetics, the wildtype variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
@@
 
-For titering virus by qPCR, vRNA was harvested from 80 $\mu$l of viral supernatant by the addition of 600 $\mu$l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini Kit protocol. 
-The cDNA was generated using SuperScript III First-Strand Synthesis SuperMix (Thermo Fisher) using the manufacturer's protocol, and using the universal vRNA primers of \citet{hoffmann2001universal} with the modifications described in \citet{xue2017parallel}.
-The number of HA molecules were then quantified by qPCR to measure genome dose.
-
-For the qPCR in \FIGSUPP[validation]{delqPCR}, A549 cells were seeded at a density of 10,000 cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
-Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at the indicated MOI.
+For the qPCR in \FIGSUPP[validation]{delqPCR} and \FIGSUPP[mutantcomparison]{cycloheximide}, A549 cells were seeded at a density of 10,000$10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
 For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml~\citep[a concentration sufficient to block secondary transcription;][]{killip2014activation} at the time of infection.
@@
 
-\jdbcomment{These details can be added once the appropriate figure is in place: ribavirin was added to cells at the time of infection to a final concentration of 100$\mu$M, respectively, which is sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.}
+For the qPCR in \FIG{mutantcomparison}B, A549 cells were infected with the NS1stop virus at a MOI of 0.1 (based on TCID50) in MDCK-SIAT1 and with the PB1del385to2163 virus at a dose that induced equivalent IFN in the absence of ribavirin.
+The ribavirin was added to cells at the time of infection to a final concentration of 100$\mu$M, which is sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.
+All other steps were performed as for the cycloheximide experiments described immediately above..
 
@@
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \FIGSUPP[validation]{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells ~\citep{hirs1941tHI}.
+For the infections in \FIGSUPP[validation]{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
@@
 This virus was titered on complementing MDCK-SIAT1 cells.
-
-\subsection{\jdbcomment{NS1 staining}}
-\jdbcomment{Most details about flow are now in figure supplements, but we do need to add details on NS1 staining after we see how that goes into paper}
-For NS1 staining, HA-stained cells were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 $\mu$g/ml, washed, stained with a goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405, washed, and analyzed.
-After channels were compensated and cells gated to exclude multiplets and debris in FlowJo, data were analyzed using custom python pipelines described in 
-For NS1, owing to sample-to-sample staining variation, thresholds were dynamically set on an individual sample to encompass an equal proportion of HA-expressing and NS1-expressing cells.
 

### 2018-10-02 paper/paper.tex
draft of Introduction

@@
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We do this by developing a new approach to determine both the cellular transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that activate an interferon response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
+We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
 However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
-Our work shows that viral genetic variation substantially contributes to but does not fully explain heterogeneity in infection outcome and immune activation in single influenza-infected cells.
+Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
 \end{abstract}
@@
 \section{Introduction}
+Infection with an acute virus such as influenza initiates a race between the virus and immune system.
+As the virus spreads, some infected cells trigger innate immunity and begin producing interferon~\citep[IFN;][]{iwasaki2014innate}.
+This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}.
 
-\citep{russell2018extreme} and \citep{steuerman2018dissection}.
+Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells~\citep{kallfass2013visualizing, killip2017single}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
+Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
-The antiviral IFNs~\citep[type I and type III IFNs;][]{kotenko2017contribution}
-Early in infection
+It remains unclear what determines if an infected cell triggers an innate-immune response.
+Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
+For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
+But for influenza, a third possible contributor also looms large: viral genetic diversity.
+Specifically, because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have genetic defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and more IFN is induced by viral stocks that contain defective virions with internal deletions in the polymerase genes~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+
+However, existing single-cell techniques are inadequate to determine how viral genetic diversity actually contributes to cell-to-cell heterogeneity during infection, since they do not systematically determine the full sequences of the virions that infect individual cells.
+Flow cytometry and fluorescent reporters only measure expression of viral proteins~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of viral RNAs~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
+None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations will be common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.
+
+Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells that do and do not activate innate immunity.
+To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
+We obtain the transcriptomes and full sequences of all expressed viral genes in 150 virus-infected cells, 49 of which express IFN.
+Even though we used a nominally ``wildtype'' stock of virus, two-thirds of the cells are infected by virions that have a mutation or defect in gene expression.
+This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
+Viral defects are especially common in IFN+ cells, and we identify at least four defects that are causally associated with IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
+However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations that we identify deterministically induce IFN.
 
@@
 \subsection{A system to identify and enrich for rare IFN+ positive cells}
-Influenza infection only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
+Influenza infection only rarely triggers expression of the anti-viral IFNs~\citep[type I and type III IFNs;][]{kotenko2017contribution} at the single-cell level~\citep{killip2017single}.
 For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
@@
 
+\jdbcomment{Section needs to be completed after remaining data on co-infections.}
+
 \section{Discussion}
 
-Add text
+\jdbcomment{Section needs to be complete.
+Add this:
+Approach builds on observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using more standard single-cell techniques, and is similar the approach used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
+}
 

### 2018-10-02 paper/paper.tex
some very minor tweaks to Intro text

@@
 
-It remains unclear what determines if an infected cell triggers an innate-immune response.
+It remains unclear why only some infected cells trigger innate-immune responses.
 Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
@@
 However, existing single-cell techniques are inadequate to determine how viral genetic diversity actually contributes to cell-to-cell heterogeneity during infection, since they do not systematically determine the full sequences of the virions that infect individual cells.
-Flow cytometry and fluorescent reporters only measure expression of viral proteins~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of viral RNAs~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
+Flow cytometry and fluorescent reporters only measure expression of viral proteins~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of viral transcripts~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations will be common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.
 
-Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells that do and do not activate innate immunity.
+Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells.
 To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
@@
 This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
-Viral defects are especially common in IFN+ cells, and we identify at least four defects that are causally associated with IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
+Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
 However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations that we identify deterministically induce IFN.
+Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
 

### 2018-10-02 paper/paper.tex
very minor tweaks to first 2 sections of Results

@@
 
-The rareness of IFN induction by influenza infection is not merely a feature of A549 cells in culture.
-Prior studies using fluorescent reporters have found that IFN expression is rare at the single-cell level in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
-In further support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
+This rareness of IFN induction is not merely a feature of A549 cells in culture.
+Prior studies using reporters have found that IFN expression is rare at the single-cell level in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
+In support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
 As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes (ISGs), which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
-But only a handful of infected cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
+But just a handful of cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
 
@@
 The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics described below).
-Therefore, for the rest of this paper, we will use ``IFN expression'' to refer to the combined expression of type I and III IFNs.
+Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
-We used the reporter cells to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
+We used the reporter cells to quantify the frequency of IFN induction in A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
 As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
-This frequency of IFN induction is roughly comparable to what we had observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B).
+This frequency of IFN induction is roughly comparable to that observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B).
 
 \subsection{Combined transcriptomics and virus-sequencing of single infected cells}
-We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
-Standard single-cell transcriptomic approaches are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
-Such measurements of transcriptomic abundance do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
+We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of infection in single cells.
+Standard single-cell transcriptomic techniques are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to measure the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}, and do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
 
@@
 
-We therefore developed the approach in \FIG{workflow} to determine both the entire transcriptome \emph{and} the full sequences of the infecting virions in single cells.
-First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain of influenza: the wildtype virus and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
+We therefore developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of the infecting virions in single cells.
+First, we generated a stock of virus that consisted of a mix of wildtype WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
 These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 
-We next used this viral stock to infect A549 cells carrying an IFN reporter (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of the cells.
+We next used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of the cells.
 From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter.
-Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to $\sim$5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}).
-To ensure the presence of IFN- cells after the MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
-We also added uninfected canine cells to $\sim$5\% of the total as an internal control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
+Based on pilot experiments, we expected this MACS to enrich IFN+ cells to $\sim$5\% to 30\% of the total (\FIGSUPP[workflow]{MACS}).
+To ensure the presence of IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
+We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
 
@@
 Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
-Combining this information with the 3'-end sequencing simultaneously determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).
+Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).
 

### 2018-10-02 paper/paper.tex
very minor tweaks 3rd section Results

@@
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}B).
-As expected, only a small fraction ($\sim$0.7\%) of the transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction.
-Using this test, we classified 290 of the 1,490 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
-We classified the other cells as uninfected, although it is possible that some of them were infected with virions that produced too little mRNA to be detected in our experiment.
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 of the 1,490 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
+We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
 The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
-As in our prior work~\citep{russell2018extreme}, the amount of viral mRNA is extremely heterogeneous: many infected cells have only a few percent of their mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
+As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
 
 We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
-Although ten transcripts are encoded by the eight viral genes, we call presence / absence at the level of genes rather than transcripts.
+Although ten transcripts are encoded by the eight viral genes, we call presence / absence of genes rather than transcripts.
 The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable in the 3'-end sequencing transcriptomic data.

### 2018-10-02 paper/paper.tex
very minor tweaks section 4 of Results

@@
 The foregoing transcriptomic analyses identify which viral genes are expressed in each cell, but not the sequences of these genes.
-This is a major shortcoming, since influenza's high mutation rate means that although the \emph{consensus} sequence of the viral stock corresponds to WSN influenza, the virions infecting individual cells will often carry mutations~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
 We therefore used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in the majority of infected cells.
@@
 By employing several PCR enrichment strategies (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange among PacBio-sequenced cDNAs (\FIGSUPP[genotypes]{StrandExchange}).
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange (\FIGSUPP[genotypes]{StrandExchange}).
 Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
@@
 For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
-This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant (estimated to be $\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
-The strategy will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
-Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is an inherent limit to our detection threshold: we can identify viral mutations that are at high frequency within single infected cells, but not mutations that occur on just a small fraction of the viral transcripts in a cell.
+This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
+Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of viral transcripts in a cell.
 

### 2018-10-02 paper/paper.tex
Immunopathology sentence in introduction

--- 2018-10-02 paper/paper.tex
@@
 For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
+Consistent with these observations, despite the beneficial antiviral effects of IFN signaling, these effects can come at a significant cost due to increased immunopathology and therefore host-cell intrinsic limitation of the IFN response would help to limit the damage done by this inflammatory process~\citep{iwasaki2014innate}.
 But for influenza, a third possible contributor also looms large: viral genetic diversity.

### 2018-10-02 paper/paper.tex
very minor tweaks Results section 5

--- 2018-10-02 paper/paper.tex
@@
 
-To more systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
+To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
 \FIG{mutations}A shows that the 49 cells infected by full wildtype viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
-Therefore, viral genetic variation is a major contributor to the heterogeneity in viral transcriptional burden among cells.
+Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
 
-Viral genetic variation also associated with IFN induction, since cells infected by full wildtype virions expressed less IFN than cells infected by incomplete or mutated virions (\FIG{mutations}B).
+Viral features also associated with IFN induction, with cells infected by full wildtype virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
 Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
-For instance, absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were more weakly enriched.
-Only the absence of NS remained significant at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are much more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
-In any case, \FIG{genotypes} and \FIG{mutations}C motivate compelling hypotheses about specific viral mutations that contribute to IFN induction, and below we pursue these hypotheses using experimental follow-up.
+Absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched.
+Only the absence of NS remained significant at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
+In any case, \FIG{genotypes} and \FIG{mutations}C motivate compelling hypotheses about specific viral defects that contribute to IFN induction, and below we pursue these hypotheses using experimental follow-up.
 
@@
 There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of the IFN induction, but the experimental follow-up below sheds more light on this question.
+The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of the IFN induction, but the experiments below shed more light on this question.
 

### 2018-10-02 paper/paper.tex
Slight changes to immunopathology sentence to make it better

--- 2018-10-02 paper/paper.tex
@@
 For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
-Consistent with these observations, despite the beneficial antiviral effects of IFN signaling, these effects can come at a significant cost due to increased immunopathology and therefore host-cell intrinsic limitation of the IFN response would help to limit the damage done by this inflammatory process~\citep{iwasaki2014innate}.
+Such self-imposed limitation of an IFN response is consistent with the biological impact of IFN as, despite the beneficial antiviral effects of IFN signaling, these effects can come at a significant cost due to increased immunopathology~\citep{iwasaki2014innate}.
 But for influenza, a third possible contributor also looms large: viral genetic diversity.

### 2018-10-02 paper/paper.tex
very minor tweaks to results section 6

--- 2018-10-02 paper/paper.tex
@@
 For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
+Such self-imposed limitation of an IFN response is consistent with the biological impact of IFN as, despite the beneficial antiviral effects of IFN signaling, these effects can come at a significant cost due to increased immunopathology~\citep{iwasaki2014innate}.
 But for influenza, a third possible contributor also looms large: viral genetic diversity.
@@
 
-\subsection{Viral features associated with infection outcome in single cells}
+\subsection{Viral defects associated with infection outcome in single cells}
 
@@
 
-Viral features also associated with IFN induction, with cells infected by full wildtype virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
+Viral defects also associated with IFN induction, with cells infected by full wildtype virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
 Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
@@
 The viral defect most strongly associated with IFN induction in single cells was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
-Although it is sometimes possible to use complementing cell lines to generate influenza viruses lacking a specific viral gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
+Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
 The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza virus's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
@@
 Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a large deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
-We therefore created a virus carrying this deletion that we were able to propagate in cells constitutively expressing the PB1 protein.
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
 
-We next tested the rate of IFN induction by each viral stock using the reporter cell lines.
-As shown in \FIG{validation}, five of the eight mutant viruses induced IFN more frequently than wildtype.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the six point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wildtype.
+We tested the rate of IFN induction by each viral stock using the reporter cells.
+\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wildtype.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wildtype.
 These results show that the viral defects in single IFN+ cells often cause increased IFN production.
@@
 It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of particular viral defects among the IFN+ cells in the single-cell data.
-However, these comparisons are confounded by the fact that the measurements in \FIG{validation} are made using \emph{bulk} stocks of each viral variant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
-For instance, our single-cell experiment used a ``wildtype'' viral stock, but the vast majority of IFN+ cells (31 of 40) still failed to express wildtype copies of all viral genes (\FIGSUPP[genotypes]{genotypes_by_ifn}).
-Therefore, although the increased IFN induction by most viral variant stocks in \FIG{validation} does indicate that these variants are causally associated with IFN, without single-cell virus-sequencing it is impossible to know how many of the IFN+ cells induced by each viral stock are infected by virions that also have other defects.
-For instance, most of the 0.67\% of cells infected with the ``wildtype'' viral stock that turned IFN+ in \FIG{validation}A are probably infected by virions that have some type of genetic defect.
+However, these comparisons are confounded by the fact that the measurements in \FIG{validation} use \emph{bulk} stocks of each viral mutant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
+For instance, our single-cell experiment used a ``wildtype'' viral stock, but most IFN+ cells (31 of 40) still had viral defects (\FIGSUPP[genotypes]{genotypes_by_ifn}). 
+So although increased IFN induction by the mutant viral stocks does indicate the mutations are causally associated with IFN, it is impossible to know how many of the IFN+ cells induced by each stock are infected by virions that also have other defects.
+The rate of IFN induction by any given viral mutant can therefore only be accurately quantified by single-cell virus-sequencing.
 

### 2018-10-02 paper/paper.tex
removed s in transcriptomic, figure 2
Changed 10x chromium to 10x chromium platform as the machine itself does not perform the RT

@@
 \caption{
-Approach for combined transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
+Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
 {\bf (A)}
@@
 {\bf (B)}
-The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes using a 10X Chromium.
+The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes using the 10X Chromium platform.
 {\bf (C)}

### 2018-10-02 paper/paper.tex
Figure 1  supplement 3 changed text slightly to make it clear this was a stable cell line rather than transduction and analysis of potentially heterogeneous cells

@@
 \figsupp[Expression of type I and type III IFN is highly correlated in A549 cells.]
-{A549 cells were dually transduced with the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
+{An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
 The cells were then infected with two different stocks of ``wildtype'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).

### 2018-10-02 paper/paper.tex
made legend to Figure 7 more succinct

--- 2018-10-02 paper/paper.tex
@@
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins.
+These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\citep{killip2014activation}.
 In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
@@
 {\bf (A)}
-Flow cytometry shows that for viruses with defects in NS1, there is increased IFN induction among infected cells that express high levels of HA protein relative to those that express low levels of HA.
-However, for wildtype virus and viruses with mutations in PB1, there is only a modest increase in IFN induction with increasing HA.
-The y-axis shows the ratio of the percent of cells that are IFN+ in the highest HA-expression quartile relative to the lowest HA-expression quartile.
-Points indicate four replicates, and lines indicate the mean.
+Flow cytometry shows that only for viruses with defects in NS1, there is more IFN induction among infected cells that express higher levels of HA protein.
+The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile.
+Points indicate replicates, and lines indicate the mean.
 See \FIGSUPP[mutantcomparison]{ha_vs_ifn} for details.
 {\bf (B)}
-Higher viral gene expression \emph{causes} more IFN induction for NS1-deficient infections, since blocking viral secondary transcription strongly decreases IFN levels in cells infected with NS1stop virus, but not cells infected with the virus carrying a deletion in PB1.
-A549 cells were infected with both viral stocks at doses that induced equivalent IFN, and ribavirin was added to half the samples at a concentration (100 $\mu$M) sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.
+Blocking viral secondary transcription strongly decreases IFN levels in cells infected with NS1stop virus, but not cells infected with virus carrying a deletion in PB1.
+A549 cells were infected with the viral stocks at doses that induced equivalent IFN, and ribavirin was added at a concentration (100 $\mu$M) sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.
 After eight hours, IFNB1 transcript was quantified relative to the housekeeping gene L32 by qPCR. 
-Points indicate four replicates, and lines indicate the mean.
 The dashed gray line indicates the limit of detection.
-\FIGSUPP[mutantcomparison]{cycloheximide} shows that similar results are obtained if secondary transcription is blocked with cycloheximide rather than ribavirin.
+\FIGSUPP[mutantcomparison]{cycloheximide} shows that results are similar if secondary transcription is instead blocked with cycloheximide.
 }
@@
 
-\figsupp[Flow cytometry analysis showing that viral burden strongly correlates with IFN induction in NS1-deficient infections.]
-{Re-analysis of the flow cytometry data in \FIGSUPP[validation]{SNPflow} indicates that a higher viral burden is associated with increased IFN-induction in NS1-deficient infections.
+\figsupp[More detailed data on the flow cytometry analysis in \FIG{mutantcomparison}A.]
+{Analysis of the flow cytometry data in \FIGSUPP[validation]{SNPflow} indicates that a higher viral burden is associated with increased IFN-induction in NS1-deficient infections.
 We re-analyzed all the infected cells (defining ``infected'' using the same HA-gating scheme in \FIGSUPP[validation]{SNPflow}) for the wildtype virus, the NS1stop virus, and the three amino-acid mutants with significantly increased IFN induction.
@@
 \figsupp[Similar to \FIG{mutantcomparison}B, but uses cycloheximide rather than ribavirin.]
-{Preventing viral secondary transcription by addition of the cycloheximide also decreases the rate of IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
+{Preventing viral secondary transcription using cycloheximide also decreases IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
 However, the results are more complex here, since cycloheximide alone increases IFN production~\citep{raj1981analysis, maroteaux1983cycloheximide, killip2014activation}.
-Therefore, we see an increase in IFN levels upon cycloheximide treatment for the uninfected cells and cells infected with the PB1 deletion virus, but a decrease in cells infected with the NS1stop virus.
+Therefore, we see an increase in IFN levels upon cycloheximide treatment for uninfected cells and cells infected with the PB1 deletion virus, but a decrease for cells infected with the NS1stop virus.
 This plot shows qPCR results for the same samples as in \FIGSUPP[validation]{cycloheximide}, but with the qPCR for expression of IFNB1 transcripts relative to the housekeeping gene L32.
-As described in \FIGSUPP[validation]{cycloheximide}, A549 cells were infected at an MOI of 0.4, such that HA expression was equivalent between viral variants when secondary transcription was blocked, and some of the samples treated with 50 $\mu$g/ml of cycloheximide.
+As described in \FIGSUPP[validation]{cycloheximide}, A549 cells were infected at an MOI of 0.4, such that HA expression was equivalent between viral variants when secondary transcription was blocked, and some of the samples treated with 50 $\mu$g/ml of cycloheximide~\citep[a concentration sufficient to block secondary transcription;][]{killip2014activation}.
 After 8 hours, transcript levels were quantified by qPCR.

### 2018-10-03 paper/paper.tex
slightly adjusted Sendai supp figure legend to align with new labeling of figure
text for last section of Results

@@
 {To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
+The name of each of reporter cell line is indicated at the top of each row of plots.
 At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
@@
 There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of the IFN induction, but the experiments below shed more light on this question.
+The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of IFN induction, but experiments below will shed more light on this question.
 
@@
 
-\subsection{Distinct IFN-inducing characteristics of different viral variants}
+\subsection{Distinct characteristics of IFN induction by different viral variants}
 
@@
 
-\jdbcomment{Section needs to be completed after remaining data on co-infections.}
+The single-cell data reveal differences among IFN+ cells depending on the defect in the virion that infects them.
+Specifically, \FIG{mutations}D shows that if a cell is infected by a virion that fails to express NS, then a higher viral transcriptional load correlates with IFN induction---but that no such association is present among infected cells that express NS.
+To validate this conclusion, we used flow cytometry to compare the frequency of IFN induction among infected cells that express low or high levels of a viral protein (HA).
+\FIG{mutantcomparison}A shows that upon infection with stocks of the NS1stop or NS1-A122V mutant, cells that express high levels of HA are much more likely to be IFN+ than those with low HA expression.
+In contrast, if cells are infected with stocks of wildtype or PB1 amino-acid mutant virus, there is only a slight increase in IFN induction among cells with high HA expression.
+
+To determine if higher viral transcriptional burden \emph{causes} IFN induction in NS-deficient infections, we treated infected cells with two different chemicals (ribavirin and cycloheximide) that inhibit secondary viral transcription, and so greatly reduce the level of viral transcripts~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg, killip2014activation}.
+As shown in \FIG{mutantcomparison}A and \FIGSUPP[mutantcomparison]{cycloheximide}, both inhibitors markedly reduced IFN induction by cells infected with the NS1stop viral stock, but not by cells infected the virus carrying a deletion in PB1.
+This result indicates that high levels of viral transcription promote IFN induction in NS-deficient infections, and suggests that there are distinct mechanisms of IFN induction by different viral variants---perhaps explaining why prior studies have not reached a consensus on the process by which influenza induces IFN~\citep{rehwinkel2010rig, weber2013incoming, killip2014activation}.
 
@@
 }
+
+\jdbcomment{It also confirms the suggestion by \citet{sjaastad2018distinct} that viral genetic variation contributes to the variability in gene expression among IFN+ cells.}
 

### 2018-10-03 paper/paper.tex
slight tweaks to abstract (moved immunopathology to first paragraph that talks about general consequences of IFN)

@@
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
-As the virus spreads, some infected cells trigger innate immunity and begin producing interferon~\citep[IFN;][]{iwasaki2014innate}.
+As the virus spreads, some infected cells trigger innate immunity and begin producing interferon (IFN).
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
-If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive prolonged immune responses later in infection can actually be associated with immunopathology and more severe disease~\citep{la2007question, iwasaki2014innate}.
 
-Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells~\citep{kallfass2013visualizing, killip2017single}.
+Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
@@
 
-It remains unclear why only some infected cells trigger innate-immune responses.
+It is unclear why only some infected cells trigger innate-immune responses.
 Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
-For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and there is evidence that the frequency of IFN induction depends on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
-Such self-imposed limitation of an IFN response is consistent with the biological impact of IFN as, despite the beneficial antiviral effects of IFN signaling, these effects can come at a significant cost due to increased immunopathology~\citep{iwasaki2014innate}.
+For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
 But for influenza, a third possible contributor also looms large: viral genetic diversity.
-Specifically, because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have genetic defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
-Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and more IFN is induced by viral stocks that contain defective virions with internal deletions in the polymerase genes~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 
-However, existing single-cell techniques are inadequate to determine how viral genetic diversity actually contributes to cell-to-cell heterogeneity during infection, since they do not systematically determine the full sequences of the virions that infect individual cells.
-Flow cytometry and fluorescent reporters only measure expression of viral proteins~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of viral transcripts~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
-None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations will be common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.
+However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection, since they do not determine full sequences of the virions that infect individual cells.
+Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
+None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.
 
@@
 To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
-We obtain the transcriptomes and full sequences of all expressed viral genes in 150 virus-infected cells, 49 of which express IFN.
-Even though we used a nominally ``wildtype'' stock of virus, two-thirds of the cells are infected by virions that have a mutation or defect in gene expression.
+We obtain the transcriptomes and full sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
+Even though we used a nominally ``wildtype'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
 This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
 Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
-However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations that we identify deterministically induce IFN.
+However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations deterministically induce IFN.
 Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
@@
 \subsection{A system to identify and enrich for rare IFN+ positive cells}
-Influenza infection only rarely triggers expression of the anti-viral IFNs~\citep[type I and type III IFNs;][]{kotenko2017contribution} at the single-cell level~\citep{killip2017single}.
+A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} at the single-cell level~\citep{killip2017single}.
 For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
 Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
-Importantly, the paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
+This paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
 

### 2018-10-03 paper/paper.tex
added draft acknowledgments

@@
 This pipeline is orchestrated by \texttt{Snakemake}~\citep{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
-The sequencing data and annotated cell-gene matrix are available on the GEO repository under \jdbcomment{add accession}.
+The sequencing data and annotated cell-gene matrix are available on the GEO repository under \jdbcomment{data have been uploaded to GEO, but accession not yet available.
+The accession number will be added to the README of the GitHub repo at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell} as soon as it's available}.
 
@@
 \section{Acknowledgments}
-Andy Marty and Genomics Core.
-Cole Trapnell.
-Jason Underwood.
-AJ Velthuis. 
-\jdbcomment{Finish this}
+We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, and AJ Velthuis for helpful scientific suggestions.
+We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
+We thank Katherine Xue, David Bacsik, and Hannah Itell for comments on the manuscript.
+This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
+ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+JDB is an Investigator of the Howard Hughes Medical Institute.
+The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
+The authors declare that no competing interests exist.
 

### 2018-10-04 paper/paper.tex
added Jacob undergrad grants
added a bit of text to Discussion

@@
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
-If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive prolonged immune responses later in infection can actually be associated with immunopathology and more severe disease~\citep{la2007question, iwasaki2014innate}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and more severe disease~\citep{la2007question, iwasaki2014innate}.
 
@@
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection, since they do not determine full sequences of the virions that infect individual cells.
-Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
+Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.
@@
 
-\clearpage
 \section{Results}
@@
 \section{Discussion}
+We have determined the full sequences of the infecting virions in single influenza-infected cells, and then examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
+Methodologically, the major advance of our work is to measure the \emph{genotypes} of viruses in single cells in addition to quantifying the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell virology studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using standard single-cell transcriptomic techniques.
+To make this genetic information complete, we have coupled a standard single-cell transcriptomic technique to long-read PacBio sequencing, a strategy analogous to that recently used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
-\jdbcomment{Section needs to be complete.
-Add this:
-Approach builds on observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using more standard single-cell techniques, and is similar the approach used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
-}
+Our results demonstrate that the single-cell viral genetic information provided by this methodological advance is extremely valuable for understanding the biology of influenza infection, especially as it relates to innate-immune detection.
 
@@
 ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
 JDB is an Investigator of the Howard Hughes Medical Institute.

### 2018-10-04 paper/paper.tex
added another section to discussion

@@
 \section{Discussion}
-We have determined the full sequences of the infecting virions in single influenza-infected cells, and then examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, the major advance of our work is to measure the \emph{genotypes} of viruses in single cells in addition to quantifying the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell virology studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using standard single-cell transcriptomic techniques.
-To make this genetic information complete, we have coupled a standard single-cell transcriptomic technique to long-read PacBio sequencing, a strategy analogous to that recently used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
+We have determined the full sequences of the viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to measuring the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using more standard single-cell transcriptomic techniques.
+To make this information complete, we have coupled a standard single-cell transcriptomic technique to long-read PacBio sequencing of viral genes, a strategy analogous to that recently used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
-Our results demonstrate that the single-cell viral genetic information provided by this methodological advance is extremely valuable for understanding the biology of influenza infection, especially as it relates to innate-immune detection.
+This viral genetic information turns out to be extremely important for understanding infection outcome and innate-immune induction.
+Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, a large majority of cells are infected by virions that have mutations or fail to express a gene.
+Although our study is certainly not the first to note that influenza virus has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
+Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experimental approach that does not directly sequence the viral genes in single cells is really averaging across a vast spectrum of hidden viral defects.
+
+\jdbcomment{Our results demonstrate that the single-cell viral genetic information provided by this methodological advance is extremely valuable for understanding the biology of influenza infection, especially as it relates to innate-immune detection.}
 

### 2018-10-04 paper/paper.tex
A few small changes up through about half of results. One or two sentences I think add a little bit of clarity. Also changed wildtype to wild-type (adjective) or wild type (noun)

--- 2018-10-04 paper/paper.tex
@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous at the cellular level, and infected cells only sometimes activate innate immunity.  
+The outcome of viral infection is extremely heterogeneous at the level of individual cells, including sparse activation of innate immunity.  
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
@@
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
-Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially as real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 But for influenza, a third possible contributor also looms large: viral genetic diversity.
-Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
 Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
@@
 Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
-None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.
+None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wild-type'' or ``mutant'' stock of virus.
 
@@
 We obtain the transcriptomes and full sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
-Even though we used a nominally ``wildtype'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
-This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
+Even though we used a nominally ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
+This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
 Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
-However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations deterministically induce IFN.
+However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wild-type virions sometimes induce IFN, and no viral mutations deterministically induce IFN.
 Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
@@
 \subsection{A system to identify and enrich for rare IFN+ positive cells}
-A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} at the single-cell level~\citep{killip2017single}.
+A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} at the level of single-cells ~\citep{killip2017single}.
 For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
@@
 This rareness of IFN induction is not merely a feature of A549 cells in culture.
-Prior studies using reporters have found that IFN expression is rare at the single-cell level in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
+Prior studies using interferon reporter constructs have found that IFN expression is rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
 In support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
@@
 Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
-Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
+Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
 As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
@@
 {An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
-The cells were then infected with two different stocks of ``wildtype'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
+The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
 After 13 hours, cells were analyzed by flow cytometry.
@@
 
-To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B).
+To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and subsequently propagated these reporters from single clones.
 These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
@@
 {\bf (A)}
-IFN reporter A549 cells are infected with a mix of wildtype and synonymously barcoded viruses.
+IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
 IFN+ cells are enriched by MACS (\FIGSUPP[workflow]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
@@
 {\bf (C)}
-The entire cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
+Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
 {\bf (D)}
 viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
-{\bf (E)}
+{\bf (E)}wild-type
 The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
@@
 Example MACS enrichments of IFN+ influenza-infected cells.
-A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
+A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
 After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
@@
 
-\figdata{Genbank files giving sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
+\figdata{Genbank files giving sequences of the wild-type and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
 \label{figdata:virus_seqs}
@@
 We therefore developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of the infecting virions in single cells.
-First, we generated a stock of virus that consisted of a mix of wildtype WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
+First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
 These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
@@
 Based on pilot experiments, we expected this MACS to enrich IFN+ cells to $\sim$5\% to 30\% of the total (\FIGSUPP[workflow]{MACS}).
-To ensure the presence of IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
+To ensure the presence of infected IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
 We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
@@
 Cells called as infected are in blue, while other cells are in green.
-The inset shows the amount of viral mRNA among the human cells called as infected.
+The inset shows the amount of viral mRNA in the human cells that are called as infected.
 {\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
@@
 {\bf (E)} Relative expression of viral genes among infected cells.
-{\bf (F)} Number of cells infected with wildtype virus, synonymously barcoded virus, or both.
+{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
 From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
@@
 Each point represents one cell, and the Pearson correlation coefficient is shown.
-IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression more often due to paracrine signaling that does not induce expression of IFN itself.
+IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
 Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
@@
 
-By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wildtype and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
+By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
 From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
@@
 This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
+This high rate of observed coinfection can also partially explain the relative completeness of many viral infections in our dataset, as absence of a segment from one viral particle can be readily complemented by a second, infecting, particle~\citep{russell2018extreme}.
 
@@
 Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
-For dual-barcode infections, mutations / indels for the wildtype and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively. 
+For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively. 
 Green boxes at left show the percent of all mRNA in that cell derived from virus.
@@
 Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences derived from the same cDNA, and cell barcodes were used to collapse duplicate sequences from the same cell.
-The bars in the plot are colored by whether the sequence is derived from the wildtype viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \FIGSUPP[genotypes]{StrandExchange} for more explanation).
+The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \FIGSUPP[genotypes]{StrandExchange} for more explanation).
 From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
@@
 A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
-We can detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wildtype virus and virus carrying synonymous barcodes near both termini of each gene.
-If there is no strand exchange, all molecules should either be wildtype or have the synonymous barcoding mutations at \emph{both} termini.
-But strand exchange will create some molecules that have wildtype nucleotides at one termini and synonymous barcoding mutations at the other termini.
+We can detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
+If there is no strand exchange, all molecules should either be wild-type or have the synonymous barcoding mutations at \emph{both} termini.
+But strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
 \FIGSUPP[genotypes]{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
@@
 Simple visual inspection of this figure reveals a wealth of information.
-For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wildtype copies all eight genes and did not induce detectable IFN.
+For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
 But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that either had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
@@
 Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
-Cells infected by full wildtype virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
+Cells infected by full wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
 {\bf (B)}
-Cells infected by full wildtype virus to express less IFN.
+Cells infected by full wild-type virus to express less IFN.
 {\bf (C)}
@@
 The major viral features that associate with IFN induction also associate with ISG expression.
-Specifically, cells that express wildtype copies of all eight viral genes tend to express ISGs less frequently and at lower levels
+Specifically, cells that express wild-type copies of all eight viral genes tend to express ISGs less frequently and at lower levels
 The absence of NS is strongly associated with higher ISG expression.
@@
 To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
-\FIG{mutations}A shows that the 49 cells infected by full wildtype viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
+\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
 
-Viral defects also associated with IFN induction, with cells infected by full wildtype virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
+Viral defects also associated with IFN induction, with cells infected by full wild-type virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
 Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
@@
 The numbers indicate the median of four measurements for each viral mutant.
-The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wildtype is indicated with a dashed blue line.
-Points are colored orange if the mutant virus stock induces IFN more frequently than the wildtype viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
+The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
+Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
 {\bf (B)}
@@
 For each viral mutant, two independently generated stocks were each assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other viral stock).
-The infections with replicate \#1 of the wildtype virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
+The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
 The HA staining indicates that the number of transcriptionally active viral units per cell was similar across most mutants, and calculating percentage IFN+ only among HA+ cells largely corrects for modest variation in titer across mutants.
@@
 A complicating factor for these data is that the virus with the deletion in PB1 (PB1del385to2163) cannot perform secondary transcription because it does not express PB1 protein, and so cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
-Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wildtype and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
+Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
 Finally, the approach in \FIGSUPP[validation]{delqPCR} was used to show that equivalent TCID50s, all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
@@
 We tested the rate of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wildtype.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wildtype.
+\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wild-type.
 These results show that the viral defects in single IFN+ cells often cause increased IFN production.
@@
 However, these comparisons are confounded by the fact that the measurements in \FIG{validation} use \emph{bulk} stocks of each viral mutant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
-For instance, our single-cell experiment used a ``wildtype'' viral stock, but most IFN+ cells (31 of 40) still had viral defects (\FIGSUPP[genotypes]{genotypes_by_ifn}). 
+For instance, our single-cell experiment used a ``wild-type'' viral stock, but most IFN+ cells (31 of 40) still had viral defects (\FIGSUPP[genotypes]{genotypes_by_ifn}). 
 So although increased IFN induction by the mutant viral stocks does indicate the mutations are causally associated with IFN, it is impossible to know how many of the IFN+ cells induced by each stock are infected by virions that also have other defects.
@@
 {Analysis of the flow cytometry data in \FIGSUPP[validation]{SNPflow} indicates that a higher viral burden is associated with increased IFN-induction in NS1-deficient infections.
-We re-analyzed all the infected cells (defining ``infected'' using the same HA-gating scheme in \FIGSUPP[validation]{SNPflow}) for the wildtype virus, the NS1stop virus, and the three amino-acid mutants with significantly increased IFN induction.
+We re-analyzed all the infected cells (defining ``infected'' using the same HA-gating scheme in \FIGSUPP[validation]{SNPflow}) for the wild-type virus, the NS1stop virus, and the three amino-acid mutants with significantly increased IFN induction.
 For each virus and replicate, we binned the infected cells into HA expression quartiles.
@@
 A similar trend also holds for the virus with an amino-acid substitution in NS1 (NS1-A122V). 
-But for wildtype and the two PB1 mutants, there is no clear trend for higher viral burden to correlate with IFN induction.
+But for wild type and the two PB1 mutants, there is no clear trend for higher viral burden to correlate with IFN induction.
 \FIG{mutantcomparison}A summarizes the data in this figure by simply showing the ratio of the percent IFN+ of the highest HA-expression quartile to the lowest HA-expression quartile for each sample and replicate.}
@@
 \FIG{mutantcomparison}A shows that upon infection with stocks of the NS1stop or NS1-A122V mutant, cells that express high levels of HA are much more likely to be IFN+ than those with low HA expression.
-In contrast, if cells are infected with stocks of wildtype or PB1 amino-acid mutant virus, there is only a slight increase in IFN induction among cells with high HA expression.
+In contrast, if cells are infected with stocks of wild-type or PB1 amino-acid mutant virus, there is only a slight increase in IFN induction among cells with high HA expression.
 
@@
 We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
-We used both the wildtype virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
+We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
 We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
 In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
-The sequences of all gene segments from the wildtype and synonymously barcoded viral strains are in \FIGDATA[workflow]{virus_seqs}.
+The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \FIGDATA[workflow]{virus_seqs}.
 These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
@@
 The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
-Specifically, after generation by reverse genetics, the wildtype variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
+Specifically, after generation by reverse genetics, the wild-type variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
 The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of \citet{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
@@
 
-The wildtype and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
+The wild-type and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
 The cells were then infected with this mixture at a dose designed to infect about half the cells (\FIG{transcriptomics}C suggests that the actual rate of detectable infection was slightly lower).
@@
 
-We generated the wildtype and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wildtype plasmid for that gene.
+We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wild-type plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
@@
 
-To generate the NS1stop mutant virus and the wildtype and PB1del385to2163 mutant viruses in \FIGSUPP[validation]{delflow}, we used slightly different procedures.
-The wildtype virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
+To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \FIGSUPP[validation]{delflow}, we used slightly different procedures.
+The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
 The NS1stop virus was similarly generated, but was passaged for 48 rather than 36 hours, since it had slower growth kinetics and so needed a longer period of time to reach high titers.
@@
 This passaging was necessary as viral titers from transfections were too low to generate sufficient virus from a single passage.
-The wildtype and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
+The wild-type and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
 The infections in \FIGSUPP[validation]{delflow} were performed at equivalent TCID50s as described in the legend to that figure.
@@
 The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
-The influenza virus transcriptome consisted of the mRNAs for the wildtype A/WSN/1933 virus strain in \FIGDATA[workflow]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wildtype and synonymously barcoded viral variants to this transcriptome).
+The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \FIGDATA[workflow]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
 The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
 The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{transcriptomics}A), and generate a matrix giving the expression of each gene in each single cell.
-We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wildtype or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.
+We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wild-type or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.
 
@@
 \item Mutations with accuracies less than 0.999 (which constitute $<$0.5\% of all mutations) were ignored.
-\item If all CCSs for a particular viral-barcode variant of a gene in a cell were wildtype, it was called as wildtype.
+\item If all CCSs for a particular viral-barcode variant of a gene in a cell were wild-type, it was called as wild type.
 \item If any CCSs for a particular viral-barcode variant of gene in a cell had a mutation, then require at least two CCSs to call the sequence.

### 2018-10-04 paper/paper.tex
added another paragraph to Discussion

--- 2018-10-04 paper/paper.tex
@@
 Although our study is certainly not the first to note that influenza virus has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
-Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experimental approach that does not directly sequence the viral genes in single cells is really averaging across a vast spectrum of hidden viral defects.
+Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experimental approach that does not directly sequence the viral genes in single cells is really averaging across a vast number of hidden viral defects.
+
+It turns out that these idiosyncratic viral defects explain much of the heterogeneity in viral transcriptional burden that has been observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
+If we examine only cells infected by truly wildtype virions, we see that the heterogeneity in viral transcriptional burden is substantially reduced (\FIG{mutations}A).
+
+Crucially, viral defects also play an important role in determining if infected cells trigger an innate-immune response.
+We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
+Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---likely act by impairing NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
+The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by \citet{velthuis2018mini} showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I.
+Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+In fact, given the extensive prior work describing such deletions, it is perhaps surprising that we did not identify more of them in our single IFN+ cells.
+There may be several reasons: first, we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI~\citep[conditions that better reflect the early stages of real human influenza infections;][]{mccrone2018stochastic, xue2018reconciling, varble2014influenza}; second, our experiments preferentially captured cells with higher viral transcriptional load; and third, most prior studies have used techniques that can detect large deletions but not subtle point mutations.
+But we emphasize that the relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it certainly remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
 

### 2018-10-04 paper/paper.tex
another Discussion paragraph

@@
 
-\jdbcomment{Our results demonstrate that the single-cell viral genetic information provided by this methodological advance is extremely valuable for understanding the biology of influenza infection, especially as it relates to innate-immune detection.}
+Our results also indicate that different types of viral defects induce IFN by somewhat different processes.
+For instance, higher viral gene transcription is associated with IFN induction only in cells that are infected by virions with deficiencies in NS.
+This finding is consistent with the suggestion by \citet{sjaastad2018distinct} that viral genetic variation contributes to cell-to-cell variability in the expression of anti-viral genes.
+Larger-scale studies similar to the one performed here should be useful for further determining the extent that virions with different defects cause cells to trigger distinct transcriptional programs.
 
-\jdbcomment{It also confirms the suggestion by \citet{sjaastad2018distinct} that viral genetic variation contributes to the variability in gene expression among IFN+ cells.}
+\jdbcomment{But results also show that non-viral factors are clearly important.}
+
+\jdbcomment{summary paragraph}
 

### 2018-10-04 paper/paper.tex
a few more tweaks to Discussion

@@
 \section{Discussion}
-We have determined the full sequences of the viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to measuring the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to measuring abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
 Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using more standard single-cell transcriptomic techniques.
-To make this information complete, we have coupled a standard single-cell transcriptomic technique to long-read PacBio sequencing of viral genes, a strategy analogous to that recently used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
+To make this information complete, we have coupled a single-cell transcriptomic technique with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
-This viral genetic information turns out to be extremely important for understanding infection outcome and innate-immune induction.
-Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, a large majority of cells are infected by virions that have mutations or fail to express a gene.
+This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
+Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, most cells are infected by virions that have mutations or fail to express a gene.
 Although our study is certainly not the first to note that influenza virus has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
-Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experimental approach that does not directly sequence the viral genes in single cells is really averaging across a vast number of hidden viral defects.
+Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experiment that does not directly sequence the viral genes in single cells is really averaging across a vast number of hidden viral defects.
 
-It turns out that these idiosyncratic viral defects explain much of the heterogeneity in viral transcriptional burden that has been observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
-If we examine only cells infected by truly wildtype virions, we see that the heterogeneity in viral transcriptional burden is substantially reduced (\FIG{mutations}A).
+It turns out that these viral defects explain much of the heterogeneity in viral transcriptional burden observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
+If we examine only cells infected by truly wildtype virions, we see that heterogeneity in viral transcriptional burden is substantially reduced (\FIG{mutations}A).
 
-Crucially, viral defects also play an important role in determining if infected cells trigger an innate-immune response.
+Viral defects also help determine if an infected cell triggers an innate-immune response.
 We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
-Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---likely act by impairing NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
+Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
 The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by \citet{velthuis2018mini} showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I.
 Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-In fact, given the extensive prior work describing such deletions, it is perhaps surprising that we did not identify more of them in our single IFN+ cells.
-There may be several reasons: first, we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI~\citep[conditions that better reflect the early stages of real human influenza infections;][]{mccrone2018stochastic, xue2018reconciling, varble2014influenza}; second, our experiments preferentially captured cells with higher viral transcriptional load; and third, most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-But we emphasize that the relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it certainly remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
+Given the extensive prior work describing such deletions, it is perhaps surprising that we did not identify more of them in our single IFN+ cells.
+There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI~\citep[conditions that better reflect the early stages of real human influenza infections;][]{mccrone2018stochastic, xue2018reconciling, varble2014influenza}; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
+But the relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
 
-Our results also indicate that different types of viral defects induce IFN by somewhat different processes.
-For instance, higher viral gene transcription is associated with IFN induction only in cells that are infected by virions with deficiencies in NS.
+Our results also indicate that different viral defects induce IFN by somewhat distinct processes.
+For instance, higher viral transcription is associated with IFN induction only in cells infected by virions with deficiencies in NS.
 This finding is consistent with the suggestion by \citet{sjaastad2018distinct} that viral genetic variation contributes to cell-to-cell variability in the expression of anti-viral genes.
-Larger-scale studies similar to the one performed here should be useful for further determining the extent that virions with different defects cause cells to trigger distinct transcriptional programs.
+Larger-scale studies similar to the one performed here should be useful for further elucidating the extent that virions with different defects cause cells to launch distinct transcriptional programs.
 
-\jdbcomment{But results also show that non-viral factors are clearly important.}
+But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
+We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wildtype virions.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency but dominant mutations that affect the behavior of cells expressing predominantly wildtype viral genes.
+However, we suspect that pure stochasticity or pre-existing cellular states also affect the outcome of viral infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.
 

### 2018-10-04 paper/paper.tex
full draft of Discussion

@@
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to measuring abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using more standard single-cell transcriptomic techniques.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques.
 To make this information complete, we have coupled a single-cell transcriptomic technique with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
@@
 This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
-Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, most cells are infected by virions that have mutations or fail to express a gene.
-Although our study is certainly not the first to note that influenza virus has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
-Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experiment that does not directly sequence the viral genes in single cells is really averaging across a vast number of hidden viral defects.
+Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, most cells are infected by non-wildtype virions.
+Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
+Visual inspection of \FIG{genotypes} makes it glaringly obvious how any experiment that does not directly sequence viral genes in single cells is really averaging across a vast number of hidden viral defects.
 
 It turns out that these viral defects explain much of the heterogeneity in viral transcriptional burden observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
-If we examine only cells infected by truly wildtype virions, we see that heterogeneity in viral transcriptional burden is substantially reduced (\FIG{mutations}A).
+If we examine only cells infected by truly wildtype virions, the heterogeneity in viral transcriptional burden is substantially reduced.
 
@@
 Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Given the extensive prior work describing such deletions, it is perhaps surprising that we did not identify more of them in our single IFN+ cells.
-There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI~\citep[conditions that better reflect the early stages of real human influenza infections;][]{mccrone2018stochastic, xue2018reconciling, varble2014influenza}; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
+Given the extensive prior work describing such deletions, why did we not identify more of them in our single IFN+ cells?
+There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
 But the relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
@@
 For instance, higher viral transcription is associated with IFN induction only in cells infected by virions with deficiencies in NS.
-This finding is consistent with the suggestion by \citet{sjaastad2018distinct} that viral genetic variation contributes to cell-to-cell variability in the expression of anti-viral genes.
+This finding is consistent with the suggestion by \citet{sjaastad2018distinct} that viral genetic variation relates to cell-to-cell variability in gene expression.
 Larger-scale studies similar to the one performed here should be useful for further elucidating the extent that virions with different defects cause cells to launch distinct transcriptional programs.
@@
 We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wildtype virions.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency but dominant mutations that affect the behavior of cells expressing predominantly wildtype viral genes.
-However, we suspect that pure stochasticity or pre-existing cellular states also affect the outcome of viral infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations that affect cells expressing predominantly wildtype viral genes.
+However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.
 
-\jdbcomment{summary paragraph}
+Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
+Real human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
+It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
+Extending our approaches to more complex systems could shed further light on how this viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
 

### 2018-10-04 paper/paper.tex
minor edits

--- 2018-10-04 paper/paper.tex
@@
 Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
-Cells infected by full wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
+Cells infected by fully wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
 {\bf (B)}
-Cells infected by full wild-type virus to express less IFN.
+Cells infected by fully wild-type virus to express IFN less frequently.
 {\bf (C)}
@@
 
-Viral defects also associated with IFN induction, with cells infected by full wild-type virions turning IFN+ less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
+Viral defects are also associated with IFN induction, with cells infected by full wild-type virions expressing IFN less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
 Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
@@
 \FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wild-type.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild-type.
 These results show that the viral defects in single IFN+ cells often cause increased IFN production.

### 2018-10-04 paper/paper.tex
tweaked a few edits and converted remaining "wildtype" to "wild-type" or "wild type"

@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous at the level of individual cells, including sparse activation of innate immunity.  
+The outcome of viral infection is extremely heterogeneous at the level of individual cells, and infected cells only sometimes activate innate immunity.  
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
@@
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
-Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially as real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 \subsection{A system to identify and enrich for rare IFN+ positive cells}
-A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} at the level of single-cells ~\citep{killip2017single}.
+A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} in single cells ~\citep{killip2017single}.
 For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
@@
 
-To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and subsequently propagated these reporters from single clones.
+To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and subsequently expanded the cell lines from single clones.
 These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
@@
 viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
-{\bf (E)}wild-type
+{\bf (E)}
 The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
@@
 Based on pilot experiments, we expected this MACS to enrich IFN+ cells to $\sim$5\% to 30\% of the total (\FIGSUPP[workflow]{MACS}).
-To ensure the presence of infected IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
+To ensure the presence of IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
 We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
@@
 This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
-This high rate of observed coinfection can also partially explain the relative completeness of many viral infections in our dataset, as absence of a segment from one viral particle can be readily complemented by a second, infecting, particle~\citep{russell2018extreme}.
+This moderately high rate of co-infection may also partially explain why relatively more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.
 
@@
 {\bf (B)}
-Cells infected by fully wild-type virus to express IFN less frequently.
+Cells infected by fully wild-type virus express IFN less frequently.
 {\bf (C)}
@@
 A complicating factor for these data is that the virus with the deletion in PB1 (PB1del385to2163) cannot perform secondary transcription because it does not express PB1 protein, and so cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
-Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
+Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
 Finally, the approach in \FIGSUPP[validation]{delqPCR} was used to show that equivalent TCID50s, all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
@@
 \FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild-type.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
 These results show that the viral defects in single IFN+ cells often cause increased IFN production.
@@
 This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
-Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, most cells are infected by non-wildtype virions.
+Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wild-type'' influenza genome, most infected cells do not express wild-type copies of all viral genes.
 Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
@@
 It turns out that these viral defects explain much of the heterogeneity in viral transcriptional burden observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
-If we examine only cells infected by truly wildtype virions, the heterogeneity in viral transcriptional burden is substantially reduced.
+If we examine only cells infected by truly wild-type virions, the heterogeneity in viral transcriptional burden is substantially reduced.
 
@@
 But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
-We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wildtype virions.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations that affect cells expressing predominantly wildtype viral genes.
+We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations that affect cells expressing predominantly wild-type viral genes.
 However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.

### 2018-10-05 paper/paper.tex
One sentence added to Figure 1–Figure supplement 3
tiny edits to Intro and first results section

@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous at the level of individual cells, and infected cells only sometimes activate innate immunity.  
+The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
@@
 
-Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
-This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
-Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+Unfortunately for the host, influenza infection only rarely triggers IFN production by infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
+Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
-For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
+For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
 But for influenza, a third possible contributor also looms large: viral genetic diversity.
@@
 
-However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection, since they do not determine full sequences of the virions that infect individual cells.
+However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
 Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
-None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wild-type'' or ``mutant'' stock of virus.
+None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
 
-Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells.
+Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
 To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
-We obtain the transcriptomes and full sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
-Even though we used a nominally ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
+We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
+Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
 This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
-Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
-However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wild-type virions sometimes induce IFN, and no viral mutations deterministically induce IFN.
+Viral defects are especially common in IFN+ cells, and we identify at least four type of defects that cause increased IFN induction.
+However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
 Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
@@
 \subsection{A system to identify and enrich for rare IFN+ positive cells}
-A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} in single cells ~\citep{killip2017single}.
+A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
 For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
@@
 This rareness of IFN induction is not merely a feature of A549 cells in culture.
-Prior studies using interferon reporter constructs have found that IFN expression is rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
+Prior studies using reporters have found that IFN expression is rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
 In support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
@@
 After 13 hours, cells were analyzed by flow cytometry.
+Cells positive for either fluorescent reporter were further analyzed.
 As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.}
@@
 
-\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1517_pHAGE2_IFNbeta_prom_LNGFR_P2A_mNeongreen.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/1767_pHAGE2_IL29_promnkozak_LNGFR_P2A_zsGreen.gb}.
+\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.}
 }
@@
 
-To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and subsequently expanded the cell lines from single clones.
+To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and expanded cell lines from single clones.
 These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.

### 2018-10-05 paper/paper.tex
Made a few more very minor edits

@@
 \centerline{
-{\bf \Large A}\includegraphics[width=0.25\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
+{\bf \Large A}\includegraphics[width=0.28\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
 \hspace{0.02\textwidth}
-{\bf \Large B}\includegraphics[width=0.32\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
+{\bf \Large B}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
 \hspace{0.02\textwidth}
-{\bf \Large C} \includegraphics[width=0.2\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
+{\bf \Large C} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
 }
@@
 
-\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.}
+\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
 }
@@
 We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of infection in single cells.
-Standard single-cell transcriptomic techniques are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to measure the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}, and do not reveal if the virions that infect individual cells have features (e.g., mutations) that contribute to heterogeneity in infection outcome.
+Standard single-cell transcriptomic techniques are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to measure the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}, and do not reveal if the virions that infect individual cells have mutations.
 
@@
 
-We processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and then reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
+We processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
 Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
@@
 Although ten transcripts are encoded by the eight viral genes, we call presence / absence of genes rather than transcripts.
-The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable in the 3'-end sequencing transcriptomic data.
-\FIG{transcriptomics}D (top panel) shows that the majority (162 of 290) of infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
+The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
+\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
 This measured frequency of infected cells expressing all eight genes is higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
-The difference is sufficiently modest that it could simply be due to methodological differences in the thresholds used to call gene presence.
-But it is also important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could alter the distribution of viral gene expression from a fully random sample of infected cells.
+However, it is important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could alter the distribution of viral gene expression from a fully random sample of infected cells.
 
@@
 From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
-Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
-This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
-This moderately high rate of co-infection may also partially explain why relatively more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.
+Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
+This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
+This moderately high rate of co-infection may also partially explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.
 
@@
 Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}C, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
-Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected but produced too few viral transcripts to be detected above background.
+Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
 Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\FIGSUPP[transcriptomics]{ISG}A).
-The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express significant IFN (\FIGSUPP[transcriptomics]{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect that they have been infected, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect viral infection~\citep{stetson2006type,honda2006type}.
+The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\FIGSUPP[transcriptomics]{ISG}B).
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.
 
@@
 \FIGSUPP[transcriptomics]{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
-Although the clustering is entirely unsupervised, cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with expression of these genes involves uninfected and IFN- cells.
+Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.
 
@@
 \figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
-{Cells for which we could determine the full sequences of all genes expressed by the infecting virus(es).
-{\bf (A)} We could call the complete genotypes of the infecting virus(es) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
+{Cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).
+{\bf (A)} We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
 {\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
@@
 
-The foregoing transcriptomic analyses identify which viral genes are expressed in each cell, but not the sequences of these genes.
-We therefore used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in the majority of infected cells.
-
-By employing several PCR enrichment strategies (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
+We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
+Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
 Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange (\FIGSUPP[genotypes]{StrandExchange}).
@@
 
-We used these CCSs to call the sequences of viral genes in single cells.
-We called a mutation if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
+We called a mutation in a viral gene if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
 For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
-This strategy reliably identifies mutations in the virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
 It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
-Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of viral transcripts in a cell.
+Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.
 
@@
 For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
-But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that either had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
+But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
 The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
@@
 
-To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not (due to gene absence, amino-acid substitution, insertion, or deletion).
+To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
 \FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
@@
 There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of IFN induction, but experiments below will shed more light on this question.
+The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of IFN induction, but experiments below shed more light on this question.
 
@@
 }
-{\includegraphics[height=0.75\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}}
+{\includegraphics[height=0.71\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}}
 \label{figsupp:SNPflow}
@@
 
-The viral defect most strongly associated with IFN induction in single cells was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
+The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
 Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
-The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza virus's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
+The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
 We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
@@
 The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
-These results show that the viral defects in single IFN+ cells often cause increased IFN production.
+Therefore, the viral defects in single IFN+ cells often cause increased IFN production.
 
-It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of particular viral defects among the IFN+ cells in the single-cell data.
+It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of viral defects among the IFN+ cells in the single-cell data.
 However, these comparisons are confounded by the fact that the measurements in \FIG{validation} use \emph{bulk} stocks of each viral mutant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
-For instance, our single-cell experiment used a ``wild-type'' viral stock, but most IFN+ cells (31 of 40) still had viral defects (\FIGSUPP[genotypes]{genotypes_by_ifn}). 
 So although increased IFN induction by the mutant viral stocks does indicate the mutations are causally associated with IFN, it is impossible to know how many of the IFN+ cells induced by each stock are infected by virions that also have other defects.
-The rate of IFN induction by any given viral mutant can therefore only be accurately quantified by single-cell virus-sequencing.
+The rate of IFN induction by any given viral mutant can only be accurately quantified by single-cell virus-sequencing.
 
@@
 Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques.
-To make this information complete, we have coupled a single-cell transcriptomic technique with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
+Our method builds on the observation by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques.
+To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
@@
 We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations that affect cells expressing predominantly wild-type viral genes.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations in cells expressing predominantly wild-type viral genes.
 However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.
@@
 It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
-Extending our approaches to more complex systems could shed further light on how this viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
+Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
+
+\clearpage
 
@@
 
-We used these constructs to generate lentiviral vectors~\citep{oconnell2010lentiviral} which were used to transduce of A549 cells in the presence of 5 $\mu$g polybrene.
+We used these constructs to generate lentiviral vectors and transduce of A549 cells in the presence of 5 $\mu$g polybrene.
 We then sorted single transduced cells and expanded them.
@@
 We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\citep{xue2016propagation}.
-The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
+The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 cells in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
 Specifically, after generation by reverse genetics, the wild-type variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
@@
 At this point, uninfected canine (MDCK-SIAT1) cells were also added to $\sim$5\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{transcriptomics}A) and background viral mRNA in uninfected cells (\FIG{transcriptomics}C).
-We begin this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
+We began this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
 The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.
@@
 However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
-We therefore needed to enrich for the viral molecules, a process made more challenging by the fact that the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end means that an enrichment PCR could only be semi-specific (the primer at the 5' end can be specific, but that the 3' end must bind the common 10X adaptor shared among all molecules).
+We therefore needed to enrich for the viral molecules, a process made challenging by the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end (the primer at the 5' end can be specific, but the one at the 3' end must bind the common 10X adaptor shared among all molecules).
 
@@
 Both the common 3' primer and the multiplex mix of eight 5' primers were added to a final concentration of 0.5$\mu$M.
-We performed 30 cycles of PCR, using an extension time of 2 minutes 15 seconds at 67$^{\circ}$C, and a melting temperature of 95$^{\circ}C$.
+We performed 30 cycles of PCR, using an extension time of 2 minutes 15 seconds at 67$^{\circ}$C, and a melting temperature of 95$^{\circ}$C.
 This melting temperature is lower than the standard 98$^{\circ}$C melting step suggested by the manufacturer for Kapa HiFi because we wanted to avoid collapse of emulsion integrity at high temperature.
@@
 The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[genotypes]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[genotypes]{StrandExchange} for an illustration of how this is determined).
-A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio run is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}).
+A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}).
 Importantly, the various PCR biases and enrichment schemes mean that the coverage of molecules by the PacBio sequencing is not proportional to their original abundance in the starting mRNA.
@@
 
-For the qPCR in \FIGSUPP[validation]{delqPCR} and \FIGSUPP[mutantcomparison]{cycloheximide}, A549 cells were seeded at a density of 10,000$10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+For the qPCR in \FIGSUPP[validation]{delqPCR} and \FIGSUPP[mutantcomparison]{cycloheximide}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
@@
 That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \FIGSUPP[validation]{delqPCR}.
-
-\subsection{\jdbcomment{Not yet used, can add when section with this virus is added}}
-To generate pseudovirus that expresses fluorescent mCherry instead of HA, a plasmid was generated similarly to a previously-described eGFP-expressing pseudovirus~\citep{russell2018extreme}. 
-To rescue this virus, we used MDCK-SIAT1 cells transduced to constituitively express HA and included an HA protein expression plasmid in our reverse-genetics transfection mixture.
-Virus was harvested from transfections at 72h and passaged for 48 at an MOI of 0.01 on MDCK-SIAT1 cells that express HA. 
-This virus was titered on complementing MDCK-SIAT1 cells.
 
@@
 \section{Acknowledgments}
-We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, and AJ Velthuis for helpful scientific suggestions.
+We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, and AJ Velthuis for helpful suggestions.
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.

### 2018-10-05 paper/paper.tex
added GEO accession

--- 2018-10-05 paper/paper.tex
@@
 This pipeline is orchestrated by \texttt{Snakemake}~\citep{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
-The sequencing data and annotated cell-gene matrix are available on the GEO repository under \jdbcomment{data have been uploaded to GEO, but accession not yet available.
-The accession number will be added to the README of the GitHub repo at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell} as soon as it's available}.
+The sequencing data and annotated cell-gene matrix are available on the GEO repository under accession GSE120839 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120839}).
 

### 2018-10-05 paper/paper.tex
added separate Competing Interests section per instructions
revisions to Intro and first section of Results (substantially shortened)

@@
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
-Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
-Viral defects are especially common in IFN+ cells, and we identify at least four type of defects that cause increased IFN induction.
+Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
 However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
@@
 A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
-Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
-This paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.
-
-This rareness of IFN induction is not merely a feature of A549 cells in culture.
-Prior studies using reporters have found that IFN expression is rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
-In support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
-As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes (ISGs), which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
-But just a handful of cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
+For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
+Prior studies using reporters have found that IFN expression is also rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection} (see \FIGSUPP[IFNrare]{mice}).
 
@@
 {\bf (A)} Number of infected and IFN+ A549 cells in our prior single-cell transcriptomics~\citep{russell2018extreme}.
-\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
 {\bf (B)} An A549 reporter cell line to identify cells expressing IFN.
@@
 An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key ISGs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one transcript detected).
+Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
 The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
@@
 
-To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and expanded cell lines from single clones.
-These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
-Cells that activate the IFN reporter can enriched by magnetic-activated cell sorting (MACS) for the cell-surface protein, or identified by flow cytometry for the fluorescent protein.
-The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics described below).
+To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}B).
+Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
+The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in our cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics below).
 Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
-We used the reporter cells to quantify the frequency of IFN induction in A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
-As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
-This frequency of IFN induction is roughly comparable to that observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B).
+We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells---a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}A).
 
@@
 The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
+
+\section{Competing Interests}
 The authors declare that no competing interests exist.

### 2018-10-05 paper/paper.tex
remove temp Word file

@@
 \subsection{Combined transcriptomics and virus-sequencing of single infected cells}
-We next developed a method to determine if features of the infecting virions are responsible for the heterogeneous outcome of infection in single cells.
-Standard single-cell transcriptomic techniques are inadequate for this purpose, since they just use 3'-end sequencing of poly-adenylated mRNAs to measure the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}, and do not reveal if the virions that infect individual cells have mutations.
 
@@
 {\bf (B)}
-The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes using the 10X Chromium platform.
+The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
 {\bf (C)}
@@
 
-We therefore developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of the infecting virions in single cells.
+To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
 First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
-These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
-
-We next used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of the cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter.
-Based on pilot experiments, we expected this MACS to enrich IFN+ cells to $\sim$5\% to 30\% of the total (\FIGSUPP[workflow]{MACS}).
-To ensure the presence of IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
+These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
+From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\FIGSUPP[workflow]{MACS}).
+To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
 We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
 
-We processed the cells on a commercially available platform~\citep[the 10X Chromium;][]{zheng2017massively} that physically isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
+We processed the cells on a commercially available platform~\citep{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
 Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.

### 2018-10-06 paper/paper.tex
remove rug plot from Figure 3B as it is too small to be useful

@@
 A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
+We previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
 Prior studies using reporters have found that IFN expression is also rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection} (see \FIGSUPP[IFNrare]{mice}).
@@
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
-Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
+Green lines show the distribution of cellular mRNAs per cell.
 Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.

### 2018-10-06 paper/paper.tex
added O'Neal 2018 reference

@@
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
-Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
+Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
@@
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
 Our method builds on the observation by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques.

### 2018-10-06 paper/paper.tex
Minor tweaks that slightly shorten 3rd section of Results

@@
 We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
-We also obtained 12 transcriptomes with a mix of human and canine transcripts; these mixes arise when multiple cells are captured in a single droplet.
-From the number of mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
-To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B).
-For the remainder of this paper, we analyze the 1,490 human cells that passed this filtering.
+We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
+To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.
 
@@
 As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 of the 1,490 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
 We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
@@
 We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
-Although ten transcripts are encoded by the eight viral genes, we call presence / absence of genes rather than transcripts.
-The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
+We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
 \FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
-This measured frequency of infected cells expressing all eight genes is higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
-However, it is important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could alter the distribution of viral gene expression from a fully random sample of infected cells.
+This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
 
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel), consistent with our prior work~\citep{russell2018extreme}.
-However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived over 50\% of their mRNA from influenza, while other cells that expressed all viral genes had only a few percent of their mRNA from influenza (\FIG{transcriptomics}D, bottom panel).
-Also consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent across cells (\FIG{transcriptomics}E) and generally matched values obtained by older bulk studies~\citep{hatada1989control}.
+The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
+However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
+Consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\citep{hatada1989control}.
 
 By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
+From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
 Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
-This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
-This moderately high rate of co-infection may also partially explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.
+This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
+This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.
 

### 2018-10-06 paper/paper.tex
slight tweaks to shorten PacBio genotypes section

@@
 Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to estimate the rate of PCR strand exchange (\FIGSUPP[genotypes]{StrandExchange}).
-Less then 3\% of the cDNAs experienced strand exchange that we were not able to identify and filter (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange (\FIGSUPP[genotypes]{StrandExchange}) was rare  (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
 
-We called a mutation in a viral gene if it was found in least two CCSs originating from different mRNAs (unique UMIs) in that cell, and was also present in at least 30\% of all CCSs for that gene in that cell.
+We called a mutation in a viral gene if it was found in least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
 For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
@@
 The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
-However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the apparently flawless virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the apparently flawless virion that infected \textit{cell 105} still induced IFN.
+However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the wild-type virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the wild-type virion that infected \textit{cell 105} still induced IFN.
 

### 2018-10-06 paper/paper.tex
tweaks to shorten viral defects section of results and add statistical test

@@
 {\bf (B)}
-Cells infected by fully wild-type virus express IFN less frequently.
+IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
 {\bf (C)}
@@
 
-Viral defects are also associated with IFN induction, with cells infected by full wild-type virions expressing IFN less often than cells infected by incomplete or mutated virions (\FIG{mutations}B).
-Specific viral defects were especially associated with IFN induction (\FIG{mutations}C).
-Absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched.
-Only the absence of NS remained significant at a false discovery rate of 10\% (\FIG{mutations}C)---although a Bayesian-minded virologist might contend that controlling for all the hypotheses in \FIG{mutations}C is too stringent, since some are more plausible \textit{a priori} given existing biological knowledge~\citep[e.g.,][]{killip2017single, velthuis2018mini, wu2014high}.
-In any case, \FIG{genotypes} and \FIG{mutations}C motivate compelling hypotheses about specific viral defects that contribute to IFN induction, and below we pursue these hypotheses using experimental follow-up.
+Cells infected by incomplete or mutated virions also expressed IFN more frequently than cells infected by full wild-type virions (\FIG{mutations}B), although this difference was not statistically significant ($P = 0.12$, Fisher's exact test).
+However, some specific viral defects were significantly associated with IFN induction: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (\FIG{mutations}C).
+Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.
 

### 2018-10-06 paper/paper.tex
only show P-values in Figure 5C and similar supplements to avoid too much text clutter

@@
 The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
-Numbers give P-values (Fisher's exact test) and associated Q-values~\citep{storey2003statistical} for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
-\FIGSUPP[mutations]{allcells} shows the trends are similar if we include infected cells with incomplete viral sequence information. 
-\FIGSUPP[mutations]{isg} shows the trends are similar if we categorize cells by ISG expression rather than IFN expression.
+Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
+\FIGSUPP[mutations]{allcells} and \FIGSUPP[mutations]{isg} show that the trends are similar if we include infected cells with incomplete viral sequence information, or categorize cells by ISG expression rather than IFN expression.
 {\bf (D)}

### 2018-10-06 paper/paper.tex
modest edits to shorten validation section of Results

@@
 There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-The statistical analysis in \FIG{mutations}D is insufficient to determine if the higher viral transcriptional burden in IFN+ NS-deficient infections is a cause or a consequence of IFN induction, but experiments below shed more light on this question.
 
@@
 The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
-Therefore, the viral defects in single IFN+ cells often cause increased IFN production.
-
-It is tempting to compare the rates of IFN induction measured in \FIG{validation} with the enrichment of viral defects among the IFN+ cells in the single-cell data.
-However, these comparisons are confounded by the fact that the measurements in \FIG{validation} use \emph{bulk} stocks of each viral mutant---and, as shown by the single-cell data in \FIG{genotypes}, many virions in each stock will also have additional mutations.
-So although increased IFN induction by the mutant viral stocks does indicate the mutations are causally associated with IFN, it is impossible to know how many of the IFN+ cells induced by each stock are infected by virions that also have other defects.
-The rate of IFN induction by any given viral mutant can only be accurately quantified by single-cell virus-sequencing.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
+But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
 

### 2018-10-06 paper/paper.tex
remove distinct mechanisms section of Results (last results section)

@@
 
-\subsection{Distinct characteristics of IFN induction by different viral variants}
-
-%%% begin mutantcomparison figure %%%
-\begin{figure}
-
-\centerline{\includegraphics[width=0.85\textwidth]{figures/MutantComparison/p_merged.pdf}}
-\caption{
-For NS1-deficient infections, higher viral gene expression causes more IFN induction.
-{\bf (A)}
-Flow cytometry shows that only for viruses with defects in NS1, there is more IFN induction among infected cells that express higher levels of HA protein.
-The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile.
-Points indicate replicates, and lines indicate the mean.
-See \FIGSUPP[mutantcomparison]{ha_vs_ifn} for details.
-{\bf (B)}
-Blocking viral secondary transcription strongly decreases IFN levels in cells infected with NS1stop virus, but not cells infected with virus carrying a deletion in PB1.
-A549 cells were infected with the viral stocks at doses that induced equivalent IFN, and ribavirin was added at a concentration (100 $\mu$M) sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.
-After eight hours, IFNB1 transcript was quantified relative to the housekeeping gene L32 by qPCR. 
-The dashed gray line indicates the limit of detection.
-\FIGSUPP[mutantcomparison]{cycloheximide} shows that results are similar if secondary transcription is instead blocked with cycloheximide.
-}
-\label{fig:mutantcomparison}
-
-\figsupp[More detailed data on the flow cytometry analysis in \FIG{mutantcomparison}A.]
-{Analysis of the flow cytometry data in \FIGSUPP[validation]{SNPflow} indicates that a higher viral burden is associated with increased IFN-induction in NS1-deficient infections.
-We re-analyzed all the infected cells (defining ``infected'' using the same HA-gating scheme in \FIGSUPP[validation]{SNPflow}) for the wild-type virus, the NS1stop virus, and the three amino-acid mutants with significantly increased IFN induction.
-For each virus and replicate, we binned the infected cells into HA expression quartiles.
-We then calculated the percent of cells that were IFN+ in each quartile (defining ``IFN+'' using the same gating scheme in \FIGSUPP[validation]{SNPflow}).
-The plots show the mean HA expression in each quartile versus the percent of cells that are IFN+.
-The results clearly indicate that for the NS1stop virus, increased viral burden (more HA) correlates with IFN induction.
-A similar trend also holds for the virus with an amino-acid substitution in NS1 (NS1-A122V). 
-But for wild type and the two PB1 mutants, there is no clear trend for higher viral burden to correlate with IFN induction.
-\FIG{mutantcomparison}A summarizes the data in this figure by simply showing the ratio of the percent IFN+ of the highest HA-expression quartile to the lowest HA-expression quartile for each sample and replicate.}
-{\includegraphics[width=0.9\textwidth]{figures/MutantComparison/p_ifn_vs_ha.pdf}}
-\label{figsupp:ha_vs_ifn}
-
-\figsupp[Similar to \FIG{mutantcomparison}B, but uses cycloheximide rather than ribavirin.]
-{Preventing viral secondary transcription using cycloheximide also decreases IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
-However, the results are more complex here, since cycloheximide alone increases IFN production~\citep{raj1981analysis, maroteaux1983cycloheximide, killip2014activation}.
-Therefore, we see an increase in IFN levels upon cycloheximide treatment for uninfected cells and cells infected with the PB1 deletion virus, but a decrease for cells infected with the NS1stop virus.
-This plot shows qPCR results for the same samples as in \FIGSUPP[validation]{cycloheximide}, but with the qPCR for expression of IFNB1 transcripts relative to the housekeeping gene L32.
-As described in \FIGSUPP[validation]{cycloheximide}, A549 cells were infected at an MOI of 0.4, such that HA expression was equivalent between viral variants when secondary transcription was blocked, and some of the samples treated with 50 $\mu$g/ml of cycloheximide~\citep[a concentration sufficient to block secondary transcription;][]{killip2014activation}.
-After 8 hours, transcript levels were quantified by qPCR.
-Points indicate four replicates, and lines indicate the mean.
-The dashed line shows the limit of detection, and points below this limit where set to that value.
-}
-{\includegraphics[width=0.6\textwidth]{figures/MutantComparison/cyclo_qPCR_plot.pdf}}
-\label{figsupp:cycloheximide}
-
-\end{figure}
-%%% end mutantcomparison figure%%%
-
-The single-cell data reveal differences among IFN+ cells depending on the defect in the virion that infects them.
-Specifically, \FIG{mutations}D shows that if a cell is infected by a virion that fails to express NS, then a higher viral transcriptional load correlates with IFN induction---but that no such association is present among infected cells that express NS.
-To validate this conclusion, we used flow cytometry to compare the frequency of IFN induction among infected cells that express low or high levels of a viral protein (HA).
-\FIG{mutantcomparison}A shows that upon infection with stocks of the NS1stop or NS1-A122V mutant, cells that express high levels of HA are much more likely to be IFN+ than those with low HA expression.
-In contrast, if cells are infected with stocks of wild-type or PB1 amino-acid mutant virus, there is only a slight increase in IFN induction among cells with high HA expression.
-
-To determine if higher viral transcriptional burden \emph{causes} IFN induction in NS-deficient infections, we treated infected cells with two different chemicals (ribavirin and cycloheximide) that inhibit secondary viral transcription, and so greatly reduce the level of viral transcripts~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg, killip2014activation}.
-As shown in \FIG{mutantcomparison}A and \FIGSUPP[mutantcomparison]{cycloheximide}, both inhibitors markedly reduced IFN induction by cells infected with the NS1stop viral stock, but not by cells infected the virus carrying a deletion in PB1.
-This result indicates that high levels of viral transcription promote IFN induction in NS-deficient infections, and suggests that there are distinct mechanisms of IFN induction by different viral variants---perhaps explaining why prior studies have not reached a consensus on the process by which influenza induces IFN~\citep{rehwinkel2010rig, weber2013incoming, killip2014activation}.
-
 \section{Discussion}
@@
 
-For the qPCR in \FIGSUPP[validation]{delqPCR} and \FIGSUPP[mutantcomparison]{cycloheximide}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+For the qPCR in \FIGSUPP[validation]{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
@@
 After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.
-
-For the qPCR in \FIG{mutantcomparison}B, A549 cells were infected with the NS1stop virus at a MOI of 0.1 (based on TCID50) in MDCK-SIAT1 and with the PB1del385to2163 virus at a dose that induced equivalent IFN in the absence of ribavirin.
-The ribavirin was added to cells at the time of infection to a final concentration of 100$\mu$M, which is sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,Reuther:2015ef,Scholtissek:1976wg}.
-All other steps were performed as for the cycloheximide experiments described immediately above..
 

### 2018-10-06 paper/paper.tex
edits to rest of Results (removed last section) and slight tweaks to Discussion

@@
 Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
-Visual inspection of \FIG{genotypes} makes it glaringly obvious how any experiment that does not directly sequence viral genes in single cells is really averaging across a vast number of hidden viral defects.
+Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
 
-It turns out that these viral defects explain much of the heterogeneity in viral transcriptional burden observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
-If we examine only cells infected by truly wild-type virions, the heterogeneity in viral transcriptional burden is substantially reduced.
-
-Viral defects also help determine if an infected cell triggers an innate-immune response.
+It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if an infected triggers an IFN response.
 We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
@@
 Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Given the extensive prior work describing such deletions, why did we not identify more of them in our single IFN+ cells?
+Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
 There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-But the relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
+The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
 
@@
 Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
-Real human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
+Natural human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
 It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.

### 2018-10-06 paper/paper.tex
adjusted acknowledgments

@@
 \author[1]{Alistair B. Russell}
-\author[1]{Jacob R. Kowalsky}
 \author[1,2,3*]{Jesse D. Bloom}
@@
 \section{Acknowledgments}
-We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, and AJ Velthuis for helpful suggestions.
+We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, and Katherine Xue for helpful suggestions.
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
-We thank Katherine Xue, David Bacsik, and Hannah Itell for comments on the manuscript.
 This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
 ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
-JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
 JDB is an Investigator of the Howard Hughes Medical Institute.

### 2018-10-06 paper/paper.tex
added three sentences at end of results to point out lack of evidence for summation of genetic defects;

@@
 But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
+Curiously, none of the mutations tested induce interferon deterministically (ie in all infected cells), indicating additional heterogeneity in response driven by unsampled heterogeneity in viral samples or host signaling pathways.
+Due to our single-cell approach, we can determine that it is unlikely that this heterogeneity is simply a summation of those immunostimulatory genetic features described here, as neither of the immunostimulatory PB1 mutations were identified in interferon-producing infections lacking the NS gene segment or possessing an internally-deleted segment, two of the NS null infections found in interferon positive cells wherein infection was  otherwise entirely wild-type in sequence, and the PB1 deletion was found in an infection that was wild-type other than a mutation in NA, unlikely to have precipitated interferon induction.
+Therefore there remains considerable heterogeneity in interferon induction beyond those genetic features identified here.
 

### 2018-10-06 paper/paper.tex
added Jacob to author list and his funding to acknowledgments
moved point about how no mutations deterministically induce IFN from results to discussion

--- 2018-10-06 paper/paper.tex
@@
 
-Our results also indicate that different viral defects induce IFN by somewhat distinct processes.
-For instance, higher viral transcription is associated with IFN induction only in cells infected by virions with deficiencies in NS.
-This finding is consistent with the suggestion by \citet{sjaastad2018distinct} that viral genetic variation relates to cell-to-cell variability in gene expression.
-Larger-scale studies similar to the one performed here should be useful for further elucidating the extent that virions with different defects cause cells to launch distinct transcriptional programs.
-
 But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
 We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency mutations in cells expressing predominantly wild-type viral genes.
-However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.
+Additionally, none of the viral defects in IFN+ cells induce IFN deterministically in our validation experiments---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that might contribute to IFN induction.
+However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\citep{oneal2018west}.
 

### 2018-10-06 paper/paper.tex
_bioRxiv_ and initial _eLife_ submission version

@@
 Additionally, none of the viral defects in IFN+ cells induce IFN deterministically in our validation experiments---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that might contribute to IFN induction.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
 However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\citep{oneal2018west}.

### 2018-10-07 paper/paper.tex
fixed a few minor typos

@@
 
-To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}B).
+To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
 As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
 We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
-The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
+The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
 As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
-Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}C, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
+Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
@@
 
-It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if an infected triggers an IFN response.
+It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
 We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
@@
 \section{Acknowledgments}
-We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, and Katherine Xue for helpful suggestions.
+We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.

### 2018-10-08 paper/paper.tex
formatting changes requested by _eLife_

@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
+The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity~\citep{russell2018extreme}.  
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
@@
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
-This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
-JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
-JDB is an Investigator of the Howard Hughes Medical Institute.
-The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
+%This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
+%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+%JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
+%JDB is an Investigator of the Howard Hughes Medical Institute.
+%The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
 

### 2018-11-15 paper/paper.tex
made the useful suggestions from eLife reviews

@@
 {\bf (C)}
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
 The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
@@
 {\bf (D)}
-viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
+viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
 {\bf (E)}
@@
 \FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected.
-{\bf (E)} Relative expression of viral genes among infected cells.
+{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
 {\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
@@
 {\bf (A)}
-Percent of cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line (see \FIGSUPP[validation]{SNPflow} for details).
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line (see \FIGSUPP[validation]{SNPflow} for details).
 The numbers indicate the median of four measurements for each viral mutant.

### 2018-11-15 paper/paper.tex
upated suppfile labels

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ELIFE ARTICLE TEMPLATE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% PREAMBLE 
-\documentclass[9pt,lineno]{elife}
-% Use the onehalfspacing option for 1.5 line spacing
-% Use the doublespacing option for 2.0 line spacing
-% Please note that these options may affect formatting.
-% Additionally, the use of the \newcommand function should be limited.
+\documentclass[9pt,lineno]{template}
 
@@
 \newfloat{suppfile}{thp}{lofsupfile}
-%\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
-\floatname{suppfile}{Supplementary file}
+\floatname{suppfile}{Supplementary File}
 
@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity~\citep{russell2018extreme}.  
+The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
 Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
@@
 Note that these sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
-(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in Supplementary~file~\ref{suppfile:pacbio_analysis}.)
+(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in \SUPPFILE{pacbio_analysis}.)
 Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
@@
 We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
-All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} for at detailed schematic).
+All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} for at detailed schematic).
 However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
@@
 The material from these eight segment-specific PCRs was then pooled with the goal of obtaining a equimolar ratio of segments, and sequenced on one SMRT Cell in a PacBio RS II and one SMRT Cell of a PacBio Sequel.
-Detailed results from the analysis of these first two sequencing runs is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and Supplementary~file~\ref{suppfile:pacbio_analysis}.
+Detailed results from the analysis of these first two sequencing runs is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and \SUPPFILE{pacbio_analysis}.
 These results showed that although the PCRs substantially enriched for influenza molecules, the relative coverage of the different viral genes was still very uneven, with the longer genes (especially the polymerase genes) severely under-sampled.
@@
 The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\citep{xue2016propagation} and are presumably preferentially amplified during PCR.
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\citep{xue2016propagation} and are presumably preferentially amplified during PCR.
 
@@
 We pooled the products of these reactions from this size-selection and sequenced on a SMRT Cell of a PacBio Sequel.
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), this sequencing yielded modestly more full-length polymerase variants, but they were still heavily undersampled compared to other viral genes.
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), this sequencing yielded modestly more full-length polymerase variants, but they were still heavily undersampled compared to other viral genes.
 
@@
 This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRT Cell of PacBio Sequel. 
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}), this process efficiently yielded many full-length polymerase variants.
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), this process efficiently yielded many full-length polymerase variants.
 
@@
 The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[genotypes]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[genotypes]{StrandExchange} for an illustration of how this is determined).
-A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also Supplementary~file~\ref{suppfile:pacbio_analysis}).
+A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
 Importantly, the various PCR biases and enrichment schemes mean that the coverage of molecules by the PacBio sequencing is not proportional to their original abundance in the starting mRNA.
@@
 The Python code has been implemented in the API of \texttt{dms\_tools2}~\citep[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
-A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:pacbio_analysis}.
+A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{pacbio_analysis}.
 We refer the reader to this notebook for a detailed description and extensive plots showing the results at each step.
@@
 \end{enumerate}
-The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or Supplementary~file~\ref{suppfile:pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
+The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
 We could call the sequences of all expressed viral genes in about half of the infected cells (\FIGSUPP[genotypes]{ncells}).
@@
 This analysis utilized a variety of R and Bioconductor~\citep{huber2015orchestrating} packages, including \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
-A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as Supplementary~file~\ref{suppfile:monocle_analysis}.
+A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{monocle_analysis}.
 We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.

### 2018-11-15 paper/paper.tex
removed Steuerman et al re-analysis

@@
 We previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
-Prior studies using reporters have found that IFN expression is also rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection} (see \FIGSUPP[IFNrare]{mice}).
+Prior studies using reporters have found that IFN expression is also rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
 
@@
 \label{fig:IFNrare}
-
-\figsupp[Few cells express detectable IFN in influenza-infected mice.]
-{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
-Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
-Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
-As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
-An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
-The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
-}
-{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
-\label{figsupp:mice}
 

### 2018-11-15 paper/paper.tex
tweaks to intro

--- 2018-11-15 paper/paper.tex
@@
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
-As the virus spreads, some infected cells trigger innate immunity and begin producing interferon (IFN).
+As the virus spreads, some cells detect infection and begin producing interferon (IFN).
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
-If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and more severe disease~\citep{la2007question, iwasaki2014innate}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.
 
-Unfortunately for the host, influenza infection only rarely triggers IFN production by infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
+Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.

### 2018-11-15 paper/paper.tex
converted Steuerman re-analysis to supp fig

@@
 \floatname{suppfile}{Supplementary File}
+
+\newfloat{suppfig}{thp}{lofsupfile}
+\floatname{suppfig}{Supplementary Figure}
 
@@
 A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-We previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
-Prior studies using reporters have found that IFN expression is also rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
+We previously performed single-cell transcriptomics on influenza-infected A549 human lung epithelial cells, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN~\citep{russell2018extreme}.
+Prior studies using reporters have found that IFN expression is also rare in influenza-infected cells~\citep{killip2017single} and mice~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).
 
@@
 
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}
+\caption{
+A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
+Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
+Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
+As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
+An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
+Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
+The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
+}
+\label{suppfig:mice}
+\end{suppfig}
+
+\clearpage
+
 \begin{suppfile}

### 2018-11-15 paper/paper.tex
moved IFN reporter to suppfile

@@
 
-\subsection{A system to identify and enrich for rare IFN+ positive cells}
+\subsection{A system to identify and enrich for rare IFN+ cells}
 A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
@@
 \centerline{
-{\bf \Large A}\includegraphics[width=0.28\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
+{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
 \hspace{0.02\textwidth}
-{\bf \Large B}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
-\hspace{0.02\textwidth}
-{\bf \Large C} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
+{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
 }
 \caption{
-IFN expression is rare in influenza-infected cells.
-{\bf (A)} Number of infected and IFN+ A549 cells in our prior single-cell transcriptomics~\citep{russell2018extreme}.
-{\bf (B)} An A549 reporter cell line to identify cells expressing IFN.
+Reporter cells to identify and enrich IFN+ cells.
+{\bf (A)} An A549 IFN reporter cell line.
 The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
-We created reporters with type I and type III IFN promoters (\FIGDATA[IFNrare]{reporter_sequences}).
+We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
 The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
-{\bf (C)}
+{\bf (B)}
 Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
@@
 \label{figsupp:flow}
-
-\figdata{Sequences of the reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
-}
-\label{figdata:reporter_sequences}
 
@@
 Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\citep{kim2011high}.
-The sequence of the last of these constructs is provided in \FIGDATA[IFNrare]{reporter_sequences}.
+The sequence of the last of these constructs is provided in \SUPPFILE{reporter_sequences}.
 
@@
 Downstream of this promoter we cloned LNGFR$\Delta$C linked to ZsGreen via a P2A linker.
-The sequence of this construct is provided in \FIGDATA[IFNrare]{reporter_sequences}.
+The sequence of this construct is provided in \SUPPFILE{reporter_sequences}.
 
@@
 \begin{suppfile}
+\caption{Sequences of the IFN reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
+}
+\label{suppfile:reporter_sequences}
+\end{suppfile}
+
+\begin{suppfile}
 \caption{\label{suppfile:pacbio_analysis}

### 2018-11-15 paper/paper.tex
changes to Figure 1

@@
 We previously performed single-cell transcriptomics on influenza-infected A549 human lung epithelial cells, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN~\citep{russell2018extreme}.
-Prior studies using reporters have found that IFN expression is also rare in influenza-infected cells~\citep{killip2017single} and mice~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).
+Prior studies using reporters have also found that IFN expression is rare in influenza-infected cells~\citep{killip2017single} and mice~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).
 
@@
 We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
-The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
+The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reporter_validation}).
 {\bf (B)}
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \SUPPFIG{flow} for details).
 The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
@@
 \label{fig:IFNrare}
-
-\figsupp[Validation of IFN reporter cell lines.]
-{To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
-The name of each of reporter cell line is indicated at the top of each row of plots.
-At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
-Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
-}
-{\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}}
-\label{figsupp:reporter_validation}
-
-\figsupp[Expression of type I and type III IFN is highly correlated in A549 cells.]
-{An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
-The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
-After 13 hours, cells were analyzed by flow cytometry.
-Cells positive for either fluorescent reporter were further analyzed.
-As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.}
-{\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}}
-\label{figsupp:type_I_vs_III}
-
-\figsupp[Full flow cytometry data for \FIG{IFNrare}C.]
-{Flow cytometry data for \FIG{IFNrare}C.
-The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
-After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
-Each condition was done in triplicate.
-The contour plots show the density of all cells, and all IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
-For the influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
-For the uninfected and Sendai-virus infected, the percentage IFN+ was calculated among all cells, since these cells do not express HA.
-}
-{\includegraphics[width=0.75\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}}
-\label{figsupp:flow}
-
 \end{figure}
@@
 
-To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}B).
+To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in our cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics below).
+The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{type_I_vs_III}; further validated by the single-cell transcriptomics below).
 Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
-We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells---a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}A).
+We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
 
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
+Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
@@
 \subsection{IFN reporter cell lines}
-We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}B).
+We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}A).
 The parental A549 cell line used to create these reporters was obtained from ATCC (CCL-185), and was tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core and authenticated using the ATCC STR profiling service.
@@
 Importantly, the cells that we retained for further use were not the same portion that were tested by poly(I:C) treatment, but rather a separate split of the same population---this avoids any selection on the cells from transient activation of IFN.
-For the dual type I / type III reporter used in \FIGSUPP[IFNrare]{type_I_vs_III}, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
+For the dual type I / type III reporter used in \SUPPFIG{type_I_vs_III}, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
 All reporter lines tested negative for mycoplasma contamination by the Fred Hutch Genomics Core.
 
-\FIGSUPP[IFNrare]{reporter_validation} shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
+\SUPPFIG{reporter_validation} shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
 For detection of the cell-surface bound LNGFR$\Delta$C, cells were stained with PE-conjugated anti-LNGFR (CD271) antibody from Miltenyi Biotec.
@@
 
+\begin{suppfig}
+\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}
+\caption{To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
+The name of each of reporter cell line is indicated at the top of each row of plots.
+At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
+Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
+}
+\label{suppfig:reporter_validation}
+\end{suppfig}
+
+\clearpage
+
+\begin{suppfig}
+\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
+\caption{
+Expression of type I and type III IFN is highly correlated in A549 cells.
+An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
+The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
+After 13 hours, cells were analyzed by flow cytometry.
+Cells positive for either fluorescent reporter were further analyzed.
+As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.}
+\label{suppfig:type_I_vs_III}
+\end{suppfig}
+
+\clearpage
+
+\begin{suppfig}
+\includegraphics[width=0.75\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
+\caption{Flow cytometry data for \FIG{IFNrare}B.
+The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
+After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
+Each condition was done in triplicate.
+The contour plots show the density of all cells, and all IFN+ cells are also indicated by orange dots.
+Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
+For the influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
+For the uninfected and Sendai-virus infected, the percentage IFN+ was calculated among all cells, since these cells do not express HA.
+}
+\label{suppfig:flow}
+\end{suppfig}
+
+\clearpage
+
 \begin{suppfile}
-\caption{Sequences of the IFN reporters in \FIG{IFNrare}B are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
+\caption{Sequences of the IFN reporters in \FIG{IFNrare}A are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
 }

### 2018-11-15 paper/paper.tex
adjusted supplements to figure 2

@@
 IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
-IFN+ cells are enriched by MACS (\FIGSUPP[workflow]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
+IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
 {\bf (B)}
@@
 
-\figsupp[MACS enrichment of IFN+ cells.]
-{
-Example MACS enrichments of IFN+ influenza-infected cells.
-A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
-After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
-{\bf (A)}
-After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
-The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
-As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
-We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
-{\bf (B)}
-Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
-The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
-The qPCR was performed in quadruplicate (hence the four points for each sample).
-}
-{\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}}
-\label{figsupp:MACS}
-
-\figdata{Genbank files giving sequences of the wild-type and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
-\label{figdata:virus_seqs}
-
 \end{fullwidth}
@@
 To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
-First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
+First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
 These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\FIGSUPP[workflow]{MACS}).
+From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
 To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
@@
 In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
-The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \FIGDATA[workflow]{virus_seqs}.
+The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
 These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
@@
 To enrich IFN+ cells, the cells were then incubated with anti-LNGFR MACSelect Microbeads (Miltenyi Biotec) and twice passed over an MS magnetic column (Miltenyi Biotec), retaining the bound (and presumably IFN-enriched) population each time.
-This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \FIGSUPP[workflow]{MACS}.
+This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \SUPPFIG{MACS}.
 The original, unsorted, population was then added back in to $\sim$10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
@@
 The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
-The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \FIGDATA[workflow]{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
+The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \SUPPFILE{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
 The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
@@
 
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}
+\caption{
+Example MACS enrichments of IFN+ influenza-infected cells.
+A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
+After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
+{\bf (A)}
+After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
+The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
+As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
+We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
+{\bf (B)}
+Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
+The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
+The qPCR was performed in quadruplicate (hence the four points for each sample).
+}
+\label{suppfig:MACS}
+\end{suppfig}
+
+\clearpage
+
 \begin{suppfile}
@@
 \label{suppfile:reporter_sequences}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+Genbank files giving sequences of the wild-type and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
+\label{suppfile:virus_seqs}
 \end{suppfile}

### 2018-11-15 paper/paper.tex
adjusted supplements for figure 3

@@
 The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
-\FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected.
+\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
 {\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
@@
 A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
-Many cells that do not express IFN still express ISGs (\FIGSUPP[transcriptomics]{ISG}).
+Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
 }
 \label{fig:transcriptomics}
-
-\figsupp[Fraction of infected cells that detectably express each influenza gene.]
-{The fraction of infected cells that are called as expressing each viral gene.
-The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
-Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
-The exception is NP, which is detected in virtually all infected cells.
-The much higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments because the number of viral mRNAs is below the detection limit.}
-{\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}}
-\label{figsupp:frac_has_gene}
-
-\figsupp[Expression of type I and type III IFN is highly correlated in single cells in our experiment.]
-{The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
-Each point represents one cell.
-The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
-Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
-}
-{\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}}
-\label{figsupp:type_I_III_correlation}
-
-\figsupp[Expression of ISGs in single infected and uninfected cells.]
-{
-For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
-{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
-We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
-Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
-This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
-{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
-Each point represents one cell, and the Pearson correlation coefficient is shown.
-IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
-Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
-}
-{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}}
-\label{figsupp:ISG}
-
-\figsupp[Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.]
-{To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
-In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
-Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA, shown on a log (top) or linear (bottom) scale.
-As is clear from this plot, expression of influenza, IFN, and ISG genes contributes substantially to the structure of the data, since cells with high expression of these genes clearly group together.
-}
-{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_tsne.pdf}}
-\label{figsupp:tSNE}
 
@@
 We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
+\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
 This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
@@
 
-We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
+We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
 Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
-Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\FIGSUPP[transcriptomics]{ISG}A).
-The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\FIGSUPP[transcriptomics]{ISG}B).
+Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{ISG}A).
+The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
 These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.
@@
 Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
-\FIGSUPP[transcriptomics]{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
+\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
 Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.
@@
 {This plot shows the same information as panels \FIG{mutations}B-D except that it shows the association of viral gene absence / mutations with ISG expression rather than IFN expression.
-Cells are classified as ISG+ or ISG- as in \FIGSUPP[transcriptomics]{ISG}.
+Cells are classified as ISG+ or ISG- as in \SUPPFIG{ISG}.
 The major viral features that associate with IFN induction also associate with ISG expression.
@@
 We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
-We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \FIGSUPP[transcriptomics]{frac_has_gene}).
+We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{frac_has_gene}).
 To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
-We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}, respectively.
+We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{ISG}, respectively.
 We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
@@
 
+\begin{suppfig}
+\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}
+\caption{The fraction of infected cells that are called as expressing each viral gene.
+The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
+Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+The exception is NP, which is detected in virtually all infected cells.
+The much higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments because the number of viral mRNAs is below the detection limit.}
+\label{suppfig:frac_has_gene}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}
+\caption{
+The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
+Each point represents one cell.
+The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
+Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
+}
+\label{suppfig:type_I_III_correlation}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}
+\caption{
+Expression of ISGs in single infected and uninfected cells.
+For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
+{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
+We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
+Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
+This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
+{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
+Each point represents one cell, and the Pearson correlation coefficient is shown.
+IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
+Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
+}
+\label{suppfig:ISG}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/single_cell_figures/p_tsne.pdf}
+\caption{
+Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.
+To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
+In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
+Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA, shown on a log (top) or linear (bottom) scale.
+As is clear from this plot, expression of influenza, IFN, and ISG genes contributes substantially to the structure of the data, since cells with high expression of these genes clearly group together.
+}
+\label{suppfig:tSNE}
+\end{suppfig}
+
+\clearpage
+
 \begin{suppfile}

### 2018-11-15 paper/paper.tex
adjusted supplements for Figure 4

@@
 {\centering
-\includegraphics[height=0.78\textheight]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 
-\figsupp[Number of PacBio CCSs and PCR strand exchange rate.]
-{
-The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
-Note that these sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
-(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in \SUPPFILE{pacbio_analysis}.)
-Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
-Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
-Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
-Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences derived from the same cDNA, and cell barcodes were used to collapse duplicate sequences from the same cell.
-The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \FIGSUPP[genotypes]{StrandExchange} for more explanation).
-From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
-Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
-Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
-A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
-}
-{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
-\label{figsupp:CCSs}
-
-\figsupp[Strategy for detecting strand exchange during sequencing of full-length viral genes.]
-{The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
-A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
-We can detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
-If there is no strand exchange, all molecules should either be wild-type or have the synonymous barcoding mutations at \emph{both} termini.
-But strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
-\FIGSUPP[genotypes]{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
-Note that since the rate of homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
-}
-{\includegraphics[width=0.6\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}}
-\label{figsupp:StrandExchange}
-
-\figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
-{Cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).
-{\bf (A)} We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
-{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
-Both facts makes sense.
-Cells with more viral mRNA are more likely to have their viral cDNA captured in the PacBio sequencing, which is only captures a small fraction of the total transcripts identified by the 3'-end sequencing transcriptomic sequencing.
-The lower calling rate for dual-barcode co-infections is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing.
-An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
-Note also that this plot is limited to the cells that were called as infected (\FIG{transcriptomics}C) and could clearly be classified as IFN- or IFN+ (\FIG{transcriptomics}G).
-}
-{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}
-\label{figsupp:ncells}
-
-\figsupp[Genotypes and infection outcomes plotted separately for IFN+ and IFN- cells.]
-{This plot shows the same data as \FIG{genotypes}, but with cells separated into columns based on whether they are IFN- (left column) or IFN+ (right column).}
-{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_genotypes_by_ifn.pdf}}
-\label{figsupp:genotypes_by_ifn}
-
-\figdata{A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
-}
-\label{figdata:primer_sequences}
-
-\figdata{A CSV file giving the genotypes is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
-\label{figdata:genotypes}
-
 \end{fullwidth}
@@
 We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
-Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\FIGSUPP[genotypes]{CCSs}).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange (\FIGSUPP[genotypes]{StrandExchange}) was rare  (\FIGSUPP[genotypes]{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange (\SUPPFIG{StrandExchange}) was rare (\SUPPFIG{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
 
@@
 
-We could call the sequences of all expressed viral genes in the majority of infected cells (\FIGSUPP[genotypes]{ncells}).
-We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\FIGSUPP[genotypes]{ncells}).
+We could call the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
 But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
@@
 We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
-In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\FIGSUPP[genotypes]{StrandExchange}).
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
 The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
@@
 We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (although some viral segments produce multiple splice forms, all mRNAs from a given segment share the same 5' end).
-The sequences of these primers are in \FIGDATA[genotypes]{primer_sequences}.
-A major concern during these PCRs is strand exchange (see \FIGSUPP[genotypes]{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
+The sequences of these primers are in \SUPPFILE{primer_sequences}.
+A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
 To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
@@
 In the downstream computational analysis, we can then determine the full sequence of the gene as well as the cell barcode of the initial molecule from which the linearized molecule is derived.
-Specifically, we first used 2.5 ng of our already-amplified segment-specific material in a 10-cycle PCR to append circularization adapters (see \FIGDATA[genotypes]{primer_sequences} for sequences), and cleaned the resultant mixture using SPRIselect beads at a volume ratio of 0.4.
+Specifically, we first used 2.5 ng of our already-amplified segment-specific material in a 10-cycle PCR to append circularization adapters (see \SUPPFILE{primer_sequences} for sequences), and cleaned the resultant mixture using SPRIselect beads at a volume ratio of 0.4.
 We then used 10 ng of this amplified material in a 20$\mu$l NEBuilder reaction using an extended reaction time of 50 minutes in order to circularize the molecules.
@@
 The circularized and digested material was then cleaned using SPRIselect beads at a volume ratio of 0.4.
-This material was then used as template for three non-emulsion PCRs specific to PB2, PB1, or PA, using two segment-specific primers that align to the central portion of each gene but in apposition to each another (see \FIGDATA[genotypes]{primer_sequences} for sequences).
+This material was then used as template for three non-emulsion PCRs specific to PB2, PB1, or PA, using two segment-specific primers that align to the central portion of each gene but in apposition to each another (see \SUPPFILE{primer_sequences} for sequences).
 These linearization reactions used 20 (PB2) or 26 (PB1 and PA) PCR cycles, and the resulting products were cleaned using SPRIselect beads at a volume ratio of 1.0.
@@
 The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
-The number of sequences obtained for each gene after pooling the data from all reactions is shown in \FIGSUPP[genotypes]{CCSs}, which also indicates that the net rate of strand exchange is very low (see \FIGSUPP[genotypes]{StrandExchange} for an illustration of how this is determined).
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{StrandExchange} for an illustration of how this is determined).
 A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
@@
 We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
-At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\FIGSUPP[genotypes]{StrandExchange}), and reassuringly found that strand exchange was rare (\FIGSUPP[genotypes]{CCSs}).
+At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{StrandExchange}), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}).
 The small number of CCSs with identifiable strand exchange were filtered from further analysis.
@@
 The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
-We could call the sequences of all expressed viral genes in about half of the infected cells (\FIGSUPP[genotypes]{ncells}).
-The mutations called using this pipeline are shown in \FIG{genotypes}, and \FIGDATA[genotypes]{genotypes} gives the number of CCSs supporting each mutation call.
+We could call the sequences of all expressed viral genes in about half of the infected cells (\SUPPFIG{ncells}).
+The mutations called using this pipeline are shown in \FIG{genotypes}, and \SUPPFILE{genotypes} gives the number of CCSs supporting each mutation call.
 The called sequences of the viral genes were added to the annotated cell-gene matrix.
@@
 
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
+\caption{
+The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
+Note that these sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
+(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in \SUPPFILE{pacbio_analysis}.)
+Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
+Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
+Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
+Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences derived from the same cDNA, and cell barcodes were used to collapse duplicate sequences from the same cell.
+The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \SUPPFIG{StrandExchange} for more explanation).
+From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
+Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
+A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
+}
+\label{suppfig:CCSs}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.6\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
+\caption{
+Strategy for detecting strand exchange during sequencing of full-length viral genes.
+The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
+A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
+We can detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
+If there is no strand exchange, all molecules should either be wild-type or have the synonymous barcoding mutations at \emph{both} termini.
+But strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
+\SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
+Note that since the rate of homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
+}
+\label{suppfig:StrandExchange}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}
+\caption{Number of cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).
+{\bf (A)} We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
+{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
+Both facts makes sense.
+Cells with more viral mRNA are more likely to have their viral cDNA captured in the PacBio sequencing, which is only captures a small fraction of the total transcripts identified by the 3'-end sequencing transcriptomic sequencing.
+The lower calling rate for dual-barcode co-infections is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing.
+An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
+Note also that this plot is limited to the cells that were called as infected (\FIG{transcriptomics}C) and could clearly be classified as IFN- or IFN+ (\FIG{transcriptomics}G).
+}
+\label{suppfig:ncells}
+\end{suppfig}
+
+\clearpage
+
 \begin{suppfile}
@@
 \label{suppfile:virus_seqs}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
+}
+\label{suppfile:primer_sequences}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{A CSV file giving the genotypes is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
+\label{suppfile:genotypes}
 \end{suppfile}

### 2018-11-15 paper/paper.tex
adjusted supp figs for Figure 5

@@
 Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
-\FIGSUPP[mutations]{allcells} and \FIGSUPP[mutations]{isg} show that the trends are similar if we include infected cells with incomplete viral sequence information, or categorize cells by ISG expression rather than IFN expression.
 {\bf (D)}
@@
 \label{fig:mutations}
-
-\figsupp[Analysis as in \FIG{mutations}C but including infected cells with incomplete viral sequence information.]
-{This plot differs from \FIG{mutations}C in that it also includes data from cells for which some viral genes were not fully sequenced.
-For incompletely sequenced cells, deletions and substitutions are included in the counts when that particular viral gene is sequenced.
-The major trends in \FIG{mutations}C are also true for the larger dataset in this figure.
-In addition, there is now a modest trend for infected cells that fail to express PB2 and PA \emph{not} to express IFN.
-}
-{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_muts_ifn_all.pdf}}
-\label{figsupp:allcells}
-
-\figsupp[Association of viral genetic variation and ISG expression.]
-{This plot shows the same information as panels \FIG{mutations}B-D except that it shows the association of viral gene absence / mutations with ISG expression rather than IFN expression.
-Cells are classified as ISG+ or ISG- as in \SUPPFIG{ISG}.
-The major viral features that associate with IFN induction also associate with ISG expression.
-Specifically, cells that express wild-type copies of all eight viral genes tend to express ISGs less frequently and at lower levels
-The absence of NS is strongly associated with higher ISG expression.
-Viral burden is not associated with ISG expression levels in NS-sufficient infections, but is in NS-deficient infections.
-}
-{\includegraphics[width=\textwidth]{figures/single_cell_figures/p_mutations_isg.pdf}}
-\label{figsupp:isg}
-
-\figdata{A CSV file giving the viral mutations and related information in \FIG{mutations} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.}
-\label{figdata:mutations}
 
@@
 \begin{suppfile}
-\caption{A CSV file giving the genotypes is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
+\caption{A CSV file giving the genotypes in \FIG{genotypes} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
 \label{suppfile:genotypes}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{A CSV file giving the viral mutations and related information in \FIG{mutations} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.}
+\label{suppfile:mutations}
 \end{suppfile}

### 2018-11-15 paper/paper.tex
adjusted supplements for Figure 6

@@
 {\bf (A)}
-Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line (see \FIGSUPP[validation]{SNPflow} for details).
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line (see \SUPPFIG{SNPflow} for details).
 The numbers indicate the median of four measurements for each viral mutant.
@@
 Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
-See \FIGSUPP[validation]{delflow} and \FIGSUPP[validation]{delqPCR} for more detailed experimental data.
+See \SUPPFIG{delflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
 The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
@@
 \label{fig:validation}
-
-\figsupp[Flow cytometry data for \FIG{validation}A.]
-{A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected (three replicates) or infected with stocks of the indicated viral mutant in quadruplicate.
-After 13 hours, cells were stained for expression of HA protein and analyzed by FACS for HA and the ZsGreen driven by the \textit{IFNL1} reporter.
-The contour plots show the density of all cells, and IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
-For all influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
-For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
-For each viral mutant, two independently generated stocks were each assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other viral stock).
-The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
-The HA staining indicates that the number of transcriptionally active viral units per cell was similar across most mutants, and calculating percentage IFN+ only among HA+ cells largely corrects for modest variation in titer across mutants.
-}
-{\includegraphics[height=0.71\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}}
-\label{figsupp:SNPflow}
-
-\figsupp[Flow cytometry data for \FIG{validation}B.]
-{This figure is similar to \FIGSUPP[validation]{SNPflow} except that it shows the flow cytometry data for \FIG{validation}B.
-Each infection was performed in triplicate.
-A complicating factor for these data is that the virus with the deletion in PB1 (PB1del385to2163) cannot perform secondary transcription because it does not express PB1 protein, and so cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
-Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
-Finally, the approach in \FIGSUPP[validation]{delqPCR} was used to show that equivalent TCID50s, all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
-All this normalization suggests that all infections were performed at a similar dose of particles active for primary transcription.
-Therefore, the percent IFN+ was calculated from the flow data in this figure for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for the PB1del385to2163 virus.
-}
-{\includegraphics[width=0.7\textwidth]{figures/Validation_Figure/del_flow_plot.pdf}}
-\label{figsupp:delflow}
-
-\figsupp[Validation of viral titer normalization for \FIG{validation}B.]
-{Validation that the infections in \FIG{validation}B and \FIGSUPP[validation]{delflow} were performed at similar doses of virions capable of initiating primary transcription.
-In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \FIGSUPP[validation]{delflow}), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
-The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\citep{killip2014activation}.
-In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
-But in the presence of cycloheximide, all viruses produced similar amounts of viral mRNA, indicating that the dose of particles active for primary transcription is roughly equivalent across variants.
-Each measurement was performed in quadruplicate.
-}
-{\includegraphics[width=0.6\textwidth]{figures/Validation_Figure/qpcr_plot.pdf}}
-\label{figsupp:delqPCR}
 
@@
 
-For the qPCR in \FIGSUPP[validation]{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+For the qPCR in \SUPPFIG{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
@@
 At 24 hours post-transfection, we changed media from D10 to influenza growth media. 
-At 50 hours post-transfection (for the replicate 1 viruses in \FIGSUPP[validation]{SNPflow}) or 72 hours (for the replicate 2 viruses in \FIGSUPP[validation]{SNPflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
+At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{SNPflow}) or 72 hours (for the replicate 2 viruses in \SUPPFIG{SNPflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \FIGSUPP[validation]{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
+For the infections in \SUPPFIG{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
 Two-fold serial dilutions of virus were added to an equal volume of diluted red blood cells, and titer was measured as the highest dilution of viral stock at which complete hemagglutination of red blood cells was observed.
-We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \FIGSUPP[validation]{SNPflow}.
+We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{SNPflow}.
 
-To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \FIGSUPP[validation]{delflow}, we used slightly different procedures.
+To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{delflow}, we used slightly different procedures.
 The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
@@
 The wild-type and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
-The infections in \FIGSUPP[validation]{delflow} were performed at equivalent TCID50s as described in the legend to that figure.
-That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \FIGSUPP[validation]{delqPCR}.
+The infections in \SUPPFIG{delflow} were performed at equivalent TCID50s as described in the legend to that figure.
+That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \SUPPFIG{delqPCR}.
 
@@
 
+\begin{suppfig}
+\includegraphics[height=0.71\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}
+\caption{Flow cytometry data for \FIG{validation}A.
+A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected (three replicates) or infected with stocks of the indicated viral mutant in quadruplicate.
+After 13 hours, cells were stained for expression of HA protein and analyzed by FACS for HA and the ZsGreen driven by the \textit{IFNL1} reporter.
+The contour plots show the density of all cells, and IFN+ cells are also indicated by orange dots.
+Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
+For all influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
+For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
+For each viral mutant, two independently generated stocks were each assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other viral stock).
+The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
+The HA staining indicates that the number of transcriptionally active viral units per cell was similar across most mutants, and calculating percentage IFN+ only among HA+ cells largely corrects for modest variation in titer across mutants.
+}
+\label{suppfig:SNPflow}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.7\textwidth]{figures/Validation_Figure/del_flow_plot.pdf}
+\caption{
+Flow cytometry data for \FIG{validation}B.
+Each infection was performed in triplicate.
+A complicating factor for these data is that the virus with the deletion in PB1 (PB1del385to2163) cannot perform secondary transcription because it does not express PB1 protein, and so cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
+Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
+Finally, the approach in \SUPPFIG{delqPCR} was used to show that equivalent TCID50s, all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
+All this normalization suggests that all infections were performed at a similar dose of particles active for primary transcription.
+Therefore, the percent IFN+ was calculated from the flow data in this figure for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for the PB1del385to2163 virus.
+}
+\label{suppfig:delflow}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.6\textwidth]{figures/Validation_Figure/qpcr_plot.pdf}
+\caption{
+Validation that the infections in \FIG{validation}B and \SUPPFIG{delflow} were performed at similar doses of virions capable of initiating primary transcription.
+In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{delflow}), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
+The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
+These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\citep{killip2014activation}.
+In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
+But in the presence of cycloheximide, all viruses produced similar amounts of viral mRNA, indicating that the dose of particles active for primary transcription is roughly equivalent across variants.
+Each measurement was performed in quadruplicate.
+}
+\label{suppfig:delqPCR}
+\end{suppfig}
+
+\clearpage
+
 \begin{suppfile}

### 2018-11-15 paper/paper.tex
merged validation flow

@@
 {\bf (A)}
-Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line (see \SUPPFIG{SNPflow} for details).
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
 The numbers indicate the median of four measurements for each viral mutant.
@@
 Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
-See \SUPPFIG{delflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
+See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
 The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
@@
 At 24 hours post-transfection, we changed media from D10 to influenza growth media. 
-At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{SNPflow}) or 72 hours (for the replicate 2 viruses in \SUPPFIG{SNPflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
+At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{validationflow}A) or 72 hours (for the replicate 2 viruses in \SUPPFIG{validationflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \SUPPFIG{SNPflow}, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
+For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
 Two-fold serial dilutions of virus were added to an equal volume of diluted red blood cells, and titer was measured as the highest dilution of viral stock at which complete hemagglutination of red blood cells was observed.
-We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{SNPflow}.
+We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{validationflow}A.
 
-To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{delflow}, we used slightly different procedures.
+To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validationflow}B, we used slightly different procedures.
 The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
@@
 The wild-type and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
-The infections in \SUPPFIG{delflow} were performed at equivalent TCID50s as described in the legend to that figure.
+The infections in \SUPPFIG{validationflow}B were performed at equivalent TCID50s as described in the legend to that figure.
 That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \SUPPFIG{delqPCR}.
@@
 \begin{suppfig}
-\includegraphics[height=0.71\textheight]{figures/Validation_Figure/SNP_flow_plot.pdf}
-\caption{Flow cytometry data for \FIG{validation}A.
-A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected (three replicates) or infected with stocks of the indicated viral mutant in quadruplicate.
-After 13 hours, cells were stained for expression of HA protein and analyzed by FACS for HA and the ZsGreen driven by the \textit{IFNL1} reporter.
-The contour plots show the density of all cells, and IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
-For all influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
+{\bf \Large A}\includegraphics[width=0.5\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
+\hspace{0.02\textwidth}
+{\bf \Large B} \includegraphics[width=0.4\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
+\caption{Flow cytometry data for \FIG{validation}.
+{\bf (A)}
+Data for \FIG{validation}A.
+A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with stocks of the indicated viral mutant.
+After 13 hours, cells were stained for HA protein and analyzed by FACS.
+Contour plots show density of all cells, and IFN+ cells are also indicated by orange dots.
+Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of uninfected cells in these populations.
+For influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
 For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
-For each viral mutant, two independently generated stocks were each assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other viral stock).
+For each viral mutant, two independent stocks were assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other).
 The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
-The HA staining indicates that the number of transcriptionally active viral units per cell was similar across most mutants, and calculating percentage IFN+ only among HA+ cells largely corrects for modest variation in titer across mutants.
+The HA staining indicates that the number of transcriptionally active viral units was similar across most mutants, and calculating percentage IFN+ only among HA+ cells corrects for modest variation in titer across mutants.
+{\bf (B)}
+Data for \FIG{validation}B.
+The virus with the deletion in PB1 cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
+Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
+\SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
+The percent IFN+ was calculated for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for the PB1del385to2163 virus.
 }
-\label{suppfig:SNPflow}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.7\textwidth]{figures/Validation_Figure/del_flow_plot.pdf}
-\caption{
-Flow cytometry data for \FIG{validation}B.
-Each infection was performed in triplicate.
-A complicating factor for these data is that the virus with the deletion in PB1 (PB1del385to2163) cannot perform secondary transcription because it does not express PB1 protein, and so cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
-Therefore, for the data in this figure all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
-Finally, the approach in \SUPPFIG{delqPCR} was used to show that equivalent TCID50s, all viral variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
-All this normalization suggests that all infections were performed at a similar dose of particles active for primary transcription.
-Therefore, the percent IFN+ was calculated from the flow data in this figure for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for the PB1del385to2163 virus.
-}
-\label{suppfig:delflow}
+\label{suppfig:validationflow}
 \end{suppfig}
@@
 \caption{
-Validation that the infections in \FIG{validation}B and \SUPPFIG{delflow} were performed at similar doses of virions capable of initiating primary transcription.
-In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{delflow}), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
+Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.
+In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{validationflow}B), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.

### 2018-11-15 paper/paper.tex
merged reporter validation supp figs

@@
 We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
-The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reporter_validation}).
+The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
 {\bf (B)}
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \SUPPFIG{flow} for details).
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \SUPPFIG{reportervalidation}C for details).
 The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
@@
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{type_I_vs_III}; further validated by the single-cell transcriptomics below).
+The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
 Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
@@
 Importantly, the cells that we retained for further use were not the same portion that were tested by poly(I:C) treatment, but rather a separate split of the same population---this avoids any selection on the cells from transient activation of IFN.
-For the dual type I / type III reporter used in \SUPPFIG{type_I_vs_III}, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
+For the dual type I / type III reporter used in \SUPPFIG{reportervalidation}B, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
 All reporter lines tested negative for mycoplasma contamination by the Fred Hutch Genomics Core.
 
-\SUPPFIG{reporter_validation} shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
+\SUPPFIG{reportervalidation}A shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
 For detection of the cell-surface bound LNGFR$\Delta$C, cells were stained with PE-conjugated anti-LNGFR (CD271) antibody from Miltenyi Biotec.
@@
 \subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
-A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}~\citep{russell2018github}. 
+A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}. 
 This pipeline is orchestrated by \texttt{Snakemake}~\citep{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
@@
 \begin{suppfig}
-\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}
-\caption{To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
+\begin{minipage}[t]{0.52\textwidth}
+{\bf \Large A}
+
+\includegraphics[width=0.9\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}
+\end{minipage}
+\begin{minipage}[t]{0.48\textwidth}
+{\bf \Large B}
+
+\includegraphics[width=0.7\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
+
+{\bf \Large C}
+
+\includegraphics[width=0.9\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
+
+\end{minipage}
+
+\caption{Details about the reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.
+{\bf (A)}
+To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
 The name of each of reporter cell line is indicated at the top of each row of plots.
@@
 Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
-}
-\label{suppfig:reporter_validation}
-\end{suppfig}
-
-\clearpage
-
-\begin{suppfig}
-\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
-\caption{
-Expression of type I and type III IFN is highly correlated in A549 cells.
+{\bf (B)}
+The type I and type III IFN reporters are highly correlated in their activation.
 An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
@@
 Cells positive for either fluorescent reporter were further analyzed.
-As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.}
-\label{suppfig:type_I_vs_III}
-\end{suppfig}
-
-\clearpage
-
-\begin{suppfig}
-\includegraphics[width=0.75\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
-\caption{Flow cytometry data for \FIG{IFNrare}B.
+As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.
+{\bf (C)}
+Flow cytometry data for \FIG{IFNrare}B.
 The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
@@
 }
-\label{suppfig:flow}
+\label{suppfig:reportervalidation}
 \end{suppfig}

### 2018-11-16 paper/paper.tex
consolidated more supplementary figures

@@
 We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
-Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange (\SUPPFIG{StrandExchange}) was rare (\SUPPFIG{CCSs}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}A).
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{CCSs}B), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
 
@@
 We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
-In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{CCSs}B).
 The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
@@
 The sequences of these primers are in \SUPPFILE{primer_sequences}.
-A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
+A major concern during these PCRs is strand exchange (see \SUPPFIG{CCSs}B) which would scramble the cell barcodes and mutations on viral cDNAs.
 To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
@@
 The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
-The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{StrandExchange} for an illustration of how this is determined).
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}A, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{CCSs}B for an illustration of how this is determined).
 A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
@@
 We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
-At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{StrandExchange}), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}).
+At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{CCSs}B), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}A).
 The small number of CCSs with identifiable strand exchange were filtered from further analysis.
@@
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
-%This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
-%JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
-%JDB is an Investigator of the Howard Hughes Medical Institute.
-%The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
+This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
+ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
+JDB is an Investigator of the Howard Hughes Medical Institute.
+The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
 
-\section{Competing Interests}
+\section{Author Contributions}
+ABR and JDB designed the project.
+ABR performed the experiments with assistance from JRK.
+ABR and JDB computationally analyzed the data and wrote the paper.
+
+\section{Declaration of Interests}
 The authors declare that no competing interests exist.
@@
 \begin{suppfig}
-\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
+{\bf \Large A}
+\includegraphics[width=0.85\textwidth, valign=t]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
+
+{\bf \Large B}
+\includegraphics[width=0.5\textwidth, valign=t]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
 \caption{
+Details about PacBio sequencing.
+{\bf (A)}
 The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
-Note that these sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
-(For exhaustive details, see the Jupyter notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and in \SUPPFILE{pacbio_analysis}.)
+These sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells (see \SUPPFILE{pacbio_analysis}).
 Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
@@
 Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
-Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences derived from the same cDNA, and cell barcodes were used to collapse duplicate sequences from the same cell.
-The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see \SUPPFIG{StrandExchange} for more explanation).
+Note that the disproportionate sequencing of the shorter internally deleted genes does not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse sequences derived from the same cDNA, and cell barcodes were used to collapse sequences from the same cell.
+The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see panel B).
 From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
-Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
-Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
+About half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
+For some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
 A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
-}
-\label{suppfig:CCSs}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.6\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
-\caption{
+{\bf (B)}
 Strategy for detecting strand exchange during sequencing of full-length viral genes.
@@
 A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
-We can detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
+We detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
 If there is no strand exchange, all molecules should either be wild-type or have the synonymous barcoding mutations at \emph{both} termini.
-But strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
-\SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
+Strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
+\SUPPFIG{CCSs}A shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
 Note that since the rate of homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
 }
-\label{suppfig:StrandExchange}
+\label{suppfig:CCSs}
 \end{suppfig}

### 2018-11-16 paper/paper.tex
a few tiny tweaks

@@
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
-If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease severity~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.
 
@@
 \subsection{A system to identify and enrich for rare IFN+ cells}
-A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-We previously performed single-cell transcriptomics on influenza-infected A549 human lung epithelial cells, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN~\citep{russell2018extreme}.
-Prior studies using reporters have also found that IFN expression is rare in influenza-infected cells~\citep{killip2017single} and mice~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).
 
 %%% start IFNrare figure %%%
-\begin{figure}
+\begin{figure}[b!]
 \centerline{
@@
 Reporter cells to identify and enrich IFN+ cells.
-{\bf (A)} An A549 IFN reporter cell line.
-The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
+{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
 We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
-The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
+In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
 {\bf (B)}
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter cell line (see \SUPPFIG{reportervalidation}C for details).
-The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (\SUPPFIG{reportervalidation}C).
+The plot also shows uninfected cells, and cells infected with Sendai virus.
 The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
@@
 %%% end IFNrare figure %%%
+
+A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
+Studies using reporters~\citep{killip2017single} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
+IFN expression also appears to be rare among cells in influenza-infected mice~\citep{kallfass2013visualizing}, a conclusion supported by our re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).
 
@@
 %%% start workflow figure
-\begin{figure}
+\begin{figure}[b!]
 \begin{fullwidth}
@@
 
-To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
+To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
 First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
@@
 This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
-This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.
+This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.
 

### 2018-11-21 paper/paper.tex
move text to PLoS template

@@
-\documentclass[9pt,lineno]{template}
+% Template for PLoS
+% Version 3.5 March 2018
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
 
-\usepackage{lipsum} % Required to insert dummy text
-\usepackage[version=4]{mhchem}
-\usepackage{siunitx}
-\usepackage[export]{adjustbox}
-\usepackage{float}
-\DeclareSIUnit\Molar{M}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
-\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
-\newcommand{\abrcomment}[1]{\emph{\color{blue} [#1]}}
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
 
-\renewcommand{\floatpagefraction}{.9}
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
 
-\newfloat{suppfile}{thp}{lofsupfile}
-\floatname{suppfile}{Supplementary File}
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
 
-\newfloat{suppfig}{thp}{lofsupfile}
-\floatname{suppfig}{Supplementary Figure}
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE SETUP
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Single-cell virus sequencing of influenza infections that trigger innate immunity}
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
 
-\author[1]{Alistair B. Russell}
-\author[1]{Jacob R. Kowalsky}
-\author[1,2,3*]{Jesse D. Bloom}
-\affil[1]{Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States}
-\affil[2]{Department of Genome Sciences, University of Washington, Seattle, United States}
-\affil[3]{Howard Hughes Medical Institute, Fred Hutchinson Cancer Research Center, Seattle, United States}
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
 
-\corr{jbloom@fredhutch.org}{JDB}
+% line numbers
+\usepackage[right]{lineno}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%% ARTICLE START
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
 
 \begin{document}
+\vspace*{0.2in}
 
-\maketitle
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Alistair B. Russell\textsuperscript{1},
+Jacob R. Kowalsky\textsuperscript{2},
+Jesse D. Bloom\textsuperscript{1,2,3*}
+\\
+\bigskip
+\textbf{1} Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States
+\\
+\textbf{2} Department of Genome Sciences, University of Washington, Seattle, United States
+\\
+\textbf{3} Howard Hughes Medical Institute, Seattle, United States
+\\
+\bigskip
 
-\begin{abstract}
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* jbloom@fredhutch.org
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
@@
 Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
-\end{abstract}
 
-\section{Introduction}
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+\section*{Author summary}
+{\color{red} add this}
+
+\linenumbers
+
+% Use "Eq" instead of "Equation" for equation citations.
+\section*{Introduction}
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
 As the virus spreads, some cells detect infection and begin producing interferon (IFN).
-This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
-If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.
+This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.
 
-Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
-This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
-Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
+Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
-For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
+For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
 But for influenza, a third possible contributor also looms large: viral genetic diversity.
-Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
-Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
-Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
+Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
@@
 A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-Studies using reporters~\citep{killip2017single} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
-IFN expression also appears to be rare among cells in influenza-infected mice~\citep{kallfass2013visualizing}, a conclusion supported by our re-analysis of \citet{steuerman2018dissection}'s single-cell transcriptomics of influenza-infected mice (\SUPPFIG{mice}).
+Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
+IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).
 
-To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
+To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\cite[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
 Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
@@
 First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
-These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
@@
 
-We processed the cells on a commercially available platform~\citep{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
-Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
-Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.
+We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
+Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
+Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.
 
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
+We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\cite[CCSs;][]{travers2010flexible}.
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
@@
 Number of cells for which transcriptomes were obtained.
-From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
+From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 {\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
-From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
+From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
 {\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
@@
 We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
-We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
+We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
 To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.
@@
 The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
-As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
+As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
 
@@
 \FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
-This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
+This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
 
@@
 However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
-Consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\citep{hatada1989control}.
+Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
 
 By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
+From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
 Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
@@
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\cite{stetson2006type,honda2006type}.
 
 Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
-\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
+\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\cite{maaten2008visualizing} of the cells.
 Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.
@@
 It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
-Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.
+Therefore, analogous to somatic variant calling in tumor sequencing~\cite{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.
 
@@
 To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
+\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
@@
 
-\figdata{Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
+\figdata{Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
 \label{figdata:sequences}
@@
 The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
-Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
-The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
+Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\cite{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
+The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
 We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
@@
 
-Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
@@
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observation by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
 To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
@@
 Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wild-type'' influenza genome, most infected cells do not express wild-type copies of all viral genes.
-Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
+Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
 Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
 
-It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
+It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
 We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
-Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
-The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by \citet{velthuis2018mini} showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I.
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
+The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini} .
+Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
-There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
+There may be several reasons: we used relatively pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
 The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
@@
 Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
-However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\citep{oneal2018west}.
+However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
 
 Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
-Natural human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
+Natural human influenza infections are established by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
 It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
 Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
-
-\clearpage
 
@@
 
-To create the type I interferon reporters, a 1kb promoter region upstream of the human IFNB1 gene were cloned into the pHAGE2 lentiviral vector~\citep{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
-Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\citep{kim2011high}.
+To create the type I interferon reporters, a 1kb promoter region upstream of the human IFNB1 gene were cloned into the pHAGE2 lentiviral vector~\cite{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
+Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\cite{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\cite{kim2011high}.
 The sequence of the last of these constructs is provided in \SUPPFILE{reporter_sequences}.
@@
 We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
-We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
+We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\cite{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
 In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{CCSs}B).
 The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
-These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
+These genes were cloned into the pHW2000~\cite{hoffmann2000dna} reverse-genetics plasmid.
 
-Both viral strains were generated by reverse genetics using the pHW18* series of bi-directional plasmids~\citep{hoffmann2000dna}.
-We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\citep{xue2016propagation}.
-The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 cells in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citet{russell2018extreme}.
+Both viral strains were generated by reverse genetics using the pHW18* series of bi-directional plasmids~\cite{hoffmann2000dna}.
+We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\cite{xue2016propagation}.
+The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 cells in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \cite{russell2018extreme}.
 Specifically, after generation by reverse genetics, the wild-type variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
-The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of \citet{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
+The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\cite{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
 
@@
 We estimated titers of transcriptionally active virions by staining for HA expression in virus-infected A549 cells.
-Specifically, we infected A549 cells (or one of the A549 reporter cell line variants as indicated) in influenza growth medium, and at 13 to 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\citep{doud2017complete}.
+Specifically, we infected A549 cells (or one of the A549 reporter cell line variants as indicated) in influenza growth medium, and at 13 to 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\cite{doud2017complete}.
 After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC, washed, fixed in 1\% formaldehyde in PBS, washed again, and then analyzed by flow cytometry to determine the fraction expressing detectable HA protein.
@@
 We began this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
-The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.
+The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\cite{zheng2017massively}, targeting capture of $\sim$1,500 cells.
 
@@
 Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
-This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\citep{Boers:2015emulsion}.
+This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{Boers:2015emulsion}.
 We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .
@@
 The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\citep{xue2016propagation} and are presumably preferentially amplified during PCR.
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\cite{xue2016propagation} and are presumably preferentially amplified during PCR.
 
@@
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
-For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml~\citep[a concentration sufficient to block secondary transcription;][]{killip2014activation} at the time of infection.
+For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml~\cite[a concentration sufficient to block secondary transcription;][]{killip2014activation} at the time of infection.
 After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.
@@
 This NS1stop virus contained six nucleotide changes resulting in the addition of five in-frame stop codons in NS1 starting 10 nucleotides downstream of the 5' splice donor site, thereby disrupting NS1 while leaving NS2 (NEP) intact.
-All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\citep{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
+All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\cite{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
  \FIGDATA[validation]{sequences} provides the full sequences for all of these plasmids.
 
-We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wild-type plasmid for that gene.
+We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\cite{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wild-type plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
@@
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
+For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\cite{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
@@
 To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validationflow}B, we used slightly different procedures.
-The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
+The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\cite{russell2018extreme}.
 The NS1stop virus was similarly generated, but was passaged for 48 rather than 36 hours, since it had slower growth kinetics and so needed a longer period of time to reach high titers.
 The viruses with deletions in the PB1 segment could not be generated in normal 293T and MDCK-SIAT1 cells, since they required the exogenous expression of the PB1 protein.
-Therefore, these viruses were generated in previously described 293T and MDCK-SIAT1 cells that had been engineered to constitutively express PB1~\citep{bloom2010permissive}.
+Therefore, these viruses were generated in previously described 293T and MDCK-SIAT1 cells that had been engineered to constitutively express PB1~\cite{bloom2010permissive}.
 These viruses were harvested from transfections at 72 hours, and passaged twice in the MDCK-SIAT1 cells constitutively expressing PB1 at a MOI of 0.001 for 72 hours and 0.01 for 48 hours. 
@@
 A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}. 
-This pipeline is orchestrated by \texttt{Snakemake}~\citep{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
+This pipeline is orchestrated by \texttt{Snakemake}~\cite{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
 The sequencing data and annotated cell-gene matrix are available on the GEO repository under accession GSE120839 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120839}).
@@
 First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
-We further processed these CCSs using custom Python code and the \texttt{minimap2}~\citep{li2018minimap2} long-read aligner (version 2.11-r797).
-The Python code has been implemented in the API of \texttt{dms\_tools2}~\citep[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
+We further processed these CCSs using custom Python code and the \texttt{minimap2}~\cite{li2018minimap2} long-read aligner (version 2.11-r797).
+The Python code has been implemented in the API of \texttt{dms\_tools2}~\cite[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
 A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{pacbio_analysis}.
@@
 Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
-This analysis utilized a variety of R and Bioconductor~\citep{huber2015orchestrating} packages, including \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
+This analysis utilized a variety of R and Bioconductor~\cite{huber2015orchestrating} packages, including \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
 A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{monocle_analysis}.
@@
 The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).
+
+% Place figure captions after the first paragraph in which they are cited.
+\begin{figure}[!h]
+\caption{{\bf Bold the figure title.}
+Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+\label{fig1}
+\end{figure}
 
@@
 
-\nolinenumbers
-
 \bibliography{references}
 
-\clearpage
+\section*{Supporting information}
 
@@
 \caption{
-A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
-Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
+A re-analysis of the data from a published single-cell transcriptomic study~\cite{steuerman2018dissection} of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
+Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
 Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
-As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
+As shown in the left-most panel, the sequencing depth of \cite{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
 An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
+Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\cite{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\cite{stetson2006type,honda2006type}.
 The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
@@
 {\bf (A)}
-To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
+To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\cite{strahle2006sendai}.
 The name of each of reporter cell line is indicated at the top of each row of plots.
@@
 Flow cytometry data for \FIG{IFNrare}B.
-The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
+The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\cite{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
 After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
@@
 The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
-Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
 The exception is NP, which is detected in virtually all infected cells.
@@
 Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
-This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
+This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\cite{stetson2006type,honda2006type}.
 {\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
@@
 Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.
-To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
+To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\cite{maaten2008visualizing} as implemented in \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
 In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
@@
 The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see panel B).
-From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+From the frequencies of these different forms, we estimate~\cite{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
 About half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
@@
 \SUPPFIG{CCSs}A shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
-Note that since the rate of homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
+Note that since the rate of homologous recombination in influenza virus in negligible~\cite{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
 }
@@
 The virus with the deletion in PB1 cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
-Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for the PB1del385to2163 virus.
+Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\cite{bloom2010permissive} for the PB1del385to2163 virus.
 \SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
@@
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\citep{killip2014activation}.
+These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\cite{killip2014activation}.
 In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
@@
 
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Fig.}
+\label{S1_Fig}
+{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+
 \end{document}

### 2018-11-21 paper/paper.tex
finished LaTex re-formatting

@@
 \noalign{\global\arrayrulewidth\savedwidth}}
+
+% additional packages imported by Jesse Bloom
+\usepackage[export]{adjustbox}
+\usepackage{float}
+
+% added by Jesse Bloom for full-width figures
+\newenvironment{fullwidth}{%
+  \begin{adjustwidth}{-2.3in}{}
+}{\end{adjustwidth}}
+
+% Added by Jesse Bloom for figures
+\renewcommand{\floatpagefraction}{.9}
+
+\newfloat{suppfile}{thp}{lofsupfile}
+\floatname{suppfile}{File}
+\renewcommand{\thesuppfile}{S\arabic{suppfile}}
+
+\newfloat{suppfig}{thp}{lofsupfile}
+\floatname{suppfig}{Fig}
+\renewcommand{\thesuppfig}{S\arabic{suppfig}}
+
+\renewcommand{\figureautorefname}{Fig}
+\newcommand{\FIG}[1]{\autoref{fig:#1}}
+
+\newcommand{\suppfileautorefname}{File}
+\newcommand{\SUPPFILE}[1]{\autoref{suppfile:#1}}
+
+\newcommand{\suppfigautorefname}{Fig}
+\newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}
 
@@
 
-\section{Results}
-
-\subsection{A system to identify and enrich for rare IFN+ cells}
+\section*{Results}
+
+\subsection*{A system to identify and enrich for rare IFN+ cells}
 
@@
 \caption{
-Reporter cells to identify and enrich IFN+ cells.
+{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
 {\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
@@
 
-To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\cite[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
+To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
@@
 
-\subsection{Combined transcriptomics and virus-sequencing of single infected cells}
+\subsection*{Combined transcriptomics and virus-sequencing of single infected cells}
 
@@
 \caption{
-Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
+{\bf Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.}
 {\bf (A)}
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\cite[CCSs;][]{travers2010flexible}.
+We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
@@
 
-\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}
+\subsection*{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}
 
@@
 \caption{
-Single-cell transcriptomics of IFN-enriched influenza-infected cells.
+{\bf Single-cell transcriptomics of IFN-enriched influenza-infected cells.}
 {\bf (A)} 
@@
 
-\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells}
+\subsection*{Full genotypes of viruses infecting single IFN+ and IFN- cells}
 
@@
 \caption{
-Viral genotypes and infection outcomes in single cells.
+{\bf Viral genotypes and infection outcomes in single cells.}
 Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
@@
 
-\subsection{Viral defects associated with infection outcome in single cells}
+\subsection*{Viral defects associated with infection outcome in single cells}
 
@@
 \caption{
-Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.
+{\bf Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.}
 {\bf (A)} 
@@
 
-\subsection{Validation that viral defects in single IFN+ cells often cause increased IFN induction}
+\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}
 
@@
 \centerline{\includegraphics[width=0.85\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
-\caption{Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.
+\caption{
+{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
 {\bf (A)}
@@
 
-\figdata{Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
-\label{figdata:sequences}
-
 \end{figure}
@@
 
-\section{Discussion}
+\section*{Discussion}
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
@@
 
-\section{Methods and Materials}
-
-\subsection{IFN reporter cell lines}
+\section*{Methods and Materials}
+
+\subsection*{IFN reporter cell lines}
 We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}A).
@@
 
-\subsection{Viruses for single-cell experiments}
+\subsection*{Viruses for single-cell experiments}
 We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
@@
 
-\subsection{Flow cytometry analyses for HA expression}
+\subsection*{Flow cytometry analyses for HA expression}
 For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
@@
 
-\subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
+\subsection*{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
 The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-P2A-mNeonGreen reporter.
@@
 
-\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
+\subsection*{Enrichment and preparation of viral cDNA for PacBio sequencing}
 We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
@@
 
-\subsection{qPCR for viral genes and IFN}
+\subsection*{qPCR for viral genes and IFN}
 We performed qPCR on reverse-transcribed mRNA for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
@@
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
-For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml~\cite[a concentration sufficient to block secondary transcription;][]{killip2014activation} at the time of infection.
+For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml (a concentration sufficient to block secondary transcription~\cite{killip2014activation}) at the time of infection.
 After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.
 
-\subsection{Viruses and experiments for validation experiments}
+\subsection*{Viruses and experiments for validation experiments}
 In \FIG{validation}, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
@@
 All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\cite{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
- \FIGDATA[validation]{sequences} provides the full sequences for all of these plasmids.
-
-We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\cite{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \FIGDATA[validation]{sequences} for the wild-type plasmid for that gene.
+ \SUPPFILE{validation_sequences} provides the full sequences for all of these plasmids.
+
+We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\cite{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{validation_sequences} for the wild-type plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
@@
 
-\subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
+\subsection*{Computational analysis of single-cell transcriptomic and viral sequence data}
 A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}. 
@@
 We further processed these CCSs using custom Python code and the \texttt{minimap2}~\cite{li2018minimap2} long-read aligner (version 2.11-r797).
-The Python code has been implemented in the API of \texttt{dms\_tools2}~\cite[][\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software} package (version 2.3.0).
+The Python code has been implemented in the API of \texttt{dms\_tools2} (\url{https://jbloomlab.github.io/dms_tools2/}~\cite{bloom2015software}) package (version 2.3.0).
 A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{pacbio_analysis}.
@@
 
-% Place figure captions after the first paragraph in which they are cited.
-\begin{figure}[!h]
-\caption{{\bf Bold the figure title.}
-Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-\label{fig1}
-\end{figure}
-
-\section{Acknowledgments}
+\section*{Acknowledgments}
 We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
@@
 
-\section{Author Contributions}
+\section*{Author Contributions}
 ABR and JDB designed the project.
@@
 
-\section{Declaration of Interests}
+\section*{Declaration of Interests}
 The authors declare that no competing interests exist.
@@
 \caption{
-A re-analysis of the data from a published single-cell transcriptomic study~\cite{steuerman2018dissection} of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
+{\bf Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
 Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
@@
 
-\caption{Details about the reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.
+\caption{
+{\bf Details about the reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.}
 {\bf (A)}
@@
 \caption{
-Example MACS enrichments of IFN+ influenza-infected cells.
+{\bf Example MACS enrichments of IFN+ influenza-infected cells.}
 A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
@@
 \includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}
-\caption{The fraction of infected cells that are called as expressing each viral gene.
+\caption{
+{\bf The fraction of infected cells that are called as expressing each viral gene.}
 The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
@@
 \caption{
-The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
+{\bf The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.}
 Each point represents one cell.
@@
 \caption{
-Expression of ISGs in single infected and uninfected cells.
+{\bf Expression of ISGs in single infected and uninfected cells.}
 For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
@@
 \caption{
-Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.
+{\bf Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.}
 To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\cite{maaten2008visualizing} as implemented in \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
@@
 \caption{
-Details about PacBio sequencing.
+{\bf Details about PacBio sequencing.}
 {\bf (A)}
@@
 \includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}
-\caption{Number of cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).
+\caption{
+{\bf Number of cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).}
 {\bf (A)} We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
@@
 {\bf \Large B} \includegraphics[width=0.4\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
-\caption{Flow cytometry data for \FIG{validation}.
+\caption{
+{\bf Flow cytometry data for \FIG{validation}.}
 {\bf (A)}
@@
 \caption{
-Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.
+{\bf Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.}
 In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{validationflow}B), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
@@
 \begin{suppfile}
-\caption{Sequences of the IFN reporters in \FIG{IFNrare}A are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
+\caption{{\bf Sequences of the IFN reporters in \FIG{IFNrare}A.}
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
 }
@@
 \caption{
-Genbank files giving sequences of the wild-type and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
+{\bf Genbank files giving sequences of the wild-type and synonymously barcoded viruses}.
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
 \label{suppfile:virus_seqs}
@@
 \begin{suppfile}
-\caption{A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
+\caption{
+{\bf A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing.}
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt} 
 }
@@
 \begin{suppfile}
-\caption{A CSV file giving the genotypes in \FIG{genotypes} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
+\caption{
+{\bf A CSV file giving the genotypes in \FIG{genotypes}.}
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
 \label{suppfile:genotypes}
@@
 \begin{suppfile}
-\caption{A CSV file giving the viral mutations and related information in \FIG{mutations} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.}
+\caption{
+{\bf A CSV file giving the viral mutations and related information in \FIG{mutations}.}
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.}
 \label{suppfile:mutations}
@@
 \begin{suppfile}
+\caption{
+{\bf Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna}.}
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
+\label{suppfile:validation_sequences}
+\end{suppfile}
+
+\begin{suppfile}
 \caption{\label{suppfile:pacbio_analysis}
-An HTML rendering of the Jupyter notebook that analyzes the PacBio data to call the viral sequences in infected cells is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
+{\bf An HTML rendering of the Jupyter notebook that analyzes the PacBio data to call the viral sequences in infected cells.}
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
 This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
@@
 \caption{\label{suppfile:monocle_analysis}
-An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
+{\bf An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper.}
+See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
 This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
@@
 
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
-
 \end{document}

### 2018-11-21 paper/paper.tex
fixed figure formatting

@@
 \renewcommand{\floatpagefraction}{.9}
+\setcounter{topnumber}{8}
+\setcounter{bottomnumber}{8}
+\setcounter{totalnumber}{8}
 
@@
 {\bf (B)}
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (\SUPPFIG{reportervalidation}C).
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
 The plot also shows uninfected cells, and cells infected with Sendai virus.
@@
 {\centering
-\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.92\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
-Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}A).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{CCSs}B), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
 
@@
 We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\cite{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
-In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{CCSs}B).
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
 The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
@@
 The sequences of these primers are in \SUPPFILE{primer_sequences}.
-A major concern during these PCRs is strand exchange (see \SUPPFIG{CCSs}B) which would scramble the cell barcodes and mutations on viral cDNAs.
+A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
 To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
@@
 The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
-The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}A, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{CCSs}B for an illustration of how this is determined).
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{StrandExchange} for an illustration of how this is determined).
 A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
@@
 We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
-At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{CCSs}B), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}A).
+At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{StrandExchange}), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}).
 The small number of CCSs with identifiable strand exchange were filtered from further analysis.
@@
 
+\clearpage
 \section*{Supporting information}
@@
 
-\clearpage
-
 \begin{suppfig}
@@
 
-{\bf \Large C}
-
-\includegraphics[width=0.9\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
-
 \end{minipage}
@@
 \caption{
-{\bf Details about the reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.}
+{\bf Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.}
 {\bf (A)}
@@
 As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.
-{\bf (C)}
-Flow cytometry data for \FIG{IFNrare}B.
+}
+\label{suppfig:reportervalidation}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.9\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
+\caption{
+{\bf Flow cytometry data for \FIG{IFNrare}B.}
 The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\cite{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
@@
 }
-\label{suppfig:reportervalidation}
-\end{suppfig}
-
-\clearpage
+\label{suppfig:IFNrare_flow}
+\end{suppfig}
 
@@
 
-\clearpage
-
-\begin{suppfig}
-{\bf \Large A}
-\includegraphics[width=0.85\textwidth, valign=t]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
-
-{\bf \Large B}
-\includegraphics[width=0.5\textwidth, valign=t]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
-\caption{
-{\bf Details about PacBio sequencing.}
-{\bf (A)}
-The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
+\caption{
+{\bf The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.}
 These sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells (see \SUPPFILE{pacbio_analysis}).
@@
 A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
-{\bf (B)}
-Strategy for detecting strand exchange during sequencing of full-length viral genes.
+}
+\label{suppfig:CCSs}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.8\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
+\caption{
+{\bf Strategy for detecting strand exchange during sequencing of full-length viral genes.}
 The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
@@
 Strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
-\SUPPFIG{CCSs}A shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
+\SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
 Note that since the rate of homologous recombination in influenza virus in negligible~\cite{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
 }
-\label{suppfig:CCSs}
+\label{suppfig:StrandExchange}
 \end{suppfig}
@@
 
-\clearpage
-
-\begin{suppfig}
-{\bf \Large A}\includegraphics[width=0.5\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
+\begin{suppfig}
+{\bf \Large A}\includegraphics[width=0.48\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
 \hspace{0.02\textwidth}
-{\bf \Large B} \includegraphics[width=0.4\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
+{\bf \Large B} \includegraphics[width=0.39\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
 \caption{
@@
 Data for \FIG{validation}A.
-A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with stocks of the indicated viral mutant.
+A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with stocks of the indicated mutant.
 After 13 hours, cells were stained for HA protein and analyzed by FACS.
@@
 Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of uninfected cells in these populations.
-For influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
+For infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
 For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
@@
 The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
-The HA staining indicates that the number of transcriptionally active viral units was similar across most mutants, and calculating percentage IFN+ only among HA+ cells corrects for modest variation in titer across mutants.
 {\bf (B)}
 Data for \FIG{validation}B.
-The virus with the deletion in PB1 cannot be effectively normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
+The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
 Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\cite{bloom2010permissive} for the PB1del385to2163 virus.
@@
 \caption{\label{suppfile:pacbio_analysis}
-{\bf An HTML rendering of the Jupyter notebook that analyzes the PacBio data to call the viral sequences in infected cells.}
+{\bf HTML rendering of Jupyter notebook that analyzes the PacBio data.}
 See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
-This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
-The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.}
+The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.}
 \end{suppfile}
@@
 \caption{\label{suppfile:monocle_analysis}
-{\bf An HTML rendering of the Jupyter notebook that analyzes the annotated cell-gene matrix to generate the figures in this paper.}
+{\bf HTML rendering of Jupyter notebook that analyzes the annotated cell-gene matrix.}
 See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
-This notebook contains detailed descriptions of each step and plots illustrating the results, and is the best way to understand this part of the analysis in detail.
-The actual Jupyter notebook rendered here is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.}
+The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.}
 \end{suppfile}

### 2018-11-21 paper/paper.tex
added author summary

@@
 \section*{Author summary}
-{\color{red} add this}
+Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
+But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not whether it has mutations.
+Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.
+Most cells are infected by virions that have a mutation or some other defect.
+This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and helps determine whether a cell activates an innate-immune response.
+However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
+Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.
 
@@
 %%% start IFNrare figure %%%
-\begin{figure}[b!]
+\begin{figure}
 \centerline{
@@
 %%% start workflow figure
-\begin{figure}[b!]
+\begin{figure}
 \begin{fullwidth}
@@
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
-This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
-JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
-JDB is an Investigator of the Howard Hughes Medical Institute.
-The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
-
-\section*{Author Contributions}
-ABR and JDB designed the project.
-ABR performed the experiments with assistance from JRK.
-ABR and JDB computationally analyzed the data and wrote the paper.
-
-\section*{Declaration of Interests}
-The authors declare that no competing interests exist.
+%This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
+%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+%JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
+%JDB is an Investigator of the Howard Hughes Medical Institute.
+%The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
 
@@
 Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of uninfected cells in these populations.
-For infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
-For the uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
-For each viral mutant, two independent stocks were assayed in duplicate (i.e., \#1a and \#1b are replicates of one viral stock, and \#2a and \#2b are replicates of the other).
+For infected cells, the percentage IFN+ was calculated among the HA+ cells (since these are the ones that are infected).
+For uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
+For each viral mutant, two independent stocks were assayed in duplicate (i.e., \#1a and \#1b are one viral stock, and \#2a and \#2b are the other).
 The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
@@
 Data for \FIG{validation}B.
-The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses lower levels of HA due to the lack of secondary transcription.
-Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for the wild-type and NS1stop viruses, and on MDCK-SIAT1 cells expressing PB1~\cite{bloom2010permissive} for the PB1del385to2163 virus.
+The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses less HA due to the lack of secondary transcription.
+Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\cite{bloom2010permissive} for PB1del385to2163.
 \SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
-The percent IFN+ was calculated for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for the PB1del385to2163 virus.
+The percent IFN+ was calculated for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for PB1del385to2163.
 }

### 2018-11-21 paper/paper.tex
Alistair edits

@@
 Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
-But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not whether it has mutations.
+But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
 Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.

### 2018-11-27 paper/paper.tex
add suppfiles per _PLoS Pathogens_ editorial staff request

@@
 \caption{{\bf Sequences of the IFN reporters in \FIG{IFNrare}A.}
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
 }
@@
 {\bf Genbank files giving sequences of the wild-type and synonymously barcoded viruses}.
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
+}
 \label{suppfile:virus_seqs}
@@
 {\bf A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing.}
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt} 
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt} 
 }
@@
 {\bf A CSV file giving the genotypes in \FIG{genotypes}.}
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.
+}
 \label{suppfile:genotypes}
@@
 {\bf A CSV file giving the viral mutations and related information in \FIG{mutations}.}
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.
+}
 \label{suppfile:mutations}
@@
 {\bf Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna}.}
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}
+}
 \label{suppfile:validation_sequences}
@@
 {\bf HTML rendering of Jupyter notebook that analyzes the PacBio data.}
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
-The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
+The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.
+}
 \end{suppfile}
@@
 {\bf HTML rendering of Jupyter notebook that analyzes the annotated cell-gene matrix.}
-See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
-The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
+The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.
+}
 \end{suppfile}

### 2018-12-24 paper/paper.tex
re-formatted for _Immunity_

@@
 
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
 % Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
+\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
+\usepackage{nameref}
 
@@
 \newfloat{suppfig}{thp}{lofsupfile}
-\floatname{suppfig}{Fig}
+\floatname{suppfig}{Figure}
 \renewcommand{\thesuppfig}{S\arabic{suppfig}}
 
-\renewcommand{\figureautorefname}{Fig}
+\renewcommand{\figureautorefname}{Figure}
 \newcommand{\FIG}[1]{\autoref{fig:#1}}
@@
 
-\newcommand{\suppfigautorefname}{Fig}
+\newcommand{\suppfigautorefname}{Figure}
 \newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}
@@
 \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
+\renewcommand{\figurename}{Figure}
+
+% Harvard-style citations
+\usepackage{natbib}
+\bibliographystyle{agsm}
 
@@
 
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
+\providecommand{\keywords}[1]{\noindent \textbf{Keywords:} #1}
 
@@
 \begin{flushleft}
-{\Large
-\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
+{\LARGE
+\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity}}
 \newline
@@
 \bigskip
-\textbf{1} Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States
+{\small
+$^1$Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center
 \\
-\textbf{2} Department of Genome Sciences, University of Washington, Seattle, United States
+$^2$Department of Genome Sciences, University of Washington
 \\
-\textbf{3} Howard Hughes Medical Institute, Seattle, United States
+$^3$Howard Hughes Medical Institute
 \\
+Seattle, WA  98109, USA
+\\
+*Correspondence: jbloom@fredhutch.org
+}
 \bigskip
 
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* jbloom@fredhutch.org
-
 \end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
+
+\section*{Summary}
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
@@
 We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
+Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
 However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
-Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
-Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
-But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
-Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.
-Most cells are infected by virions that have a mutation or some other defect.
-This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and helps determine whether a cell activates an innate-immune response.
-However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
-Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.
+Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
+
+\vspace{0.2in}
+\keywords{PacBio, influenza, interferon, single-cell RNA-seq, NS1, stochastic}
 
@@
 As the virus spreads, some cells detect infection and begin producing interferon (IFN).
-This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
-If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.
-
-Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
-This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
-Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.
+
+Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
+Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
-For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
+For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
 But for influenza, a third possible contributor also looms large: viral genetic diversity.
-Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
-Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
-Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
+Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
@@
 A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
-IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).
-
-To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
+Studies using reporters~\citep{killip2017single} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
+IFN expression also appears to be rare among cells in influenza-infected mice~\citep{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\citep{steuerman2018dissection} (\SUPPFIG{mice}).
+
+To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\citep{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
 Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
@@
 First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
-These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
@@
 
-We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
-Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
-Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.
+We processed the cells on a commercially available platform~\citep{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
+Because influenza virus mRNAs are polyadenylated~\citep{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
+Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\citep{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\citep{robertson1981polyadenylation} in each of the eight viral gene segments.
 
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
+We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\citep{travers2010flexible}).
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
@@
 Number of cells for which transcriptomes were obtained.
-From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
+From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 {\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
-From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
+From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
 {\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
@@
 We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
-We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
+We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
 To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.
@@
 The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
-As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
+As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
 
@@
 \FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
-This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
+This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
 
@@
 However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
-Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
+Consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\citep{hatada1989control}.
 
 By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
+From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
 Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
@@
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\cite{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.
 
 Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
-\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\cite{maaten2008visualizing} of the cells.
+\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
 Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.
@@
 It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
-Therefore, analogous to somatic variant calling in tumor sequencing~\cite{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.
+Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.
 
@@
 To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
+\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
@@
 The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
-Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\cite{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
-The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
+Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
+The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
 We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
@@
 
-Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
@@
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
-To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\citep{saikia2018simultaneous, zanini2018virus}.
+To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \citep{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
@@
 Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wild-type'' influenza genome, most infected cells do not express wild-type copies of all viral genes.
-Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
+Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
 Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
 
-It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
+It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
 We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
-Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
-The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini} .
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
+The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini} .
+Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
-There may be several reasons: we used relatively pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
+There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
 The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
@@
 Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
-However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
+However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\citep{oneal2018west}.
 
 Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
-Natural human influenza infections are established by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
+Natural human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
 It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
@@
 
-\section*{Methods and Materials}
+\small
+\subsection*{Acknowledgments}
+We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
+We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
+This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
+ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
+JDB is an Investigator of the Howard Hughes Medical Institute.
+
+\subsection*{Author Contributions}
+ABR and JDB designed the study.
+ABR performed the experiments with assistance from JRK.
+ABR and JDB computationally analyzed the data and wrote the paper.
+
+\subsection*{Declaration of Interests}
+The authors declare no competing interests.
+
+\bibliography{references}
+
+\section*{Methods}
 
@@
 
-To create the type I interferon reporters, a 1kb promoter region upstream of the human IFNB1 gene were cloned into the pHAGE2 lentiviral vector~\cite{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
-Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\cite{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\cite{kim2011high}.
+To create the type I interferon reporters, a 1kb promoter region upstream of the human IFNB1 gene were cloned into the pHAGE2 lentiviral vector~\citep{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
+Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\citep{kim2011high}.
 The sequence of the last of these constructs is provided in \SUPPFILE{reporter_sequences}.
@@
 We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
-We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\cite{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
+We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
 In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
 The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
-These genes were cloned into the pHW2000~\cite{hoffmann2000dna} reverse-genetics plasmid.
-
-Both viral strains were generated by reverse genetics using the pHW18* series of bi-directional plasmids~\cite{hoffmann2000dna}.
-We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\cite{xue2016propagation}.
-The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 cells in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \cite{russell2018extreme}.
+These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
+
+Both viral strains were generated by reverse genetics using the pHW18* series of bi-directional plasmids~\citep{hoffmann2000dna}.
+We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\citep{xue2016propagation}.
+The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 cells in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citep{russell2018extreme}.
 Specifically, after generation by reverse genetics, the wild-type variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
-The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\cite{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
+The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\citep{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
 
@@
 We estimated titers of transcriptionally active virions by staining for HA expression in virus-infected A549 cells.
-Specifically, we infected A549 cells (or one of the A549 reporter cell line variants as indicated) in influenza growth medium, and at 13 to 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\cite{doud2017complete}.
+Specifically, we infected A549 cells (or one of the A549 reporter cell line variants as indicated) in influenza growth medium, and at 13 to 14 hours post-infection, we trypsinized cells, re-suspended in phosphate-buffered saline (PBS) supplemented with 2\% heat-inactivated fetal bovine serum (FBS), and stained with 10 $\mu$g/ml of H17-L19, a mouse monoclonal antibody previously shown to bind to the HA from the A/WSN/1933 strain of virus~\citep{doud2017complete}.
 After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC, washed, fixed in 1\% formaldehyde in PBS, washed again, and then analyzed by flow cytometry to determine the fraction expressing detectable HA protein.
@@
 We began this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
-The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\cite{zheng2017massively}, targeting capture of $\sim$1,500 cells.
+The final cell suspension was counted using a disposable hemocytometer and loaded on the 10x Genomics Chromium instrument~\citep{zheng2017massively}, targeting capture of $\sim$1,500 cells.
 
@@
 Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
-This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{Boers:2015emulsion}.
+This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\citep{Boers:2015emulsion}.
 We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .
@@
 The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\cite{xue2016propagation} and are presumably preferentially amplified during PCR.
+As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\citep{xue2016propagation} and are presumably preferentially amplified during PCR.
 
@@
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
-For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml (a concentration sufficient to block secondary transcription~\cite{killip2014activation}) at the time of infection.
+For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml (a concentration sufficient to block secondary transcription~\citep{killip2014activation}) at the time of infection.
 After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.
@@
 This NS1stop virus contained six nucleotide changes resulting in the addition of five in-frame stop codons in NS1 starting 10 nucleotides downstream of the 5' splice donor site, thereby disrupting NS1 while leaving NS2 (NEP) intact.
-All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\cite{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
+All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\citep{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
  \SUPPFILE{validation_sequences} provides the full sequences for all of these plasmids.
 
-We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\cite{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{validation_sequences} for the wild-type plasmid for that gene.
+We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{validation_sequences} for the wild-type plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
@@
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\cite{hirst1942quantitative}.
+For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
@@
 To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validationflow}B, we used slightly different procedures.
-The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\cite{russell2018extreme}.
+The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
 The NS1stop virus was similarly generated, but was passaged for 48 rather than 36 hours, since it had slower growth kinetics and so needed a longer period of time to reach high titers.
 The viruses with deletions in the PB1 segment could not be generated in normal 293T and MDCK-SIAT1 cells, since they required the exogenous expression of the PB1 protein.
-Therefore, these viruses were generated in previously described 293T and MDCK-SIAT1 cells that had been engineered to constitutively express PB1~\cite{bloom2010permissive}.
+Therefore, these viruses were generated in previously described 293T and MDCK-SIAT1 cells that had been engineered to constitutively express PB1~\citep{bloom2010permissive}.
 These viruses were harvested from transfections at 72 hours, and passaged twice in the MDCK-SIAT1 cells constitutively expressing PB1 at a MOI of 0.001 for 72 hours and 0.01 for 48 hours. 
@@
 A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}. 
-This pipeline is orchestrated by \texttt{Snakemake}~\cite{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
+This pipeline is orchestrated by \texttt{Snakemake}~\citep{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
 The sequencing data and annotated cell-gene matrix are available on the GEO repository under accession GSE120839 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120839}).
@@
 First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
-We further processed these CCSs using custom Python code and the \texttt{minimap2}~\cite{li2018minimap2} long-read aligner (version 2.11-r797).
-The Python code has been implemented in the API of \texttt{dms\_tools2} (\url{https://jbloomlab.github.io/dms_tools2/}~\cite{bloom2015software}) package (version 2.3.0).
+We further processed these CCSs using custom Python code and the \texttt{minimap2}~\citep{li2018minimap2} long-read aligner (version 2.11-r797).
+The Python code has been implemented in the API of \texttt{dms\_tools2} (\url{https://jbloomlab.github.io/dms_tools2/}~\citep{bloom2015software}) package (version 2.3.0).
 A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{pacbio_analysis}.
@@
 Finally, we process the annotated cell-gene matrix in R to generate the plots shown in this paper.
-This analysis utilized a variety of R and Bioconductor~\cite{huber2015orchestrating} packages, including \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
+This analysis utilized a variety of R and Bioconductor~\citep{huber2015orchestrating} packages, including \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
 A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{monocle_analysis}.
@@
 
-\section*{Acknowledgments}
-We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
-We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
-%This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
-%JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
-%JDB is an Investigator of the Howard Hughes Medical Institute.
-%The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
-
-\bibliography{references}
-
 \clearpage
+\nolinenumbers
+\normalsize
+
 \section*{Supporting information}
@@
 \caption{
-{\bf Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
-Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
+{\bf Re-analysis of a published single-cell transcriptomic study~\citep{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
+Specifically, we downloaded the data from \citep{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citep{steuerman2018dissection}.
 Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
-As shown in the left-most panel, the sequencing depth of \cite{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
+As shown in the left-most panel, the sequencing depth of \citep{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
 An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\cite{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\cite{stetson2006type,honda2006type}.
+Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
 The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
@@
 {\bf (A)}
-To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\cite{strahle2006sendai}.
+To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
 The name of each of reporter cell line is indicated at the top of each row of plots.
@@
 {\bf Flow cytometry data for \FIG{IFNrare}B.}
-The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\cite{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
+The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
 After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
@@
 The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
-Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
 The exception is NP, which is detected in virtually all infected cells.
@@
 Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
-This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\cite{stetson2006type,honda2006type}.
+This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
 {\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
@@
 {\bf Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.}
-To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\cite{maaten2008visualizing} as implemented in \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
+To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
 In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
@@
 The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see panel B).
-From the frequencies of these different forms, we estimate~\cite{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
 About half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
@@
 \SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
-Note that since the rate of homologous recombination in influenza virus in negligible~\cite{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
+Note that since the rate of homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
 }
@@
 The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses less HA due to the lack of secondary transcription.
-Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\cite{bloom2010permissive} for PB1del385to2163.
+Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for PB1del385to2163.
 \SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
@@
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\cite{killip2014activation}.
+These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\citep{killip2014activation}.
 In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
@@
 \caption{
-{\bf Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna}.}
+{\bf Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna}.}
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}

### 2018-12-24 paper/paper.tex
removed tSNE plot (have too many supp figs and need to reduce)

@@
 These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.
-
-Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
-\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\citep{maaten2008visualizing} of the cells.
-Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.
 
@@
 \begin{suppfig}
-\includegraphics[width=\textwidth]{figures/single_cell_figures/p_tsne.pdf}
-\caption{
-{\bf Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.}
-To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\citep{maaten2008visualizing} as implemented in \texttt{Monocle}~\citep{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
-In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
-Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA, shown on a log (top) or linear (bottom) scale.
-As is clear from this plot, expression of influenza, IFN, and ISG genes contributes substantially to the structure of the data, since cells with high expression of these genes clearly group together.
-}
-\label{suppfig:tSNE}
-\end{suppfig}
-
-\begin{suppfig}
 \includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}

### 2018-12-24 paper/paper.tex
removed Steuerman reanalysis (have too many supp figs and need to reduce)

@@
 A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-Studies using reporters~\citep{killip2017single} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
-IFN expression also appears to be rare among cells in influenza-infected mice~\citep{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\citep{steuerman2018dissection} (\SUPPFIG{mice}).
-
-To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\citep{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
+Studies using reporters~\citep{killip2017single,kallfass2013visualizing} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare among influenza-infected cells, at least early in infection prior to extensive paracrine signaling and the accumulation of defective viral particles.
+Therefore, to efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\citep{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
 The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
-Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
+For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
@@
 Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
-The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini} .
+The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini}.
 Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
@@
 \begin{suppfig}
-\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}
-\caption{
-{\bf Re-analysis of a published single-cell transcriptomic study~\citep{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
-Specifically, we downloaded the data from \citep{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citep{steuerman2018dissection}.
-Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
-As shown in the left-most panel, the sequencing depth of \citep{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
-An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
-The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
-}
-\label{suppfig:mice}
-\end{suppfig}
-
-\begin{suppfig}
 \begin{minipage}[t]{0.52\textwidth}

### 2018-12-24 paper/paper.tex
updated reference style

@@
 \usepackage{natbib}
-\bibliographystyle{agsm}
+\bibliographystyle{cell}
 
@@
 Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
-This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\citep{Boers:2015emulsion}.
+This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\citep{boers2015micelle}.
 We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .

### 2018-12-24 paper/paper.tex
_Immunity_ formatting

@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
 \documentclass[10pt,letterpaper]{article}
@@
 
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
 % Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
+\definecolor{linkblue}{RGB}{0,76,153}
+\usepackage[colorlinks=true,linkcolor=linkblue,citecolor=linkblue,urlcolor=linkblue]{hyperref}
 \usepackage{nameref}
@@
 \DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
 
@@
 % Text layout
-\raggedright
+%\raggedright
 \setlength{\parindent}{0.5cm}
@@
 
+% section header formats
+\usepackage{sectsty}
+\definecolor{darkred}{RGB}{140,0,0}
+\sectionfont{\normalsize \color{darkred}}
+\subsectionfont{\normalsize \color{darkred}}
+\subsubsectionfont{\normalsize \color{darkred}}
+\paragraphfont{\normalsize \color{darkred}}
+
 %% Include all macros below
@@
 {\LARGE
-\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity}}
+\textbf{Single-cell virus sequencing of influenza infections that trigger innate immunity}}
 \newline
@@
 \\
+{\bf
 Alistair B. Russell\textsuperscript{1},
@@
 Jesse D. Bloom\textsuperscript{1,2,3*}
+}
 \\
@@
 
-\section*{Summary}
+\section*{SUMMARY}
+{\bf
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
@@
 Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
+}
 
@@
 % Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
+\section*{INTRODUCTION}
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
@@
 
-\section*{Results}
+\section*{RESULTS}
 
@@
 
-\section*{Discussion}
+\section*{DISCUSSION}
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
@@
 \small
-\subsection*{Acknowledgments}
+\subsection*{ACKNOWLEDGMENTS}
 We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
@@
 
-\subsection*{Author Contributions}
+\subsection*{AUTHOR CONTRIBUTIONS}
 ABR and JDB designed the study.
@@
 
-\subsection*{Declaration of Interests}
+\subsection*{DECLARATION OF INTERESTS}
 The authors declare no competing interests.
@@
 
-\section*{Methods}
+\section*{METHODS}
 
@@
 
-\section*{Supporting information}
+\section*{SUPPORTING INFORMATION}
 

### 2018-12-24 paper/paper.tex
re-worked abstract to emphasize stochasticity and indicate that most defects are not statistically significantly associated with IFN

@@
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
-Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
-However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
-Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
+Here we develop a new approach to assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
+We do this by determining both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Most cells are infected by virions with defects such as amino-acid mutations, internal deletions, or failure to express a gene.
+We identify instances of each type of defect that increase the likelihood that a cell activates an innate-immune response.
+However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even when a cell is infected by a virion that expresses unmutated copies of all genes.
+Our work shows that viral genetic variation substantially contributes to but does not fully explain the heterogeneity in single influenza-infected cells.
 }
@@
 \vspace{0.2in}
-\keywords{PacBio, influenza, interferon, single-cell RNA-seq, NS1, stochastic}
+\keywords{PacBio, single-cell RNA-seq, influenza, interferon, NS1, stochastic}
 

### 2018-12-24 paper/paper.tex
revision to Intro: emphasize stochasticity and do not overstate correlation of mutations and IFN

@@
 \documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\usepackage[top=0.85in,left=2.5in,footskip=0.75in]{geometry}
 
@@
 \newenvironment{fullwidth}{%
-  \begin{adjustwidth}{-2.3in}{}
+  \begin{adjustwidth}{-2.05in}{}
 }{\end{adjustwidth}}
@@
 \setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
+\textwidth 5.5in 
 \textheight 8.75in
@@
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
-As the virus spreads, some cells detect infection and begin producing interferon (IFN).
+As the virus spreads, some cells detect infection and produce interferon (IFN).
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
@@
 For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
-But for influenza, a third possible contributor also looms large: viral genetic diversity.
-Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+But for influenza, a third possible contributor looms large: viral genetic diversity.
+Because influenza has a high mutation rate, individual virions often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, bloom2014experimentally, pauly2017novel}.
 Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
@@
 Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
-None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
-
-Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
-To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
+None of these techniques reliably reveal if there are mutations in the virion infecting any given single cell.
+
+Here we develop a new approach to determine both the transcriptome and full sequences of all viral genes in single influenza-infected cells.
+To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
 We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
-Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
-This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
-Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
-However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
+Two-thirds of cells are infected by virions with a mutation or defect in gene expression.
+This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
+We identify several types of viral defects that increase IFN induction.
+However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
 Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

### 2018-12-24 paper/paper.tex
combined all plasmid maps in one suppfile to reduce number of supplementary files

@@
 
-\subsection*{A system to identify and enrich for rare IFN+ cells}
+\subsection*{A system to identify and enrich rare IFN+ cells}
 
@@
 \centerline{
-{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
+{\bf \Large A}\includegraphics[width=0.34\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
 \hspace{0.02\textwidth}
-{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
+{\bf \Large B} \includegraphics[width=0.22\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
 }
@@
 {\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
-We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
+We created reporters with type I and type III IFN promoters (\SUPPFILE{plasmids}).
 In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
@@
 Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
-The plot also shows uninfected cells, and cells infected with Sendai virus.
+The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
 The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
@@
 
-A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-Studies using reporters~\citep{killip2017single,kallfass2013visualizing} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that IFN induction is rare among influenza-infected cells, at least early in infection prior to extensive paracrine signaling and the accumulation of defective viral particles.
-Therefore, to efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\citep{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
+A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
+Studies using reporters~\citep{killip2017single,kallfass2013visualizing} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that at early timepoints IFN induction occurs in $<$1\% of influenza-infected cells.
+To efficiently identify and enrich these rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein ~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
 For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
-We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
+We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells~\citep{russell2018extreme}.
 
@@
 All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\citep{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
- \SUPPFILE{validation_sequences} provides the full sequences for all of these plasmids.
-
-We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{validation_sequences} for the wild-type plasmid for that gene.
+ \SUPPFILE{plasmids} provides the full sequences for all of these plasmids.
+
+We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{plasmids} for the wild-type plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
@@
 \begin{suppfile}
-\caption{{\bf Sequences of the IFN reporters in \FIG{IFNrare}A.}
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
-}
-\label{suppfile:reporter_sequences}
+\caption{{\bf Sequences of the plasmids used in this study.}
+The plasmid maps are for the IFN reporters in \FIG{IFNrare}A and the mutant viral genes cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna}.
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}.
+}
+\label{suppfile:plasmids}
 \end{suppfile}
@@
 \label{suppfile:mutations}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
-{\bf Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna}.}
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}
-}
-\label{suppfile:validation_sequences}
 \end{suppfile}

### 2018-12-24 paper/paper.tex
merged supporting figures for Fig 1 to reduce amount of supp material

@@
 % Added by Jesse Bloom for figures
-\renewcommand{\floatpagefraction}{.9}
+\renewcommand{\floatpagefraction}{0.99}
+\renewcommand{\textfraction}{0.01}
 \setcounter{topnumber}{8}
@@
 {\bf (B)}
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{reportervalidation}C for full data).
 The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
@@
 \begin{suppfig}
-\begin{minipage}[t]{0.52\textwidth}
+\begin{minipage}[t]{0.46\textwidth}
 {\bf \Large A}
-
 \includegraphics[width=0.9\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}
 \end{minipage}
-\begin{minipage}[t]{0.48\textwidth}
+\begin{minipage}[t]{0.55\textwidth}
 {\bf \Large B}
-
-\includegraphics[width=0.7\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
+\includegraphics[width=0.57\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
+\vspace{0.1in}
+
+{\bf \Large C}
+\includegraphics[width=0.9\textwidth, valign=t]{figures/IFN_stochastic/Flow/flow_plot.pdf}
 
@@
 \caption{
-{\bf Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.}
+{\bf Validation of IFN reporter, related to \FIG{IFNrare}.}
 {\bf (A)}
-To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\citep{strahle2006sendai}.
-The name of each of reporter cell line is indicated at the top of each row of plots.
-At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
-Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
+IFN reporter cell lines were infected at high MOI with the Cantell strain of Sendai virus~\citep{strahle2006sendai}.
+The name of the reporter is indicated above the plots.
+At 13 hours post-infection, activation of the reporter was monitored by flow cytometry using the marker indicated below each plot (a fluorescent protein or antibody staining for the LNGFR$\Delta$C using PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
+Sendai infection efficiently activated the IFN reporter in all cases.
 {\bf (B)}
 The type I and type III IFN reporters are highly correlated in their activation.
-An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
-The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
+An A549 cell line was generated by transduction with both IFN-$\beta$ and IFN-$\lambda$ reporters driving mCherry and ZsGreen, respectively.
+The cells were infected with two different stocks of WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1.
 After 13 hours, cells were analyzed by flow cytometry.
-Cells positive for either fluorescent reporter were further analyzed.
-As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.
+Cells positive for either reporter were further analyzed.
+As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated.
+{\bf (C)}
+Flow cytometry data for \FIG{IFNrare}B.
+A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus, or infected at a target MOI of 0.3 with the stock of influenza virus used in the single-cell experiment.
+At 13 hours, cells were analyzed by FACS for HA and ZsGreen expression.
+Each condition was done in triplicate.
+Contour plots show density of all cells, and IFN+ cells are also indicated by orange dots.
+Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
+For influenza-infected cells, the percentage IFN+ was calculated only among HA+ cells (since these are the ones that are infected).
+For uninfected and Sendai-infected cells, the percentage IFN+ was calculated among all cells, since the cells do not express HA.
 }
 \label{suppfig:reportervalidation}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.9\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
-\caption{
-{\bf Flow cytometry data for \FIG{IFNrare}B.}
-The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\citep{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
-After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
-Each condition was done in triplicate.
-The contour plots show the density of all cells, and all IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
-For the influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
-For the uninfected and Sendai-virus infected, the percentage IFN+ was calculated among all cells, since these cells do not express HA.
-}
-\label{suppfig:IFNrare_flow}
 \end{suppfig}

### 2018-12-24 paper/paper.tex
consolidated plasmid maps and virus sequences supp files to reduce number of supplementary files

@@
 
-To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
-First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
+To determine if mutations or other defects in the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
+First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{plasmids}).
 These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
@@
 In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
-The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
+The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{plasmids}.
 These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
@@
 The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
-The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \SUPPFILE{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
+The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \SUPPFILE{plasmids} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
 The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
@@
 \begin{suppfile}
-\caption{{\bf Sequences of the plasmids used in this study.}
+\caption{{\bf Sequences of plasmids and viruses used in this study.}
 The plasmid maps are for the IFN reporters in \FIG{IFNrare}A and the mutant viral genes cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna}.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}.
+In addition, there are Genbank files giving the sequences of the wild-type and synonymously barcoded WSN viruses.
+A README file explains the contents of the ZIP.
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}, \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}, \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb}, and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}
 }
 \label{suppfile:plasmids}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{
-{\bf Genbank files giving sequences of the wild-type and synonymously barcoded viruses}.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
-}
-\label{suppfile:virus_seqs}
 \end{suppfile}

### 2018-12-24 paper/paper.tex
tiny edits to second section of results

@@
 \caption{
-{\bf Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.}
+{\bf Approach for combined transcriptomics and virus sequencing of single influenza-infected cells that express IFN.}
 {\bf (A)}
@@
 First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{plasmids}).
-These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\citep{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+These viral barcodes allow us to identify co-infections~\citep{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
+From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
 To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\citep{travers2010flexible}).
+We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.

### 2018-12-24 paper/paper.tex
tweaks to transcriptomics section

@@
 Number of cells for which transcriptomes were obtained.
-From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
+%From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
-To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.
+To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\citep{haque2017practical}.
+After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B).
 
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
-We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
-The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets prior to reverse-transcription.
+We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected by virus (\FIG{transcriptomics}C).
+We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
+The distribution of the amount of viral mRNA across infected cells is in the inset in \FIG{transcriptomics}C.
 As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
 
-We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
-We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
-This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
-
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
-However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
-Consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\citep{hatada1989control}.
+We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
+%We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
+\FIG{transcriptomics}D (top panel) shows that a slight majority (162 of 290) of infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
+This measured frequency of infected cells expressing all eight genes is slightly higher than in our prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which estimated that only 13\% to 50\% of infected cells express all genes.
+
+The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
+However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus.
+Consistent with prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and similar to values from older bulk studies~\citep{hatada1989control}.
 
@@
 From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
-Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
+This co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
 This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
+Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
@@
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not directly detect infection~\citep{stetson2006type,honda2006type}.
 
@@
 \caption{
-{\bf Example MACS enrichments of IFN+ influenza-infected cells.}
+{\bf Example MACS enrichments of IFN+ influenza-infected cells, related to \FIG{workflow}.}
 A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.

### 2018-12-26 paper/paper.tex
merged supp figures for transcriptomics figure to reduce number of supp figures

@@
 The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
-\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
+\SUPPFIG{transcriptomics_supp}A shows the frequency that each viral gene is detected.
 {\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
@@
 A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
-Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
+Many cells that do not express IFN still express ISGs (\SUPPFIG{transcriptomics_supp}C,D).
 }
@@
 %We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that a slight majority (162 of 290) of infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
+\FIG{transcriptomics}D (top panel) shows that a slight majority (162 of 290) of infected cells express all eight genes (see \SUPPFIG{transcriptomics_supp}A for frequencies for individual genes).
 This measured frequency of infected cells expressing all eight genes is slightly higher than in our prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which estimated that only 13\% to 50\% of infected cells express all genes.
@@
 
-We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
+We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{transcriptomics_supp}B-D).
 Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
-Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{ISG}A).
-The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not directly detect infection~\citep{stetson2006type,honda2006type}.
+Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{transcriptomics_supp}C).
+The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{transcriptomics_supp}D).
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\citep{stetson2006type,honda2006type}.
 
@@
 Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\citep{kim2011high}.
-The sequence of the last of these constructs is provided in \SUPPFILE{reporter_sequences}.
+The sequence of the last of these constructs is provided in \SUPPFILE{plasmids}.
 
@@
 Downstream of this promoter we cloned LNGFR$\Delta$C linked to ZsGreen via a P2A linker.
-The sequence of this construct is provided in \SUPPFILE{reporter_sequences}.
+The sequence of this construct is provided in \SUPPFILE{plasmids}.
 
@@
 We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
-We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{frac_has_gene}).
+We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{transcriptomics_supp}A).
 To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
-We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{ISG}, respectively.
-We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
+We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{transcriptomics_supp}C,D, respectively.
+We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts, since type I and type III IFN were highly correlated (\SUPPFIG{transcriptomics_supp}B).
 We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
@@
 \begin{suppfig}
-\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}
+\includegraphics[width=0.9\textwidth]{figures/single_cell_figures/p_cell_summary_supp.pdf}
 \caption{
-{\bf The fraction of infected cells that are called as expressing each viral gene.}
-The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
-Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+{\bf Additional plots from analysis of transcriptomic data, related to \FIG{transcriptomics}.}
+{\bf (A)}
+Fraction of infected cells that are called as expressing each viral gene.
+The gray dashed line is at one.
+Each gene is detected in $\sim$80-90\% of cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
 The exception is NP, which is detected in virtually all infected cells.
-The much higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments because the number of viral mRNAs is below the detection limit.}
-\label{suppfig:frac_has_gene}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}
-\caption{
-{\bf The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.}
+The higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have lower viral gene expression and so are not reliably called as infected in our experiments because the number of viral mRNAs is below the detection limit.
+{\bf (B)}
+Correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
 Each point represents one cell.
 The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
-Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
+{\bf (C)}
+Distribution of ISGs expression across single infected and uninfected cells. 
+For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
+We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from these ISGs, and color these cells red.
+Comparison to \FIG{transcriptomics}G shows that more cells are ISG+ than IFN+, probably because paracrine signaling can induce ISG expression in cells that do not themselves express IFN~\citep{stetson2006type,honda2006type}.
+{\bf (D)} 
+Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
+Each point is one cell, and the Pearson correlation coefficient is shown.
+IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
+Among both the infected and uninfected populations, there are many cells that express ISGs but not IFN, but no cells that express IFN but not ISGs.
 }
-\label{suppfig:type_I_III_correlation}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}
-\caption{
-{\bf Expression of ISGs in single infected and uninfected cells.}
-For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
-{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
-We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
-Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
-This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
-{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
-Each point represents one cell, and the Pearson correlation coefficient is shown.
-IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
-Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
-}
-\label{suppfig:ISG}
+\label{suppfig:transcriptomics_supp}
 \end{suppfig}

### 2018-12-26 paper/paper.tex
merged PacBio supp figures into one to reduce total number of supp figures

@@
 We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
-Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{PacBio_supp}A).
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{PacBio_supp}B), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
 
@@
 We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
-In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{PacBio_supp}B).
 The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{plasmids}.
@@
 The sequences of these primers are in \SUPPFILE{primer_sequences}.
-A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
+A major concern during these PCRs is strand exchange (see \SUPPFIG{PacBio_supp}B) which would scramble the cell barcodes and mutations on viral cDNAs.
 To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
@@
 The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
-The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{StrandExchange} for an illustration of how this is determined).
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{PacBio_supp}A, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{PacBio_supp}B for an illustration of how this is determined).
 A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
@@
 We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
-At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{StrandExchange}), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}).
+At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange, and reassuringly found that strand exchange was rare (\SUPPFIG{PacBio_supp}).
 The small number of CCSs with identifiable strand exchange were filtered from further analysis.
@@
 \begin{suppfig}
-\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
+{\bf \Large A}
+\includegraphics[width=0.9\textwidth, valign=t]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
+\vspace{0.05in}
+
+{\bf \Large B}
+\includegraphics[width=0.52\textwidth, valign=t]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
 \caption{
-{\bf The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.}
-These sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells (see \SUPPFILE{pacbio_analysis}).
-Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
-Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
-Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
-Note that the disproportionate sequencing of the shorter internally deleted genes does not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse sequences derived from the same cDNA, and cell barcodes were used to collapse sequences from the same cell.
-The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see panel B).
-From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
-About half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
-For some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
-A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
+{\bf PacBio CCS quality control, related to \FIG{genotypes}.}
+{\bf (A)}
+Number of PacBio CCSs that passed quality-control and aligned to each viral gene.
+CCSs were obtained using several PacBio runs which were loaded with different amounts of each gene to balance coverage (see \SUPPFILE{pacbio_analysis}).
+Therefore, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs is \emph{not} an indicator of a transcript's abundance.
+Especially for the polymerase genes, many CCSs corresponded to genes with internal deletions, since these shorter forms are preferentially amplified by PCR.
+Therefore, the plot is faceted by CCSs for any length of the gene, and for full-length genes.
+The disproportionate sequencing of the shorter internally deleted genes does not affect the genotype calling since UMIs were used to collapse sequences from the same cDNA, and cell barcodes were used to collapse sequences from the same cell.
+The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or is a mixed-barcode molecule (see panel B).
+We estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+About half of these chimeras can identified by the mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
+For some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the viral barcode could only be partially called.
+{\bf (B)}
+Strategy for detecting strand exchange in PacBio sequencing.
+Preparation for PacBio sequencing required many PCR cycles, which can cause strand exchange that scrambles mutations and 10X cell barcodes / UMIs from different molecules.
+If there is no strand exchange, all molecules are either wild-type or have the synonymous barcoding mutations at \emph{both} termini.
+Strand exchange creates molecules with wild-type nucleotides at one termini and synonymous barcoding mutations at the other.
+Panel A shows the frequencies of these molecules.
+Since homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, mixed-barcode molecules are \emph{not} generated naturally during co-infection.
 }
-\label{suppfig:CCSs}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.8\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
-\caption{
-{\bf Strategy for detecting strand exchange during sequencing of full-length viral genes.}
-The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
-A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
-We detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
-If there is no strand exchange, all molecules should either be wild-type or have the synonymous barcoding mutations at \emph{both} termini.
-Strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
-\SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
-Note that since the rate of homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
-}
-\label{suppfig:StrandExchange}
+\label{suppfig:PacBio_supp}
 \end{suppfig}

### 2018-12-26 paper/paper.tex
slightly shortened legend to Figure S6

@@
 Data for \FIG{validation}A.
-A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with stocks of the indicated mutant.
+A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with the indicated mutant.
 After 13 hours, cells were stained for HA protein and analyzed by FACS.
 Contour plots show density of all cells, and IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of uninfected cells in these populations.
+Cells were classified as HA+ or IFN+ based on gates that put 0.05\% of uninfected cells in these populations.
 For infected cells, the percentage IFN+ was calculated among the HA+ cells (since these are the ones that are infected).
@@
 For each viral mutant, two independent stocks were assayed in duplicate (i.e., \#1a and \#1b are one viral stock, and \#2a and \#2b are the other).
-The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
+Infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
 {\bf (B)}
@@
 The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses less HA due to the lack of secondary transcription.
-Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for PB1del385to2163.
+Therefore, all cells were infected at a MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for PB1del385to2163.
 \SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.

### 2018-12-26 paper/paper.tex
revised genotypes section

@@
 {\centering
-\includegraphics[height=0.92\textheight]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 {\bf Viral genotypes and infection outcomes in single cells.}
-Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
-Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
+Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
+The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
+Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
+Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
+The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
 For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
-Green boxes at left show the percent of all mRNA in that cell derived from virus.
-Orange boxes show the percent of cellular mRNA derived from IFN, with orange box frames indicating cells classified as IFN+ in \FIG{transcriptomics}G.
+For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1. 
 }
@@
 
-We called a mutation in a viral gene if it was found in least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
-For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
-This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+After calling the presence / absence of each viral gene in each cell using the transcriptomic data as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
+For cells co-infected with both viral barcode variants, we called mutations separately for each variant.
+This strategy identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
 It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
@@
 
-We could call the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
-We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
-But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
+We called the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
+We were most effective at calling viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
+But we also called genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
 
-The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
-Simple visual inspection of this figure reveals a wealth of information.
-For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
-But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
-The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
-However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the wild-type virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the wild-type virion that infected \textit{cell 105} still induced IFN.
+The 150 cells for which we called the viral genotypes are shown in \FIG{genotypes}.
+Inspection of this figure reveals a wealth of anecdotal relationships between viral genotype and infection outcome.
+For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all genes and did not induce detectable IFN.
+But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
+As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\citep{garcia1998influenza, hale2008multifunctional}, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+
+However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
+This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
+As an example of the former case, expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only expressed viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
+But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
+For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
 

### 2018-12-27 paper/paper.tex
revised legend to Figure 5

@@
 \caption{
-{\bf Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.}
+{\bf Viral features associated with heterogeneity among cells for which we determined viral genotypes.}
 {\bf (A)} 
 Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
-Cells infected by fully wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
+Cells infected by unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals indicated).
 {\bf (B)}
-IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
+IFN expression among cells expressing unmutated copies of all genes, and cells with mutations or missing genes.
 {\bf (C)}
@@
 The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
-The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
-Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
+The middle and bottom panels show the percent of IFN- and IFN+ cells with a deletion or amino-acid substitution in each gene, conditioned on the gene being expressed.
+Numbers give P-values (Fisher's exact test) for rejecting null hypothesis that percents are equal among IFN- and IFN+ cells. 
 {\bf (D)}
-There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
-Throughout this figure, we only consider substitutions that are non-synonymous.
-Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
+There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN among cells that lack NS.
+This figure only shows non-synonymous substitutions, and disregards insertions as they are very rare.
 }

### 2018-12-27 paper/paper.tex
partially added text about stochasticity to last results section

@@
 \providecommand{\keywords}[1]{\noindent \textbf{Keywords:} #1}
+
+\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
 
@@
 To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
+\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
@@
 Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.
+However, the validation experiments in the next section show that the lack of statistical significance for many of the trends is due to the modest number of infected cells that were sequenced rather than a lack of true association between viral defects and IFN induction.
 
 One other interesting trend emerges from the single-cell data.
-There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
+There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS (\FIG{mutations}D).
+But among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
+\jdbcomment{Perhaps elaborate a bit after adding additional data relevant to this point to next section.}
 
-\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}
+\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction}
 
@@
 \caption{
-{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
+{\bf Validation that IFN induction is increased by mutations identified in the single-cell virus sequencing of IFN+ cells.}
 {\bf (A)}
-Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
-The numbers indicate the median of four measurements for each viral mutant.
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated mutant, as determined using a reporter cell line.
+Numbers give the median of four measurements for each mutant.
 The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
-Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
+Points are orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
 {\bf (B)}
@@
 See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
-The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
+The experiments in the two panels were performed on different days, and so numerical values can be compared within but not between panels.
 }
@@
 
-To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
+To test if the viral defects identified in single IFN+ cells directly increase IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
 
@@
 The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
-We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
+We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.
 
-The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of the amino-acid substitutions found in these genes among IFN+ cells.
-Overall, we created six such point-mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
+The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
 
-Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
-We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
+Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variants has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+We created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
 
 We tested the rate of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
+\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild type viral stock.
 The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
@@
 But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
+
+However, IFN induction remains stochastic even for the most strongly IFN-inducing viral mutants.
+\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the viral genotype.
+All the viral mutants induce IFN in less than a fifth of cells
 

### 2018-12-27 paper/paper.tex
finished updating last section of Results to emphasize stochasticity

@@
 We tested the rate of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild type viral stock.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
-The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
-But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
-
-However, IFN induction remains stochastic even for the most strongly IFN-inducing viral mutants.
-\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the viral genotype.
-All the viral mutants induce IFN in less than a fifth of cells
+\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild-type viral stock.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN more frequently than wild type.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we expect some mutations to be found in IFN+ cells by chance.
+Overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often increase IFN induction.
+
+However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
+\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the genotype.
+These flow cytometry data reveal that no mutant virus stock induces IFN in more than a fraction of cells.
+Of course, these mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to that revealed by our single-cell sequencing of the ``wild-type'' viral stock.
+But IFN induction is stochastic at the single-cell level even for infections that share the same defect,such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
+Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction is stochastic even with these defects.
+
+\jdbcomment{Perhaps also add NS+ / NS- data here}
 

### 2018-12-27 paper/paper.tex
tweaks to Discussion to emphasize stochasticity

@@
 Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\citep{saikia2018simultaneous, zanini2018virus}.
+Our method builds on the observation by that fragmentary viral genetic information can be obtained by Illumina-based single-cell transcriptomic techniques~\citep{saikia2018simultaneous, zanini2018virus}.
 To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \citep{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
-This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
+This viral genetic information is crucial for understanding infection outcome and innate-immune induction.
 Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wild-type'' influenza genome, most infected cells do not express wild-type copies of all viral genes.
-Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
-Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
-
-It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
-We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
-Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
+Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of defects across single cells.
+Inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
+
+These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
+We identified four types of defects in single IFN+ cells that we validated increase IFN induction. 
+Two types of defects---absence of NS and mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
 The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini}.
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
 There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
+The relative importance of different types of viral defects also likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
 
 But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
-We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
-Additionally, none of the viral defects in IFN+ cells induce IFN deterministically in our validation experiments---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
+We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
+Additionally, none of the viral defects in IFN+ cells induce IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
 Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
-However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\citep{oneal2018west}.
+However, we suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\citep{oneal2018west}.
 
 Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
-Natural human influenza infections are established by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
-It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
-Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
+Natural human influenza infections are initiated by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
+It is therefor eplausible that early heterogeneity could have a large effect on downstream events.
+Extending our approaches to more complex systems could shed further light on how viral variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
 

### 2018-12-27 paper/paper.tex
combined supp figures for validation experiments to reduce number of supp figs

@@
 Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
-See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
+See \SUPPFIG{validation_supp} for more detailed experimental data.
 The experiments in the two panels were performed on different days, and so numerical values can be compared within but not between panels.
@@
 However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
-\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the genotype.
+\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validation_supp}), which is itself a single-cell measurement, albeit one that does not report the genotype.
 These flow cytometry data reveal that no mutant virus stock induces IFN in more than a fraction of cells.
@@
 
-For the qPCR in \SUPPFIG{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+For the qPCR in \SUPPFIG{validation_supp}C, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
@@
 At 24 hours post-transfection, we changed media from D10 to influenza growth media. 
-At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{validationflow}A) or 72 hours (for the replicate 2 viruses in \SUPPFIG{validationflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
+At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{validation_supp}A) or 72 hours (for the replicate 2 viruses in \SUPPFIG{validation_supp}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
+For the infections in \SUPPFIG{validation_supp}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
 Two-fold serial dilutions of virus were added to an equal volume of diluted red blood cells, and titer was measured as the highest dilution of viral stock at which complete hemagglutination of red blood cells was observed.
-We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{validationflow}A.
-
-To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validationflow}B, we used slightly different procedures.
+We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{validation_supp}A.
+
+To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validation_supp}B, we used slightly different procedures.
 The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
@@
 The wild-type and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
-The infections in \SUPPFIG{validationflow}B were performed at equivalent TCID50s as described in the legend to that figure.
-That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \SUPPFIG{delqPCR}.
+The infections in \SUPPFIG{validation_supp}B were performed at equivalent TCID50s as described in the legend to that figure.
+That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \SUPPFIG{validation_supp}C.
 
@@
 \begin{suppfig}
-{\bf \Large A}\includegraphics[width=0.48\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
+\begin{minipage}[t]{0.46\textwidth}
+{\bf \Large A}
+\includegraphics[width=0.9\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
+\end{minipage}
 \hspace{0.02\textwidth}
-{\bf \Large B} \includegraphics[width=0.39\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
-\caption{
-{\bf Flow cytometry data for \FIG{validation}.}
+\begin{minipage}[t]{0.5\textwidth}
+{\bf \Large B}
+\includegraphics[width=0.75\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
+\vspace{0.1in}
+
+{\bf \Large C}
+\includegraphics[width=0.75\textwidth, valign=t]{figures/Validation_Figure/qpcr_plot.pdf}
+
+\end{minipage}
+\caption{
+{\bf Additional data for validation experiments, related to \FIG{validation}.}
 {\bf (A)}
-Data for \FIG{validation}A.
-A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with the indicated mutant.
-After 13 hours, cells were stained for HA protein and analyzed by FACS.
-Contour plots show density of all cells, and IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates that put 0.05\% of uninfected cells in these populations.
-For infected cells, the percentage IFN+ was calculated among the HA+ cells (since these are the ones that are infected).
-For uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
-For each viral mutant, two independent stocks were assayed in duplicate (i.e., \#1a and \#1b are one viral stock, and \#2a and \#2b are the other).
-Infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
+Flow cytometry data for \FIG{validation}A.
+Data were collected and analyzed as in \SUPPFIG{reportervalidation}C.
+For each viral mutant, two independent stocks were assayed in duplicate (\#1a and \#1b are one viral stock; \#2a and \#2b are the other).
+Infections with replicate \#1 of wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and other viruses were infected at an equivalent particle number as determined by HI assay. 
 {\bf (B)}
-Data for \FIG{validation}B.
+Flow cytometry data for \FIG{validation}B.
 The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses less HA due to the lack of secondary transcription.
-Therefore, all cells were infected at a MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for PB1del385to2163.
-\SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
+Therefore, cells were infected at a MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for PB1del385to2163.
+Panel C shows that at these TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
 The percent IFN+ was calculated for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for PB1del385to2163.
-}
-\label{suppfig:validationflow}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.6\textwidth]{figures/Validation_Figure/qpcr_plot.pdf}
-\caption{
-{\bf Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.}
-In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{validationflow}B), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
+{\bf (C)}
+Validation that infections in panel B were performed at similar doses of virions capable of initiating primary transcription.
+A549 cells were infected at MOI of 0.4 (based on TCID50 as described panel B), and after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse-transcription products.
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\citep{killip2014activation}.
-In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
+Infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription~\citep{killip2014activation}.
+In the absence of cycloheximide, viruses with deletions in PB1 produced less viral mRNA because they could not produce PB1 protein for secondary transcription.
 But in the presence of cycloheximide, all viruses produced similar amounts of viral mRNA, indicating that the dose of particles active for primary transcription is roughly equivalent across variants.
@@
 }
-\label{suppfig:delqPCR}
+\label{suppfig:validation_supp}
 \end{suppfig}

### 2018-12-27 paper/paper.tex
NS referenced before properly defined, added a few words to make the text a bit clearer

@@
 But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
-For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
+For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS, the viral gene segment encoding the major interferon antagonist, NS1.
 
@@
 Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
-The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
+The NS gene encodes two proteins (NS1 and NS2), the first of which, as mentioned, is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
 We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.

### 2018-12-27 paper/paper.tex
Changed wording of helping in third paragraph of discussion. It feels an odd word to use here. Hope that the changed language is a bit better.

@@
 
-These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
+These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including the probability that a cell triggers an IFN response.
 We identified four types of defects in single IFN+ cells that we validated increase IFN induction. 

### 2018-12-27 paper/paper.tex
Viral variation feels colloquial, tried a slightly different wording
just fixing a minor typo

@@
 
-But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
+But although variation within viral populations is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
 We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
@@
 Natural human influenza infections are initiated by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
-It is therefor eplausible that early heterogeneity could have a large effect on downstream events.
+It is therefore plausible that early heterogeneity could have a large effect on downstream events.
 Extending our approaches to more complex systems could shed further light on how viral variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.

### 2018-12-27 paper/paper.tex
Added Damon Runyon grant number before I forget
added plots on IFN induction by NS+ versus NS-

--- 2018-12-27 paper/paper.tex
@@
 But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
-For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS, the viral gene segment encoding the major interferon antagonist, NS1.
+For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
 
@@
 Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
-The NS gene encodes two proteins (NS1 and NS2), the first of which, as mentioned, is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
+The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
 We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.
@@
 Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction is stochastic even with these defects.
+
+%%% begin mutantcomparison figure %%%
+\begin{figure}
+\centerline{\includegraphics[width=0.85\textwidth]{figures/MutantComparison/p_merged.pdf}}
+\caption{
+Higher viral gene expression is strongly associated with more IFN induction only in NS1-deficient infections.
+{\bf (A)}
+For viruses with defects in NS1, there is more IFN induction among infected cells that express higher levels of HA protein.
+The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile as determined by flow cytometry.
+Points indicate replicates, and lines indicate the mean.
+See \SUPPFIG{mutantcomparison_supp}A for details.
+{\bf (B)}
+Blocking viral secondary transcription strongly decreases IFN in cells infected with NS1stop virus, but not cells infected with virus carrying a deletion in PB1.
+A549 cells were infected at doses that induced equivalent IFN, and ribavirin was added at a concentration (100 $\mu$M) sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg}.
+After eight hours, IFNB1 transcript was quantified relative to the housekeeping gene L32 by qPCR. 
+The dashed gray line indicates the limit of detection.
+\SUPPFIG{mutantcomparison_supp}B shows that results are similar if secondary transcription is instead blocked with cycloheximide.
+}
+\label{fig:mutantcomparison}
+\end{figure}
+%%% end mutantcomparison figure%%%
 
@@
 After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.
+
+For the qPCR in \FIG{mutantcomparison}B, A549 cells were infected with the NS1stop virus at a MOI of 0.1 (based on TCID50) in MDCK-SIAT1 and with the PB1del385to2163 virus at a dose that induced equivalent IFN in the absence of ribavirin.
+The ribavirin was added to cells at the time of infection to a final concentration of 100$\mu$M, which is sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg}.
+All other steps were performed as for the cycloheximide experiments described immediately above..
 
@@
 
+\begin{suppfig}
+\begin{minipage}[t]{0.55\textwidth}
+{\bf A}
+\includegraphics[width=0.9\textwidth, valign=t]{figures/MutantComparison/p_ifn_vs_ha.pdf}
+\end{minipage}
+\hspace{0.02\textwidth}
+\begin{minipage}[t]{0.43\textwidth}
+{\bf B}
+\includegraphics[width=0.9\textwidth, valign=t]{figures/MutantComparison/cyclo_qPCR_plot.pdf}
+\end{minipage}
+
+\caption{
+{\bf Additional data for comparison of IFN induction as a function of viral load, related to \FIG{mutantcomparison}.}
+{\bf (A)}
+Detailed analysis of the relationship among HA expression and IFN induction for different viral mutants that is summarized in \FIG{mutantcomparison}A.
+We re-analyzed all the infected cells in \SUPPFIG{validation_supp}A (defining ``infected'' using the same HA-gating scheme in \SUPPFIG{validation_supp}A) for the wild-type virus, the NS1stop virus, and the three amino-acid mutants with significantly increased IFN induction.
+For each virus and replicate, we binned the infected cells into HA expression quartiles.
+We then calculated the percent of cells that were IFN+ in each quartile (defining ``IFN+'' using the same gating scheme in \SUPPFIG{validation_supp}A).
+The plots show the mean HA expression in each quartile versus the percent of cells that are IFN+.
+The results clearly indicate that for the NS1stop virus, increased viral burden (more HA) correlates with IFN induction.
+A similar trend also holds for the virus with an amino-acid substitution in NS1 (NS1-A122V). 
+But for wild type and the two PB1 mutants, there is no clear trend for higher viral burden to correlate with IFN induction.
+\FIG{mutantcomparison}A summarizes the data in this figure by simply showing the ratio of the percent IFN+ of the highest HA-expression quartile to the lowest HA-expression quartile for each sample and replicate.
+{\bf (B)}
+Similar to \FIG{mutantcomparison}B, but uses cycloheximide rather than ribavirin.
+Preventing viral secondary transcription using cycloheximide also decreases IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
+However, the results are more complex here, since cycloheximide alone increases IFN production~\citep{raj1981analysis, maroteaux1983cycloheximide, killip2014activation}.
+Therefore, we see an increase in IFN levels upon cycloheximide treatment for uninfected cells and cells infected with the PB1 deletion virus, but a decrease for cells infected with the NS1stop virus.
+This plot shows qPCR results for the same samples as in \SUPPFIG{validation_supp}C, but with the qPCR for expression of IFNB1 transcripts relative to the housekeeping gene L32.
+As described in \SUPPFIG{validation_supp}C, A549 cells were infected at an MOI of 0.4, such that HA expression was equivalent between viral variants when secondary transcription was blocked, and some of the samples treated with 50 $\mu$g/ml of cycloheximide~\citep[a concentration sufficient to block secondary transcription;][]{killip2014activation}.
+After 8 hours, transcript levels were quantified by qPCR.
+Points indicate four replicates, and lines indicate the mean.
+The dashed line shows the limit of detection, and points below this limit where set to that value.
+}
+\label{suppfig:mutantcomparison_supp}
+\end{suppfig}
+
 \clearpage

### 2018-12-27 paper/paper.tex

@@
 This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation (DRG-2227-15).
 JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.

### 2018-12-27 paper/paper.tex
tweaks merging Alistair edits

@@
 As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
-The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\citep{garcia1998influenza, hale2008multifunctional}, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist, NS1~\citep{garcia1998influenza, hale2008multifunctional}.
+Many other IFN+ cells also lacked NS or had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
 
@@
 Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
-The NS gene encodes two proteins (NS1 and NS2), the first of which, as mentioned, is influenza's primary innate-immune antagonist~\citep{garcia1998influenza, hale2008multifunctional}.
+%The NS gene encodes influenza's primary innate-immune antagonist, NS1~\citep{garcia1998influenza, hale2008multifunctional}.
 We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.
@@
 The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini}.
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
+Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Given the extensive prior work describing such deletions, it is surprising we did not identify more of them in our single IFN+ cells.
 There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-The relative importance of different types of viral defects also likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
-
-But although variation within viral populations is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
+The relative importance of different defects also likely varies across conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
+
+But although variation in viral populations is clearly important, it only partially explains heterogeneity among influenza-infected cells.
 We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
-Additionally, none of the viral defects in IFN+ cells induce IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
+Additionally, no viral defect induces IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at high frequency.
 However, we suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\citep{oneal2018west}.

### 2018-12-27 paper/paper.tex
added NS / viral burden data

@@
 %%% begin mutations figure %%%
-\begin{figure}
+\begin{figure}[b!]
 \begin{fullwidth}
@@
 
-To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
+To systematically assess viral features associated with infection outcome, we divided the 150 cells with viral genotypes into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
+\FIG{mutations}A shows that the 49 cells infected by unmutated virions had a tighter distribution of the viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
@@
 But among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-\jdbcomment{Perhaps elaborate a bit after adding additional data relevant to this point to next section.}
+This finding is elaborated on in the validation experiments below.
 
@@
 %%% begin validation figure %%%
-\begin{figure}
+\begin{figure}[b!]
 
@@
 
-The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
+The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in IFN+ cells: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
 
 Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variants has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
 We created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
 
-We tested the rate of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild-type viral stock.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN more frequently than wild type.
+We tested the frequency of IFN induction by each viral stock using the reporter cells.
+\FIG{validation} shows that five of eight mutant viral stocks induced IFN more often than a wild-type viral stock.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN more than wild type.
 The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we expect some mutations to be found in IFN+ cells by chance.
-Overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often increase IFN induction.
 
 However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
-\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validation_supp}), which is itself a single-cell measurement, albeit one that does not report the genotype.
-These flow cytometry data reveal that no mutant virus stock induces IFN in more than a fraction of cells.
-Of course, these mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to that revealed by our single-cell sequencing of the ``wild-type'' viral stock.
-But IFN induction is stochastic at the single-cell level even for infections that share the same defect,such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
-Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction is stochastic even with these defects.
+\FIG{validation} shows flow cytometry data, which is itself a single-cell measurement, albeit one that does not report the genotype.
+These data reveal that no mutant virus stock induces IFN in more than a fraction of cells.
+Of course, the mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to those revealed by our single-cell sequencing of the ``wild-type'' viral stock.
+But our single-cell data show that IFN induction is stochastic even for infections with the same defect, such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
+Therefore, the experiments in \FIG{validation} not only validate some viral defects that increase IFN induction, but also that induction is stochastic even with these defects.
 
 %%% begin mutantcomparison figure %%%
-\begin{figure}
+\begin{figure}[b!]
 \centerline{\includegraphics[width=0.85\textwidth]{figures/MutantComparison/p_merged.pdf}}
@@
 
-\jdbcomment{Perhaps also add NS+ / NS- data here}
+The single-cell data also suggest heterogeneity in the process by which different viral variants induce IFN.
+Specifically, these data show that NS-deficient virions are much more likely to induce IFN when they transcribe more RNA, but that there is no such association for other virions (\FIG{mutations}D).
+To validate this conclusion, we used flow cytometry to compare IFN induction among cells that express low or high levels of the viral HA protein under the assumption that protein levels correlate with viral transcription.
+Consistent with the single-cell data, the NS1stop or NS1-A122V viruses are much more likely to induce IFN if they express high levels of HA (\FIG{mutantcomparison}A and \SUPPFIG{mutantcomparison_supp}A).
+In contrast, wild-type or PB1-mutant viruses are only slightly more likely to induce IFN if they express more HA. 
+To confirm that higher viral transcription or genome replication \emph{cause} more IFN induction by NS-deficient virions, we treated infected cells with chemicals (ribavirin or cycloheximide) that inhibit secondary viral transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg, killip2014activation}.
+Both inhibitors strongly reduced IFN induction by cells infected with NS1stop virus, but not cells infected by virus with a deletion in PB1 (\FIG{mutantcomparison}B and \SUPPFIG{mutantcomparison_supp}B).
+Therefore, different viral variants induce IFN by different processes.
 

### 2018-12-28 paper/paper.tex
made minor changes to results paragraph that I think clean up the concepts without overstating. Changed cycloheximide sentence a little in supplement to point out its issue with regards to IFN a bit more bluntly and added another citation

@@
 In contrast, wild-type or PB1-mutant viruses are only slightly more likely to induce IFN if they express more HA. 
-To confirm that higher viral transcription or genome replication \emph{cause} more IFN induction by NS-deficient virions, we treated infected cells with chemicals (ribavirin or cycloheximide) that inhibit secondary viral transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg, killip2014activation}.
-Both inhibitors strongly reduced IFN induction by cells infected with NS1stop virus, but not cells infected by virus with a deletion in PB1 (\FIG{mutantcomparison}B and \SUPPFIG{mutantcomparison_supp}B).
-Therefore, different viral variants induce IFN by different processes.
+To confirm that higher viral transcription or genome replication \emph{drive} IFN induction by NS-deficient virions, we treated infected cells with either ribavirin or cycloheximide, inhibiting secondary viral transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg, killip2014activation}.
+Both inhibitors strongly reduced IFN induction by cells infected with NS1stop virus, but not cells infected by virus with a deletion in PB1---a virus for which induction of interferon could not be dependent on secondary viral transcription or replication due to its fundamental inability to engage in these processes (\FIG{mutantcomparison}B and \SUPPFIG{mutantcomparison_supp}B).
+These data support a more complex model of interferon induction by influenza, wherein different viral variants trigger the production of IFN by different processes.
 
@@
 Preventing viral secondary transcription using cycloheximide also decreases IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
-However, the results are more complex here, since cycloheximide alone increases IFN production~\citep{raj1981analysis, maroteaux1983cycloheximide, killip2014activation}.
+However, the results are more complex here, since cycloheximide alone increases IFN production, even in response to non-viral challenges such as poly I:C ~\citep{raj1981analysis, maroteaux1983cycloheximide, ringold1984cycloheximide, killip2014activation}.
 Therefore, we see an increase in IFN levels upon cycloheximide treatment for uninfected cells and cells infected with the PB1 deletion virus, but a decrease for cells infected with the NS1stop virus.

### 2018-12-28 paper/paper.tex
final round of small-scale edits

--- 2018-12-27 paper/paper.tex
@@
 \fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
+\fancyfootoffset[L]{2in}
 \lfoot{\today}
@@
 Alistair B. Russell\textsuperscript{1},
-Jacob R. Kowalsky\textsuperscript{2},
+Jacob R. Kowalsky\textsuperscript{1},
 Jesse D. Bloom\textsuperscript{1,2,3*}
@@
 
-Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
+Unfortunately for the host, influenza only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
@@
 
-A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
-Studies using reporters~\citep{killip2017single,kallfass2013visualizing} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that at early timepoints IFN induction occurs in $<$1\% of influenza-infected cells.
-To efficiently identify and enrich these rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein ~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
-Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells~\citep{killip2017single,kallfass2013visualizing,russell2018extreme}.
+To identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein ~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
+Cells that activate the reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
+The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
 For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
-We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells~\citep{russell2018extreme}.
+We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that reported by prior studies that have examined IFN induction by influenza in single cells~\citep{killip2017single,russell2018extreme,kallfass2013visualizing}.
 
@@
 First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{plasmids}).
-These viral barcodes allow us to identify co-infections~\citep{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+These viral barcodes allow us to identify co-infections, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
+We performed PacBio sequencing on these viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
 Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
-Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).
+Combining this information with the 3'-end sequencing determines the entire transcriptome and full sequences of the infecting virions in single cells (\FIG{workflow}E).
 
@@
 To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\citep{haque2017practical}.
-After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B).
+This filtering left 1,490 human cells for further analysis (\FIG{transcriptomics}B).
 
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets prior to reverse-transcription.
-We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected by virus (\FIG{transcriptomics}C).
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets prior to reverse-transcription.
+We tested if each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected (\FIG{transcriptomics}C).
 We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
@@
 
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
+The amount of viral mRNA was lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
 However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus.
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{transcriptomics_supp}B-D).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
+Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B).
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
@@
 For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
-For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1. 
+For instance, \textit{cell 5} was co-infected such that it expressed both unmutated and internally deleted PB1. 
 }
@@
 But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
-For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
+For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that do not activate IFN despite being infected by virions that fail to express NS.
 
@@
 The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN more than wild type.
-The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we expect some mutations to be found in IFN+ cells by chance.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we expect some mutations in IFN+ cells by chance.
 
 However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
-\FIG{validation} shows flow cytometry data, which is itself a single-cell measurement, albeit one that does not report the genotype.
+\FIG{validation} shows flow cytometry data, which is itself a single-cell measurement, albeit one that does not report the viral genotype.
 These data reveal that no mutant virus stock induces IFN in more than a fraction of cells.
@@
 In contrast, wild-type or PB1-mutant viruses are only slightly more likely to induce IFN if they express more HA. 
-To confirm that higher viral transcription or genome replication \emph{cause} more IFN induction by NS-deficient virions, we treated infected cells with chemicals (ribavirin or cycloheximide) that inhibit secondary viral transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg, killip2014activation}.
+To confirm that higher viral transcription or genome replication drive IFN induction by NS-deficient virions, we treated infected cells with ribavirin or cycloheximide, which inhibit secondary viral transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg, killip2014activation}.
 Both inhibitors strongly reduced IFN induction by cells infected with NS1stop virus, but not cells infected by virus with a deletion in PB1 (\FIG{mutantcomparison}B and \SUPPFIG{mutantcomparison_supp}B).
-Therefore, different viral variants induce IFN by different processes.
+These data indicate that different viral variants trigger production of IFN by different processes.
 
@@
 Our method builds on the observation by that fragmentary viral genetic information can be obtained by Illumina-based single-cell transcriptomic techniques~\citep{saikia2018simultaneous, zanini2018virus}.
-To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \citep{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
+To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \citet{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
@@
 Preventing viral secondary transcription using cycloheximide also decreases IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
-However, the results are more complex here, since cycloheximide alone increases IFN production~\citep{raj1981analysis, maroteaux1983cycloheximide, killip2014activation}.
+However, the results are more complex here, since cycloheximide alone increases IFN production, even in response to non-viral challenges such as poly(I:C)~\citep{raj1981analysis, maroteaux1983cycloheximide, ringold1984cycloheximide, killip2014activation}.
 Therefore, we see an increase in IFN levels upon cycloheximide treatment for uninfected cells and cells infected with the PB1 deletion virus, but a decrease for cells infected with the NS1stop virus.

### 2019-03-04 paper/paper.tex
updated contact
updated sentence in abstract

--- 2018-12-27 paper/paper.tex
@@
+% Template for PLoS
+% Version 3.5 March 2018
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
 \documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.5in,footskip=0.75in]{geometry}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
@@
 
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
 
 % Use nameref to cite supporting information files (see Supporting Information section for more info)
-\definecolor{linkblue}{RGB}{0,76,153}
-\usepackage[colorlinks=true,linkcolor=linkblue,citecolor=linkblue,urlcolor=linkblue]{hyperref}
-\usepackage{nameref}
+\usepackage{nameref,hyperref}
 
@@
 \DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
 
@@
 \newenvironment{fullwidth}{%
-  \begin{adjustwidth}{-2.05in}{}
+  \begin{adjustwidth}{-2.3in}{}
 }{\end{adjustwidth}}
@@
 % Added by Jesse Bloom for figures
-\renewcommand{\floatpagefraction}{0.99}
-\renewcommand{\textfraction}{0.01}
+\renewcommand{\floatpagefraction}{.9}
 \setcounter{topnumber}{8}
@@
 \newfloat{suppfig}{thp}{lofsupfile}
-\floatname{suppfig}{Figure}
+\floatname{suppfig}{Fig}
 \renewcommand{\thesuppfig}{S\arabic{suppfig}}
 
-\renewcommand{\figureautorefname}{Figure}
+\renewcommand{\figureautorefname}{Fig}
 \newcommand{\FIG}[1]{\autoref{fig:#1}}
@@
 
-\newcommand{\suppfigautorefname}{Figure}
+\newcommand{\suppfigautorefname}{Fig}
 \newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}
@@
 % Text layout
-%\raggedright
+\raggedright
 \setlength{\parindent}{0.5cm}
-\textwidth 5.5in 
+\textwidth 5.25in 
 \textheight 8.75in
@@
 \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Figure}
-
-% Harvard-style citations
-\usepackage{natbib}
-\bibliographystyle{cell}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
 
@@
 
-% section header formats
-\usepackage{sectsty}
-\definecolor{darkred}{RGB}{140,0,0}
-\sectionfont{\normalsize \color{darkred}}
-\subsectionfont{\normalsize \color{darkred}}
-\subsubsectionfont{\normalsize \color{darkred}}
-\paragraphfont{\normalsize \color{darkred}}
-
 %% Include all macros below
 
-\providecommand{\keywords}[1]{\noindent \textbf{Keywords:} #1}
-
-\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
 
@@
 \begin{flushleft}
-{\LARGE
-\textbf{Single-cell virus sequencing of influenza infections that trigger innate immunity}}
+{\Large
+\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
 \newline
@@
 \\
-{\bf
 Alistair B. Russell\textsuperscript{1},
@@
 Jesse D. Bloom\textsuperscript{1,2,3*}
-}
 \\
 \bigskip
-{\small
-$^1$Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center
+\textbf{1} Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States
 \\
-$^2$Department of Genome Sciences, University of Washington
+\textbf{2} Department of Genome Sciences, University of Washington, Seattle, United States
 \\
-$^3$Howard Hughes Medical Institute
+\textbf{3} Howard Hughes Medical Institute, Seattle, United States
 \\
-Seattle, WA  98109, USA
-\\
-*Correspondence: jbloom@fredhutch.org
-}
 \bigskip
 
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* jbloom@fredhutch.org
+
 \end{flushleft}
-
-\section*{SUMMARY}
-{\bf
+% Please keep the abstract below 300 words
+\section*{Abstract}
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
-Here we develop a new approach to assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We do this by determining both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Most cells are infected by virions with defects such as amino-acid mutations, internal deletions, or failure to express a gene.
-We identify instances of each type of defect that increase the likelihood that a cell activates an innate-immune response.
-However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even when a cell is infected by a virion that expresses unmutated copies of all genes.
-Our work shows that viral genetic variation substantially contributes to but does not fully explain the heterogeneity in single influenza-infected cells.
-}
-
-\vspace{0.2in}
-\keywords{PacBio, single-cell RNA-seq, influenza, interferon, NS1, stochastic}
+Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
+We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
+However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
+Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
+
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+\section*{Author summary}
+Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
+But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
+Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.
+Most cells are infected by virions that have a mutation or some other defect.
+This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and helps determine whether a cell activates an innate-immune response.
+However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
+Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.
 
@@
 % Use "Eq" instead of "Equation" for equation citations.
-\section*{INTRODUCTION}
+\section*{Introduction}
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
-As the virus spreads, some cells detect infection and produce interferon (IFN).
-This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\citep{stetson2006type,honda2006type}.
-If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\citep{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\citep{la2007question, iwasaki2014innate}.
-
-Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\citep{kallfass2013visualizing, killip2017single}.
-This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
-Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+As the virus spreads, some cells detect infection and begin producing interferon (IFN).
+This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
+If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.
+
+Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
+Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
-For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
-But for influenza, a third possible contributor looms large: viral genetic diversity.
-Because influenza has a high mutation rate, individual virions often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, bloom2014experimentally, pauly2017novel}.
-Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
+But for influenza, a third possible contributor also looms large: viral genetic diversity.
+Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
-Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
-None of these techniques reliably reveal if there are mutations in the virion infecting any given single cell.
-
-Here we develop a new approach to determine both the transcriptome and full sequences of all viral genes in single influenza-infected cells.
-To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
+Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
+None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
+
+Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
+To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
 We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
-Two-thirds of cells are infected by virions with a mutation or defect in gene expression.
-This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
-We identify several types of viral defects that increase IFN induction.
-However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
+Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
+This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
+Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
+However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
 Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
 
-\section*{RESULTS}
-
-\subsection*{A system to identify and enrich rare IFN+ cells}
+\section*{Results}
+
+\subsection*{A system to identify and enrich for rare IFN+ cells}
 
@@
 \centerline{
-{\bf \Large A}\includegraphics[width=0.34\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
+{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
 \hspace{0.02\textwidth}
-{\bf \Large B} \includegraphics[width=0.22\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
+{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
 }
@@
 {\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
-We created reporters with type I and type III IFN promoters (\SUPPFILE{plasmids}).
+We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
 In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
 {\bf (B)}
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{reportervalidation}C for full data).
-The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
+The plot also shows uninfected cells, and cells infected with Sendai virus.
 The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
@@
 
-A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
-Studies using reporters~\citep{killip2017single,kallfass2013visualizing} and our own previous single-cell transcriptomics~\citep{russell2018extreme} have found that at early timepoints IFN induction occurs in $<$1\% of influenza-infected cells.
-To efficiently identify and enrich these rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein ~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}A).
+A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
+Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
+IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).
+
+To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
-For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
-
-We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells~\citep{russell2018extreme}.
+The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
+
+We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
 
@@
 \caption{
-{\bf Approach for combined transcriptomics and virus sequencing of single influenza-infected cells that express IFN.}
+{\bf Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.}
 {\bf (A)}
@@
 
-To determine if mutations or other defects in the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
-First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{plasmids}).
-These viral barcodes allow us to identify co-infections~\citep{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
+First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
+These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
+From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
 To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences~\citep[CCSs;][]{travers2010flexible}.
+We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
@@
 Number of cells for which transcriptomes were obtained.
-%From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
+From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
-\SUPPFIG{transcriptomics_supp}A shows the frequency that each viral gene is detected.
+\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
 {\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
 {\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
-From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
+From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
 {\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
@@
 A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
-Many cells that do not express IFN still express ISGs (\SUPPFIG{transcriptomics_supp}C,D).
+Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
 }
@@
 We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
-We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
-To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\citep{haque2017practical}.
-After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B).
+We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
+To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.
 
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets prior to reverse-transcription.
-We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected by virus (\FIG{transcriptomics}C).
-We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
-The distribution of the amount of viral mRNA across infected cells is in the inset in \FIG{transcriptomics}C.
-As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
-
-We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
-%We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that a slight majority (162 of 290) of infected cells express all eight genes (see \SUPPFIG{transcriptomics_supp}A for frequencies for individual genes).
-This measured frequency of infected cells expressing all eight genes is slightly higher than in our prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which estimated that only 13\% to 50\% of infected cells express all genes.
-
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
-However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus.
-Consistent with prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and similar to values from older bulk studies~\citep{hatada1989control}.
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
+We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
+The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
+As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
+
+We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
+We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
+\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
+This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
+
+The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
+However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
+Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
 
 By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
-This co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
+From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
+Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
 This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
@@
 
-We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{transcriptomics_supp}B-D).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
+We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
+Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
-Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{transcriptomics_supp}C).
-The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{transcriptomics_supp}D).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\citep{stetson2006type,honda2006type}.
+Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{ISG}A).
+The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\cite{stetson2006type,honda2006type}.
+
+Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
+\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\cite{maaten2008visualizing} of the cells.
+Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.
 
@@
 {\bf Viral genotypes and infection outcomes in single cells.}
-Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
-The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
-Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
-Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
-The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
+Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
+Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
 For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively. 
-For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1. 
+Green boxes at left show the percent of all mRNA in that cell derived from virus.
+Orange boxes show the percent of cellular mRNA derived from IFN, with orange box frames indicating cells classified as IFN+ in \FIG{transcriptomics}G.
 }
@@
 We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
-Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{PacBio_supp}A).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{PacBio_supp}B), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
-
-After calling the presence / absence of each viral gene in each cell using the transcriptomic data as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
-For cells co-infected with both viral barcode variants, we called mutations separately for each variant.
-This strategy identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
+Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+
+We called a mutation in a viral gene if it was found in least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
+For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
+This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
 It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
-Therefore, analogous to somatic variant calling in tumor sequencing~\citep{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.
-
-We called the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
-We were most effective at calling viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
-But we also called genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
-
-The 150 cells for which we called the viral genotypes are shown in \FIG{genotypes}.
-Inspection of this figure reveals a wealth of anecdotal relationships between viral genotype and infection outcome.
-For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all genes and did not induce detectable IFN.
-But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
-As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
-
-The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist, NS1~\citep{garcia1998influenza, hale2008multifunctional}.
-Many other IFN+ cells also lacked NS or had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
-
-However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
-This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
-As an example of the former case, expressing unmutated copies of all eight genes does not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only expressed viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
-But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
-For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
+Therefore, analogous to somatic variant calling in tumor sequencing~\cite{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.
+
+We could call the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
+But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
+
+The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
+Simple visual inspection of this figure reveals a wealth of information.
+For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
+But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the wild-type virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the wild-type virion that infected \textit{cell 105} still induced IFN.
 
@@
 %%% begin mutations figure %%%
-\begin{figure}[b!]
+\begin{figure}
 \begin{fullwidth}
@@
 \caption{
-{\bf Viral features associated with heterogeneity among cells for which we determined viral genotypes.}
+{\bf Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.}
 {\bf (A)} 
 Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
-Cells infected by unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals indicated).
+Cells infected by fully wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
 {\bf (B)}
-IFN expression among cells expressing unmutated copies of all genes, and cells with mutations or missing genes.
+IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
 {\bf (C)}
@@
 The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
-The middle and bottom panels show the percent of IFN- and IFN+ cells with a deletion or amino-acid substitution in each gene, conditioned on the gene being expressed.
-Numbers give P-values (Fisher's exact test) for rejecting null hypothesis that percents are equal among IFN- and IFN+ cells. 
+The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
+Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
 {\bf (D)}
-There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN among cells that lack NS.
-This figure only shows non-synonymous substitutions, and disregards insertions as they are very rare.
+There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
+Throughout this figure, we only consider substitutions that are non-synonymous.
+Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
 }
@@
 
-To systematically assess viral features associated with infection outcome, we divided the 150 cells with viral genotypes into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by unmutated virions had a tighter distribution of the viral mRNA per cell than the other 101 cells as quantified by the Gini index~\citep{gini1921measurement}.
+To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
+\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
@@
 Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.
-However, the validation experiments in the next section show that the lack of statistical significance for many of the trends is due to the modest number of infected cells that were sequenced rather than a lack of true association between viral defects and IFN induction.
 
 One other interesting trend emerges from the single-cell data.
-There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS (\FIG{mutations}D).
-But among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-This finding is elaborated on in the validation experiments below.
-
-\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction}
+There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
+
+\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}
 
 %%% begin validation figure %%%
-\begin{figure}[b!]
+\begin{figure}
 
@@
 \caption{
-{\bf Validation that IFN induction is increased by mutations identified in the single-cell virus sequencing of IFN+ cells.}
+{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
 {\bf (A)}
-Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated mutant, as determined using a reporter cell line.
-Numbers give the median of four measurements for each mutant.
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
+The numbers indicate the median of four measurements for each viral mutant.
 The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
-Points are orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
+Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
 {\bf (B)}
 Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
-See \SUPPFIG{validation_supp} for more detailed experimental data.
-The experiments in the two panels were performed on different days, and so numerical values can be compared within but not between panels.
+See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
+The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
 }
@@
 
-To test if the viral defects identified in single IFN+ cells directly increase IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
+To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
 
 The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
-Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\citep{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
-%The NS gene encodes influenza's primary innate-immune antagonist, NS1~\citep{garcia1998influenza, hale2008multifunctional}.
-We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.
-
-The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in IFN+ cells: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
-
-Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
-We created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
-
-We tested the frequency of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of eight mutant viral stocks induced IFN more often than a wild-type viral stock.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN more than wild type.
-The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we expect some mutations to be found in IFN+ cells by chance.
-
-However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
-\FIG{validation} shows flow cytometry data, which is itself a single-cell measurement, albeit one that does not report the genotype.
-These data reveal that no mutant virus stock induces IFN in more than a fraction of cells.
-Of course, the mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to those revealed by our single-cell sequencing of the ``wild-type'' viral stock.
-But our single-cell data show that IFN induction is stochastic even for infections with the same defect, such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
-Therefore, the experiments in \FIG{validation} not only validate some viral defects that increase IFN induction, but also that induction is stochastic even with these defects.
-
-%%% begin mutantcomparison figure %%%
-\begin{figure}[b!]
-\centerline{\includegraphics[width=0.85\textwidth]{figures/MutantComparison/p_merged.pdf}}
-\caption{
-Higher viral gene expression is strongly associated with more IFN induction only in NS1-deficient infections.
-{\bf (A)}
-For viruses with defects in NS1, there is more IFN induction among infected cells that express higher levels of HA protein.
-The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile as determined by flow cytometry.
-Points indicate replicates, and lines indicate the mean.
-See \SUPPFIG{mutantcomparison_supp}A for details.
-{\bf (B)}
-Blocking viral secondary transcription strongly decreases IFN in cells infected with NS1stop virus, but not cells infected with virus carrying a deletion in PB1.
-A549 cells were infected at doses that induced equivalent IFN, and ribavirin was added at a concentration (100 $\mu$M) sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg}.
-After eight hours, IFNB1 transcript was quantified relative to the housekeeping gene L32 by qPCR. 
-The dashed gray line indicates the limit of detection.
-\SUPPFIG{mutantcomparison_supp}B shows that results are similar if secondary transcription is instead blocked with cycloheximide.
-}
-\label{fig:mutantcomparison}
-\end{figure}
-%%% end mutantcomparison figure%%%
-
-The single-cell data also suggest heterogeneity in the process by which different viral variants induce IFN.
-Specifically, these data show that NS-deficient virions are much more likely to induce IFN when they transcribe more RNA, but that there is no such association for other virions (\FIG{mutations}D).
-To validate this conclusion, we used flow cytometry to compare IFN induction among cells that express low or high levels of the viral HA protein under the assumption that protein levels correlate with viral transcription.
-Consistent with the single-cell data, the NS1stop or NS1-A122V viruses are much more likely to induce IFN if they express high levels of HA (\FIG{mutantcomparison}A and \SUPPFIG{mutantcomparison_supp}A).
-In contrast, wild-type or PB1-mutant viruses are only slightly more likely to induce IFN if they express more HA. 
-To confirm that higher viral transcription or genome replication \emph{cause} more IFN induction by NS-deficient virions, we treated infected cells with chemicals (ribavirin or cycloheximide) that inhibit secondary viral transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg, killip2014activation}.
-Both inhibitors strongly reduced IFN induction by cells infected with NS1stop virus, but not cells infected by virus with a deletion in PB1 (\FIG{mutantcomparison}B and \SUPPFIG{mutantcomparison_supp}B).
-Therefore, different viral variants induce IFN by different processes.
-
-\section*{DISCUSSION}
+Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\cite{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
+The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
+We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
+
+The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of the amino-acid substitutions found in these genes among IFN+ cells.
+Overall, we created six such point-mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
+
+Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
+
+We tested the rate of IFN induction by each viral stock using the reporter cells.
+\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
+But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
+
+\section*{Discussion}
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observation by that fragmentary viral genetic information can be obtained by Illumina-based single-cell transcriptomic techniques~\citep{saikia2018simultaneous, zanini2018virus}.
-To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \citep{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
-
-This viral genetic information is crucial for understanding infection outcome and innate-immune induction.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
+Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
+To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
+
+This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
 Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wild-type'' influenza genome, most infected cells do not express wild-type copies of all viral genes.
-Although our study is certainly not the first to note that influenza has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of defects across single cells.
-Inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
-
-These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including the probability that a cell triggers an IFN response.
-We identified four types of defects in single IFN+ cells that we validated increase IFN induction. 
-Two types of defects---absence of NS and mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
-The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\citep{velthuis2018mini}.
-Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Given the extensive prior work describing such deletions, it is surprising we did not identify more of them in our single IFN+ cells.
-There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-The relative importance of different defects also likely varies across conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
-
-But although variation in viral populations is clearly important, it only partially explains heterogeneity among influenza-infected cells.
-We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
-Additionally, no viral defect induces IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at high frequency.
-However, we suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\citep{oneal2018west}.
+Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
+Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
+
+It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
+We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
+Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
+The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini} .
+Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
+There may be several reasons: we used relatively pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
+The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
+
+But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
+We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
+Additionally, none of the viral defects in IFN+ cells induce IFN deterministically in our validation experiments---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
+Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
+However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
 
 Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
-Natural human influenza infections are initiated by just a handful of virions~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\citep{stetson2006type,honda2006type}.
-It is therefore plausible that early heterogeneity could have a large effect on downstream events.
-Extending our approaches to more complex systems could shed further light on how viral variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
-
-\small
-\subsection*{ACKNOWLEDGMENTS}
-We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
-We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
-This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation (DRG-2227-15).
-JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
-JDB is an Investigator of the Howard Hughes Medical Institute.
-
-\subsection*{AUTHOR CONTRIBUTIONS}
-ABR and JDB designed the study.
-ABR performed the experiments with assistance from JRK.
-ABR and JDB computationally analyzed the data and wrote the paper.
-
-\subsection*{DECLARATION OF INTERESTS}
-The authors declare no competing interests.
-
-\bibliography{references}
-
-\section*{METHODS}
+Natural human influenza infections are established by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
+It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
+Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
+
+\section*{Methods and Materials}
 
@@
 
-To create the type I interferon reporters, a 1kb promoter region upstream of the human IFNB1 gene were cloned into the pHAGE2 lentiviral vector~\citep{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
-Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\citep{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\citep{kim2011high}.
-The sequence of the last of these constructs is provided in \SUPPFILE{plasmids}.
+To create the type I interferon reporters, a 1kb promoter region upstream of the human IFNB1 gene were cloned into the pHAGE2 lentiviral vector~\cite{oconnell2010lentiviral}, with a NotI site immediately downstream of the promoter serving as an artificial Kozak sequence.
+Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\cite{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\cite{kim2011high}.
+The sequence of the last of these constructs is provided in \SUPPFILE{reporter_sequences}.
 
@@
 Downstream of this promoter we cloned LNGFR$\Delta$C linked to ZsGreen via a P2A linker.
-The sequence of this construct is provided in \SUPPFILE{plasmids}.
+The sequence of this construct is provided in \SUPPFILE{reporter_sequences}.
 
@@
 We used both the wild-type virus and a variant of the virus where synonymous mutations were added within a few 100 nucleotides of each termini of each gene segment.
-We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\citep{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
-In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{PacBio_supp}B).
-The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{plasmids}.
-These genes were cloned into the pHW2000~\citep{hoffmann2000dna} reverse-genetics plasmid.
-
-Both viral strains were generated by reverse genetics using the pHW18* series of bi-directional plasmids~\citep{hoffmann2000dna}.
-We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\citep{xue2016propagation}.
-The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 cells in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \citep{russell2018extreme}.
+We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\cite{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
+The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
+These genes were cloned into the pHW2000~\cite{hoffmann2000dna} reverse-genetics plasmid.
+
+Both viral strains were generated by reverse genetics using the pHW18* series of bi-directional plasmids~\cite{hoffmann2000dna}.
+We controlled the durations and MOI during viral passaging since these factors can greatly affect the accumulation of defective viral particles~\cite{xue2016propagation}.
+The viruses were generated by reverse genetics in co-cultures of 293T and MDCK-SIAT1 cells in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100 $\mu$g of streptomycin/ml, and 100 $\mu$g of calcium chloride/ml) and then propagated in MDCK-SIAT1 cells in influenza growth media using the same basic procedures detailed in \cite{russell2018extreme}.
 Specifically, after generation by reverse genetics, the wild-type variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
-The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\citep{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
+The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\cite{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
 
@@
 The sequences of these primers are in \SUPPFILE{primer_sequences}.
-A major concern during these PCRs is strand exchange (see \SUPPFIG{PacBio_supp}B) which would scramble the cell barcodes and mutations on viral cDNAs.
+A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
 To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
 Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
-This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\citep{boers2015micelle}.
+This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{Boers:2015emulsion}.
 We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .
@@
 The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
-The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{PacBio_supp}A, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{PacBio_supp}B for an illustration of how this is determined).
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{StrandExchange} for an illustration of how this is determined).
 A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
@@
 
-For the qPCR in \SUPPFIG{validation_supp}C, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+For the qPCR in \SUPPFIG{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
-For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml (a concentration sufficient to block secondary transcription~\citep{killip2014activation}) at the time of infection.
+For the cells with cycloheximide added to block protein expression (and hence secondary transcription), cycloheximide was added to a final concentration of 50 $\mu$g/ml (a concentration sufficient to block secondary transcription~\cite{killip2014activation}) at the time of infection.
 After 8 hours, mRNA was harvested using the CellAmp Direct RNA Prep Kit for RT-PCR, reverse-transcribed using an oligo-dT primer, and qPCR was performed as described above.
-
-For the qPCR in \FIG{mutantcomparison}B, A549 cells were infected with the NS1stop virus at a MOI of 0.1 (based on TCID50) in MDCK-SIAT1 and with the PB1del385to2163 virus at a dose that induced equivalent IFN in the absence of ribavirin.
-The ribavirin was added to cells at the time of infection to a final concentration of 100$\mu$M, which is sufficient to block secondary transcription~\citep{Vanderlinden:2016ec,reuther2015generation,Scholtissek:1976wg}.
-All other steps were performed as for the cycloheximide experiments described immediately above..
 
@@
 This NS1stop virus contained six nucleotide changes resulting in the addition of five in-frame stop codons in NS1 starting 10 nucleotides downstream of the 5' splice donor site, thereby disrupting NS1 while leaving NS2 (NEP) intact.
-All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\citep{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
- \SUPPFILE{plasmids} provides the full sequences for all of these plasmids.
-
-We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\citep{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{plasmids} for the wild-type plasmid for that gene.
+All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\cite{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
+ \SUPPFILE{validation_sequences} provides the full sequences for all of these plasmids.
+
+We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\cite{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{validation_sequences} for the wild-type plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
 At 24 hours post-transfection, we changed media from D10 to influenza growth media. 
-At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{validation_supp}A) or 72 hours (for the replicate 2 viruses in \SUPPFIG{validation_supp}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
+At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{validationflow}A) or 72 hours (for the replicate 2 viruses in \SUPPFIG{validationflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \SUPPFIG{validation_supp}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\citep{hirst1942quantitative}.
+For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\cite{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
 Two-fold serial dilutions of virus were added to an equal volume of diluted red blood cells, and titer was measured as the highest dilution of viral stock at which complete hemagglutination of red blood cells was observed.
-We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{validation_supp}A.
-
-To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validation_supp}B, we used slightly different procedures.
-The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\citep{russell2018extreme}.
+We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{validationflow}A.
+
+To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validationflow}B, we used slightly different procedures.
+The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\cite{russell2018extreme}.
 The NS1stop virus was similarly generated, but was passaged for 48 rather than 36 hours, since it had slower growth kinetics and so needed a longer period of time to reach high titers.
 The viruses with deletions in the PB1 segment could not be generated in normal 293T and MDCK-SIAT1 cells, since they required the exogenous expression of the PB1 protein.
-Therefore, these viruses were generated in previously described 293T and MDCK-SIAT1 cells that had been engineered to constitutively express PB1~\citep{bloom2010permissive}.
+Therefore, these viruses were generated in previously described 293T and MDCK-SIAT1 cells that had been engineered to constitutively express PB1~\cite{bloom2010permissive}.
 These viruses were harvested from transfections at 72 hours, and passaged twice in the MDCK-SIAT1 cells constitutively expressing PB1 at a MOI of 0.001 for 72 hours and 0.01 for 48 hours. 
@@
 The wild-type and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
-The infections in \SUPPFIG{validation_supp}B were performed at equivalent TCID50s as described in the legend to that figure.
-That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \SUPPFIG{validation_supp}C.
+The infections in \SUPPFIG{validationflow}B were performed at equivalent TCID50s as described in the legend to that figure.
+That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \SUPPFIG{delqPCR}.
 
@@
 The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
-The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \SUPPFILE{plasmids} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
+The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \SUPPFILE{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
 The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
@@
 First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
-We further processed these CCSs using custom Python code and the \texttt{minimap2}~\citep{li2018minimap2} long-read aligner (version 2.11-r797).
-The Python code has been implemented in the API of \texttt{dms\_tools2} (\url{https://jbloomlab.github.io/dms_tools2/}~\citep{bloom2015software}) package (version 2.3.0).
+We further processed these CCSs using custom Python code and the \texttt{minimap2}~\cite{li2018minimap2} long-read aligner (version 2.11-r797).
+The Python code has been implemented in the API of \texttt{dms\_tools2} (\url{https://jbloomlab.github.io/dms_tools2/}~\cite{bloom2015software}) package (version 2.3.0).
 A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{pacbio_analysis}.
@@
 We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
-At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange, and reassuringly found that strand exchange was rare (\SUPPFIG{PacBio_supp}).
+At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{StrandExchange}), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}).
 The small number of CCSs with identifiable strand exchange were filtered from further analysis.
@@
 We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
-We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{transcriptomics_supp}A).
+We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{frac_has_gene}).
 To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
-We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{transcriptomics_supp}C,D, respectively.
-We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts, since type I and type III IFN were highly correlated (\SUPPFIG{transcriptomics_supp}B).
+We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{ISG}, respectively.
+We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
 We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
@@
 
+\section*{Acknowledgments}
+We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
+We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
+%This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
+%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+%JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
+%JDB is an Investigator of the Howard Hughes Medical Institute.
+%The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
+
+\bibliography{references}
+
 \clearpage
-\nolinenumbers
-\normalsize
-
-\section*{SUPPORTING INFORMATION}
-
-\begin{suppfig}
-\begin{minipage}[t]{0.46\textwidth}
+\section*{Supporting information}
+
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}
+\caption{
+{\bf Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
+Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
+Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
+As shown in the left-most panel, the sequencing depth of \cite{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
+An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
+Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\cite{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
+However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\cite{stetson2006type,honda2006type}.
+The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
+}
+\label{suppfig:mice}
+\end{suppfig}
+
+\begin{suppfig}
+\begin{minipage}[t]{0.52\textwidth}
 {\bf \Large A}
+
 \includegraphics[width=0.9\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}
 \end{minipage}
-\begin{minipage}[t]{0.55\textwidth}
+\begin{minipage}[t]{0.48\textwidth}
 {\bf \Large B}
-\includegraphics[width=0.57\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
-\vspace{0.1in}
-
-{\bf \Large C}
-\includegraphics[width=0.9\textwidth, valign=t]{figures/IFN_stochastic/Flow/flow_plot.pdf}
+
+\includegraphics[width=0.7\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
 
@@
 \caption{
-{\bf Validation of IFN reporter, related to \FIG{IFNrare}.}
+{\bf Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.}
 {\bf (A)}
-IFN reporter cell lines were infected at high MOI with the Cantell strain of Sendai virus~\citep{strahle2006sendai}.
-The name of the reporter is indicated above the plots.
-At 13 hours post-infection, activation of the reporter was monitored by flow cytometry using the marker indicated below each plot (a fluorescent protein or antibody staining for the LNGFR$\Delta$C using PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
-Sendai infection efficiently activated the IFN reporter in all cases.
+To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\cite{strahle2006sendai}.
+The name of each of reporter cell line is indicated at the top of each row of plots.
+At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
+Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
 {\bf (B)}
 The type I and type III IFN reporters are highly correlated in their activation.
-An A549 cell line was generated by transduction with both IFN-$\beta$ and IFN-$\lambda$ reporters driving mCherry and ZsGreen, respectively.
-The cells were infected with two different stocks of WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1.
+An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
+The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
 After 13 hours, cells were analyzed by flow cytometry.
-Cells positive for either reporter were further analyzed.
-As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated.
-{\bf (C)}
-Flow cytometry data for \FIG{IFNrare}B.
-A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus, or infected at a target MOI of 0.3 with the stock of influenza virus used in the single-cell experiment.
-At 13 hours, cells were analyzed by FACS for HA and ZsGreen expression.
+Cells positive for either fluorescent reporter were further analyzed.
+As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.
+}
+\label{suppfig:reportervalidation}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.9\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
+\caption{
+{\bf Flow cytometry data for \FIG{IFNrare}B.}
+The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\cite{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
+After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
 Each condition was done in triplicate.
-Contour plots show density of all cells, and IFN+ cells are also indicated by orange dots.
+The contour plots show the density of all cells, and all IFN+ cells are also indicated by orange dots.
 Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
-For influenza-infected cells, the percentage IFN+ was calculated only among HA+ cells (since these are the ones that are infected).
-For uninfected and Sendai-infected cells, the percentage IFN+ was calculated among all cells, since the cells do not express HA.
-}
-\label{suppfig:reportervalidation}
+For the influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
+For the uninfected and Sendai-virus infected, the percentage IFN+ was calculated among all cells, since these cells do not express HA.
+}
+\label{suppfig:IFNrare_flow}
 \end{suppfig}
@@
 \caption{
-{\bf Example MACS enrichments of IFN+ influenza-infected cells, related to \FIG{workflow}.}
+{\bf Example MACS enrichments of IFN+ influenza-infected cells.}
 A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
@@
 \begin{suppfig}
-\includegraphics[width=0.9\textwidth]{figures/single_cell_figures/p_cell_summary_supp.pdf}
-\caption{
-{\bf Additional plots from analysis of transcriptomic data, related to \FIG{transcriptomics}.}
-{\bf (A)}
-Fraction of infected cells that are called as expressing each viral gene.
-The gray dashed line is at one.
-Each gene is detected in $\sim$80-90\% of cells, roughly in line with prior estimates~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}
+\caption{
+{\bf The fraction of infected cells that are called as expressing each viral gene.}
+The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
+Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
 The exception is NP, which is detected in virtually all infected cells.
-The higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have lower viral gene expression and so are not reliably called as infected in our experiments because the number of viral mRNAs is below the detection limit.
-{\bf (B)}
-Correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
+The much higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments because the number of viral mRNAs is below the detection limit.}
+\label{suppfig:frac_has_gene}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}
+\caption{
+{\bf The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.}
 Each point represents one cell.
 The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
-{\bf (C)}
-Distribution of ISGs expression across single infected and uninfected cells. 
+Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
+}
+\label{suppfig:type_I_III_correlation}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}
+\caption{
+{\bf Expression of ISGs in single infected and uninfected cells.}
 For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
-We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from these ISGs, and color these cells red.
-Comparison to \FIG{transcriptomics}G shows that more cells are ISG+ than IFN+, probably because paracrine signaling can induce ISG expression in cells that do not themselves express IFN~\citep{stetson2006type,honda2006type}.
-{\bf (D)} 
-Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
-Each point is one cell, and the Pearson correlation coefficient is shown.
+{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
+We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
+Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
+This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\cite{stetson2006type,honda2006type}.
+{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
+Each point represents one cell, and the Pearson correlation coefficient is shown.
 IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
-Among both the infected and uninfected populations, there are many cells that express ISGs but not IFN, but no cells that express IFN but not ISGs.
-}
-\label{suppfig:transcriptomics_supp}
-\end{suppfig}
-
-\begin{suppfig}
-{\bf \Large A}
-\includegraphics[width=0.9\textwidth, valign=t]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
-\vspace{0.05in}
-
-{\bf \Large B}
-\includegraphics[width=0.52\textwidth, valign=t]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
-\caption{
-{\bf PacBio CCS quality control, related to \FIG{genotypes}.}
-{\bf (A)}
-Number of PacBio CCSs that passed quality-control and aligned to each viral gene.
-CCSs were obtained using several PacBio runs which were loaded with different amounts of each gene to balance coverage (see \SUPPFILE{pacbio_analysis}).
-Therefore, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs is \emph{not} an indicator of a transcript's abundance.
-Especially for the polymerase genes, many CCSs corresponded to genes with internal deletions, since these shorter forms are preferentially amplified by PCR.
-Therefore, the plot is faceted by CCSs for any length of the gene, and for full-length genes.
-The disproportionate sequencing of the shorter internally deleted genes does not affect the genotype calling since UMIs were used to collapse sequences from the same cDNA, and cell barcodes were used to collapse sequences from the same cell.
-The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or is a mixed-barcode molecule (see panel B).
-We estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
-About half of these chimeras can identified by the mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
-For some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the viral barcode could only be partially called.
-{\bf (B)}
-Strategy for detecting strand exchange in PacBio sequencing.
-Preparation for PacBio sequencing required many PCR cycles, which can cause strand exchange that scrambles mutations and 10X cell barcodes / UMIs from different molecules.
-If there is no strand exchange, all molecules are either wild-type or have the synonymous barcoding mutations at \emph{both} termini.
-Strand exchange creates molecules with wild-type nucleotides at one termini and synonymous barcoding mutations at the other.
-Panel A shows the frequencies of these molecules.
-Since homologous recombination in influenza virus in negligible~\citep{boni2008homologous}, mixed-barcode molecules are \emph{not} generated naturally during co-infection.
-}
-\label{suppfig:PacBio_supp}
+Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
+}
+\label{suppfig:ISG}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/single_cell_figures/p_tsne.pdf}
+\caption{
+{\bf Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.}
+To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\cite{maaten2008visualizing} as implemented in \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
+In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
+Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA, shown on a log (top) or linear (bottom) scale.
+As is clear from this plot, expression of influenza, IFN, and ISG genes contributes substantially to the structure of the data, since cells with high expression of these genes clearly group together.
+}
+\label{suppfig:tSNE}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
+\caption{
+{\bf The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.}
+These sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells (see \SUPPFILE{pacbio_analysis}).
+Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
+Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
+Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
+Note that the disproportionate sequencing of the shorter internally deleted genes does not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse sequences derived from the same cDNA, and cell barcodes were used to collapse sequences from the same cell.
+The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see panel B).
+From the frequencies of these different forms, we estimate~\cite{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+About half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
+For some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
+A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
+}
+\label{suppfig:CCSs}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.8\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
+\caption{
+{\bf Strategy for detecting strand exchange during sequencing of full-length viral genes.}
+The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
+A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
+We detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
+If there is no strand exchange, all molecules should either be wild-type or have the synonymous barcoding mutations at \emph{both} termini.
+Strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
+\SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
+Note that since the rate of homologous recombination in influenza virus in negligible~\cite{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
+}
+\label{suppfig:StrandExchange}
 \end{suppfig}
@@
 \begin{suppfig}
-\begin{minipage}[t]{0.46\textwidth}
-{\bf \Large A}
-\includegraphics[width=0.9\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
-\end{minipage}
+{\bf \Large A}\includegraphics[width=0.48\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
 \hspace{0.02\textwidth}
-\begin{minipage}[t]{0.5\textwidth}
-{\bf \Large B}
-\includegraphics[width=0.75\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
-\vspace{0.1in}
-
-{\bf \Large C}
-\includegraphics[width=0.75\textwidth, valign=t]{figures/Validation_Figure/qpcr_plot.pdf}
-
-\end{minipage}
-\caption{
-{\bf Additional data for validation experiments, related to \FIG{validation}.}
+{\bf \Large B} \includegraphics[width=0.39\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
+\caption{
+{\bf Flow cytometry data for \FIG{validation}.}
 {\bf (A)}
-Flow cytometry data for \FIG{validation}A.
-Data were collected and analyzed as in \SUPPFIG{reportervalidation}C.
-For each viral mutant, two independent stocks were assayed in duplicate (\#1a and \#1b are one viral stock; \#2a and \#2b are the other).
-Infections with replicate \#1 of wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and other viruses were infected at an equivalent particle number as determined by HI assay. 
+Data for \FIG{validation}A.
+A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with stocks of the indicated mutant.
+After 13 hours, cells were stained for HA protein and analyzed by FACS.
+Contour plots show density of all cells, and IFN+ cells are also indicated by orange dots.
+Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of uninfected cells in these populations.
+For infected cells, the percentage IFN+ was calculated among the HA+ cells (since these are the ones that are infected).
+For uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
+For each viral mutant, two independent stocks were assayed in duplicate (i.e., \#1a and \#1b are one viral stock, and \#2a and \#2b are the other).
+The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
 {\bf (B)}
-Flow cytometry data for \FIG{validation}B.
+Data for \FIG{validation}B.
 The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses less HA due to the lack of secondary transcription.
-Therefore, cells were infected at a MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\citep{bloom2010permissive} for PB1del385to2163.
-Panel C shows that at these TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
+Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\cite{bloom2010permissive} for PB1del385to2163.
+\SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
 The percent IFN+ was calculated for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for PB1del385to2163.
-{\bf (C)}
-Validation that infections in panel B were performed at similar doses of virions capable of initiating primary transcription.
-A549 cells were infected at MOI of 0.4 (based on TCID50 as described panel B), and after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse-transcription products.
+}
+\label{suppfig:validationflow}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.6\textwidth]{figures/Validation_Figure/qpcr_plot.pdf}
+\caption{
+{\bf Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.}
+In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{validationflow}B), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-Infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription~\citep{killip2014activation}.
-In the absence of cycloheximide, viruses with deletions in PB1 produced less viral mRNA because they could not produce PB1 protein for secondary transcription.
+These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\cite{killip2014activation}.
+In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
 But in the presence of cycloheximide, all viruses produced similar amounts of viral mRNA, indicating that the dose of particles active for primary transcription is roughly equivalent across variants.
@@
 }
-\label{suppfig:validation_supp}
-\end{suppfig}
-
-\begin{suppfig}
-\begin{minipage}[t]{0.55\textwidth}
-{\bf A}
-\includegraphics[width=0.9\textwidth, valign=t]{figures/MutantComparison/p_ifn_vs_ha.pdf}
-\end{minipage}
-\hspace{0.02\textwidth}
-\begin{minipage}[t]{0.43\textwidth}
-{\bf B}
-\includegraphics[width=0.9\textwidth, valign=t]{figures/MutantComparison/cyclo_qPCR_plot.pdf}
-\end{minipage}
-
-\caption{
-{\bf Additional data for comparison of IFN induction as a function of viral load, related to \FIG{mutantcomparison}.}
-{\bf (A)}
-Detailed analysis of the relationship among HA expression and IFN induction for different viral mutants that is summarized in \FIG{mutantcomparison}A.
-We re-analyzed all the infected cells in \SUPPFIG{validation_supp}A (defining ``infected'' using the same HA-gating scheme in \SUPPFIG{validation_supp}A) for the wild-type virus, the NS1stop virus, and the three amino-acid mutants with significantly increased IFN induction.
-For each virus and replicate, we binned the infected cells into HA expression quartiles.
-We then calculated the percent of cells that were IFN+ in each quartile (defining ``IFN+'' using the same gating scheme in \SUPPFIG{validation_supp}A).
-The plots show the mean HA expression in each quartile versus the percent of cells that are IFN+.
-The results clearly indicate that for the NS1stop virus, increased viral burden (more HA) correlates with IFN induction.
-A similar trend also holds for the virus with an amino-acid substitution in NS1 (NS1-A122V). 
-But for wild type and the two PB1 mutants, there is no clear trend for higher viral burden to correlate with IFN induction.
-\FIG{mutantcomparison}A summarizes the data in this figure by simply showing the ratio of the percent IFN+ of the highest HA-expression quartile to the lowest HA-expression quartile for each sample and replicate.
-{\bf (B)}
-Similar to \FIG{mutantcomparison}B, but uses cycloheximide rather than ribavirin.
-Preventing viral secondary transcription using cycloheximide also decreases IFN induction by the NS1stop virus relative to the PB1 deletion virus, corroborating the result in \FIG{mutantcomparison}B with a different inhibitor.
-However, the results are more complex here, since cycloheximide alone increases IFN production~\citep{raj1981analysis, maroteaux1983cycloheximide, killip2014activation}.
-Therefore, we see an increase in IFN levels upon cycloheximide treatment for uninfected cells and cells infected with the PB1 deletion virus, but a decrease for cells infected with the NS1stop virus.
-This plot shows qPCR results for the same samples as in \SUPPFIG{validation_supp}C, but with the qPCR for expression of IFNB1 transcripts relative to the housekeeping gene L32.
-As described in \SUPPFIG{validation_supp}C, A549 cells were infected at an MOI of 0.4, such that HA expression was equivalent between viral variants when secondary transcription was blocked, and some of the samples treated with 50 $\mu$g/ml of cycloheximide~\citep[a concentration sufficient to block secondary transcription;][]{killip2014activation}.
-After 8 hours, transcript levels were quantified by qPCR.
-Points indicate four replicates, and lines indicate the mean.
-The dashed line shows the limit of detection, and points below this limit where set to that value.
-}
-\label{suppfig:mutantcomparison_supp}
+\label{suppfig:delqPCR}
 \end{suppfig}
@@
 \begin{suppfile}
-\caption{{\bf Sequences of plasmids and viruses used in this study.}
-The plasmid maps are for the IFN reporters in \FIG{IFNrare}A and the mutant viral genes cloned into the pHW* bi-directional reverse genetics plasmid~\citep{hoffmann2000dna}.
-In addition, there are Genbank files giving the sequences of the wild-type and synonymously barcoded WSN viruses.
-A README file explains the contents of the ZIP.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}, \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}, \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb}, and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}
-}
-\label{suppfile:plasmids}
+\caption{{\bf Sequences of the IFN reporters in \FIG{IFNrare}A.}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
+}
+\label{suppfile:reporter_sequences}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+{\bf Genbank files giving sequences of the wild-type and synonymously barcoded viruses}.
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
+}
+\label{suppfile:virus_seqs}
 \end{suppfile}
@@
 \label{suppfile:mutations}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{
+{\bf Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna}.}
+%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}
+}
+\label{suppfile:validation_sequences}
 \end{suppfile}

### 2019-03-04 paper/paper.tex
fixed minor typo

@@
 Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
-The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini} .
+The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini}.
 Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

### 2019-03-04 paper/paper.tex
cleaned up citations

@@
 Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
-This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{Boers:2015emulsion}.
+This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{boers2015micelle}.
 We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .

### 2019-03-04 paper/paper.tex
tweak abstract

@@
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.
-Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Here we develop a new method to assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
+We do this by determining both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
 Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
@@
 Most cells are infected by virions that have a mutation or some other defect.
-This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and helps determine whether a cell activates an innate-immune response.
+This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and some viral mutations make a cell more likely to activate an innate-immune response.
 However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.

### 2019-03-04 paper/paper.tex
revision to Intro: emphasize stochasticity and do not overstate correlation of mutations and IFN

@@
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
-As the virus spreads, some cells detect infection and begin producing interferon (IFN).
+As the virus spreads, some cells detect infection and produce interferon (IFN).
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
@@
 For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
-But for influenza, a third possible contributor also looms large: viral genetic diversity.
-Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+But for influenza, a third possible contributor looms large: viral genetic diversity.
+Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
 Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
@@
 Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
-None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
-
-Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
-To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
+None of these techniques reliably reveal if there are mutations in the virion infecting any given cell.
+
+Here we develop a new approach to measure both the transcriptome and full sequences of all viral genes in single influenza-infected cells.
+To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
 We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
-Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
-This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
-Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
-However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
+Two-thirds of cells are infected by virions with a mutation or defect in gene expression.
+This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
+We identify several types of viral defects that increase IFN induction.
+However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
 Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

### 2019-03-04 paper/paper.tex
minor tweaks to text

@@
 
-A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
+A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
+Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction occurs at a frequency of $<$1\% in cell lines infected with influenza.
 IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).
@@
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
-Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
-
-We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
+The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
+
+We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
 

### 2019-03-04 paper/paper.tex
2 tiny textual changes

@@
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
+From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
 To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.

### 2019-03-04 paper/paper.tex
tweaks to transcriptomics section

@@
 Number of cells for which transcriptomes were obtained.
-From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
+%From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
-To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.
+To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\cite{haque2017practical}.
+After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B)
 
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
-We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
-The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
+We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
+The distribution of the amount of viral mRNA across infected cells is in the inset in \FIG{transcriptomics}C.
 As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
 
-We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
+We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
 We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
+\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
 This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
 
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
-However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
-Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
+The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
+However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
+Consistent with prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
 
@@
 From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
-Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
+This co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
 This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
@@
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\cite{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not directly detect infection~\cite{stetson2006type,honda2006type}.
 

### 2019-03-05 paper/paper.tex
revised genotypes section

@@
 {\bf Viral genotypes and infection outcomes in single cells.}
-Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
-Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
+Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
+The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
+Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
+Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
+The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
 For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
-Green boxes at left show the percent of all mRNA in that cell derived from virus.
-Orange boxes show the percent of cellular mRNA derived from IFN, with orange box frames indicating cells classified as IFN+ in \FIG{transcriptomics}G.
+For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1.
 }
@@
 We could call the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
-We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
-But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
-
-The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
-Simple visual inspection of this figure reveals a wealth of information.
-For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
-But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
-The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
-However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the wild-type virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the wild-type virion that infected \textit{cell 105} still induced IFN.
+We were most effective at calling viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
+But we also called genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
+
+The 150 cells for which we called the viral genotypes are shown in \FIG{genotypes}.
+Inspection of this figure reveals a wealth of anecdotal relationships between viral genotype and infection outcome.
+For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
+But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
+As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+
+However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
+This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
+As an example of the former case, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
+But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
+For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
 

### 2019-03-05 paper/paper.tex
revisions to `p_mutations` figure legend

@@
 \caption{
-{\bf Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.}
+{\bf Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.}
 {\bf (A)} 
 Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
-Cells infected by fully wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
+Cells infected by fully unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
 {\bf (B)}
-IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
+IFN expression among cells expressing unmutated copies of all genes, and cells with mutations or missing genes.
 {\bf (C)}
@@
 The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
-The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
+The middle and bottom panels show the percent of IFN- and IFN+ cells with a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
 Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
 {\bf (D)}
-There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
+There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN among cells that lack NS.
 Throughout this figure, we only consider substitutions that are non-synonymous.

### 2019-03-05 paper/paper.tex
added text about stochasticity to last results section

@@
 Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.
+However, the validation experiments in the next section show that the lack of statistical significance for many of the trends is due to the modest number of infected cells that were sequenced rather than a lack of true association between viral defects and IFN induction.
 
@@
 
-\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}
+\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction}
 
@@
 See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
-The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
+The experiments in the two panels were performed on different days, and so numerical values can be compared within panels but not between panels.
 }
@@
 The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
-We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
-
-The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of the amino-acid substitutions found in these genes among IFN+ cells.
-Overall, we created six such point-mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
-
-Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
-We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
+We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.
+
+The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells:
+PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
+
+Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+We created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
 
 We tested the rate of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
-The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
-But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
+\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild-type viral stock.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations to be found in IFN+ cells by chance.
+Overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often increase IFN production.
+
+However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
+\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the viral genotype.
+As can be seen from these data, none of the mutant viral stocks induce IFN in more than 20\% of infected cells. 
+Of course, these mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to that revealed by our single-cell sequencing of the ``wild-type'' viral stock.
+But IFN induction is stochastic at the single-cell level even for infections that share the same defect,such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
+Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction remains stochastic even with these defects.
 

### 2019-03-05 paper/paper.tex
tweaks to Discussion to emphasize stochasticity

@@
 Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
+Our method builds on the observation that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
 To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
-This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
+This viral genetic information is crucial for understanding infection outcome and innate-immune induction.
 Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wild-type'' influenza genome, most infected cells do not express wild-type copies of all viral genes.
-Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
-Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
-
-It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
-We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
-Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
+Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of defects across single cells.
+Inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
+
+These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
+We identified four types of viral defects in single IFN+ cells that we validated increase IFN induction. 
+Two types of defects---absence of the NS gene and mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
 The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini}.
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with prior work showing that such deletions can be immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
 There may be several reasons: we used relatively pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
+The relative importance of different types of viral defects also likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
 
 But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
-We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
-Additionally, none of the viral defects in IFN+ cells induce IFN deterministically in our validation experiments---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
+We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
+Additionally, none of the viral defects in IFN+ cells induce IFN deterministically---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
 Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
-However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
+However, we suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\cite{oneal2018west}.
 
 Perhaps the most intriguing question is how the heterogeneity that we have described ultimately affects the macroscopic outcome of infection.
-Natural human influenza infections are established by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
-It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
-Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
+Natural human influenza infections are initiated by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
+It is therefore plausible that early heterogeneity could have a large effect on downstream events.
+Extending our approaches to more complex systems could shed further light on how viral variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
 

### 2019-03-05 paper/paper.tex
a few more tweaks to text

@@
 
-Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
+Unfortunately for the host, influenza only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
@@
 
-To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
-Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+To identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
+Cells that activate the reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
+The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
 For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
+We performed PacBio sequencing on these viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
 Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
-Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).
+Combining this information with the 3'-end sequencing determines the entire transcriptome and full sequences of the infecting virions in single cells (\FIG{workflow}E).
 
@@
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+We tested if each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
 We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
@@
 
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
+The amount of viral mRNA was lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
 However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
+Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
@@
 As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
-The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
+Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
 
@@
 The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini}.
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with prior work showing that such deletions can be immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
+Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing that such deletions can be immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Given the extensive prior work describing such deletions, it is surprising we did not identify more of them in our single IFN+ cells.
 There may be several reasons: we used relatively pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-The relative importance of different types of viral defects also likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
-
-But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
+The relative importance of different defects also likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
+
+But although variation in viral populations is clearly important, it only partially explains heterogeneity among influenza-infected cells.
 We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
-Additionally, none of the viral defects in IFN+ cells induce IFN deterministically---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
-However, we suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\cite{oneal2018west}.
+Additionally, no viral defect induces IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
+We suspect that stochasticity or pre-existing cellular states explain most remaining heterogeneity, since IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell} or other viruses~\cite{oneal2018west}.
 
@@
 %This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
+%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation (DRG-2227-15).
 %JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.

### 2019-03-06 paper/paper_orig_submission.tex
PLoS Pathogens reviewer response

--- 2019-03-05 paper/paper.tex
@@
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
-Here we develop a new method to assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We do this by determining both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
+Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
+We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
 However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
-Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
+Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
 
@@
 Most cells are infected by virions that have a mutation or some other defect.
-This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and some viral mutations make a cell more likely to activate an innate-immune response.
+This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and helps determine whether a cell activates an innate-immune response.
 However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
@@
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
-As the virus spreads, some cells detect infection and produce interferon (IFN).
+As the virus spreads, some cells detect infection and begin producing interferon (IFN).
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
@@
 For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
-But for influenza, a third possible contributor looms large: viral genetic diversity.
-Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+But for influenza, a third possible contributor also looms large: viral genetic diversity.
+Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
 Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
@@
 Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
-None of these techniques reliably reveal if there are mutations in the virion infecting any given cell.
-
-Here we develop a new approach to measure both the transcriptome and full sequences of all viral genes in single influenza-infected cells.
-To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
+None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.
+
+Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
+To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
 We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
-Two-thirds of cells are infected by virions with a mutation or defect in gene expression.
-This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
-We identify several types of viral defects that increase IFN induction.
-However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
+Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
+This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
+Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
+However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
 Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
@@
 
-\subsection*{A system to identify and enrich rare IFN+ cells}
+\subsection*{A system to identify and enrich for rare IFN+ cells}
 
@@
 Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
-The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
+The plot also shows uninfected cells, and cells infected with Sendai virus.
 The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
@@
 
-A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
-Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction occurs at a frequency of $<$1\% in cell lines infected with influenza.
+A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
+Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
 IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).
@@
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
-For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
-
-We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
+The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
+
+We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
 
@@
 We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
+From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
 To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
@@
 Number of cells for which transcriptomes were obtained.
-%From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
+From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
 {\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
-To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\cite{haque2017practical}.
-After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B)
+To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.
 
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
-We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
-The distribution of the amount of viral mRNA across infected cells is in the inset in \FIG{transcriptomics}C.
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
+We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
+The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
 As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
 
-We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
+We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
 We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
+\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
 This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
 
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
-However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
-Consistent with prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
+The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
+However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
+Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
 
@@
 From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
-This co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
+Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
 This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
+Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
@@
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not directly detect infection~\cite{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\cite{stetson2006type,honda2006type}.
 
@@
 {\bf Viral genotypes and infection outcomes in single cells.}
-Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
-The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
-Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
-Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
-The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
+Arrows indicate presence of a viral gene from one (light blue) or both viral barcode variants (dark blue).
+Circles and boxes indicate mutations or indels as described in the legend, with circle areas and box heights proportional to the fraction of CCSs with that mutation.
 For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively. 
-For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1.
+Green boxes at left show the percent of all mRNA in that cell derived from virus.
+Orange boxes show the percent of cellular mRNA derived from IFN, with orange box frames indicating cells classified as IFN+ in \FIG{transcriptomics}G.
 }
@@
 
-After calling the presence / absence of each viral gene in each cell using the transcriptomic data as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
-For cells co-infected with both viral barcode variants, we called mutations separately for each variant.
-This strategy identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+We called a mutation in a viral gene if it was found in least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
+For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
+This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
 It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
@@
 We could call the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
-We were most effective at calling viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
-But we also called genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
-
-The 150 cells for which we called the viral genotypes are shown in \FIG{genotypes}.
-Inspection of this figure reveals a wealth of anecdotal relationships between viral genotype and infection outcome.
-For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
-But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
-As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
-The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
-
-However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
-This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
-As an example of the former case, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
-But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
-For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
+But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
+
+The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
+Simple visual inspection of this figure reveals a wealth of information.
+For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed wild-type copies all eight genes and did not induce detectable IFN.
+But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express at least one gene, and five of these cells expressed IFN.
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene, and many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
+However, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the wild-type virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the wild-type virion that infected \textit{cell 105} still induced IFN.
 
@@
 \caption{
-{\bf Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.}
+{\bf Viral features associated with heterogeneity in infection outcome among cells for which we sequenced all expressed viral genes.}
 {\bf (A)} 
 Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
-Cells infected by fully unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
+Cells infected by fully wild-type virus exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
 {\bf (B)}
-IFN expression among cells expressing unmutated copies of all genes, and cells with mutations or missing genes.
+IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
 {\bf (C)}
@@
 The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
-The middle and bottom panels show the percent of IFN- and IFN+ cells with a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
+The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
 Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
 {\bf (D)}
-There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN among cells that lack NS.
+There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
 Throughout this figure, we only consider substitutions that are non-synonymous.
@@
 Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.
-However, the validation experiments in the next section show that the lack of statistical significance for many of the trends is due to the modest number of infected cells that were sequenced rather than a lack of true association between viral defects and IFN induction.
 
@@
 
-\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction}
+\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}
 
@@
 \caption{
-{\bf Validation that IFN induction is increased by mutations identified in the single-cell virus sequencing of IFN+ cells.}
+{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
 {\bf (A)}
-Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated mutant, as determined using a reporter cell line.
-Numbers give the median of four measurements for each mutant.
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
+The numbers indicate the median of four measurements for each viral mutant.
 The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
-Points are orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
+Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
 {\bf (B)}
@@
 See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
-The experiments in the two panels were performed on different days, and so numerical values can be compared within panels but not between panels.
+The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
 }
@@
 
-To test if the viral defects identified in single IFN+ cells directly increase IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
+To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
 
@@
 The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
-We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.
-
-The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells:
-PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
-
-Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
-We created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
+We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
+
+The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with many of the amino-acid substitutions found in these genes among IFN+ cells.
+Overall, we created six such point-mutant viruses: PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
+
+Finally, prior work has suggested that virions carrying internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is an IFN+ cell that has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
 
 We tested the rate of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild-type viral stock.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
-The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations to be found in IFN+ cells by chance.
-Overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often increase IFN production.
-
-However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
-\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the viral genotype.
-As can be seen from these data, none of the mutant viral stocks induce IFN in more than 20\% of infected cells. 
-Of course, these mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to that revealed by our single-cell sequencing of the ``wild-type'' viral stock.
-But IFN induction is stochastic at the single-cell level even for infections that share the same defect,such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
-Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction remains stochastic even with these defects.
+\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
+But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
 
@@
 Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
-The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini}.
+The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini} .
 Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

### 2019-03-18 paper/paper.tex
re-format for _Journal of Virology_

--- 2019-03-05 paper/paper.tex
@@
-% Template for PLoS
-% Version 3.5 March 2018
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended 
-% to minimize problems and delays during our production 
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Once your paper is accepted for publication, 
-% PLEASE REMOVE ALL TRACKED CHANGES in this file 
-% and leave only the final text of your manuscript. 
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that 
-% no packages listed in the template may be deleted.
-%
-% Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file. 
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% Please add line breaks to long display equations when possible in order to fit size of the column. 
-%
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
-%
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
-%
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % % 
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-
-\documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
+\documentclass[lineno]{asm-article}
+
+\usepackage{blindtext}
+
+\usepackage[version=4]{mhchem}
+\usepackage{siunitx}
 
@@
 \usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
 
@@
 \usepackage{float}
-
-% added by Jesse Bloom for full-width figures
-\newenvironment{fullwidth}{%
-  \begin{adjustwidth}{-2.3in}{}
-}{\end{adjustwidth}}
 
@@
 \newfloat{suppfile}{thp}{lofsupfile}
-\floatname{suppfile}{File}
-\renewcommand{\thesuppfile}{S\arabic{suppfile}}
+\floatname{suppfile}{\bf FILE}
+\renewcommand{\thesuppfile}{\bf S\arabic{suppfile}}
 
 \newfloat{suppfig}{thp}{lofsupfile}
-\floatname{suppfig}{Fig}
-\renewcommand{\thesuppfig}{S\arabic{suppfig}}
-
-\renewcommand{\figureautorefname}{Fig}
+\floatname{suppfig}{\bf FIG}
+\renewcommand{\thesuppfig}{\bf S\arabic{suppfig}}
+
+\renewcommand{\figureautorefname}{Fig.}
 \newcommand{\FIG}[1]{\autoref{fig:#1}}
@@
 
-\newcommand{\suppfigautorefname}{Fig}
+\newcommand{\suppfigautorefname}{Fig.}
 \newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}
 
-% Remove comment for double spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-
-%% Include all macros below
-
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
-
-%% END MACROS SECTION
+\title{Single-cell virus sequencing of influenza infections that trigger innate immunity}
+
+\author{
+  Alistair B. Russell,\afn{a}
+  Jacob R. Kowalsky,\afn{a}
+  Jesse D. Bloom\afn{a,b,c,\authfn{1}}
+}
+
+\affil{
+  Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States\afn{a};
+  Department of Genome Sciences, University of Washington, Seattle, United States\afn{b};
+  Howard Hughes Medical Institute, Seattle, United States\afn{c}
+}
+
+\runningauthor{Russell et al.}
+\runningtitle{Single-cell virus sequencing of influenza infections that trigger innate immunity}
+
+\corraddress{Jesse D. Bloom, jbloom@fredhutch.org}
+
+\papertype{Research Article}
 
 \begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Single-cell virus sequencing of influenza infections that trigger innate immunity} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-\\
-Alistair B. Russell\textsuperscript{1},
-Jacob R. Kowalsky\textsuperscript{2},
-Jesse D. Bloom\textsuperscript{1,2,3*}
-\\
-\bigskip
-\textbf{1} Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States
-\\
-\textbf{2} Department of Genome Sciences, University of Washington, Seattle, United States
-\\
-\textbf{3} Howard Hughes Medical Institute, Seattle, United States
-\\
-\bigskip
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* jbloom@fredhutch.org
-
-\end{flushleft}
-% Please keep the abstract below 300 words
-\section*{Abstract}
+
+\maketitle
+
+\begin{abstract}
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.
@@
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step. 
-% Author Summary not valid for PLOS ONE submissions.   
-\section*{Author summary}
+\begin{importance}
 Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
@@
 Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.
-
-\linenumbers
-
-% Use "Eq" instead of "Equation" for equation citations.
-\section*{Introduction}
+\end{importance}
+
+\end{abstract}
+
+\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; mini viral RNA; heterogeneity}
+
+\section{Introduction}
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
-As the virus spreads, some cells detect infection and produce interferon (IFN).
+As the virus spreads, some cells detect infection and begin producing interferon (IFN).
 This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
@@
 For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
-But for influenza, a third possible contributor looms large: viral genetic diversity.
+But for influenza, a third possible contributor also looms large: viral genetic diversity.
 Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
@@
 Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
-None of these techniques reliably reveal if there are mutations in the virion infecting any given cell.
-
-Here we develop a new approach to measure both the transcriptome and full sequences of all viral genes in single influenza-infected cells.
-To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
+None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation.
+
+Here we develop a new approach measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
+To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
 We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
-Two-thirds of cells are infected by virions with a mutation or defect in gene expression.
+Two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
 This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
@@
 
-\section*{Results}
-
-\subsection*{A system to identify and enrich rare IFN+ cells}
+\section{Results}
+
+\subsection{A system to identify and enrich rare IFN+ cells}
 
@@
 \centerline{
-{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
-\hspace{0.02\textwidth}
-{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
+{\bf \Large A}\includegraphics[width=0.4\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
+\hspace{0.03\textwidth}
+{\bf \Large B} \includegraphics[width=0.26\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
 }
@@
 
-A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
-Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction occurs at a frequency of $<$1\% in cell lines infected with influenza.
+A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
+Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
 IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).
 
-To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
+To identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
 Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
 The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
-For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
+Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
@@
 
-\subsection*{Combined transcriptomics and virus-sequencing of single infected cells}
+\subsection{Combined transcriptomics and virus-sequencing of single infected cells}
 
@@
 
-\subsection*{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}
+\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}
 
@@
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
-We tested whether each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
-We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
-The distribution of the amount of viral mRNA across infected cells is in the inset in \FIG{transcriptomics}C.
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
+We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
+The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
 As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
@@
 We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
+\FIG{transcriptomics}D (top panel) shows that 162 of 290 infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
 This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
 
-The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
-However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
-Consistent with prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
+The amount of viral mRNA was lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
+However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
+Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
 
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
+Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
@@
 
-\subsection*{Full genotypes of viruses infecting single IFN+ and IFN- cells}
+\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells}
 
@@
 {\centering
-\includegraphics[height=0.9\textheight]{figures/single_cell_figures/p_genotypes.pdf}
+\includegraphics[height=0.85\textheight]{figures/single_cell_figures/p_genotypes.pdf}
 }
@@
 After calling the presence / absence of each viral gene in each cell using the transcriptomic data as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
-For cells co-infected with both viral barcode variants, we called mutations separately for each variant.
-This strategy identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
+For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
+This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
 It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
@@
 We could call the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
-We were most effective at calling viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
-But we also called genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
-
-The 150 cells for which we called the viral genotypes are shown in \FIG{genotypes}.
-Inspection of this figure reveals a wealth of anecdotal relationships between viral genotype and infection outcome.
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
+But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
+
+The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
+Simple visual inspection of this figure reveals a wealth of information.
 For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
@@
 But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
-For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that fail to activate IFN despite being infected by virions that fail to express NS.
-
-\subsection*{Viral defects associated with infection outcome in single cells}
+For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that do not activate IFN despite being infected by virions that fail to express NS.
+
+\subsection{Viral defects associated with infection outcome in single cells}
 
@@
 {\bf (B)}
-IFN expression among cells expressing unmutated copies of all genes, and cells with mutations or missing genes.
+IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
 {\bf (C)}
@@
 However, some specific viral defects were significantly associated with IFN induction: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (\FIG{mutations}C).
-Due to the low number of cells and large number of hypotheses in \FIG{mutations}C, the only trend that remained significant at a false discovery rate of 10\% was absence of NS.
-However, the validation experiments in the next section show that the lack of statistical significance for many of the trends is due to the modest number of infected cells that were sequenced rather than a lack of true association between viral defects and IFN induction.
+The only trend that remained significant at a false discovery rate (FDR) of 10\% was absence of NS.
+This lack of statistical significance after FDR correction could be due to the relatively modest number of fully sequenced infected cells (just 150).
+The validation experiments in the next section show that many of the viral mutations in IFN+ cells do in fact increase the rate of IFN induction.
 
@@
 
-\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction}
+\subsection{Validation that viral defects in single IFN+ cells often increase IFN induction}
 
@@
 
-\centerline{\includegraphics[width=0.85\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
-\caption{
-{\bf Validation that IFN induction is increased by mutations identified in the single-cell virus sequencing of IFN+ cells.}
+\centerline{\includegraphics[width=0.9\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
+\caption{
+{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
 {\bf (A)}
-Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated mutant, as determined using a reporter cell line.
-Numbers give the median of four measurements for each mutant.
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
+The numbers indicate the median of four measurements for each viral mutant.
 The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
-Points are orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
+Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
 {\bf (B)}
@@
 The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
-We therefore mimicked the effect of absence of NS by creating a mutant virus (NS1stop) that had multiple stop codons early in the NS1 coding sequence.
-
-The single-cell data also showed that amino-acid substitutions in PB1 and NS were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells:
+We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
+
+The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells:
 PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
@@
 Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
-We created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
+We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
 
@@
 \FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild-type viral stock.
-The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
-The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations to be found in IFN+ cells by chance.
-Overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often increase IFN production.
+The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
+But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
 
@@
 Of course, these mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to that revealed by our single-cell sequencing of the ``wild-type'' viral stock.
-But IFN induction is stochastic at the single-cell level even for infections that share the same defect,such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
+But IFN induction is stochastic at the single-cell level even for infections that share the same defect, such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
 Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction remains stochastic even with these defects.
 
-\section*{Discussion}
+A detailed analysis of the flow cytometry underlying \FIG{validation} also confirms the trend in the single-cell data noted in \FIG{mutations}D: namely that IFN induction is positively associated with total viral gene expression only when the virus has defects in NS1.
+Specifically, \SUPPFIG{IFN_vs_HA} shows the relative frequency of IFN induction for each viral mutant as a function of HA protein expression.
+For the NS1stop and NS1-A122V mutants, there is a strong association between HA protein expression and IFN induction---but not for the other viral variants.
+Taken together with the single-cell data, this fact suggests that only in the absence of functional NS1 does higher viral gene expression predispose an infected cell to induce IFN.
+Therefore, viral variation influences both the rate of IFN induction and which factors contribute to this induction.
+
+\section{Discussion}
 We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
 Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observation by that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
+Our method builds on the observation that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
 To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
@@
 
-It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
+These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
 We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
-Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
+Two types of defects---absence of the NS gene and amino-acid mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
 The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini}.
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
@@
 But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
-We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wild-type virions.
-Additionally, none of the viral defects in IFN+ cells induce IFN deterministically in our validation experiments---and the apparently most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
+We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
+Additionally, no viral defect induces IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
 Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
@@
 Natural human influenza infections are established by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
-It is therefore plausible that genetic variation in the founding viral population could have a large effect on downstream events.
+It is therefore plausible that early heterogeneity could have a large effect on downstream events.
 Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
 
-\section*{Methods and Materials}
-
-\subsection*{IFN reporter cell lines}
+\section{Materials and Methods}
+\small
+\subsection{IFN reporter cell lines}
 We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}A).
@@
 
-\subsection*{Viruses for single-cell experiments}
+\subsection{Viruses for single-cell experiments}
 We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
@@
 
-\subsection*{Flow cytometry analyses for HA expression}
+\subsection{Flow cytometry analyses for HA expression}
 For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
@@
 
-\subsection*{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
+\subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
 The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-P2A-mNeonGreen reporter.
@@
 
-\subsection*{Enrichment and preparation of viral cDNA for PacBio sequencing}
+\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
 We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
@@
 
-\subsection*{qPCR for viral genes and IFN}
+\subsection{qPCR for viral genes and IFN}
 We performed qPCR on reverse-transcribed mRNA for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
@@
 
-\subsection*{Viruses and experiments for validation experiments}
+\subsection{Viruses and experiments for validation experiments}
 In \FIG{validation}, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
@@
 
-\subsection*{Computational analysis of single-cell transcriptomic and viral sequence data}
+\subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
 A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}. 
@@
 
-\section*{Acknowledgments}
+\normalsize
+
+\section{Acknowledgments}
 We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
-%This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
-%ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
-%JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
-%JDB is an Investigator of the Howard Hughes Medical Institute.
-%The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
+This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
+ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation (DRG-2227-15).
+JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
+JDB is an Investigator of the Howard Hughes Medical Institute.
+The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
 
@@
 \clearpage
-\section*{Supporting information}
+\section{Supplemental Material}
 
@@
 
+\begin{suppfig}
+\centerline{
+{\bf \Large A}
+\includegraphics[width=0.32\textwidth, valign=t]{figures/MutantComparison/p_foldchange.pdf}
+\hspace{0.03\textwidth}
+{\bf \Large B}
+\includegraphics[width=0.55\textwidth, valign=t]{figures/MutantComparison/p_ifn_vs_ha.pdf}
+}
+\caption{
+Higher viral gene expression is strongly associated with more IFN induction only in NS1-deficient infections.
+This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in \SUPPFIG{validationflow}.
+{\bf (A)}
+For viruses with defects in NS1, there is more IFN induction among infected cells that express higher levels of HA protein.
+The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile as determined by flow cytometry.
+Points indicate replicates, and lines indicate the mean.
+{\bf (B)}
+A more detailed version of the data summarized in (A).
+For each virus and replicate, we binned the infected cells into HA expression quartiles.
+We then calculated the percent of cells that were IFN+ in each quartile.
+The plots show the mean HA expression of the quartile versus the precent of cells that are IFN+.
+The results clearly show that for the NS1stop and to a lesser extend the NS1-A122V variants, more viral protein (higher HA signal) correlations with IFN induction.
+Note that panel (A) summarizes these same data by simply showing the ratio of percent IFN+ between the highest and lowest quartile.
+}
+\label{suppfig:IFN_vs_HA}
+\end{suppfig}
+
 \clearpage

### 2019-03-18 paper/paper.tex
added te Velthuis PB1 data *without* editing

@@
   Alistair B. Russell,\afn{a}
+  Elizaveta Elshina,\afn{b}
   Jacob R. Kowalsky,\afn{a}
-  Jesse D. Bloom\afn{a,b,c,\authfn{1}}
+  Aartjan J. W. te Velthuis,\afn{b}
+  Jesse D. Bloom\afn{a,c,d,\authfn{1}}
 }
@@
   Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States\afn{a};
-  Department of Genome Sciences, University of Washington, Seattle, United States\afn{b};
-  Howard Hughes Medical Institute, Seattle, United States\afn{c}
+  Division of Virology, Department of Pathology, Addenbrooke's Hospital, University of Cambridge, Cambridge, United Kingdom;
+  Department of Genome Sciences, University of Washington, Seattle, United States\afn{c};
+  Howard Hughes Medical Institute, Seattle, United States\afn{d}
 }
@@
 \caption{
-{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
-{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
+Reporter cells to identify and enrich infections that activate IFN expression.
+(A) The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
 We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
 In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
-{\bf (B)}
+(B)
 Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
@@
 \caption{
-{\bf Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.}
-{\bf (A)}
+Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
+(A)
 IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
 IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
-{\bf (B)}
+(B)
 The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
-{\bf (C)}
+(C)
 Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
-{\bf (D)}
+(D)
 viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
-{\bf (E)}
+(E)
 The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
@@
 \caption{
-{\bf Single-cell transcriptomics of IFN-enriched influenza-infected cells.}
-{\bf (A)} 
+Single-cell transcriptomics of IFN-enriched influenza-infected cells.
+(A) 
 Number of cells for which transcriptomes were obtained.
 From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
-{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
+(B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
 Green lines show the distribution of cellular mRNAs per cell.
 Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
-{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
+(C) Distribution across cells of the fraction of all mRNA derived from influenza.
 Cells called as infected are in blue, while other cells are in green.
 The inset shows the amount of viral mRNA in the human cells that are called as infected.
-{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
+(D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
 The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
 \SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
-{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
-{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
+(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
+(F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
 From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
-{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
+(G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
 Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
@@
 \caption{
-{\bf Viral genotypes and infection outcomes in single cells.}
+Viral genotypes and infection outcomes in single cells.
 Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
@@
 \caption{
-{\bf Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.}
-{\bf (A)} 
+Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.
+(A) 
 Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
 Cells infected by fully unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
-{\bf (B)}
+(B)
 IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
-{\bf (C)}
+(C)
 Specific viral defects associated with IFN induction.
@@
 Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
-{\bf (D)}
+(D)
 There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
@@
 \caption{
-{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
-{\bf (A)}
+Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.
+(A)
 Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
@@
 Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
-{\bf (B)}
+(B)
 Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
@@
 Therefore, viral variation influences both the rate of IFN induction and which factors contribute to this induction.
+
+\subsection{PB1 mutants}
+
+%%% begin PB1 figure %%%
+\begin{figure}
+
+\centerline{\includegraphics[width=0.7\textwidth]{figures/PB1_figure.png}}
+\caption{
+Mutations D27N and T677A in the PB1 subunit affect polymerase activity in vitro. 
+(A) Structural model of bat influenza A virus RNA polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the structure of the influenza B virus RNA polymerase (PDB 5MSG)~\cite{reich2017vitro}.
+The locations of PB1 D27 and T677 (both red) of the influenza A virus polymerase relative to the 5' (blue) and 3' (orange) termini of the template and the PB1 active site (grey; PB1 act) are indicated.
+The PA endonuclease (green; PA endo) and PB2 cap binding domain (pink; PB2 cap) are also indicated for reference.
+Part of the fingers subdomain of PB1 is not shown to reveal the template in the entry channel.
+(B) Influenza A virus RNA polymerase activity on a neuraminidase (NA) encoding vRNA template in HEK 293T cells.
+Steady state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging.
+A PB1 active site control (PB1a) was used as negative control and background correction. 
+The 5S rRNA signal was used as loading control. 
+Other panels show western blot analysis of PB1, NP and GAPDH expression. 
+Graph shows the average steady state RNA levels of three biological repeats. 
+(C) Panel shows PAGE analysis of RT-PCR of short (17-200 nt long) influenza virus RNAs or mini viral RNAs (mvRNAs).
+The length of the mini viral RNAs was extended by the adapters (adapt) present in the primers.
+Graph shows the IFN-beta promoter activity in transfected HEK 293T cells as measured using a dual luciferase reporter assay.
+(D) Influenza A virus polymerase activity on a segment 5-derived 246 nt long vRNA template. 
+The top panel shows the steady state levels of vRNA template were analysed by primer extension and denaturing PAGE. 
+The other two panels show the PB1 and tubulin expression levels analysed by western blot. 
+The top graph shows the average steady state RNA levels of three biological repeats. 
+The second graph shows the IFN-beta promoter activity in transfected HEK 293T cells as measured using a dual luciferase reporter assay. 
+Error bars indicate standard deviation. 
+P-values were determined by two-sided t-test.
+}
+\label{fig:PB1}
+
+\end{figure}
+%%% end PB1 figure %%%
+
+Previous research has shown that aberrant RNA products, or mini viral RNAs (mvRNAs) {te Velthuis 2018}, produced during influenza A virus replication can trigger RIG-I activation and IFN production.
+To investigate if the PB1 mutations D27N and T667A could affect influenza A virus RNA polymerase activity we first identified their location and interactions with the template by superposing the pre-initiation complex of the bat influenza A virus RNA polymerase with the transcription initiation complex of the influenza B virus RNA polymerase (Fig.
+7A).
+The model shows how the 3' terminus of the template enters the template entry channel above the PB1 active site and that T667A sits at tip of a helix that interacts with the 3' terminus as it enters the channel.
+By contrast, D27N resides deeper in the polymerase and close to the binding pocket of the 5' terminus.
+These observations suggest that both mutations may affect polymerase activity by changing the interactions of the RNA polymerase with the template.
+To investigate how the two PB1 mutations affect polymerase activity on a full-length viral RNA template, we expressed the wildtype or mutant PB1 subunits along with PB2, PA, NP and the segment 6 vRNA in HEK 293T cells.
+Twenty-four hours after transfection, we extracted the total RNA and analysed the replication (vRNA level) and transcriptional (mRNA level) activity by primer extension.
+Interestingly, both mutations showed activities that were significantly different from wildtype (Fig.
+7B), with D27N and T667A showing lower and higher viral replication and transcription, respectively.
+Both PB1 mutants showed higher activities than the PB1 active site control (PB1a) that we used to subtract background RNA levels.
+Moreover, both mutants were expressed to similar levels, as assessed by western blot.
+To investigate if the mutants produced any mvRNAs, we fractionated the total RNA to enrich for small RNA species (17-200 nt long) and performed an RT-PCR that amplified RNA species that contained both the 3' and 5' terminus of the template on equal amounts of RNA.
+This revealed that the D27N expressing cells contained higher levels of mvRNAs than the wildtype control and the T667A mutant (Fig.
+7C).
+Analysis of the IFN levels showed that...
+The low activity on the full-length segment 6 vRNA and higher mvRNA level produced on this template suggest that the D27N mutation in PB1 impaired polymerase processivity on full-length vRNAs.
+To confirm this, we replaced the segment 6 template with a segment 5-derived 246-nt long template (Fig.
+7D).
+Under these conditions, the activity of the D27N mutant was similar to wildtype, while the activity of the T677A mutant proved again higher than wildtype, although not significantly higher in three biological repeats.
+Analysis of the IFN expression level in response to viral replication on the short template showed higher IFN induction by both PB1 mutants.
 
@@
 This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{boers2015micelle}.
-We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .
+We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume of 100$\mu$l.
 Both the common 3' primer and the multiplex mix of eight 5' primers were added to a final concentration of 0.5$\mu$M.
@@
 
+\subsection{Structural analysis of PB1 mutants}
+To locate the PB1 mutations in the influenza A virus RNA polymerase structure relative to the template and active site in \FIG{PB1}A, we superposed the bat influenza A virus RNA polymerase structure (PDB 4WSB)~\cite{reich2014structural}, which shows the 3' terminus of the template on the surface of the RNA polymerase, with the influenza B virus transcription initiation complex (PDB 5MSG)~\cite{reich2017vitro}, which shows the 3' terminus of the template in the template entry channel that leads towards the active site.
+The structural alignment was performed in Pymol 1.8.7 using motifs A and C.
+
+\subsection{Experimental analysis of PB1 mutants}
+For the experimental analysis of the PB1 mutants in \FIG{PB1}, we used plasmids pcDNA-PB1, pcDNA-PA, pcDNA-PB2, and pcDNA-NP, which encode the WSN proteins that compose the polymerase complex~\cite{fodor2002single}; pPolI-NA, which encodes the viral RNA for the WSN NA~\cite{fodor2002single}; and pcDNA-PB1a, which encodes an inactive version of the WSN PB1 polymerase protein~\cite{vreede2004model}.
+To construct plasmids expressing mutant PB1 proteins D27N and T677A, the plasmid pcDNA-PB1 was subjected to site-directed mutagenesis. 
+PB1 expression was analysed by western blot using antibody GTX125923 (GeneTex).
+
+To analyse the activity of the PB1 mutants in cell culture, the plasmids expressing the WSN PA, PB2, NP and PB1 proteins were transfected into 293T cells together with the plasmid expressing the wildtype NA vRNA or a 246-nt long segment 5-based template~\cite{velthuis2018mini}.
+Twenty-four hours post transfection, the RNA was extracted using Trizol (Invitrogen), and the steady state RNA levels assessed using reverse transcription with 32P-labelled oligonucleotides against the viral RNA species and ribosomal 5S RNA as described previously~\cite{velthuis2018mini, velthuis2016role}.
+32P-derived signals were imaged using phosphorimaging on a Typhoon scanner and analysed using Prism (GraphPad).
+In all experiments, the apparent RNA levels were background corrected using the PB1a mutant and normalized to the 5S rRNA loading control.
+
+To measure the induction of the IFN-beta promoter during these RNP reconstitution assays, they were carried out in the presence of a plasmid expressing Renilla luciferase from a CMV promoter and a plasmid expressing Firefly luciferase from the IFN-beta promoter~\cite{velthuis2018mini}.
+Twenty-four hours post transfection, cells were harvested, lysed and analysed using a DualGlo luciferase kit (Promega) according to the manufacturer’s instructions.
+Samples were analysed using a GloMax (Promega).
+
 \normalsize
@@
 \section{Acknowledgments}
-We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
+We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, Katherine Xue, and Hannah Itell for helpful suggestions.
 We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
@@
 \caption{
-{\bf Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
+Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
 Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
@@
 \caption{
-{\bf Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.}
-{\bf (A)}
+Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.
+(A)
 To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\cite{strahle2006sendai}.
@@
 Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
-{\bf (B)}
+(B)
 The type I and type III IFN reporters are highly correlated in their activation.
@@
 \caption{
-{\bf Flow cytometry data for \FIG{IFNrare}B.}
+Flow cytometry data for \FIG{IFNrare}B.
 The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\cite{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
@@
 \caption{
-{\bf Example MACS enrichments of IFN+ influenza-infected cells.}
+Example MACS enrichments of IFN+ influenza-infected cells.
 A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
 After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
-{\bf (A)}
+(A)
 After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
@@
 We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
-{\bf (B)}
+(B)
 Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
@@
 \caption{
-{\bf The fraction of infected cells that are called as expressing each viral gene.}
+The fraction of infected cells that are called as expressing each viral gene.
 The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
@@
 \caption{
-{\bf The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.}
+The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
 Each point represents one cell.
@@
 \caption{
-{\bf Expression of ISGs in single infected and uninfected cells.}
+Expression of ISGs in single infected and uninfected cells.
 For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
-{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
+(A) The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
 We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
@@
 This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\cite{stetson2006type,honda2006type}.
-{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
+(B) Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
 Each point represents one cell, and the Pearson correlation coefficient is shown.
@@
 \caption{
-{\bf Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.}
+Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.
 To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\cite{maaten2008visualizing} as implemented in \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
@@
 \caption{
-{\bf The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.}
+The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
 These sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells (see \SUPPFILE{pacbio_analysis}).
@@
 \caption{
-{\bf Strategy for detecting strand exchange during sequencing of full-length viral genes.}
+Strategy for detecting strand exchange during sequencing of full-length viral genes.
 The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
@@
 \caption{
-{\bf Number of cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).}
-{\bf (A)} We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
-{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
+Number of cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).
+(A) We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
+(B) The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
 Both facts makes sense.
@@
 \caption{
-{\bf Flow cytometry data for \FIG{validation}.}
-{\bf (A)}
+Flow cytometry data for \FIG{validation}.
+(A)
 Data for \FIG{validation}A.
@@
 The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
-{\bf (B)}
+(B)
 Data for \FIG{validation}B.
@@
 \caption{
-{\bf Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.}
+Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.
 In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{validationflow}B), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
@@
 This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in \SUPPFIG{validationflow}.
-{\bf (A)}
+(A)
 For viruses with defects in NS1, there is more IFN induction among infected cells that express higher levels of HA protein.
@@
 Points indicate replicates, and lines indicate the mean.
-{\bf (B)}
+(B)
 A more detailed version of the data summarized in (A).
@@
 \begin{suppfile}
-\caption{{\bf Sequences of the IFN reporters in \FIG{IFNrare}A.}
+\caption{
+Sequences of the IFN reporters in \FIG{IFNrare}A.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
@@
 \caption{
-{\bf Genbank files giving sequences of the wild-type and synonymously barcoded viruses}.
+Genbank files giving sequences of the wild-type and synonymously barcoded viruses.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
@@
 \caption{
-{\bf A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing.}
+A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt} 
@@
 \caption{
-{\bf A CSV file giving the genotypes in \FIG{genotypes}.}
+A CSV file giving the genotypes in \FIG{genotypes}.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.
@@
 \caption{
-{\bf A CSV file giving the viral mutations and related information in \FIG{mutations}.}
+A CSV file giving the viral mutations and related information in \FIG{mutations}.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.
@@
 \caption{
-{\bf Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna}.}
+Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna}.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}
@@
 \caption{\label{suppfile:pacbio_analysis}
-{\bf HTML rendering of Jupyter notebook that analyzes the PacBio data.}
+HTML rendering of Jupyter notebook that analyzes the PacBio data.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
@@
 \caption{\label{suppfile:monocle_analysis}
-{\bf HTML rendering of Jupyter notebook that analyzes the annotated cell-gene matrix.}
+HTML rendering of Jupyter notebook that analyzes the annotated cell-gene matrix.
 %See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.

### 2019-03-18 paper/paper.tex
revise abstract

@@
 
-\author{
+\author{%
 Alistair B. Russell,\afn{a}
@@
 
-\affil{
-  Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States\afn{a};
-  Division of Virology, Department of Pathology, Addenbrooke's Hospital, University of Cambridge, Cambridge, United Kingdom;
-  Department of Genome Sciences, University of Washington, Seattle, United States\afn{c};
-  Howard Hughes Medical Institute, Seattle, United States\afn{d}
+\affil{%
+\afn{a}Basic Sciences and Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, United States
+
+\afn{b} Division of Virology, Department of Pathology, University of Cambridge, Cambridge, United Kingdom
+
+\afn{c} Department of Genome Sciences, University of Washington, Seattle, United States
+
+\afn{d} Howard Hughes Medical Institute, Seattle, United States
 }
@@
 The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.
-Here we develop a new method to assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
-We do this by determining both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
-Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
-However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
-Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
+Here we develop a new method to assess how the genetic variation in viral populations contributes to this heterogeneity.
+We do this by determining the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+Most cells are infected by virions with defects, some of which increase the frequency of innate-immune activation.
+These immunostimulatory defects are diverse, and include mutations that perturb the function of the viral polymerase protein PB1, large internal deletions in viral genes, and failure to express the virus's interferon antagonist NS1.
+However, immune activation remains stochastic in cells infected by virions with these defects, and occasionally is triggered even by virions that express unmutated copies of all genes.
+Our work shows that the diverse spectrum of defects in influenza virus populations contributes to---but does not completely explain---the heterogeneity in infection outcome and immune activation in single infected cells.
 
@@
 Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
-But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
-Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.
-Most cells are infected by virions that have a mutation or some other defect.
-This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and some viral mutations make a cell more likely to activate an innate-immune response.
-However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
-Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.
+But so far, it has been impossible to fully characterize the sequence of the virion infecting any given cell, since conventional techniques like flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
+Here we develop a new approach that uses long-read PacBio sequencing to determine the sequences of virions infecting single cells.
+We show that viral genetic variation explains some but not all of the cell-to-cell variability in infection outcome.
+Overall, our study provides the first complete picture of how viral mutations affect the course of infection in single cells.
 \end{importance}
@@
 
-\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; mini viral RNA; heterogeneity}
+\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; mini viral RNA; mvRNA; heterogeneity}
 

### 2019-03-18 paper/paper.tex
revise Intro and Results

@@
 \floatname{suppfile}{\bf FILE}
-\renewcommand{\thesuppfile}{\bf S\arabic{suppfile}}
+\renewcommand{\thesuppfile}{S\arabic{suppfile}}
 
@@
 \floatname{suppfig}{\bf FIG}
-\renewcommand{\thesuppfig}{\bf S\arabic{suppfig}}
+\renewcommand{\thesuppfig}{S\arabic{suppfig}}
 
@@
 \newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}
+
+% Allow big figures on pages
+\renewcommand{\topfraction}{0.9}
+\renewcommand{\bottomfraction}{0.9}
+\renewcommand{\textfraction}{0.1}
+\renewcommand{\floatpagefraction}{0.9}
 
@@
 Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
-This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell}.
 Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
@@
 Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
-Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{killip2017single, velthuis2018mini, du2018genome, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
-Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
+Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, steuerman2018dissection, wang2018cell, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation.
@@
 Here we develop a new approach measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
-To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
-We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
+To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
 Two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
-This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
-We identify several types of viral defects that increase IFN induction.
-However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
-Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.
+We identify several types of viral defects that increase IFN induction, but also show that viral diversity is insufficient to fully explain cell-to-cell heterogeneity during influenza infection.
 
@@
 
-\subsection{A system to identify and enrich rare IFN+ cells}
+\subsection{A system to identify and enrich rare IFN+ cells.}
 
@@
 
-A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
-Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
-IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).
-
-To identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
-Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+Influenza virus only rarely triggers IFN expression in infected cells~\cite{killip2017single, russell2018extreme, kallfass2013visualizing}---a fact that poses a challenge for the study IFN induction in single cells.
+Therefore, we developed a method to identify and enrich rare IFN+ cells by creating A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
+Cells that activate the IFN reporters can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
+The reporters were efficiently activated by infection with a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
 Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
 
-We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.
-
-\subsection{Combined transcriptomics and virus-sequencing of single infected cells}
+We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
+We also validated that MACS for the cell-surface protein driven by the IFN reporter enriched the IFN+ cells by $>$50-fold (\SUPPFIG{MACS}).
+
+\subsection{Combined transcriptomics and virus-sequencing of single infected cells.}
 
@@
 \caption{
-Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
+Approach for combined transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
 (A)
 IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
-IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
+IFN+ cells are enriched by MACS, and pooled with non-enriched cells and uninfected canine cells that serve as a control for multiplets and mRNA leakage.
 (B)
@@
 (D)
-viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
+viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are sequenced by PacBio).
 (E)
@@
 
-To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
+We developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
 First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
-These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
-We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
+These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells.
+From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
 To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
@@
 We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
-Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
+Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to viral as well as cellular mRNAs.
 Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.
@@
 But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
-We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
+We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs)~\cite{travers2010flexible}.
 These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
@@
 
-\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}
+\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells.}
 
@@
 Number of cells for which transcriptomes were obtained.
-From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
+From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are derived from multiple cells. 
 (B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
@@
 (D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
-The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
 \SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
-(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
+(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell derived from each gene.
 (F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
@@
 (G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
-Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
+Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line as IFN+.
 A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
-Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
 }
@@
 To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
-As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
+As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA.
 We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
-We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
+We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
 The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
@@
 We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
-We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
+We called presence / absence of genes rather than transcripts, since the two genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
 \FIG{transcriptomics}D (top panel) shows that 162 of 290 infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
-This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.
+This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work using the WSN strain~\cite{russell2018extreme}, and slightly to substantially higher than that reported in studies by others using different viral strains or methodologies~\cite{heldt2015single, brooke2013most, dou2017analysis, jacobs2019incomplete}.
 
@@
 We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
-Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
+Over 20\% of infected cells were IFN+, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
-Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{ISG}A).
+Many more cells expressed ISGs than IFN itself (\SUPPFIG{ISG}A).
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not directly detect infection~\cite{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\cite{stetson2006type,honda2006type}.
 
@@
 
-\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells}
+\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells.}
 
@@
 For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
-For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1.
+For instance, \textit{cell 5} was co-infected by a virion with one unmutated and one internally deleted copy of PB1.
 }
@@
 
-We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
-Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
-Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
-
-After calling the presence / absence of each viral gene in each cell using the transcriptomic data as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
+We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes in single infected cells.
+We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
+The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+
+After calling the presence / absence of each viral gene in each cell as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
 For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
@@
 The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
-Simple visual inspection of this figure reveals a wealth of information.
+Visual inspection of this figure reveals a wealth of information.
 For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
-But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
+But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
 As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
-The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
+The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
 Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).
@@
 As an example of the former case, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
-But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
+But in other cases, the stochasticity allows a defective virus to still escape immune recognition.
 For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that do not activate IFN despite being infected by virions that fail to express NS.
 
-\subsection{Viral defects associated with infection outcome in single cells}
+\subsection{Viral defects associated with infection outcome in single cells.}
 
@@
 Throughout this figure, we only consider substitutions that are non-synonymous.
-Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
 }
@@
 To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by full wild-type viruses had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
-Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden among cells.
+\FIG{mutations}A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
+Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.
 
@@
 One other interesting trend emerges from the single-cell data.
-There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS---but among cells that lack NS, the IFN+ ones have significantly more viral mRNA (\FIG{mutations}D).
-
-\subsection{Validation that viral defects in single IFN+ cells often increase IFN induction}
+There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS.
+But among cells that lack NS, cells with more viral mRNA are significantly more likely to be IFN+ (\FIG{mutations}D).
+This finding is elaborated on in the validation experiments below.
+
+\subsection{Validation that viral defects in single IFN+ cells often increase IFN induction.}
 
@@
 
-To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
+To test if the viral defects identified in single IFN+ cells cause increased IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.
 
@@
 The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
-We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
-
-The single-cell data also showed that amino-acid substitutions in the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of the substitutions found in these genes among IFN+ cells:
+We therefore mimicked the absence of NS by creating a mutant virus (which we term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
+
+The single-cell data also showed that amino-acid substitutions in proteins encoded by the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of these substitutions:
 PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
 
-Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce higher levels of IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce higher levels of IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
@@
 The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
-The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we also expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
-But overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
+The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
+Overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
 
@@
 Of course, these mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to that revealed by our single-cell sequencing of the ``wild-type'' viral stock.
-But IFN induction is stochastic at the single-cell level even for infections that share the same defect, such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
+But our single-cell data show that IFN induction is stochastic even for infections that share the same defect, such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
 Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction remains stochastic even with these defects.
 
-A detailed analysis of the flow cytometry underlying \FIG{validation} also confirms the trend in the single-cell data noted in \FIG{mutations}D: namely that IFN induction is positively associated with total viral gene expression only when the virus has defects in NS1.
-Specifically, \SUPPFIG{IFN_vs_HA} shows the relative frequency of IFN induction for each viral mutant as a function of HA protein expression.
-For the NS1stop and NS1-A122V mutants, there is a strong association between HA protein expression and IFN induction---but not for the other viral variants.
-Taken together with the single-cell data, this fact suggests that only in the absence of functional NS1 does higher viral gene expression predispose an infected cell to induce IFN.
-Therefore, viral variation influences both the rate of IFN induction and which factors contribute to this induction.
-
-\subsection{PB1 mutants}
+\subsection{The IFN-inducing viral defects act by diverse mechanisms.}
+Some of the viral defects in IFN+ are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are known to be prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Other defects are more surprising: for instance, it is not obvious why amino-acid mutations in PB1 increase IFN induction.
+We therefore designed experiments to interrogate the mechanisms of some of these defects.
+
+We first focused on the PB1 mutations.
+Recent research has shown that mutations to the PB2 polymerase subunit of the WSN viral strain can enhance IFN induction by leading to the production of aberrant mini viral RNAs (mvRNAs) that trigger RIG-I~\cite{velthuis2018mini}.
+To investigate if the PB1 mutations act by a similar mechanism, we first examined their location in a structural model of the polymerase complex (\FIG{PB1}A).
+The IFN-enhancing PB1 mutation T677A occurs at the tip of a helix that interacts with the 3' terminus of the RNA template as it enters the channel above the PB1 active site, whereas the IFN-enhancing D27N mutation is deeper in the polymerase and close to the binding pocket of the 5' terminus of the template.
+Therefore, both mutations could plausibly affect the polymerase's activity by changing its interactions with the RNA template.
 
@@
 \caption{
-Mutations D27N and T677A in the PB1 subunit affect polymerase activity in vitro. 
-(A) Structural model of bat influenza A virus RNA polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the structure of the influenza B virus RNA polymerase (PDB 5MSG)~\cite{reich2017vitro}.
-The locations of PB1 D27 and T677 (both red) of the influenza A virus polymerase relative to the 5' (blue) and 3' (orange) termini of the template and the PB1 active site (grey; PB1 act) are indicated.
-The PA endonuclease (green; PA endo) and PB2 cap binding domain (pink; PB2 cap) are also indicated for reference.
-Part of the fingers subdomain of PB1 is not shown to reveal the template in the entry channel.
-(B) Influenza A virus RNA polymerase activity on a neuraminidase (NA) encoding vRNA template in HEK 293T cells.
-Steady state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging.
+IFN-inducing mutations D27N and T677A in the PB1 protein affect polymerase activity \textit{in vitro}. 
+(A) Model of bat influenza A virus RNA polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the influenza B virus RNA polymerase (PDB 5MSG)~\cite{reich2017vitro}.
+The locations of PB1 D27 and T677 (both red) relative to the 5' (blue) and 3' (orange) termini of the RNA template and the PB1 active site (grey; PB1 act) are indicated.
+The PA endonuclease (green; PA endo) and PB2 cap binding domain (pink; PB2 cap) are also indicated. 
+Part of the fingers subdomain of PB1 is hidden to reveal the template in the entry channel. 
+(B) Polymerase activity on full-length NA vRNA template in 293T cells.
+Steady-state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging. 
 A PB1 active site control (PB1a) was used as negative control and background correction. 
 The 5S rRNA signal was used as loading control. 
-Other panels show western blot analysis of PB1, NP and GAPDH expression. 
-Graph shows the average steady state RNA levels of three biological repeats. 
-(C) Panel shows PAGE analysis of RT-PCR of short (17-200 nt long) influenza virus RNAs or mini viral RNAs (mvRNAs).
-The length of the mini viral RNAs was extended by the adapters (adapt) present in the primers.
-Graph shows the IFN-beta promoter activity in transfected HEK 293T cells as measured using a dual luciferase reporter assay.
-(D) Influenza A virus polymerase activity on a segment 5-derived 246 nt long vRNA template. 
-The top panel shows the steady state levels of vRNA template were analysed by primer extension and denaturing PAGE. 
-The other two panels show the PB1 and tubulin expression levels analysed by western blot. 
-The top graph shows the average steady state RNA levels of three biological repeats. 
+Other panels show Western blot analysis of PB1, NP and GAPDH protein expression. 
+Graph shows the average of three biological repeats. 
+(C) Panel shows PAGE analysis of RT-PCR of mvRNAs.
+The same amount of total input RNA was used for each sample.
+The length of the mmvRNAs was extended by the adapters (adapt) present in the primers.
+Graph shows the IFN-beta promoter activity in transfected 293T cells as measured using a dual luciferase reporter assay.
+(D) Polymerase activity on a short 246-nucleotide vRNA template. 
+The top panel shows the steady state levels of vRNA template as determined by primer extension and denaturing PAGE. 
+The other two panels show the PB1 and tubulin expression levels analyzed by western blot. 
+The top graph shows the average of three biological repeats. 
 The second graph shows the IFN-beta promoter activity in transfected HEK 293T cells as measured using a dual luciferase reporter assay. 
@@
 P-values were determined by two-sided t-test.
+{\color{red} Legend will be further edited and figure possibly re-arranged after all data are added.}
 }
@@
 
-Previous research has shown that aberrant RNA products, or mini viral RNAs (mvRNAs) {te Velthuis 2018}, produced during influenza A virus replication can trigger RIG-I activation and IFN production.
-To investigate if the PB1 mutations D27N and T667A could affect influenza A virus RNA polymerase activity we first identified their location and interactions with the template by superposing the pre-initiation complex of the bat influenza A virus RNA polymerase with the transcription initiation complex of the influenza B virus RNA polymerase (Fig.
-7A).
-The model shows how the 3' terminus of the template enters the template entry channel above the PB1 active site and that T667A sits at tip of a helix that interacts with the 3' terminus as it enters the channel.
-By contrast, D27N resides deeper in the polymerase and close to the binding pocket of the 5' terminus.
-These observations suggest that both mutations may affect polymerase activity by changing the interactions of the RNA polymerase with the template.
-To investigate how the two PB1 mutations affect polymerase activity on a full-length viral RNA template, we expressed the wildtype or mutant PB1 subunits along with PB2, PA, NP and the segment 6 vRNA in HEK 293T cells.
-Twenty-four hours after transfection, we extracted the total RNA and analysed the replication (vRNA level) and transcriptional (mRNA level) activity by primer extension.
-Interestingly, both mutations showed activities that were significantly different from wildtype (Fig.
-7B), with D27N and T667A showing lower and higher viral replication and transcription, respectively.
-Both PB1 mutants showed higher activities than the PB1 active site control (PB1a) that we used to subtract background RNA levels.
-Moreover, both mutants were expressed to similar levels, as assessed by western blot.
-To investigate if the mutants produced any mvRNAs, we fractionated the total RNA to enrich for small RNA species (17-200 nt long) and performed an RT-PCR that amplified RNA species that contained both the 3' and 5' terminus of the template on equal amounts of RNA.
-This revealed that the D27N expressing cells contained higher levels of mvRNAs than the wildtype control and the T667A mutant (Fig.
-7C).
-Analysis of the IFN levels showed that...
-The low activity on the full-length segment 6 vRNA and higher mvRNA level produced on this template suggest that the D27N mutation in PB1 impaired polymerase processivity on full-length vRNAs.
-To confirm this, we replaced the segment 6 template with a segment 5-derived 246-nt long template (Fig.
-7D).
-Under these conditions, the activity of the D27N mutant was similar to wildtype, while the activity of the T677A mutant proved again higher than wildtype, although not significantly higher in three biological repeats.
-Analysis of the IFN expression level in response to viral replication on the short template showed higher IFN induction by both PB1 mutants.
+To test if the PB1 mutations affect polymerase activity on a full-length viral RNA template, we expressed wild-type or mutant PB1 protein along with the other proteins in the polymerase complex (PB2, PA, and NP) and the viral RNA (vRNA) for the NA segment in 293T cells.
+Twenty-four hours after transfection, we extracted total RNA and analyzed replication (vRNA level) and transcription (mRNA level) activity by primer extension.
+Both mutations had activities that were significantly different from wild type, despite being expressed at wild-type protein levels (\FIG{PB1}B).
+Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both activities---although D27N still retained activity far in excess of a control active-site mutant (\FIG{PB1}B).
+To test if these activity changes increased production of immunostimulatory mvRNAs~\cite{velthuis2018mini}, we fractionated the total RNA to enrich for small species (17-200 nt long) and used reverse transcription followed by PCR (RT-PCR) to amplify species with both the 3' and 5' vRNA termini.
+The D27N mutant produced more mvRNAs than wild type, while the T667A mutant produced the same low levels of mvRNAs as wild type (\FIG{PB1}C).
+{\color{red} Analysis of the IFN levels in these cells showed that... \textit{(Still need to figure out how to work this in after experiments done.)} }
+
+Because the D27N PB1 mutant might produced more mvRNAs and had reduced activity on a full-length vRNA template, we speculated that it might have reduced polymerase processivity.
+To confirm this, we repeated the \textit{in vitro} polymerase activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (\FIG{PB1}D).
+On this shorter template, the activity of the D27N mutant was similar to wild type, while the activity of the T677A mutant remained higher than wildtype (although not significantly higher in three biological repeats).
+{\color{red} Analysis of the IFN expression level in response to viral replication on the short template showed higher IFN induction by both PB1 mutants.
+\textit{(Need to work in prior sentence and add a wrap up sentence; will wait to do this until we have all the IFN results.
+End with something about how even PB1 point mutants act by different mechanisms (more activity versus more mvRNAs) to tie into next paragraph.)}}
+
+%%%%% begin IFN_vs_HA figure %%%%%%%%
+\begin{figure}
+\centerline{\includegraphics[width=0.35\textwidth, valign=t]{figures/MutantComparison/p_foldchange.pdf}}
+\caption{
+Cells that express higher levels of HA protein are much more likely to induce IFN expression only if they are infected by virus with defects in NS1.
+The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile among infected cells as determined by flow cytometry.
+Points indicate replicates, and lines indicate the mean.
+This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in \SUPPFIG{validationflow}; see \SUPPFIG{IFN_vs_HA} for a more detailed breakdown of the frequency of IFN induction as a function of HA expression.
+}
+\label{fig:IFN_vs_HA}
+\end{figure}
+%%%%% end IFN_vs_HA figure %%%%%%%%
+
+The diversity of mechanisms of IFN induction extends to other types of viral defects.
+For instance, the single-cell data indicates that increased viral gene expression is only associated with increased IFN induction when the infecting virion fails to express NS (\FIG{mutations}D).
+To confirm this observation by an independent method, we performed a detailed analysis of the flow cytometry underlying \FIG{validation} to determine relationship between IFN induction and expression of the viral HA protein.
+\FIG{IFN_vs_HA} shows that cells that express more HA are much more likely to turn IFN+ if they are infected with the NS1stop or NS1-A122V mutants, but not if they are infected with any of the other viral variants.
+Taken together with the single-cell data, this observation suggests that only in the absence of functional NS1 does higher viral gene expression predispose an infected cell to induce IFN.
+Therefore, the diverse range of viral defects that induce IFN act by diverse processes, showing that viral variation influences not only the rate of IFN induction but also the factors that contribute to this induction.
 
@@
 \small
-\subsection{IFN reporter cell lines}
+\subsection{IFN reporter cell lines.}
 We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}A).
@@
 
-\subsection{Viruses for single-cell experiments}
+\subsection{Viruses for single-cell experiments.}
 We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
@@
 
-\subsection{Flow cytometry analyses for HA expression}
+\subsection{Flow cytometry analyses for HA expression.}
 For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
@@
 
-\subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium}
+\subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium.}
 The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-P2A-mNeonGreen reporter.
@@
 
-\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing}
+\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing.}
 We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
@@
 
-\subsection{qPCR for viral genes and IFN}
+\subsection{qPCR for viral genes and IFN.}
 We performed qPCR on reverse-transcribed mRNA for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
@@
 
-\subsection{Viruses and experiments for validation experiments}
+\subsection{Viruses and experiments for validation experiments.}
 In \FIG{validation}, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
@@
 
-\subsection{Computational analysis of single-cell transcriptomic and viral sequence data}
+\subsection{Computational analysis of single-cell transcriptomic and viral sequence data.}
 A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}. 
@@
 
-\subsection{Structural analysis of PB1 mutants}
+\subsection{Structural analysis of PB1 mutants.}
 To locate the PB1 mutations in the influenza A virus RNA polymerase structure relative to the template and active site in \FIG{PB1}A, we superposed the bat influenza A virus RNA polymerase structure (PDB 4WSB)~\cite{reich2014structural}, which shows the 3' terminus of the template on the surface of the RNA polymerase, with the influenza B virus transcription initiation complex (PDB 5MSG)~\cite{reich2017vitro}, which shows the 3' terminus of the template in the template entry channel that leads towards the active site.
@@
 
-\subsection{Experimental analysis of PB1 mutants}
+\subsection{Experimental analysis of PB1 mutants.}
 For the experimental analysis of the PB1 mutants in \FIG{PB1}, we used plasmids pcDNA-PB1, pcDNA-PA, pcDNA-PB2, and pcDNA-NP, which encode the WSN proteins that compose the polymerase complex~\cite{fodor2002single}; pPolI-NA, which encodes the viral RNA for the WSN NA~\cite{fodor2002single}; and pcDNA-PB1a, which encodes an inactive version of the WSN PB1 polymerase protein~\cite{vreede2004model}.
@@
 JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
+AtV was supported by joint Wellcome Trust and Royal Society grant 206579/Z/17/Z and Isaac Newton Trust grant 17.37(r).
+EE was supported by a Vice-Chancellor’s scholarship from the University of Cambridge.
 JDB is an Investigator of the Howard Hughes Medical Institute.
@@
 \section{Supplemental Material}
-
-\begin{suppfig}
-\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}
-\caption{
-Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
-Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
-Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
-As shown in the left-most panel, the sequencing depth of \cite{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
-An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
-Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\cite{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
-However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\cite{stetson2006type,honda2006type}.
-The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
-}
-\label{suppfig:mice}
-\end{suppfig}
 
@@
 \centerline{
-{\bf \Large A}
-\includegraphics[width=0.32\textwidth, valign=t]{figures/MutantComparison/p_foldchange.pdf}
-\hspace{0.03\textwidth}
-{\bf \Large B}
-\includegraphics[width=0.55\textwidth, valign=t]{figures/MutantComparison/p_ifn_vs_ha.pdf}
-}
-\caption{
-Higher viral gene expression is strongly associated with more IFN induction only in NS1-deficient infections.
-This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in \SUPPFIG{validationflow}.
-(A)
-For viruses with defects in NS1, there is more IFN induction among infected cells that express higher levels of HA protein.
-The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile as determined by flow cytometry.
-Points indicate replicates, and lines indicate the mean.
-(B)
-A more detailed version of the data summarized in (A).
-For each virus and replicate, we binned the infected cells into HA expression quartiles.
+\includegraphics[width=0.6\textwidth, valign=t]{figures/MutantComparison/p_ifn_vs_ha.pdf}
+}
+\caption{
+A more detailed version of the data summarized in \FIG{IFN_vs_HA}.
+For each virus and replicate, we binned the infected cells in \SUPPFIG{validationflow} into HA expression quartiles based on the flow cytometry measurements.
 We then calculated the percent of cells that were IFN+ in each quartile.
@@
 The results clearly show that for the NS1stop and to a lesser extend the NS1-A122V variants, more viral protein (higher HA signal) correlations with IFN induction.
-Note that panel (A) summarizes these same data by simply showing the ratio of percent IFN+ between the highest and lowest quartile.
+\FIG{IFN_vs_HA} summarizes these same data by simply showing the ratio of percent IFN+ between the highest and lowest quartile.
 }

### 2019-03-21 paper/paper.tex
revisions to Discussion

@@
 
-\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; mini viral RNA; mvRNA; heterogeneity}
+\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; mini viral RNA; heterogeneity}
 
@@
 \section{Discussion}
-We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
-Our method builds on the observation that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{saikia2018simultaneous, zanini2018virus}.
+We have determined the full sequences of all viral genes in single influenza-infected cells.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, wang2018cell, brooke2013most, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high}.
+Our method builds on the observation that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{wang2018cell, zanini2018virus, saikia2018simultaneous}.
 To make this information complete, we have coupled single-cell transcriptomics with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.
 
-This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
-Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wild-type'' influenza genome, most infected cells do not express wild-type copies of all viral genes.
-Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
-Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a vast number of hidden viral defects.
-
-These viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
-We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
-Two types of defects---absence of the NS gene and amino-acid mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\cite{garcia1998influenza, hale2008multifunctional}.
-The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I~\cite{velthuis2018mini}.
-Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with prior work showing that such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
-Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
-There may be several reasons: we used relatively pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.
-
-But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
-We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
-Additionally, no viral defect induces IFN deterministically---and the most immunostimulatory defect (absence of NS) also occurs in multiple IFN- cells in our single-cell dataset.
-Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, and so could overlook low-frequency mutations that contribute to IFN induction.
-However, we suspect that stochasticity or pre-existing cellular states also affect the outcome of infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
+This viral genetic information is crucial for understanding infection outcome and innate-immune induction.
+Despite the fact that we used a low-passage viral stock generated from plasmids, most infected cells do not express unmutated copies of all viral genes.
+Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{russell2018extreme, heldt2015single, brooke2013most, dou2017analysis, jacobs2019incomplete}, it is the first to directly observe the full spectrum of these defects across single cells.
+Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a diverse spectrum of viral defects.
+
+We identified four types of defects that we validated to increase IFN induction. 
+Two types of defects---absence of the NS gene and amino-acid mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate immunity~\cite{garcia1998influenza, hale2008multifunctional}.
+A third type of defect, amino-acid mutations in PB1, was more surprising since this protein has not been described as a major player in innate-immune detection.
+We characterized two IFN-inducing PB1 mutations, and showed that one impaired polymerase processivity whereas the other increased overall activity.
+We speculate that these alterations increase production of aberrant immunostimulatory products such as mini-viral RNAs~\cite{velthuis2018mini} or other deleted forms of vRNAs.
+Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+In fact, given the extensive prior work on deletions, we were surprised not to identify more of them in our IFN+ cells.
+There may be several reasons: we used pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
+Additionally, the relative importance of different defects likely varies across infection conditions, viral strains, and cell types: it remains an open question which defects are most relevant for immune detection during actual human infections.
+
+However, the greatest value of our work is not as a screen for IFN-inducing defects, but rather as a relatively unbiased survey of the breadth of viral variation in individual infected cells.
+This survey shows that no single type of viral defect determines whether a cell induces IFN: even the most immunostimulatory defect (absence of NS) occurs in only about a quarter of IFN+ cells.
+Therefore, innate-immune detection of influenza is a multi-faceted process that cannot be ascribed a single dominant viral genetic cause.
+
+Our results further show that viral genetic defects only partially explain the heterogeneity among influenza-infected cells.
+There is substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
+Additionally, no viral defect induces IFN deterministically: every type of immunostimulatory defect we characterize is also observed in IFN- cells in our single-cell dataset.
+This fact shows that stochasticity or pre-existing cellular states play a major role in affecting the outcome of infection---a finding that concords with the fact that IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
 
@@
 Natural human influenza infections are established by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
-It is therefore plausible that early heterogeneity could have a large effect on downstream events.
+It is therefore plausible that early heterogeneity could affect downstream events.
 Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.

### 2019-03-21 paper/paper.tex
added revised PB1 data

@@
 \subsection{The IFN-inducing viral defects act by diverse mechanisms.}
-Some of the viral defects in IFN+ are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are known to be prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Some of the viral defects in IFN+ are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
 Other defects are more surprising: for instance, it is not obvious why amino-acid mutations in PB1 increase IFN induction.
-We therefore designed experiments to interrogate the mechanisms of some of these defects.
+We therefore designed experiments to interrogate some of these defects.
 
 We first focused on the PB1 mutations.
-Recent research has shown that mutations to the PB2 polymerase subunit of the WSN viral strain can enhance IFN induction by leading to the production of aberrant mini viral RNAs (mvRNAs) that trigger RIG-I~\cite{velthuis2018mini}.
-To investigate if the PB1 mutations act by a similar mechanism, we first examined their location in a structural model of the polymerase complex (\FIG{PB1}A).
-The IFN-enhancing PB1 mutation T677A occurs at the tip of a helix that interacts with the 3' terminus of the RNA template as it enters the channel above the PB1 active site, whereas the IFN-enhancing D27N mutation is deeper in the polymerase and close to the binding pocket of the 5' terminus of the template.
-Therefore, both mutations could plausibly affect the polymerase's activity by changing its interactions with the RNA template.
+Recent research has shown that mutations to the PB2 polymerase subunit of the WSN viral strain can enhance IFN induction by leading to production of aberrant mini viral RNAs (mvRNAs) that trigger RIG-I~\cite{velthuis2018mini}.
+To investigate if the PB1 mutations might similarly perturb polymerase activity, we examined their location in a structural model of the polymerase complex (\FIG{PB1}A).
+The IFN-enhancing PB1 mutation T677A occurs at the tip of a helix that interacts with the 3' terminus of the RNA template as it enters the channel above the active site, whereas the IFN-enhancing D27N mutation is deeper in the polymerase close to the binding pocket of the 5' terminus of the template.
+Therefore, both mutations could plausibly alter the polymerase's interactions with the RNA template.
 
@@
 
-\centerline{\includegraphics[width=0.7\textwidth]{figures/PB1_figure.png}}
-\caption{
-IFN-inducing mutations D27N and T677A in the PB1 protein affect polymerase activity \textit{in vitro}. 
-(A) Model of bat influenza A virus RNA polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the influenza B virus RNA polymerase (PDB 5MSG)~\cite{reich2017vitro}.
+\centerline{\includegraphics[width=\textwidth]{figures/PB1_figure/PB1_figure.pdf}}
+\caption{
+IFN-inducing mutations D27N and T677A in the PB1 protein affect polymerase activity. 
+(A) Model of bat influenza A virus polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the influenza B virus polymerase (PDB 5MSG)~\cite{reich2017vitro}.
 The locations of PB1 D27 and T677 (both red) relative to the 5' (blue) and 3' (orange) termini of the RNA template and the PB1 active site (grey; PB1 act) are indicated.
@@
 Part of the fingers subdomain of PB1 is hidden to reveal the template in the entry channel. 
-(B) Polymerase activity on full-length NA vRNA template in 293T cells.
+(B) IFN-beta promoter activity measured using a dual luciferase reporter assay in 293T cells transfected with plasmids expressing the indicated PB1 protein, the other polymerase complex proteins (PB2, PA, and NP), and a full-length NA vRNA template.
+PB1a is a catalytically inactive PB1 active site control.
+Points show three biological replicates; ``n.d.'' indicates not detectable and points in orange were significantly greater than wild type by a two-sided t-test.
+(C) Polymerase activity in 293T cells transfected as in panel (B).
 Steady-state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging. 
-A PB1 active site control (PB1a) was used as negative control and background correction. 
+PB1a was used as negative control and background correction. 
 The 5S rRNA signal was used as loading control. 
@@
 Graph shows the average of three biological repeats. 
-(C) Panel shows PAGE analysis of RT-PCR of mvRNAs.
-The same amount of total input RNA was used for each sample.
-The length of the mmvRNAs was extended by the adapters (adapt) present in the primers.
-Graph shows the IFN-beta promoter activity in transfected 293T cells as measured using a dual luciferase reporter assay.
+Error bars indicate standard deviation.
+P-values were determined by two-sided t-test.
 (D) Polymerase activity on a short 246-nucleotide vRNA template. 
@@
 The other two panels show the PB1 and tubulin expression levels analyzed by western blot. 
-The top graph shows the average of three biological repeats. 
-The second graph shows the IFN-beta promoter activity in transfected HEK 293T cells as measured using a dual luciferase reporter assay. 
-Error bars indicate standard deviation. 
-P-values were determined by two-sided t-test.
-{\color{red} Legend will be further edited and figure possibly re-arranged after all data are added.}
+Graph shows the average of three biological repeats. 
 }
@@
 
-To test if the PB1 mutations affect polymerase activity on a full-length viral RNA template, we expressed wild-type or mutant PB1 protein along with the other proteins in the polymerase complex (PB2, PA, and NP) and the viral RNA (vRNA) for the NA segment in 293T cells.
-Twenty-four hours after transfection, we extracted total RNA and analyzed replication (vRNA level) and transcription (mRNA level) activity by primer extension.
-Both mutations had activities that were significantly different from wild type, despite being expressed at wild-type protein levels (\FIG{PB1}B).
-Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both activities---although D27N still retained activity far in excess of a control active-site mutant (\FIG{PB1}B).
-To test if these activity changes increased production of immunostimulatory mvRNAs~\cite{velthuis2018mini}, we fractionated the total RNA to enrich for small species (17-200 nt long) and used reverse transcription followed by PCR (RT-PCR) to amplify species with both the 3' and 5' vRNA termini.
-The D27N mutant produced more mvRNAs than wild type, while the T667A mutant produced the same low levels of mvRNAs as wild type (\FIG{PB1}C).
-{\color{red} Analysis of the IFN levels in these cells showed that... \textit{(Still need to figure out how to work this in after experiments done.)} }
-
-Because the D27N PB1 mutant might produced more mvRNAs and had reduced activity on a full-length vRNA template, we speculated that it might have reduced polymerase processivity.
-To confirm this, we repeated the \textit{in vitro} polymerase activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (\FIG{PB1}D).
-On this shorter template, the activity of the D27N mutant was similar to wild type, while the activity of the T677A mutant remained higher than wildtype (although not significantly higher in three biological repeats).
-{\color{red} Analysis of the IFN expression level in response to viral replication on the short template showed higher IFN induction by both PB1 mutants.
-\textit{(Need to work in prior sentence and add a wrap up sentence; will wait to do this until we have all the IFN results.
-End with something about how even PB1 point mutants act by different mechanisms (more activity versus more mvRNAs) to tie into next paragraph.)}}
+To test if the PB1 mutations affect activity, we transfected 293T cells with plasmids that express wild-type or mutant PB1 protein along with the other proteins in the polymerase complex (PB2, PA, and NP) and full-length viral RNA (vRNA) for the NA segment.
+Both polymerase mutations increased IFN expression in this assay (\FIG{PB1}B), indicating that they have an immunostimulatory effect in the context of an active viral polymerase even when other viral components are absent.
+We next directly measured polymerase activity on the full-length vRNA template by extracting total RNA and quantifying replication (vRNA) and transcription (mRNA) products by primer extension.
+Both immunostimulatory PB1 mutations had activities that were significantly different from wild type, despite being expressed at wild-type protein levels (\FIG{PB1}B).
+Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both---although D27N still retained activity far in excess of a control active-site mutant (\FIG{PB1}C).
+We speculated that the mutations might reduce polymerase processivity, leading to accumulation of mvRNAs~\cite{velthuis2018mini} or other aberrant short RNA products that might activate the innate immune system~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+We therefore repeated the activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (\FIG{PB1}D).
+On this shorter template, the activity of the D27N mutant was now similar to wild type, while the activity of the T677A mutant remained higher than wildtype (although not significantly so in three biological repeats).
+Therefore, the two immunostimulatory PB1 mutations have distinct effects on the polymerase: D27N reduces processivity thereby favoring shorter RNA products, whereas T677A increases overall activity which could also lead to accumulation of aberrant RNA products.
 
@@
 \caption{
-Cells that express higher levels of HA protein are much more likely to induce IFN expression only if they are infected by virus with defects in NS1.
+Infected cells that express higher levels of HA protein are much more likely to induce IFN expression only if they are infected by virus with defects in NS1.
 The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile among infected cells as determined by flow cytometry.
@@
 
-To analyse the activity of the PB1 mutants in cell culture, the plasmids expressing the WSN PA, PB2, NP and PB1 proteins were transfected into 293T cells together with the plasmid expressing the wildtype NA vRNA or a 246-nt long segment 5-based template~\cite{velthuis2018mini}.
+To analyze the activity of the PB1 mutants in cell culture, the plasmids expressing the WSN PA, PB2, NP and PB1 proteins were transfected into 293T cells together with the plasmid expressing the wildtype NA vRNA or a 246-nt long segment NP-based template~\cite{velthuis2018mini}.
 Twenty-four hours post transfection, the RNA was extracted using Trizol (Invitrogen), and the steady state RNA levels assessed using reverse transcription with 32P-labelled oligonucleotides against the viral RNA species and ribosomal 5S RNA as described previously~\cite{velthuis2018mini, velthuis2016role}.

### 2019-03-21 paper/paper.tex
add a few refs

@@
 Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
-This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell,sjaastad2018distinct}.
 Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
@@
 Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
-Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{killip2017single, velthuis2018mini, du2018genome, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{killip2017single, velthuis2018mini, du2018genome, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
-Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, steuerman2018dissection, wang2018cell, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west}.
+Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single, sjaastad2018distinct}, and current single-cell transcriptomic techniques measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, steuerman2018dissection, wang2018cell, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation.
@@
 
-Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce higher levels of IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce higher levels of IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
@@
 \subsection{The IFN-inducing viral defects act by diverse mechanisms.}
-Some of the viral defects in IFN+ are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Some of the viral defects in IFN+ are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 Other defects are more surprising: for instance, it is not obvious why amino-acid mutations in PB1 increase IFN induction.
@@
 Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both---although D27N still retained activity far in excess of a control active-site mutant (\FIG{PB1}C).
-We speculated that the mutations might reduce polymerase processivity, leading to accumulation of mvRNAs~\cite{velthuis2018mini} or other aberrant short RNA products that might activate the innate immune system~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+We speculated that the mutations might reduce polymerase processivity, leading to accumulation of mvRNAs~\cite{velthuis2018mini} or other aberrant short RNA products that might activate the innate immune system~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 We therefore repeated the activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (\FIG{PB1}D).
@@
 We have determined the full sequences of all viral genes in single influenza-infected cells.
-Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, wang2018cell, brooke2013most, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high}.
+Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\cite{russell2018extreme, steuerman2018dissection, heldt2015single, wang2018cell, sjaastad2018distinct, brooke2013most, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west, zhu2009growth, schulte2014single, akpinar2016high}.
 Our method builds on the observation that fragmentary viral genetic information can be obtained by more standard single-cell transcriptomic techniques~\cite{wang2018cell, zanini2018virus, saikia2018simultaneous}.
@@
 We speculate that these alterations increase production of aberrant immunostimulatory products such as mini-viral RNAs~\cite{velthuis2018mini} or other deleted forms of vRNAs.
-Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
+Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 In fact, given the extensive prior work on deletions, we were surprised not to identify more of them in our IFN+ cells.

### 2019-03-21 paper/paper.tex
shortened PacBio methods

@@
 Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
-This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell,sjaastad2018distinct}.
+This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell}.
 Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
@@
 However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
-Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single, sjaastad2018distinct}, and current single-cell transcriptomic techniques measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, steuerman2018dissection, wang2018cell, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west}.
+Flow cytometry and fluorescent reporters only measure protein levels~\cite{sjaastad2018distinct, brooke2013most, guo2017single}, and current single-cell transcriptomic techniques measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, steuerman2018dissection, wang2018cell, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west}.
 None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation.
@@
 There may be several reasons: we used pure viral stocks~\cite{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
-Additionally, the relative importance of different defects likely varies across infection conditions, viral strains, and cell types: it remains an open question which defects are most relevant for immune detection during actual human infections.
+Additionally, the relative importance of different defects likely varies across infection conditions, viral strains, and cell types: it is an open question which defects are most relevant for immune detection during actual human infections.
 
@@
 There is substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
-Additionally, no viral defect induces IFN deterministically: every type of immunostimulatory defect we characterize is also observed in IFN- cells in our single-cell dataset.
+Additionally, no viral defect induces IFN deterministically: every type of immunostimulatory defect that we characterize is also observed in IFN- cells in our single-cell dataset.
 This fact shows that stochasticity or pre-existing cellular states play a major role in affecting the outcome of infection---a finding that concords with the fact that IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
@@
 \subsection{Enrichment and preparation of viral cDNA for PacBio sequencing.}
-We amplified virus-derived molecules from the small amount of cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
-All molecules in this cDNA have at their 3' end the cell barcode and UMI plus the constant adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, or the detailed analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} for at detailed schematic).
-However, we only wanted to PacBio sequence cDNA molecules derived from virus, since it would be prohibitively expensive to sequence all molecules.
-We therefore needed to enrich for the viral molecules, a process made challenging by the need to also retain the 10X adaptor / UMI / cell barcode at the 3' end (the primer at the 5' end can be specific, but the one at the 3' end must bind the common 10X adaptor shared among all molecules).
-
-We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (although some viral segments produce multiple splice forms, all mRNAs from a given segment share the same 5' end).
-The sequences of these primers are in \SUPPFILE{primer_sequences}.
+We amplified virus-derived molecules from cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
+These cDNA have at their 3' end the cell barcode and UMI plus the adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, and \SUPPFILE{pacbio_analysis} for more details).
+We only wanted to PacBio sequence cDNA molecules derived from virus.
+We therefore needed to enrich for the viral molecules while retaining the 10X adaptor / UMI / cell barcode at the 3' end.
+
+We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (\SUPPFILE{primer_sequences}).
 A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
-To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
-Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
-This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{boers2015micelle}.
+To reduce strand-exchange and obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon), which physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents~\cite{boers2015micelle}.
 We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume of 100$\mu$l.
@@
 This melting temperature is lower than the standard 98$^{\circ}$C melting step suggested by the manufacturer for Kapa HiFi because we wanted to avoid collapse of emulsion integrity at high temperature.
-We performed 30 cycles in order to saturate the reagents in each emulsion---unlike for open PCR reactions, the physical occlusion of amplified material in emulsion PCR reduces artifacts at high cycle numbers.
-
-Because were were concerned that the multiplex PCR would still result in highly uneven amplification of different influenza cDNAs due to differences in their expression levels and lengths, the product of this multiplex PCR was subjected to eight additional individual emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 3' primer, using 1 ng of material in each reaction.
+
+The product of this multiplex PCR was subjected to eight additional individual emulsion PCR reactions, each using only a single segment-specific 5' primer as well as the common 3' primer, using 1 ng of material in each reaction.
 The material from these eight segment-specific PCRs was then pooled with the goal of obtaining a equimolar ratio of segments, and sequenced on one SMRT Cell in a PacBio RS II and one SMRT Cell of a PacBio Sequel.
-Detailed results from the analysis of these first two sequencing runs is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} and \SUPPFILE{pacbio_analysis}.
-These results showed that although the PCRs substantially enriched for influenza molecules, the relative coverage of the different viral genes was still very uneven, with the longer genes (especially the polymerase genes) severely under-sampled.
-
-To try to improve coverage of the polymerase genes, we produced two new sequencing pools: one consisting of the five shortest viral segments (HA, NP, NA, M, and NS) from the aforementioned segment-specific emulsion PCRs, and the other consisting of the three longer polymerase segments (PB2, PB1, and PA).
+Detailed results from the analysis of these first two sequencing runs is shown in \SUPPFILE{pacbio_analysis}.
+These results showed that although the PCRs substantially enriched for influenza molecules, the relative coverage of the different viral genes was still uneven, with the longer genes under-sampled.
+
+To improve coverage of the polymerase genes, we produced two new sequencing pools: one consisting of the five shortest viral segments (HA, NP, NA, M, and NS) from the aforementioned segment-specific emulsion PCRs, and the other consisting of the three longer polymerase segments (PB2, PB1, and PA).
 The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), even with these new sequencing runs the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes, which arise commonly during influenza replication~\cite{xue2016propagation} and are presumably preferentially amplified during PCR.
-
-To obtain more reads for longer full-length polymerase variants, we therefore subjected 10 ng of our amplified material for each polymerase segment to a bead selection using SPRIselect beads at a volume ratio of 0.4 to select for larger molecules.
+As is shown \SUPPFILE{pacbio_analysis}, the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes~\cite{xue2016propagation} which are preferentially amplified during PCR.
+
+To obtain more reads for longer full-length polymerase variants, we therefore subjected 10 ng of our amplified material for each polymerase segment to a bead selection using SPRIselect beads at a volume ratio of 0.4.
 This selection removes most low-molecular weight DNA species including internally-deleted defective segments.
-Material from this selection was then amplified using 16 (PB1) or 14 (PB2 and PA) cycles of a non-emulsion PCR using the standard conditions recommended by the Kapa HiFi Hotstart ReadyMix (extension at 67$^{\circ}$C for 2 minutes 15 seconds, and melting at 98$^{\circ}C$).
-The use of relatively few PCR cycles was designed to prevent the occurrence of the artifacts (including strand exchange) that occur in non-emulsion PCRs as the reactions approach saturation.
+Material from this selection was amplified using 16 (PB1) or 14 (PB2 and PA) cycles of a non-emulsion PCR using the standard conditions recommended by the Kapa HiFi Hotstart ReadyMix (extension at 67$^{\circ}$C for 2 minutes 15 seconds, and melting at 98$^{\circ}C$).
+The use of relatively few PCR cycles was designed to prevent the occurrence of the artifacts (including strand exchange) that occur in non-emulsion PCRs.
 We pooled the products of these reactions from this size-selection and sequenced on a SMRT Cell of a PacBio Sequel.
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), this sequencing yielded modestly more full-length polymerase variants, but they were still heavily undersampled compared to other viral genes.
-
-To further to improve recovery of full-length PB1, PB2, and PA, we therefore took an alternative approach that allowed us to perform a specific PCR for full-length polymerase variants.
-Specifically, we circularized the template molecules (including the full length of genes plus the cell barcode, UMI, and 10X adaptor), and then used two segment-specific primers that annealed in apposition near the center of each polymerase gene to linearize these circular molecules.
+As is shown in \SUPPFILE{pacbio_analysis}, this sequencing yielded more full-length polymerase variants, but they were still undersampled compared to other viral genes.
+
+To further to improve recovery of full-length PB1, PB2, and PA, we use an approach that allowed us to perform a specific PCR for full-length polymerase variants.
+We circularized the template molecules, and then used two segment-specific primers that annealed in apposition near the center of each polymerase gene to linearize these circular molecules.
 Only molecules that contain the middle of the polymerase genes (which are typically full-length) are linearized by this process.
@@
 This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRT Cell of PacBio Sequel. 
-As is shown in the PacBio analysis notebook at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}), this process efficiently yielded many full-length polymerase variants.
-
-The computational analyses of the full-length viral gene sequences described below used the combination of the data from all of these reactions.
+As is shown in \SUPPFILE{pacbio_analysis}, this process yielded many full-length polymerase variants.
+
+The computational analyses of the full-length viral gene sequences described below combined the data from all these reactions.
 The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{StrandExchange} for an illustration of how this is determined).
-A more detailed breakdown of the coverage of each gene and data showing a consistently low rate of strand exchange for all PacBio runs is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} (see also \SUPPFILE{pacbio_analysis}).
-Importantly, the various PCR biases and enrichment schemes mean that the coverage of molecules by the PacBio sequencing is not proportional to their original abundance in the starting mRNA.
+A detailed breakdown of the coverage of each gene and PacBio run is in \SUPPFILE{pacbio_analysis}.
+Importantly, the PCR biases mean that the coverage of molecules by the PacBio sequencing is not proportional to their abundance in the starting mRNA.
 However, as described in the computational analysis section below, the final analyses use the cell barcodes and UMIs in conjunction with the standard 10X Illumina sequencing to ensure that none of the conclusions are affected by the disproportionate amplification of some molecules during the PacBio library preparation (for instance, duplicate UMIs are removed from the PacBio data, and all conclusions about gene abundance or absence are based on the Illumina data).
@@
 \end{enumerate}
-The plots in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb} or \SUPPFILE{pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
+The plots in \SUPPFILE{pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
 We could call the sequences of all expressed viral genes in about half of the infected cells (\SUPPFIG{ncells}).

### 2019-03-23 paper/paper.tex
minor comments from AJ te Velthuis

@@
 Here we develop a new method to assess how the genetic variation in viral populations contributes to this heterogeneity.
-We do this by determining the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
+We do this by determining the transcriptome and full-length sequences of all viral genes in single cells infected with a well-defined stock of influenza virus.
 Most cells are infected by virions with defects, some of which increase the frequency of innate-immune activation.
@@
 
-\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; mini viral RNA; heterogeneity}
+\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; defective virus; heterogeneity}
 
@@
 
-Here we develop a new approach measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
+Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
 To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
@@
 
-We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
+We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN'') directly from reverse-genetics plasmids~\cite{hoffmann2000dna}, and passaged this stock at low multiplicity of infection (MOI).
+This process ensures that the viral stock that is relatively ``pure'', with only low levels of the large internal deletions and other defects that arise in stocks passaged at high MOI~\cite{xue2016propagation}.
+As described in the next subsection, our stock actually consisted of a mix of two viruses: wild-type WSN and a variant of this virus that carries synonymous viral ``barcodes'' near the termini of each gene.
+This viral stock activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
 We also validated that MACS for the cell-surface protein driven by the IFN reporter enriched the IFN+ cells by $>$50-fold (\SUPPFIG{MACS}).
@@
 \subsection{Combined transcriptomics and virus-sequencing of single infected cells.}
+
+We developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
+First, we generated the viral stock described in the previous subsection, which consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
+These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
+We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells.
+From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
+To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
+We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
 
@@
 %%% end workflow figure
-
-We developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
-First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
-These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
-We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells.
-From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
-To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
-We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
 
@@
 We first focused on the PB1 mutations.
-Recent research has shown that mutations to the PB2 polymerase subunit of the WSN viral strain can enhance IFN induction by leading to production of aberrant mini viral RNAs (mvRNAs) that trigger RIG-I~\cite{velthuis2018mini}.
+Recent research has shown that mutations to the PB2 polymerase subunit of the WSN viral strain can enhance IFN induction by leading to production of aberrant RNAs that trigger RIG-I~\cite{velthuis2018mini, du2018genome}.
 To investigate if the PB1 mutations might similarly perturb polymerase activity, we examined their location in a structural model of the polymerase complex (\FIG{PB1}A).
@@
 PB1a is a catalytically inactive PB1 active site control.
-Points show three biological replicates; ``n.d.'' indicates not detectable and points in orange were significantly greater than wild type by a two-sided t-test.
-(C) Polymerase activity in 293T cells transfected as in panel (B).
+In this panel and the next two panels, points show three biological replicates; ``n.d.'' indicates not detectable, and orange indicates a variant was significantly different than wild type by a two-sided t-test.
+(C) Polymerase activity on full-length vRNA template in 293T cells transfected as in panel (B).
 Steady-state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging. 
@@
 The 5S rRNA signal was used as loading control. 
-Other panels show Western blot analysis of PB1, NP and GAPDH protein expression. 
-Graph shows the average of three biological repeats. 
-Error bars indicate standard deviation.
-P-values were determined by two-sided t-test.
+Other panels show Western blot analysis of PB1, NP and GAPDH protein expression, and the graph at the bottom shows quantification by phosphoroimaging. 
 (D) Polymerase activity on a short 246-nucleotide vRNA template. 
 The top panel shows the steady state levels of vRNA template as determined by primer extension and denaturing PAGE. 
-The other two panels show the PB1 and tubulin expression levels analyzed by western blot. 
-Graph shows the average of three biological repeats. 
+The other two panels show the PB1 and tubulin expression levels analyzed by Western blot, and the graph shows quantification.
 }
@@
 Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both---although D27N still retained activity far in excess of a control active-site mutant (\FIG{PB1}C).
-We speculated that the mutations might reduce polymerase processivity, leading to accumulation of mvRNAs~\cite{velthuis2018mini} or other aberrant short RNA products that might activate the innate immune system~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
+We speculated that the mutations might alter polymerase processivity, leading to accumulation of aberrant RNA products that might activate the innate immune system~\cite{velthuis2018mini, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 We therefore repeated the activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (\FIG{PB1}D).
@@
 We characterized two IFN-inducing PB1 mutations, and showed that one impaired polymerase processivity whereas the other increased overall activity.
-We speculate that these alterations increase production of aberrant immunostimulatory products such as mini-viral RNAs~\cite{velthuis2018mini} or other deleted forms of vRNAs.
+We speculate that these alterations increase production of aberrant immunostimulatory RNA products~\cite{velthuis2018mini}.
 Finally, we found an internal deletion in PB1 that enhances IFN induction, consistent with prior work showing such deletions are immunostimulatory~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
@@
 The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\cite{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
+After being passaged independently, the two viral stocks were combined at equivalent numbers of infectious units to make the stock used in the single-cell experiments.
 

### 2019-03-24 paper/paper.tex
re-ordered last section of Results

@@
 Here we develop a new method to assess how the genetic variation in viral populations contributes to this heterogeneity.
-We do this by determining the transcriptome and full-length sequences of all viral genes in single cells infected with a well-defined stock of influenza virus.
+We do this by determining the transcriptome and full-length sequences of all viral genes in single cells infected with a nominally ``pure'' stock of influenza virus.
 Most cells are infected by virions with defects, some of which increase the frequency of innate-immune activation.
@@
 \subsection{The IFN-inducing viral defects act by diverse mechanisms.}
-Some of the viral defects in IFN+ are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
+Some of the viral defects in IFN+ cells are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 Other defects are more surprising: for instance, it is not obvious why amino-acid mutations in PB1 increase IFN induction.
-We therefore designed experiments to interrogate some of these defects.
-
-We first focused on the PB1 mutations.
-Recent research has shown that mutations to the PB2 polymerase subunit of the WSN viral strain can enhance IFN induction by leading to production of aberrant RNAs that trigger RIG-I~\cite{velthuis2018mini, du2018genome}.
-To investigate if the PB1 mutations might similarly perturb polymerase activity, we examined their location in a structural model of the polymerase complex (\FIG{PB1}A).
+We therefore designed experiments to interrogate some of these defects in more detail.
+
+%%%%% begin IFN_vs_HA figure %%%%%%%%
+\begin{figure}
+\centerline{\includegraphics[width=0.35\textwidth, valign=t]{figures/MutantComparison/p_foldchange.pdf}}
+\caption{
+Infected cells that express higher levels of HA protein are much more likely to induce IFN expression only if they are infected by virus with defects in NS1.
+The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile.
+Points indicate replicates, and lines indicate the mean.
+This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in \SUPPFIG{validationflow}; see \SUPPFIG{IFN_vs_HA} for more details.
+}
+\label{fig:IFN_vs_HA}
+\end{figure}
+%%%%% end IFN_vs_HA figure %%%%%%%%
+
+We first focused on one of the strongest trends from the single cell data: increased viral gene expression is associated with increased IFN induction when the infecting virion fails to express NS, but not otherwise (\FIG{mutations}D).
+To confirm this observation, we performed a flow cytometry analysis of the reporter cells infected by different immunostimulatory viral mutants to examine the association between expression of a viral gene product (HA protein) and IFN induction.
+Consistent with the single-cell data, cells that expressed more HA were much more likely to turn IFN+ when infected with the NS1stop or NS1-A122V mutants, but not when infected with any of the other viral variants (\FIG{IFN_vs_HA}).
+This fact suggests that when there are high levels of viral transcription, NS1 becomes more important as a buffer against detection of viral products.
+
+We hypothesized that the immunostimulatory mutations to PB1 might cause the viral polymerase to produce aberrant products, in line with recent work showing that mutations to PB2 can lead to the generation of aberrant RNAs that trigger RIG-I~\cite{velthuis2018mini, du2018genome}.
+To investigate if the PB1 mutations might perturb polymerase activity, we examined their location in a structural model of the polymerase complex (\FIG{PB1}A).
 The IFN-enhancing PB1 mutation T677A occurs at the tip of a helix that interacts with the 3' terminus of the RNA template as it enters the channel above the active site, whereas the IFN-enhancing D27N mutation is deeper in the polymerase close to the binding pocket of the 5' terminus of the template.
@@
 Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both---although D27N still retained activity far in excess of a control active-site mutant (\FIG{PB1}C).
-We speculated that the mutations might alter polymerase processivity, leading to accumulation of aberrant RNA products that might activate the innate immune system~\cite{velthuis2018mini, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
+We speculated that the mutations might alter polymerase processivity, leading to accumulation of aberrant RNA products that activate the innate immune system~\cite{velthuis2018mini, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
 We therefore repeated the activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (\FIG{PB1}D).
@@
 
-%%%%% begin IFN_vs_HA figure %%%%%%%%
-\begin{figure}
-\centerline{\includegraphics[width=0.35\textwidth, valign=t]{figures/MutantComparison/p_foldchange.pdf}}
-\caption{
-Infected cells that express higher levels of HA protein are much more likely to induce IFN expression only if they are infected by virus with defects in NS1.
-The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile among infected cells as determined by flow cytometry.
-Points indicate replicates, and lines indicate the mean.
-This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in \SUPPFIG{validationflow}; see \SUPPFIG{IFN_vs_HA} for a more detailed breakdown of the frequency of IFN induction as a function of HA expression.
-}
-\label{fig:IFN_vs_HA}
-\end{figure}
-%%%%% end IFN_vs_HA figure %%%%%%%%
-
-The diversity of mechanisms of IFN induction extends to other types of viral defects.
-For instance, the single-cell data indicates that increased viral gene expression is only associated with increased IFN induction when the infecting virion fails to express NS (\FIG{mutations}D).
-To confirm this observation by an independent method, we performed a detailed analysis of the flow cytometry underlying \FIG{validation} to determine relationship between IFN induction and expression of the viral HA protein.
-\FIG{IFN_vs_HA} shows that cells that express more HA are much more likely to turn IFN+ if they are infected with the NS1stop or NS1-A122V mutants, but not if they are infected with any of the other viral variants.
-Taken together with the single-cell data, this observation suggests that only in the absence of functional NS1 does higher viral gene expression predispose an infected cell to induce IFN.
-Therefore, the diverse range of viral defects that induce IFN act by diverse processes, showing that viral variation influences not only the rate of IFN induction but also the factors that contribute to this induction.
+Overall, the results in this section show that the diverse range immunostimulatory viral defects identified in single cells act by diverse processes, showing that viral variation influences not only the rate of IFN induction but also the factors that contribute to this induction.
 

### 2019-03-24 paper/paper.tex
Alistair edits

@@
 (B)
-Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
-See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
+Similar to the first panel, but validates increased IFN induction for a large internal deletion in the PB1 gene, and normalizes infecting virion dose rather than calling IFN+ percentage only among infected cells.
+See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for details.
 The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
@@
 (B) The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
-Both facts makes sense.
+Both facts make sense.
 Cells with more viral mRNA are more likely to have their viral cDNA captured in the PacBio sequencing, which is only captures a small fraction of the total transcripts identified by the 3'-end sequencing transcriptomic sequencing.
@@
 The plots show the mean HA expression of the quartile versus the precent of cells that are IFN+.
-The results clearly show that for the NS1stop and to a lesser extend the NS1-A122V variants, more viral protein (higher HA signal) correlations with IFN induction.
+The results clearly show that for the NS1stop and to a lesser extent the NS1-A122V variants, more viral protein (higher HA signal) correlates with IFN induction.
 \FIG{IFN_vs_HA} summarizes these same data by simply showing the ratio of percent IFN+ between the highest and lowest quartile.

### 2019-04-25 paper/paper_JVI_orig_submission.tex
made file with original submission for track changes diffs

--- 2019-03-24 paper/paper.tex
@@
 AtV was supported by joint Wellcome Trust and Royal Society grant 206579/Z/17/Z and Isaac Newton Trust grant 17.37(r).
-EE was supported by a Vice-Chancellor’s scholarship from the University of Cambridge.
+EE was supported by a Vice-Chancellor's scholarship from the University of Cambridge.
 JDB is an Investigator of the Howard Hughes Medical Institute.
@@
 \caption{
-Sequences of the IFN reporters in \FIG{IFNrare}A.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
+Sequences of the IFN reporters in \FIG{IFNrare}A are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
 }
@@
 \caption{
-Genbank files giving sequences of the wild-type and synonymously barcoded viruses.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
+Genbank files giving sequences of the wild-type and synonymously barcoded viruses are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
 }
@@
 \caption{
-A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt} 
+A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
 }
@@
 \caption{
-A CSV file giving the genotypes in \FIG{genotypes}.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.
+A CSV file giving the genotypes in \FIG{genotypes} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.
 }
@@
 \caption{
-A CSV file giving the viral mutations and related information in \FIG{mutations}.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.
+A CSV file giving the viral mutations and related information in \FIG{mutations} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.
 }
@@
 \caption{
-Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna}.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}
+Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}.
 }
@@
 \caption{\label{suppfile:pacbio_analysis}
-HTML rendering of Jupyter notebook that analyzes the PacBio data.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/pacbio_single_cell_figures/pacbio_analysis.html}.
-The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.
+The Jupyter notebook that analyzes the PacBio data is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.
 }
@@
 \caption{\label{suppfile:monocle_analysis}
-HTML rendering of Jupyter notebook that analyzes the annotated cell-gene matrix.
-%See \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/raw/master/paper/figures/single_cell_figures/monocle_analysis.html}.
-The actual Jupyter notebook is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.
+The Jupyter notebook that analyzes the annotated cell-gene matrix is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.
 }

### 2019-04-25 paper/paper.tex
added responses to Reviewer 1

@@
 \begin{abstract}
-The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.
+Influenza-infected cells vary widely in their expression of viral genes, and only occasionally activate innate immunity.
 Here we develop a new method to assess how the genetic variation in viral populations contributes to this heterogeneity.
@@
 However, immune activation remains stochastic in cells infected by virions with these defects, and occasionally is triggered even by virions that express unmutated copies of all genes.
-Our work shows that the diverse spectrum of defects in influenza virus populations contributes to---but does not completely explain---the heterogeneity in infection outcome and immune activation in single infected cells.
+Our work shows that the diverse spectrum of defects in influenza virus populations contributes to---but does not completely explain---the heterogeneity in viral gene expression and immune activation in single infected cells.
 
@@
 Here we develop a new approach that uses long-read PacBio sequencing to determine the sequences of virions infecting single cells.
-We show that viral genetic variation explains some but not all of the cell-to-cell variability in infection outcome.
+We show that viral genetic variation explains some but not all of the cell-to-cell variability in viral gene expression and innate-immune induction.
 Overall, our study provides the first complete picture of how viral mutations affect the course of infection in single cells.
@@
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell}.
-Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+Because viral growth and the IFN response are both feed-forward processes that amplify themselves, early variation in the initiation of these events could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically founded by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 
@@
 But for influenza, a third possible contributor also looms large: viral genetic diversity.
-Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
+The virus has evolved mechanisms to avoid IFN induction, including expressing proteins that interfere with innate-immune induction~\cite{garcia1998influenza, hale2008multifunctional,hayashi2015influenza,vreede2010mechanisms,dudek2011influenza} and sequestering immunogenic viral RNA~\cite{killip2015influenza}.
+However, because influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}, individual virions often have genetic defects that could impair these immune-evasion strategies.
 Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{killip2017single, velthuis2018mini, du2018genome, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
@@
 From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
-To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
+To ensure the presence of IFN-negative cells, we added back non-enriched cells to $\sim$10\% of the total.
 We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.
@@
 This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
-This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.
+The first explanation seems unlikely, as there is no tendency for co-infected cells to express more IFN (\SUPPFIG{coinf_IFN}); therefore we favor the latter two explanations.
+The moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.
 
@@
 Over 20\% of infected cells were IFN+, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
-Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
+The expression of type I and type III IFN was highly correlated in single cells, justifying our decision to collapse both classes under the single label of ``IFN'' in the analyses that follow (\SUPPFIG{type_I_III_correlation}).
+Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that activated IFN in response to non-viral ligands~\cite{dhir2018mitochondrial, benitez2015vivo, chiang2018viral} or because some cells that we classified as uninfected were actually infected at low levels.
+The difference in the frequency of IFN positivity among infected and uninfected cells in \FIG{transcriptomics}G was highly significant ($P < 10^{-5}$, Fisher's exact test).
 Many more cells expressed ISGs than IFN itself (\SUPPFIG{ISG}A).
 The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
-These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\cite{stetson2006type,honda2006type}.
+These results are consistent with the established knowledge that IFN is expressed primarily in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\cite{stetson2006type,honda2006type}.
 
@@
 This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
-As an example of the former case, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
+As an example of the former case, expressing unmutated copies of all eight genes did not guarantee high viral gene expression and successful innate-immune evasion: for instance, the unmutated virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
 But in other cases, the stochasticity allows a defective virus to still escape immune recognition.
@@
 
-\subsection{Viral defects associated with infection outcome in single cells.}
+\subsection{Viral defects associated with viral gene expression and IFN induction in single cells.}
 
@@
 
-Cells infected by incomplete or mutated virions also expressed IFN more frequently than cells infected by full wild-type virions (\FIG{mutations}B), although this difference was not statistically significant ($P = 0.12$, Fisher's exact test).
-However, some specific viral defects were significantly associated with IFN induction: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (\FIG{mutations}C).
+Some viral defects also contribute to IFN induction.
+Specifically, cells infected by incomplete or mutated virions expressed IFN more frequently than cells infected by virions that expressed unmutated copies of all genes (\FIG{mutations}B), although this difference was not statistically significant ($P = 0.12$, Fisher's exact test).
+However, the association was significant for certain classes of viral defects: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (\FIG{mutations}C).
 The only trend that remained significant at a false discovery rate (FDR) of 10\% was absence of NS.
@@
 The numbers indicate the median of four measurements for each viral mutant.
-The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
+The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for the wild type viral stock is indicated with a dashed blue line.
 Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
@@
 
-This viral genetic information is crucial for understanding infection outcome and innate-immune induction.
+This viral genetic information helps explain cell-to-cell variation in viral gene expression and innate-immune induction.
 Despite the fact that we used a low-passage viral stock generated from plasmids, most infected cells do not express unmutated copies of all viral genes.
@@
 Two types of defects---absence of the NS gene and amino-acid mutations to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate immunity~\cite{garcia1998influenza, hale2008multifunctional}.
+But although the general role of NS in innate-immune antagonism has long been appreciated, our work represents the first direct demonstration that stochastic absence and mutations to this gene are a major contributor to IFN induction in single cells.
 A third type of defect, amino-acid mutations in PB1, was more surprising since this protein has not been described as a major player in innate-immune detection.
@@
 
-Our results further show that viral genetic defects only partially explain the heterogeneity among influenza-infected cells.
+Our results further show that viral genetic defects do not fully explain the heterogeneity among influenza-infected cells.
 There is substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with unmutated virions.
 Additionally, no viral defect induces IFN deterministically: every type of immunostimulatory defect that we characterize is also observed in IFN- cells in our single-cell dataset.
-This fact shows that stochasticity or pre-existing cellular states play a major role in affecting the outcome of infection---a finding that concords with the fact that IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
+Therefore, stochasticity or pre-existing cellular states also play a major role in affecting innate-immune induction---a finding that concords with the fact that IFN induction is heterogeneous even among cells treated with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}, as well as for other viruses~\cite{oneal2018west}.
 
@@
 Natural human influenza infections are established by just a handful of virions~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza} that then undergo exponential growth, and early IFN responses are amplified by paracrine signaling~\cite{stetson2006type,honda2006type}.
-It is therefore plausible that early heterogeneity could affect downstream events.
-Extending our approaches to more complex systems could shed further light on how viral genetic variation interacts with cell-to-cell heterogeneity to shape the race between virus and immune system.
+It is therefore plausible that early heterogeneity in innate-immune induction could affect the entire course of infection.
+Extending our approaches to more complex systems could shed further light on how viral genetic variation and stochasticity interact to shape the race between virus and immune system.
 
@@
 \begin{suppfig}
-\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}
-\caption{
-The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
-Each point represents one cell.
-The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
-Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
-}
-\label{suppfig:type_I_III_correlation}
+\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_ifn_coinfect.pdf}
+\caption{
+There is no association between viral co-infection and expression of IFN in our experiments using low MOI infections with a relatively ``pure'' viral stock.
+Histograms show the fraction of all cellular mRNA derived from IFN among cells expressing viral mRNA from just a single viral barcode variant, or cells expressing viral mRNA from both the wildtype and synonymously barcoded viral variants.
+The dual-barcode cells represent known co-infections, whereas the single-barcode cells represent a mix of singly infected cells and co-infections with the same viral barcode.
+There is no significant difference in the frequency of IFN induction among the two classes of cells ($P = 0.53$, Fisher's exact test). 
+}
+\label{suppfig:coinf_IFN}
 \end{suppfig}
@@
 \label{suppfig:ISG}
+\end{suppfig}
+
+\begin{suppfig}
+\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}
+\caption{
+The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
+Each point represents one cell.
+The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
+Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
+}
+\label{suppfig:type_I_III_correlation}
 \end{suppfig}

### 2019-04-25 paper/paper.tex
responses to Reviewer 2

@@
 These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
-We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells.
+We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells (this moderately low MOI reasonably balances our desire to limit the number of co-infections with the cost of performing transcriptomics on uninfected cells).
 From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
@@
 By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
+From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
 Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
-This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
-The first explanation seems unlikely, as there is no tendency for co-infected cells to express more IFN (\SUPPFIG{coinf_IFN}); therefore we favor the latter two explanations.
+This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection complements otherwise transcriptionally defective virions to increase the likelihood that we identify a cell as infected.
+The first explanation seems unlikely, as there is no tendency for co-infected cells to express more IFN (\SUPPFIG{coinf_IFN}). 
+Therefore we favor the latter two explanations, both of which have been demonstrated for other viruses~\cite{aguilera2017plaques, combe2015single}.
 The moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.
@@
 One other interesting trend emerges from the single-cell data.
-There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS.
-But among cells that lack NS, cells with more viral mRNA are significantly more likely to be IFN+ (\FIG{mutations}D).
-This finding is elaborated on in the validation experiments below.
+There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS (\FIG{mutations}D).
+But among cells that lack NS, cells with more viral mRNA are significantly more likely to be IFN+ (\FIG{mutations}D); this finding is elaborated on in the validation experiments below.
+Overall, the lack of reduced viral gene expression in IFN+ cells suggests that autocrine IFN signaling typically occurs too late to suppress viral transcription, and the well-known inhibitory effect of IFN against influenza depends mainly on paracrine signaling.
 

### 2019-04-25 paper/paper.tex
remove "feed-forward"

@@
 This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell}.
-Because viral growth and the IFN response are both feed-forward processes that amplify themselves, early variation in the initiation of these events could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically founded by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
+Because viral growth and the IFN response both amplify themselves, early variation in the initiation of these events could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically founded by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.
 

### 2019-04-26 paper/paper.tex
added DataDryad links for supp files, remove citations in supp

@@
 (A)
-To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\cite{strahle2006sendai}.
+To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression.
 The name of each of reporter cell line is indicated at the top of each row of plots.
@@
 Flow cytometry data for \FIG{IFNrare}B.
-The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus~\cite{strahle2006sendai}, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
+The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
 After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
@@
 The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
-Each viral gene is detected in $\sim$80-90\% of the infected cells, roughly in line with prior estimates~\cite{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}.
+Each viral gene is detected in $\sim$80-90\% of the infected cells.
 The exception is NP, which is detected in virtually all infected cells.
@@
 Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
-This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\cite{stetson2006type,honda2006type}.
+This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN.
 (B) Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
@@
 Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.
-To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering~\cite{maaten2008visualizing} as implemented in \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} to generate a two-dimensional representation of the data.
+To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering as implemented in \texttt{Monocle} to generate a two-dimensional representation of the data.
 In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
@@
 The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see panel B).
-From the frequencies of these different forms, we estimate~\cite{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
+From the frequencies of these different forms, we estimate that 5.7\% of molecules are chimeric due to PCR strand exchange.
 About half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
@@
 \SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
-Note that since the rate of homologous recombination in influenza virus in negligible~\cite{boni2008homologous}, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
+Note that since the rate of homologous recombination in influenza virus in negligible, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
 }
@@
 The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses less HA due to the lack of secondary transcription.
-Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1~\cite{bloom2010permissive} for PB1del385to2163.
+Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1 for PB1del385to2163.
 \SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
@@
 The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins~\cite{killip2014activation}.
+These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins.
 In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
@@
 Sequences of the IFN reporters in \FIG{IFNrare}A are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
 }
@@
 Genbank files giving sequences of the wild-type and synonymously barcoded viruses are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
 }
@@
 \caption{
-A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}
+A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}. 
 }
@@
 A CSV file giving the genotypes in \FIG{genotypes} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
 }
@@
 A CSV file giving the viral mutations and related information in \FIG{mutations} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
 }
@@
 \caption{
-Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid~\cite{hoffmann2000dna} are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}.
+Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
 }
@@
 The Jupyter notebook that analyzes the PacBio data is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
 }
@@
 The Jupyter notebook that analyzes the annotated cell-gene matrix is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.
+This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
 }

### 2019-04-26 paper/paper_final_version_JVI_no-figs-supp.tex
created version with no figs / supp to convert to Word

--- 2019-04-26 paper/paper.tex
@@
-\documentclass[lineno]{asm-article}
+\documentclass[]{article}
 
-\usepackage{blindtext}
-
-\usepackage[version=4]{mhchem}
-\usepackage{siunitx}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% additional packages imported by Jesse Bloom
-\usepackage[export]{adjustbox}
-\usepackage{float}
-
-% Added by Jesse Bloom for figures
-\renewcommand{\floatpagefraction}{.9}
-\setcounter{topnumber}{8}
-\setcounter{bottomnumber}{8}
-\setcounter{totalnumber}{8}
-
-\newfloat{suppfile}{thp}{lofsupfile}
-\floatname{suppfile}{\bf FILE}
-\renewcommand{\thesuppfile}{S\arabic{suppfile}}
-
-\newfloat{suppfig}{thp}{lofsupfile}
-\floatname{suppfig}{\bf FIG}
-\renewcommand{\thesuppfig}{S\arabic{suppfig}}
-
-\renewcommand{\figureautorefname}{Fig.}
-\newcommand{\FIG}[1]{\autoref{fig:#1}}
-
-\newcommand{\suppfileautorefname}{File}
-\newcommand{\SUPPFILE}[1]{\autoref{suppfile:#1}}
-
-\newcommand{\suppfigautorefname}{Fig.}
-\newcommand{\SUPPFIG}[1]{\autoref{suppfig:#1}}
-
-% Allow big figures on pages
-\renewcommand{\topfraction}{0.9}
-\renewcommand{\bottomfraction}{0.9}
-\renewcommand{\textfraction}{0.1}
-\renewcommand{\floatpagefraction}{0.9}
-
-\title{Single-cell virus sequencing of influenza infections that trigger innate immunity}
-
-\author{%
-Alistair B. Russell,\afn{a}
-Elizaveta Elshina,\afn{b}
-Jacob R. Kowalsky,\afn{a}
-Aartjan J. W. te Velthuis,\afn{b}
-Jesse D. Bloom\afn{a,c,d,\authfn{1}}
-}
-
-\affil{%
-\afn{a}Basic Sciences and Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, United States
-
-\afn{b} Division of Virology, Department of Pathology, University of Cambridge, Cambridge, United Kingdom
-
-\afn{c} Department of Genome Sciences, University of Washington, Seattle, United States
-
-\afn{d} Howard Hughes Medical Institute, Seattle, United States
-}
-
-\runningauthor{Russell et al.}
-\runningtitle{Single-cell virus sequencing of influenza infections that trigger innate immunity}
-
-\corraddress{Jesse D. Bloom, jbloom@fredhutch.org}
-
-\papertype{Research Article}
+\input{asm-biblatex.tex}
+\bibliography{references}
 
@@
 
-\maketitle
+Single-cell virus sequencing of influenza infections that trigger innate immunity
 
-\begin{abstract}
+Alistair B. Russell,$^{a}$
+Elizaveta Elshina,$^{b}$
+Jacob R. Kowalsky,$^{a}$
+Aartjan J. W. te Velthuis,$^{b}$
+Jesse D. Bloom$^{a,c,d,1}$
+
+$^{a}$ Basic Sciences and Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, United States
+
+$^{b}$ Division of Virology, Department of Pathology, University of Cambridge, Cambridge, United Kingdom
+
+$^{c}$ Department of Genome Sciences, University of Washington, Seattle, United States
+
+$^{d}$ Howard Hughes Medical Institute, Seattle, United States
+
+$^1$ Jesse D. Bloom, jbloom@fredhutch.org
+
+Running title: Single-cell virus sequencing of influenza infections that trigger innate immunity
+
+Keywords: influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; defective virus; heterogeneity
+
+\clearpage
+
+\section*{Abstract}
 Influenza-infected cells vary widely in their expression of viral genes, and only occasionally activate innate immunity.
@@
 
-\begin{importance}
+\section*{Importance}
 Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
@@
 Overall, our study provides the first complete picture of how viral mutations affect the course of infection in single cells.
-\end{importance}
 
-\end{abstract}
-
-\keywords{influenza virus; interferon; PacBio; single-cell RNAseq; 10X Chromium; NS1; PB1; defective virus; heterogeneity}
-
-\section{Introduction}
+\section*{Introduction}
 Infection with an acute virus such as influenza initiates a race between the virus and immune system.
@@
 
-\section{Results}
+\section*{Results}
 
-\subsection{A system to identify and enrich rare IFN+ cells.}
-
-%%% start IFNrare figure %%%
-\begin{figure}
-\centerline{
-{\bf \Large A}\includegraphics[width=0.4\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
-\hspace{0.03\textwidth}
-{\bf \Large B} \includegraphics[width=0.26\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
-}
-\caption{
-Reporter cells to identify and enrich infections that activate IFN expression.
-(A) The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
-We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
-In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
-(B)
-Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
-The plot also shows uninfected cells, and cells infected with Sendai virus.
-The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
-}
-\label{fig:IFNrare}
-\end{figure}
-%%% end IFNrare figure %%%
+\subsection*{A system to identify and enrich rare IFN+ cells.}
 
 Influenza virus only rarely triggers IFN expression in infected cells~\cite{killip2017single, russell2018extreme, kallfass2013visualizing}---a fact that poses a challenge for the study IFN induction in single cells.
-Therefore, we developed a method to identify and enrich rare IFN+ cells by creating A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
+Therefore, we developed a method to identify and enrich rare IFN+ cells by creating A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (Fig.
+1A).
 Cells that activate the IFN reporters can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
-The reporters were efficiently activated by infection with a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
+The reporters were efficiently activated by infection with a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (Fig.
+S1A), and activation of the type I and type III IFN reporters was highly correlated in our cells (Fig.
+S1B; further validated by the single-cell transcriptomics below).
 Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.
@@
 As described in the next subsection, our stock actually consisted of a mix of two viruses: wild-type WSN and a variant of this virus that carries synonymous viral ``barcodes'' near the termini of each gene.
-This viral stock activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
-We also validated that MACS for the cell-surface protein driven by the IFN reporter enriched the IFN+ cells by $>$50-fold (\SUPPFIG{MACS}).
+This viral stock activated the IFN reporter in $\sim$0.5\% of infected cells  (Fig. 1B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
+We also validated that MACS for the cell-surface protein driven by the IFN reporter enriched the IFN+ cells by $>$50-fold (Fig. S3).
 
-\subsection{Combined transcriptomics and virus-sequencing of single infected cells.}
+\subsection*{Combined transcriptomics and virus-sequencing of single infected cells.}
 
-We developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
-First, we generated the viral stock described in the previous subsection, which consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
+We developed the approach in Fig. 2 to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
+First, we generated the viral stock described in the previous subsection, which consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (File S2).
 These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
-We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells (this moderately low MOI reasonably balances our desire to limit the number of co-infections with the cost of performing transcriptomics on uninfected cells).
+We used this viral stock to infect A549 IFN reporter cells (Fig.
+2A) at a dose that led to detectable viral transcription in $\sim$25\% of cells (this moderately low MOI reasonably balances our desire to limit the number of co-infections with the cost of performing transcriptomics on uninfected cells).
 From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
@@
 
-%%% start workflow figure
-\begin{figure}
-\begin{fullwidth}
-
-\includegraphics[width=\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
-
-\caption{
-Approach for combined transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
-(A)
-IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
-IFN+ cells are enriched by MACS, and pooled with non-enriched cells and uninfected canine cells that serve as a control for multiplets and mRNA leakage.
-(B)
-The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
-(C)
-Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
-(D)
-viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are sequenced by PacBio).
-(E)
-The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
-}
-\label{fig:workflow}
-
-\end{fullwidth}
-\end{figure}
-%%% end workflow figure
-
-We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
+We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (Fig.
+2B).
 Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to viral as well as cellular mRNAs.
@@
 
-We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
-But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
+We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (Fig. 2C).
+But we also took a portion and enriched for full-length viral molecules by PCR (Fig. 2D).
 We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs)~\cite{travers2010flexible}.
@@
 Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
-Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).
+Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (Fig. 2E).
 
-\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells.}
+\subsection*{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells.}
 
-%%% start transcriptomics figure
-\begin{figure}
-\begin{fullwidth}
-
-\includegraphics[width=\linewidth, clip=false]{figures/single_cell_figures/p_cell_summary.pdf}
-
-\caption{
-Single-cell transcriptomics of IFN-enriched influenza-infected cells.
-(A) 
-Number of cells for which transcriptomes were obtained.
-From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are derived from multiple cells. 
-(B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
-Green lines show the distribution of cellular mRNAs per cell.
-Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
-(C) Distribution across cells of the fraction of all mRNA derived from influenza.
-Cells called as infected are in blue, while other cells are in green.
-The inset shows the amount of viral mRNA in the human cells that are called as infected.
-(D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
-\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
-(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell derived from each gene.
-(F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
-From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
-(G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
-Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line as IFN+.
-A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
-}
-\label{fig:transcriptomics}
-
-\end{fullwidth}
-\end{figure}
-%%% end transcriptomics figure
-
-We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
+We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (Fig. 3A).
 We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
 To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\cite{haque2017practical}.
-After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B)
+After this filtering, we retained 1,490 human cells for further analysis (Fig. 3B)
 
-To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
+To identify infected cells, we examined the fraction of each transcriptome derived from virus (Fig. 3C).
 As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA.
-We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
+We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (Fig.
+3C).
 We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
-The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
+The distribution of the amount of viral mRNA across infected cells is shown in the inset in Fig. 3C.
 As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.
@@
 We called presence / absence of genes rather than transcripts, since the two genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
-\FIG{transcriptomics}D (top panel) shows that 162 of 290 infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
+Fig. 3D (top panel) shows that 162 of 290 infected cells express all eight genes (see Fig. S4 for frequencies for individual genes).
 This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work using the WSN strain~\cite{russell2018extreme}, and slightly to substantially higher than that reported in studies by others using different viral strains or methodologies~\cite{heldt2015single, brooke2013most, dou2017analysis, jacobs2019incomplete}.
 
-The amount of viral mRNA was lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
-However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
-Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.
+The amount of viral mRNA was lower in cells that failed to express viral genes (Fig. 3D, bottom panel).
+However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (Fig.
+3D, bottom panel).
+Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (Fig.
+3E) and generally matched values from older bulk studies~\cite{hatada1989control}.
 
-By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
-From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
-Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
+By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (Fig.
+3F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
+From Fig. 3F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
+Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (Fig. 3C) if infection is Poisson.
 This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection complements otherwise transcriptionally defective virions to increase the likelihood that we identify a cell as infected.
-The first explanation seems unlikely, as there is no tendency for co-infected cells to express more IFN (\SUPPFIG{coinf_IFN}). 
+The first explanation seems unlikely, as there is no tendency for co-infected cells to express more IFN (Fig. S5). 
 Therefore we favor the latter two explanations, both of which have been demonstrated for other viruses~\cite{aguilera2017plaques, combe2015single}.
@@
 
-We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
+We next examined expression of IFN and ISGs (Fig. 3G and Fig. S6).
 Over 20\% of infected cells were IFN+, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
-The expression of type I and type III IFN was highly correlated in single cells, justifying our decision to collapse both classes under the single label of ``IFN'' in the analyses that follow (\SUPPFIG{type_I_III_correlation}).
+The expression of type I and type III IFN was highly correlated in single cells, justifying our decision to collapse both classes under the single label of ``IFN'' in the analyses that follow (Fig.
+S7).
 Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that activated IFN in response to non-viral ligands~\cite{dhir2018mitochondrial, benitez2015vivo, chiang2018viral} or because some cells that we classified as uninfected were actually infected at low levels.
-The difference in the frequency of IFN positivity among infected and uninfected cells in \FIG{transcriptomics}G was highly significant ($P < 10^{-5}$, Fisher's exact test).
-Many more cells expressed ISGs than IFN itself (\SUPPFIG{ISG}A).
-The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
+The difference in the frequency of IFN positivity among infected and uninfected cells in Fig. 3G was highly significant ($P < 10^{-5}$, Fisher's exact test).
+Many more cells expressed ISGs than IFN itself (Fig. S6A).
+The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (Fig. S6B).
 These results are consistent with the established knowledge that IFN is expressed primarily in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\cite{stetson2006type,honda2006type}.
@@
 Finally, we qualitatively examined how expression of viral genes, IFN, and ISGs relate to the overall structure of the high-dimensional transcriptomic data.
-\SUPPFIG{tSNE} shows unsupervised t-SNE clustering~\cite{maaten2008visualizing} of the cells.
+Fig. S8 shows unsupervised t-SNE clustering~\cite{maaten2008visualizing} of the cells.
 Cells expressing high levels of viral genes, IFN, and ISGs cluster together---and most of the structure in the t-SNE plot that is not associated with these genes involves uninfected and IFN- cells.
 
-\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells.}
+\subsection*{Full genotypes of viruses infecting single IFN+ and IFN- cells.}
 
-%%% start genotypes figure %%%
-\begin{figure}
-\begin{fullwidth}
-{\centering
-\includegraphics[height=0.85\textheight]{figures/single_cell_figures/p_genotypes.pdf}
-}
-\caption{
-Viral genotypes and infection outcomes in single cells.
-Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
-The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
-Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
-Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
-The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
-For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
-For instance, \textit{cell 5} was co-infected by a virion with one unmutated and one internally deleted copy of PB1.
-}
-\label{fig:genotypes}
-
-\end{fullwidth}
-\end{figure}
-%%% end genotypes figure %%%
-
-We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes in single infected cells.
-We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
-The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
+We next used PacBio sequencing (Fig. 2D) to determine the full sequences of the viral genes in single infected cells.
+We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (Fig. S9).
+The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (Fig.
+S10), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.
 
@@
 
-We could call the sequences of all expressed viral genes in the majority of infected cells (\SUPPFIG{ncells}).
-We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (\SUPPFIG{ncells}).
+We could call the sequences of all expressed viral genes in the majority of infected cells (Fig. S11).
+We were most effective at calling full viral genotypes in cells that expressed high amounts of viral mRNA and were infected by only one viral barcode variant (Fig. S11).
 But we also called full genotypes for many cells that had low viral burden or were co-infected by both viral barcode variants.
 
-The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
+The 150 cells for which we called the full viral genotypes are shown in Fig. 4.
 Visual inspection of this figure reveals a wealth of information.
-For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
+For instance, the cell with the highest viral burden (\textit{cell 1} in Fig.
+4, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
 But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
@@
 
-However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
+However, Fig. 4 also reveals stochasticity that is independent of viral genotype.
 This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
@@
 
-\subsection{Viral defects associated with viral gene expression and IFN induction in single cells.}
+\subsection*{Viral defects associated with viral gene expression and IFN induction in single cells.}
 
-%%% begin mutations figure %%%
-\begin{figure}
-\begin{fullwidth}
-{\centering
-\includegraphics[width=\linewidth]{figures/single_cell_figures/p_mutations.pdf}
-}
-\caption{
-Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.
-(A) 
-Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
-Cells infected by fully unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
-(B)
-IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
-(C)
-Specific viral defects associated with IFN induction.
-The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
-The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
-Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
-(D)
-There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
-Throughout this figure, we only consider substitutions that are non-synonymous.
-}
-\label{fig:mutations}
-
-\end{fullwidth}
-\end{figure}
-%%% end mutations figure %%%
-
-To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
+To systematically assess viral features associated with infection outcome, we divided the 150 cells in Fig.
+4 into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
+Fig.
+5A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.
@@
 Some viral defects also contribute to IFN induction.
-Specifically, cells infected by incomplete or mutated virions expressed IFN more frequently than cells infected by virions that expressed unmutated copies of all genes (\FIG{mutations}B), although this difference was not statistically significant ($P = 0.12$, Fisher's exact test).
-However, the association was significant for certain classes of viral defects: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (\FIG{mutations}C).
+Specifically, cells infected by incomplete or mutated virions expressed IFN more frequently than cells infected by virions that expressed unmutated copies of all genes (Fig.
+5B), although this difference was not statistically significant ($P = 0.12$, Fisher's exact test).
+However, the association was significant for certain classes of viral defects: absence of NS and amino-acid mutations in PB1 were significantly enriched in IFN+ cells, and amino-acid mutations in NS and deletions in HA were weakly enriched (Fig.
+5C).
 The only trend that remained significant at a false discovery rate (FDR) of 10\% was absence of NS.
@@
 One other interesting trend emerges from the single-cell data.
-There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS (\FIG{mutations}D).
-But among cells that lack NS, cells with more viral mRNA are significantly more likely to be IFN+ (\FIG{mutations}D); this finding is elaborated on in the validation experiments below.
+There is no difference in the amount of viral mRNA between IFN+ and IFN- cells that express NS (Fig. 5D).
+But among cells that lack NS, cells with more viral mRNA are significantly more likely to be IFN+ (Fig. 5D); this finding is elaborated on in the validation experiments below.
 Overall, the lack of reduced viral gene expression in IFN+ cells suggests that autocrine IFN signaling typically occurs too late to suppress viral transcription, and the well-known inhibitory effect of IFN against influenza depends mainly on paracrine signaling.
 
-\subsection{Validation that viral defects in single IFN+ cells often increase IFN induction.}
-
-%%% begin validation figure %%%
-\begin{figure}
-
-\centerline{\includegraphics[width=0.9\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
-\caption{
-Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.
-(A)
-Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
-The numbers indicate the median of four measurements for each viral mutant.
-The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for the wild type viral stock is indicated with a dashed blue line.
-Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
-(B)
-Similar to the first panel, but validates increased IFN induction for a large internal deletion in the PB1 gene, and normalizes infecting virion dose rather than calling IFN+ percentage only among infected cells.
-See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for details.
-The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
-}
-\label{fig:validation}
-
-\end{figure}
-%%% end validation figure %%%
+\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction.}
 
@@
 
-The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
+The viral defect most strongly associated with IFN induction was failure to express the NS gene (Fig. 4, Fig. 5C).
 Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\cite{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
@@
 
-The single-cell data also showed that amino-acid substitutions in proteins encoded by the PB1 and NS genes were enriched in IFN+ cells (\FIG{genotypes}, \FIG{mutations}C), so we created mutant viruses with some of these substitutions:
+The single-cell data also showed that amino-acid substitutions in proteins encoded by the PB1 and NS genes were enriched in IFN+ cells (Fig.
+4, Fig.
+5C), so we created mutant viruses with some of these substitutions:
 PB1-D27N, PB1-G206S, PB1-K279R, PB1-T677A, NS1-A122V, and NS2-E47G.
@@
 Finally, prior work has suggested that virions with internal deletions in the polymerase genes can induce higher levels of IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
-Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (\FIG{mutations}C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in \FIG{genotypes}).
+Although such deletions are not significantly enriched among IFN+ cells in our single-cell data (Fig.
+5C), there is a co-infected IFN+ cell where one viral variant has a deletion in PB1 spanning nucleotides 385 to 2163 (\textit{cell 5} in Fig.
+4).
 We therefore created a virus carrying this deletion, and propagated it in cells constitutively expressing PB1 protein.
@@
 We tested the rate of IFN induction by each viral stock using the reporter cells.
-\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than a wild-type viral stock.
+Fig. 6 shows that five of the eight mutant viral stocks induced IFN more frequently than a wild-type viral stock.
 The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced IFN significantly more frequently than wild type.
 The other three point mutants (PB1-G206S, PB1-K279R, and NS2-E47G) did not increase IFN induction---an unsurprising finding, since we expect some mutations without an IFN-enhancing effect to be found in IFN+ cells by chance.
-Overall, the results in \FIG{validation} validate that the viral defects in single IFN+ cells often cause increased IFN production.
+Overall, the results in Fig. 6 validate that the viral defects in single IFN+ cells often cause increased IFN production.
 
 However, IFN induction remains stochastic even for the most potently IFN-inducing viral mutants.
-\FIG{validation} shows flow cytometry data (see also \SUPPFIG{validationflow}), which is itself a single-cell measurement, albeit one that does not report the viral genotype.
+Fig. 6 shows flow cytometry data (see also Fig. S12), which is itself a single-cell measurement, albeit one that does not report the viral genotype.
 As can be seen from these data, none of the mutant viral stocks induce IFN in more than 20\% of infected cells. 
 Of course, these mutant virus stocks are themselves genetically heterogeneous, as many virions will have additional defects similar to that revealed by our single-cell sequencing of the ``wild-type'' viral stock.
-But our single-cell data show that IFN induction is stochastic even for infections that share the same defect, such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in \FIG{genotypes}).
+But our single-cell data show that IFN induction is stochastic even for infections that share the same defect, such as absence of NS (e.g., compare \textit{cell 62} and \textit{cell 69} in Fig. 4).
 Therefore, the experiments in this section not only validate some specific viral defects that increase IFN induction, but also show that induction remains stochastic even with these defects.
 
-\subsection{The IFN-inducing viral defects act by diverse mechanisms.}
+\subsection*{The IFN-inducing viral defects act by diverse mechanisms.}
 Some of the viral defects in IFN+ cells are easy to reconcile with existing knowledge: for instance, NS1 is the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}, and internal deletions are prevalent in immunostimulatory viral stocks~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
@@
 
-%%%%% begin IFN_vs_HA figure %%%%%%%%
-\begin{figure}
-\centerline{\includegraphics[width=0.35\textwidth, valign=t]{figures/MutantComparison/p_foldchange.pdf}}
-\caption{
-Infected cells that express higher levels of HA protein are much more likely to induce IFN expression only if they are infected by virus with defects in NS1.
-The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile.
-Points indicate replicates, and lines indicate the mean.
-This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in \SUPPFIG{validationflow}; see \SUPPFIG{IFN_vs_HA} for more details.
-}
-\label{fig:IFN_vs_HA}
-\end{figure}
-%%%%% end IFN_vs_HA figure %%%%%%%%
-
-We first focused on one of the strongest trends from the single cell data: increased viral gene expression is associated with increased IFN induction when the infecting virion fails to express NS, but not otherwise (\FIG{mutations}D).
+We first focused on one of the strongest trends from the single cell data: increased viral gene expression is associated with increased IFN induction when the infecting virion fails to express NS, but not otherwise (Fig.
+5D).
 To confirm this observation, we performed a flow cytometry analysis of the reporter cells infected by different immunostimulatory viral mutants to examine the association between expression of a viral gene product (HA protein) and IFN induction.
-Consistent with the single-cell data, cells that expressed more HA were much more likely to turn IFN+ when infected with the NS1stop or NS1-A122V mutants, but not when infected with any of the other viral variants (\FIG{IFN_vs_HA}).
+Consistent with the single-cell data, cells that expressed more HA were much more likely to turn IFN+ when infected with the NS1stop or NS1-A122V mutants, but not when infected with any of the other viral variants (Fig.
+7).
 This fact suggests that when there are high levels of viral transcription, NS1 becomes more important as a buffer against detection of viral products.
@@
 We hypothesized that the immunostimulatory mutations to PB1 might cause the viral polymerase to produce aberrant products, in line with recent work showing that mutations to PB2 can lead to the generation of aberrant RNAs that trigger RIG-I~\cite{velthuis2018mini, du2018genome}.
-To investigate if the PB1 mutations might perturb polymerase activity, we examined their location in a structural model of the polymerase complex (\FIG{PB1}A).
+To investigate if the PB1 mutations might perturb polymerase activity, we examined their location in a structural model of the polymerase complex (Fig. 8A).
 The IFN-enhancing PB1 mutation T677A occurs at the tip of a helix that interacts with the 3' terminus of the RNA template as it enters the channel above the active site, whereas the IFN-enhancing D27N mutation is deeper in the polymerase close to the binding pocket of the 5' terminus of the template.
@@
 
-%%% begin PB1 figure %%%
-\begin{figure}
-
-\centerline{\includegraphics[width=\textwidth]{figures/PB1_figure/PB1_figure.pdf}}
-\caption{
-IFN-inducing mutations D27N and T677A in the PB1 protein affect polymerase activity. 
-(A) Model of bat influenza A virus polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the influenza B virus polymerase (PDB 5MSG)~\cite{reich2017vitro}.
-The locations of PB1 D27 and T677 (both red) relative to the 5' (blue) and 3' (orange) termini of the RNA template and the PB1 active site (grey; PB1 act) are indicated.
-The PA endonuclease (green; PA endo) and PB2 cap binding domain (pink; PB2 cap) are also indicated. 
-Part of the fingers subdomain of PB1 is hidden to reveal the template in the entry channel. 
-(B) IFN-beta promoter activity measured using a dual luciferase reporter assay in 293T cells transfected with plasmids expressing the indicated PB1 protein, the other polymerase complex proteins (PB2, PA, and NP), and a full-length NA vRNA template.
-PB1a is a catalytically inactive PB1 active site control.
-In this panel and the next two panels, points show three biological replicates; ``n.d.'' indicates not detectable, and orange indicates a variant was significantly different than wild type by a two-sided t-test.
-(C) Polymerase activity on full-length vRNA template in 293T cells transfected as in panel (B).
-Steady-state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging. 
-PB1a was used as negative control and background correction. 
-The 5S rRNA signal was used as loading control. 
-Other panels show Western blot analysis of PB1, NP and GAPDH protein expression, and the graph at the bottom shows quantification by phosphoroimaging. 
-(D) Polymerase activity on a short 246-nucleotide vRNA template. 
-The top panel shows the steady state levels of vRNA template as determined by primer extension and denaturing PAGE. 
-The other two panels show the PB1 and tubulin expression levels analyzed by Western blot, and the graph shows quantification.
-}
-\label{fig:PB1}
-
-\end{figure}
-%%% end PB1 figure %%%
-
 To test if the PB1 mutations affect activity, we transfected 293T cells with plasmids that express wild-type or mutant PB1 protein along with the other proteins in the polymerase complex (PB2, PA, and NP) and full-length viral RNA (vRNA) for the NA segment.
-Both polymerase mutations increased IFN expression in this assay (\FIG{PB1}B), indicating that they have an immunostimulatory effect in the context of an active viral polymerase even when other viral components are absent.
+Both polymerase mutations increased IFN expression in this assay (Fig.
+8B), indicating that they have an immunostimulatory effect in the context of an active viral polymerase even when other viral components are absent.
 We next directly measured polymerase activity on the full-length vRNA template by extracting total RNA and quantifying replication (vRNA) and transcription (mRNA) products by primer extension.
-Both immunostimulatory PB1 mutations had activities that were significantly different from wild type, despite being expressed at wild-type protein levels (\FIG{PB1}B).
-Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both---although D27N still retained activity far in excess of a control active-site mutant (\FIG{PB1}C).
+Both immunostimulatory PB1 mutations had activities that were significantly different from wild type, despite being expressed at wild-type protein levels (Fig. 8B).
+Specifically, T677A had higher levels of both activities, whereas D27N had reduced levels of both---although D27N still retained activity far in excess of a control active-site mutant (Fig. 8C).
 We speculated that the mutations might alter polymerase processivity, leading to accumulation of aberrant RNA products that activate the innate immune system~\cite{velthuis2018mini, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned, liu2019inhibition}.
-We therefore repeated the activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (\FIG{PB1}D).
+We therefore repeated the activity assays using a short 246-nucleotide template~\cite{velthuis2018mini} in place of the full-length NA vRNA (Fig. 8D).
 On this shorter template, the activity of the D27N mutant was now similar to wild type, while the activity of the T677A mutant remained higher than wildtype (although not significantly so in three biological repeats).
@@
 
-\section{Discussion}
+\section*{Discussion}
 We have determined the full sequences of all viral genes in single influenza-infected cells.
@@
 Although our study is certainly not the first to note that influenza has a high mutation rate~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\cite{russell2018extreme, heldt2015single, brooke2013most, dou2017analysis, jacobs2019incomplete}, it is the first to directly observe the full spectrum of these defects across single cells.
-Visual inspection of \FIG{genotypes} shows how any experiment that does not sequence viral genes in single cells is averaging across a diverse spectrum of viral defects.
+Visual inspection of Fig. 4 shows how any experiment that does not sequence viral genes in single cells is averaging across a diverse spectrum of viral defects.
 
@@
 
-\section{Materials and Methods}
+\section*{Materials and Methods}
 \small
-\subsection{IFN reporter cell lines.}
-We created IFN reporter variants of the A549 human lung epithelial cell line (\FIG{IFNrare}A).
+\subsection*{IFN reporter cell lines.}
+We created IFN reporter variants of the A549 human lung epithelial cell line (Fig. 1A).
 The parental A549 cell line used to create these reporters was obtained from ATCC (CCL-185), and was tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core and authenticated using the ATCC STR profiling service.
@@
 Downstream of this NotI site, each of the following reporter constructs was cloned: mCherry, mNeonGreen, and low-affinity nerve growth factor lacking the C-terminal signaling domain (LNGFR$\Delta$C)~\cite{bonini1997hsv,ruggieri1997cell} linked to mNeonGreen by a P2A linker~\cite{kim2011high}.
-The sequence of the last of these constructs is provided in \SUPPFILE{reporter_sequences}.
+The sequence of the last of these constructs is provided in File S1.
 
@@
 Downstream of this promoter we cloned LNGFR$\Delta$C linked to ZsGreen via a P2A linker.
-The sequence of this construct is provided in \SUPPFILE{reporter_sequences}.
+The sequence of this construct is provided in File S1.
 
@@
 Importantly, the cells that we retained for further use were not the same portion that were tested by poly(I:C) treatment, but rather a separate split of the same population---this avoids any selection on the cells from transient activation of IFN.
-For the dual type I / type III reporter used in \SUPPFIG{reportervalidation}B, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
+For the dual type I / type III reporter used in Fig.
+S1B, a single-cell clone of the type III reporter cell line was transduced with the type I reporter bearing the mCherry fluorescent marker, and then isolated and propagated as a single cell clone for the other cell lines.
 All reporter lines tested negative for mycoplasma contamination by the Fred Hutch Genomics Core.
 
-\SUPPFIG{reportervalidation}A shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
+Fig. S1A shows validation of the reporter cell lines using infection with saturating amounts of the Cantell strain of Sendai virus (obtained from Charles River Laboratories).
 For detection of the cell-surface bound LNGFR$\Delta$C, cells were stained with PE-conjugated anti-LNGFR (CD271) antibody from Miltenyi Biotec.
 
-\subsection{Viruses for single-cell experiments.}
+\subsection*{Viruses for single-cell experiments.}
 We performed the single-cell experiments using the A/WSN/1933 (H1N1) strain of influenza virus.
@@
 We have used a similar synonymous viral barcoding strategy in our prior single-cell work~\cite{russell2018extreme} as it allows us to detect about half of co-infected cells based on the expression of both viral barcode variants.
-In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (\SUPPFIG{StrandExchange}).
-The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in \SUPPFILE{virus_seqs}.
+In the current work, we extended this approach by placing synonymous barcodes near \emph{both} termini of the gene segments in order to quantify strand exchange during PacBio sequencing (Fig. S10).
+The sequences of all gene segments from the wild-type and synonymously barcoded viral strains are in File S2.
 These genes were cloned into the pHW2000~\cite{hoffmann2000dna} reverse-genetics plasmid.
@@
 Specifically, after generation by reverse genetics, the wild-type variant was expanded at an MOI of 0.001 for 72 hours twice in MDCK-SIAT1 cells, and the synonymously barcoded variant was expanded once at an MOI of 0.01 for 60 hours.
-The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\cite{reed1938simple} as implemented at \url{https://github.com/jbloomlab/reedmuenchcalculator}.
+The MOIs for this passaging are based on titers determined using TCID50 assays via the formula of Reed and Muench~\cite{reed1938simple} as implemented at https://github.com/jbloomlab/reedmuenchcalculator.
 After being passaged independently, the two viral stocks were combined at equivalent numbers of infectious units to make the stock used in the single-cell experiments.
 
-\subsection{Flow cytometry analyses for HA expression.}
+\subsection*{Flow cytometry analyses for HA expression.}
 For the single-cell experiments (which only examine the transcriptional results of a single cycle of infection), we were most interested in the titer of viral particles that are transcriptionally active for a single round of infection of A549 cells.
@@
 
-\subsection{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium.}
+\subsection*{Single-cell transcriptomics of IFN-enriched infected cells using 10X Chromium.}
 The single-cell transcriptomics and virus sequencing was performed using the A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-P2A-mNeonGreen reporter.
-A schematic of the experiment is shown in \FIG{workflow}.
+A schematic of the experiment is shown in Fig. 2.
 
 The wild-type and synonymously barcoded viruses were mixed with the goal of adding equal numbers of transcriptionally active HA-expressing virions of each virus strain.
-The cells were then infected with this mixture at a dose designed to infect about half the cells (\FIG{transcriptomics}C suggests that the actual rate of detectable infection was slightly lower).
+The cells were then infected with this mixture at a dose designed to infect about half the cells (Fig. 3C suggests that the actual rate of detectable infection was slightly lower).
 Infections were allowed to proceed for 12 hours.
@@
 To enrich IFN+ cells, the cells were then incubated with anti-LNGFR MACSelect Microbeads (Miltenyi Biotec) and twice passed over an MS magnetic column (Miltenyi Biotec), retaining the bound (and presumably IFN-enriched) population each time.
-This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in \SUPPFIG{MACS}.
+This MACS sorting is expected to give approximately the enrichment for IFN+ cells shown in Fig. S3.
 The original, unsorted, population was then added back in to $\sim$10\% of the final cell fraction in order to ensure the presence of interferon negative cells. 
-At this point, uninfected canine (MDCK-SIAT1) cells were also added to $\sim$5\% of the final cell fraction to enable quantification of the cell multiplet rate (\FIG{transcriptomics}A) and background viral mRNA in uninfected cells (\FIG{transcriptomics}C).
+At this point, uninfected canine (MDCK-SIAT1) cells were also added to $\sim$5\% of the final cell fraction to enable quantification of the cell multiplet rate (Fig.
+3A) and background viral mRNA in uninfected cells (Fig.
+3C).
 We began this entire process of cell collection and enrichment at 12 hours post-infection, but the process (which was performed at room temperature) took about an hour, and thus we consider the cells to have been analyzed at 13 hours post-infection.
@@
 
-\subsection{Enrichment and preparation of viral cDNA for PacBio sequencing.}
+\subsection*{Enrichment and preparation of viral cDNA for PacBio sequencing.}
 We amplified virus-derived molecules from cDNA retained from the 10X Genomics protocol for PacBio sequencing of the full-length cDNA.
-These cDNA have at their 3' end the cell barcode and UMI plus the adaptor sequence that is added during the 10X protocol (see \FIG{workflow} for simple schematic, and \SUPPFILE{pacbio_analysis} for more details).
+These cDNA have at their 3' end the cell barcode and UMI plus the adaptor sequence that is added during the 10X protocol (see Fig. 2 for simple schematic, and File S7 for more details). 
 We only wanted to PacBio sequence cDNA molecules derived from virus.
@@
 
-We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (\SUPPFILE{primer_sequences}).
-A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
+We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (File S3).
+A major concern during these PCRs is strand exchange (see Fig. S10) which would scramble the cell barcodes and mutations on viral cDNAs.
 To reduce strand-exchange and obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon), which physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents~\cite{boers2015micelle}.
@@
 The material from these eight segment-specific PCRs was then pooled with the goal of obtaining a equimolar ratio of segments, and sequenced on one SMRT Cell in a PacBio RS II and one SMRT Cell of a PacBio Sequel.
-Detailed results from the analysis of these first two sequencing runs is shown in \SUPPFILE{pacbio_analysis}.
+Detailed results from the analysis of these first two sequencing runs is shown in File S7.
 These results showed that although the PCRs substantially enriched for influenza molecules, the relative coverage of the different viral genes was still uneven, with the longer genes under-sampled.
@@
 The former was sequenced on one cell of a single SMRT Cell of a PacBio Sequel, and the latter on two additional SMRT Cells of a PacBio Sequel. 
-As is shown \SUPPFILE{pacbio_analysis}, the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes~\cite{xue2016propagation} which are preferentially amplified during PCR.
+As is shown File S7, the coverage remained relatively low for the polymerase genes---and most of the reads we did obtain were dominated by shorter internally deleted variants of the polymerase genes~\cite{xue2016propagation} which are preferentially amplified during PCR.
 
@@
 We pooled the products of these reactions from this size-selection and sequenced on a SMRT Cell of a PacBio Sequel.
-As is shown in \SUPPFILE{pacbio_analysis}, this sequencing yielded more full-length polymerase variants, but they were still undersampled compared to other viral genes.
+As is shown in File S7, this sequencing yielded more full-length polymerase variants, but they were still undersampled compared to other viral genes.
 
@@
 In the downstream computational analysis, we can then determine the full sequence of the gene as well as the cell barcode of the initial molecule from which the linearized molecule is derived.
-Specifically, we first used 2.5 ng of our already-amplified segment-specific material in a 10-cycle PCR to append circularization adapters (see \SUPPFILE{primer_sequences} for sequences), and cleaned the resultant mixture using SPRIselect beads at a volume ratio of 0.4.
+Specifically, we first used 2.5 ng of our already-amplified segment-specific material in a 10-cycle PCR to append circularization adapters (see File S3 for sequences), and cleaned the resultant mixture using SPRIselect beads at a volume ratio of 0.4.
 We then used 10 ng of this amplified material in a 20$\mu$l NEBuilder reaction using an extended reaction time of 50 minutes in order to circularize the molecules.
@@
 The circularized and digested material was then cleaned using SPRIselect beads at a volume ratio of 0.4.
-This material was then used as template for three non-emulsion PCRs specific to PB2, PB1, or PA, using two segment-specific primers that align to the central portion of each gene but in apposition to each another (see \SUPPFILE{primer_sequences} for sequences).
+This material was then used as template for three non-emulsion PCRs specific to PB2, PB1, or PA, using two segment-specific primers that align to the central portion of each gene but in apposition to each another (see File S3 for sequences).
 These linearization reactions used 20 (PB2) or 26 (PB1 and PA) PCR cycles, and the resulting products were cleaned using SPRIselect beads at a volume ratio of 1.0.
 This material was pooled to produce an equimolar mixture of full-length PB1, PA, and PB2 and sequenced in an additional SMRT Cell of PacBio Sequel. 
-As is shown in \SUPPFILE{pacbio_analysis}, this process yielded many full-length polymerase variants.
+As is shown in File S7, this process yielded many full-length polymerase variants.
 
 The computational analyses of the full-length viral gene sequences described below combined the data from all these reactions.
-The number of sequences obtained for each gene after pooling the data from all reactions is shown in \SUPPFIG{CCSs}, which also indicates that the net rate of strand exchange is very low (see \SUPPFIG{StrandExchange} for an illustration of how this is determined).
-A detailed breakdown of the coverage of each gene and PacBio run is in \SUPPFILE{pacbio_analysis}.
+The number of sequences obtained for each gene after pooling the data from all reactions is shown in Fig.
+S9, which also indicates that the net rate of strand exchange is very low (see Fig.
+S10 for an illustration of how this is determined).
+A detailed breakdown of the coverage of each gene and PacBio run is in File S7.
 Importantly, the PCR biases mean that the coverage of molecules by the PacBio sequencing is not proportional to their abundance in the starting mRNA.
@@
 
-\subsection{qPCR for viral genes and IFN.}
+\subsection*{qPCR for viral genes and IFN.}
 We performed qPCR on reverse-transcribed mRNA for influenza HA (to quantify viral transcription), IFNB1 (to quantify IFN induction), and L32 (a cellular housekeeping gene for normalization).
@@
 
-For the qPCR in \SUPPFIG{delqPCR}, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
+For the qPCR in Fig.
+S13, A549 cells were seeded at a density of $10^4$ cells/well in a 96-well plate in D10 media 24 hours prior to infection, with four independent wells seeded per experimental treatment.
 Immediately prior to infection D10 media was removed and replaced with influenza growth media and infected with the indicated influenza strains at a MOI of 0.4 based on TCID50 in MDCK-SIAT1 cells.
@@
 
-\subsection{Viruses and experiments for validation experiments.}
-In \FIG{validation}, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
+\subsection*{Viruses and experiments for validation experiments.}
+In Fig. 6, we tested the IFN inducing capacity of a variety of viral mutants identified in the single-cell experiments.
 For point-mutant viruses, we created variants for all amino-acid substitutions found in PB1 and NS among IFN+ cells that did not also lack NS.
@@
 All of these mutants were cloned into the pHW2000 bi-directional reverse-genetics plasmid~\cite{hoffmann2000dna} in order to enable generation of viruses encoding the mutant genes.
- \SUPPFILE{validation_sequences} provides the full sequences for all of these plasmids.
+ File S6 provides the full sequences for all of these plasmids.
 
-We generated the wild-type and point-mutant viruses for the validation experiments in \FIG{validation}A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\cite{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in \SUPPFILE{validation_sequences} for the wild-type plasmid for that gene.
+We generated the wild-type and point-mutant viruses for the validation experiments in Fig.
+6A by reverse genetics using the pHW18* series of WSN reverse genetics plasmids~\cite{hoffmann2000dna}, but substituting the appropriate mutant plasmid listed in File S6 for the wild-type plasmid for that gene.
 To generate the viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into co-cultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1.
 At 24 hours post-transfection, we changed media from D10 to influenza growth media. 
-At 50 hours post-transfection (for the replicate 1 viruses in \SUPPFIG{validationflow}A) or 72 hours (for the replicate 2 viruses in \SUPPFIG{validationflow}), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
+At 50 hours post-transfection (for the replicate 1 viruses in Fig.
+S12A) or 72 hours (for the replicate 2 viruses in Fig.
+S12), we harvested the virus-containing supernatant, clarified this supernatant by centrifugation at 300$\times$g for 4 min, and stored aliquots of the clarified viral supernatant at -80$^{\circ}$C.
 We then thawed aliquots and titered by TCID50 on MDCK-SIAT1 cells.
-For the infections in \SUPPFIG{validationflow}A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\cite{hirst1942quantitative}.
+For the infections in Fig.
+S12A, we wanted to use equivalent particle counts, so we normalized all viruses to an equivalent hemagglutination titer on turkey red blood cells~\cite{hirst1942quantitative}.
  Briefly, a solution of 10\% v/v red blood cells (LAMPIRE Biological Laboratories, Fisher Scientific catalogue number 50412942)  was washed in PBS and diluted to a final concentration of 0.5\% v/v. 
 Two-fold serial dilutions of virus were added to an equal volume of diluted red blood cells, and titer was measured as the highest dilution of viral stock at which complete hemagglutination of red blood cells was observed.
-We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in \SUPPFIG{validationflow}A.
+We then performed infections of the A549 reporter cell line at equivalent hemagglutination titer and analyzed the data as described in Fig. S12A.
 
-To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in \SUPPFIG{validationflow}B, we used slightly different procedures.
+To generate the NS1stop mutant virus and the wild-type and PB1del385to2163 mutant viruses in Fig. S12B, we used slightly different procedures.
 The wild-type virus was generated by reverse genetics as described for the point-mutant viruses above, harvested at 48 hours post-transfection, and then passaged on MDCK-SIAT1 cells for 36 hours at an MOI of 0.05---conditions that we previously validated to lead to relatively little accumulation of defective particles~\cite{russell2018extreme}.
@@
 The wild-type and NS1stop viruses were titered by TCID50 on MDCK-SIAT1 cells, and the PB1 deletion viruses were titered on the MDCK-SIAT1 cells constitutively expressing PB1.
-The infections in \SUPPFIG{validationflow}B were performed at equivalent TCID50s as described in the legend to that figure.
-That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in \SUPPFIG{delqPCR}.
+The infections in Fig. S12B were performed at equivalent TCID50s as described in the legend to that figure.
+That these equivalent TCID50s were also roughly equivalent in terms of particles capable of undergoing primary transcription is shown in Fig. S13.
 
-\subsection{Computational analysis of single-cell transcriptomic and viral sequence data.}
-A computational pipeline that performs all steps in the data analysis is available at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell}. 
+\subsection*{Computational analysis of single-cell transcriptomic and viral sequence data.}
+A computational pipeline that performs all steps in the data analysis is available at https://github.com/jbloomlab/IFNsorted\_flu\_single\_cell. 
 This pipeline is orchestrated by \texttt{Snakemake}~\cite{koster2012snakemake}, and begins with the raw sequencing data and ends by generating the figures shown in this paper.
-The sequencing data and annotated cell-gene matrix are available on the GEO repository under accession GSE120839 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120839}).
+The sequencing data and annotated cell-gene matrix are available on the GEO repository under accession GSE120839 (https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120839).
 
@@
 The human transcriptome was generated by filtering genome assembly GRCh38 for protein-coding genes defined in GTF file GRCh38.87.
-The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in \SUPPFILE{virus_seqs} (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
+The influenza virus transcriptome consisted of the mRNAs for the wild-type A/WSN/1933 virus strain in File S2 (the \texttt{cellranger} alignment is sufficiently permissive that it aligns sequences from both the wild-type and synonymously barcoded viral variants to this transcriptome).
 The canine transcriptome was generated by filtering genome assembly CanFam3.1 for protein-coding genes defined in GTF file CanFam3.1.87.
-The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (\FIG{transcriptomics}A), and generate a matrix giving the expression of each gene in each single cell.
+The \texttt{cellranger} software was used to align the Illumina 3'-end sequencing reads to this multi-species transcriptome, call human+influenza and canine cells (Fig.
+3A), and generate a matrix giving the expression of each gene in each single cell.
 We used a custom Python script to determine the number of influenza virus reads that could be assigned to the wild-type or synonymously barcoded virus, and added this information to the annotated the cell-gene matrix.
@@
 The PacBio sequences of the full-length viral genes were analyzed as follows.
-First, we used version 3.1.0 of PacBio's \texttt{ccs} program (\url{https://github.com/PacificBiosciences/unanimity}) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
+First, we used version 3.1.0 of PacBio's \texttt{ccs} program (https://github.com/PacificBiosciences/unanimity) to build circular consensus sequences (CCSs) from the subreads files, requiring at least 3 passes and a minimum accuracy of 0.999.
 We further processed these CCSs using custom Python code and the \texttt{minimap2}~\cite{li2018minimap2} long-read aligner (version 2.11-r797).
-The Python code has been implemented in the API of \texttt{dms\_tools2} (\url{https://jbloomlab.github.io/dms_tools2/}~\cite{bloom2015software}) package (version 2.3.0).
-A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{pacbio_analysis}.
+The Python code has been implemented in the API of \texttt{dms\_tools2} (https://jbloomlab.github.io/dms\_tools2/) software package (version 2.3.0)~\cite{bloom2015software}.
+A Jupyter notebook that performs these analyses is at https://github.com/jbloomlab/IFNsorted\_flu\_single\_cell/blob/master/pacbio\_analysis.ipynb, and is also provided as File S7.
 We refer the reader to this notebook for a detailed description and extensive plots showing the results at each step.
@@
 We aligned the cDNAs flanked by these termini to the influenza transcriptome, and performed a variety of quality control steps.
-At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (\SUPPFIG{StrandExchange}), and reassuringly found that strand exchange was rare (\SUPPFIG{CCSs}).
+At this point, we examined whether cDNAs had the synonymous viral barcodes at both ends or neither end as expected in the absence of strand exchange (Fig.
+S10), and reassuringly found that strand exchange was rare (Fig.
+S9).
 The small number of CCSs with identifiable strand exchange were filtered from further analysis.
@@
 \end{enumerate}
-The plots in \SUPPFILE{pacbio_analysis} indicate that these are reasonable mutation-calling criteria.
-We could call the sequences of all expressed viral genes in about half of the infected cells (\SUPPFIG{ncells}).
-The mutations called using this pipeline are shown in \FIG{genotypes}, and \SUPPFILE{genotypes} gives the number of CCSs supporting each mutation call.
+The plots in File S7 indicate that these are reasonable mutation-calling criteria.
+We could call the sequences of all expressed viral genes in about half of the infected cells (Fig. S11).
+The mutations called using this pipeline are shown in Fig. 4, and File S4 gives the number of CCSs supporting each mutation call.
 The called sequences of the viral genes were added to the annotated cell-gene matrix.
@@
 This analysis utilized a variety of R and Bioconductor~\cite{huber2015orchestrating} packages, including \texttt{Monocle}~\cite{qiu2017reversed, trapnell2014dynamics} and \texttt{ggplot2}.
-A Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}, and is also provided in HTML form as \SUPPFILE{monocle_analysis}.
+A Jupyter notebook that performs these analyses is at https://github.com/jbloomlab/IFNsorted\_flu\_single\_cell/blob/master/monocle\_analysis.ipynb, and is also provided as File S8.
 We refer the reader to this notebook for a detailed description and a variety of additional plots not included in the paper.
-Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in \FIG{transcriptomics}B.
-We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (\FIG{transcriptomics}C).
-We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (\FIG{transcriptomics}D and \SUPPFIG{frac_has_gene}).
-To identify cells that were co-infected with both viral barcodes (\FIG{transcriptomics}F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
-We called IFN+ and ISG+ cells using the heuristic thresholds shown in \FIG{transcriptomics}G and \SUPPFIG{ISG}, respectively.
+Briefly, we first filtered cells that were extreme outliers in the amount of mRNA as shown in Fig. 3B.
+We used the uninfected canine cells to estimate the percentage of total mRNA in a cell that would come from influenza purely due to background (e.g., from cell lysis) in the absence of infection, and called as infected the human cells for which significantly more than this amount of mRNA was derived from influenza under a Poisson model (Fig.
+3C).
+We next used a Poisson model parameterized by the amount of expected background mRNA for each influenza gene to call the presence or absence of each influenza gene in each infected cell (Fig.
+3D and Fig.
+S4).
+To identify cells that were co-infected with both viral barcodes (Fig.
+3F), we used a binomial test to identify cells for which we could reject the null hypothesis that at least 95\% of viral mRNA was derived from the more common viral barcode.
+We called IFN+ and ISG+ cells using the heuristic thresholds shown in Fig. 3G and Fig. S6, respectively.
 We counted IFN mRNAs as any IFN-$\alpha$, IFN-$\beta$, or IFN-$\lambda$ transcripts.
 We counted ISG mRNAs as any of CCL5, IFIT1, ISG15, or Mx1.
-The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).
+The plot in Fig. 4 summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (https://github.com/wilkox/gggenes).
 
-\subsection{Structural analysis of PB1 mutants.}
-To locate the PB1 mutations in the influenza A virus RNA polymerase structure relative to the template and active site in \FIG{PB1}A, we superposed the bat influenza A virus RNA polymerase structure (PDB 4WSB)~\cite{reich2014structural}, which shows the 3' terminus of the template on the surface of the RNA polymerase, with the influenza B virus transcription initiation complex (PDB 5MSG)~\cite{reich2017vitro}, which shows the 3' terminus of the template in the template entry channel that leads towards the active site.
+\subsection*{Structural analysis of PB1 mutants.}
+To locate the PB1 mutations in the influenza A virus RNA polymerase structure relative to the template and active site in Fig.
+8A, we superposed the bat influenza A virus RNA polymerase structure (PDB 4WSB)~\cite{reich2014structural}, which shows the 3' terminus of the template on the surface of the RNA polymerase, with the influenza B virus transcription initiation complex (PDB 5MSG)~\cite{reich2017vitro}, which shows the 3' terminus of the template in the template entry channel that leads towards the active site.
 The structural alignment was performed in Pymol 1.8.7 using motifs A and C.
 
-\subsection{Experimental analysis of PB1 mutants.}
-For the experimental analysis of the PB1 mutants in \FIG{PB1}, we used plasmids pcDNA-PB1, pcDNA-PA, pcDNA-PB2, and pcDNA-NP, which encode the WSN proteins that compose the polymerase complex~\cite{fodor2002single}; pPolI-NA, which encodes the viral RNA for the WSN NA~\cite{fodor2002single}; and pcDNA-PB1a, which encodes an inactive version of the WSN PB1 polymerase protein~\cite{vreede2004model}.
+\subsection*{Experimental analysis of PB1 mutants.}
+For the experimental analysis of the PB1 mutants in Fig.
+8, we used plasmids pcDNA-PB1, pcDNA-PA, pcDNA-PB2, and pcDNA-NP, which encode the WSN proteins that compose the polymerase complex~\cite{fodor2002single}; pPolI-NA, which encodes the viral RNA for the WSN NA~\cite{fodor2002single}; and pcDNA-PB1a, which encodes an inactive version of the WSN PB1 polymerase protein~\cite{vreede2004model}.
 To construct plasmids expressing mutant PB1 proteins D27N and T677A, the plasmid pcDNA-PB1 was subjected to site-directed mutagenesis. 
@@
 
-\section{Acknowledgments}
+\section*{Acknowledgments}
 We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, Katherine Xue, and Hannah Itell for helpful suggestions.
@@
 
-\bibliography{references}
-
-\clearpage
-\section{Supplemental Material}
-
-\begin{suppfig}
-\begin{minipage}[t]{0.52\textwidth}
-{\bf \Large A}
-
-\includegraphics[width=0.9\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/Sendai_validation.pdf}
-\end{minipage}
-\begin{minipage}[t]{0.48\textwidth}
-{\bf \Large B}
-
-\includegraphics[width=0.7\textwidth, valign=t]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}
-
-\end{minipage}
-
-\caption{
-Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.
-(A)
-To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression.
-The name of each of reporter cell line is indicated at the top of each row of plots.
-At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
-Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
-(B)
-The type I and type III IFN reporters are highly correlated in their activation.
-An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
-The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
-After 13 hours, cells were analyzed by flow cytometry.
-Cells positive for either fluorescent reporter were further analyzed.
-As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.
-}
-\label{suppfig:reportervalidation}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.9\textwidth]{figures/IFN_stochastic/Flow/flow_plot.pdf}
-\caption{
-Flow cytometry data for \FIG{IFNrare}B.
-The A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were not infected, infected with saturating amounts of the Cantell strain of Sendai virus, or infected the same stock of influenza virus used in the single-cell experiment at a target MOI of 0.3.
-After 13 hours, the cells were stained for expression of HA protein and analyzed by FACS for HA and expression of the ZsGreen driven by the \textit{IFNL1} reporter.
-Each condition was done in triplicate.
-The contour plots show the density of all cells, and all IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of the uninfected cells in these populations.
-For the influenza-infected cells, the percentage IFN+ was calculated only among the HA+ cells (since these are the ones that are infected).
-For the uninfected and Sendai-virus infected, the percentage IFN+ was calculated among all cells, since these cells do not express HA.
-}
-\label{suppfig:IFNrare_flow}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}
-\caption{
-Example MACS enrichments of IFN+ influenza-infected cells.
-A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
-After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
-(A)
-After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
-The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
-As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
-We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
-(B)
-Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
-The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
-The qPCR was performed in quadruplicate (hence the four points for each sample).
-}
-\label{suppfig:MACS}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_has_gene.pdf}
-\caption{
-The fraction of infected cells that are called as expressing each viral gene.
-The gray dashed line is at one (the fraction that would be observed if all viral genes are expressed in all infected cells).
-Each viral gene is detected in $\sim$80-90\% of the infected cells.
-The exception is NP, which is detected in virtually all infected cells.
-The much higher frequency of detecting NP could reflect a biological phenomenon, but we suspect it is more likely that cells lacking NP tend to have much lower viral gene expression overall and so are not reliably called as being infected in our experiments because the number of viral mRNAs is below the detection limit.}
-\label{suppfig:frac_has_gene}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.5\textwidth]{figures/single_cell_figures/p_frac_ifn_coinfect.pdf}
-\caption{
-There is no association between viral co-infection and expression of IFN in our experiments using low MOI infections with a relatively ``pure'' viral stock.
-Histograms show the fraction of all cellular mRNA derived from IFN among cells expressing viral mRNA from just a single viral barcode variant, or cells expressing viral mRNA from both the wildtype and synonymously barcoded viral variants.
-The dual-barcode cells represent known co-infections, whereas the single-barcode cells represent a mix of singly infected cells and co-infections with the same viral barcode.
-There is no significant difference in the frequency of IFN induction among the two classes of cells ($P = 0.53$, Fisher's exact test). 
-}
-\label{suppfig:coinf_IFN}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}
-\caption{
-Expression of ISGs in single infected and uninfected cells.
-For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
-(A) The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
-We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
-Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
-This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN.
-(B) Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
-Each point represents one cell, and the Pearson correlation coefficient is shown.
-IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
-Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
-}
-\label{suppfig:ISG}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.7\textwidth]{figures/single_cell_figures/p_ifn_genes_corr.pdf}
-\caption{
-The correlation between the fraction of cellular mRNA derived from type I and type III IFN in the A549 cells in our single-cell transcriptomics.
-Each point represents one cell.
-The plots are faceted by whether the cells are called as infected, and the Pearson correlation coefficient is shown.
-Because type I and type III IFN expression are highly correlated, for the remainder of the paper we group them together and refer to their combined expression as the level of IFN.
-}
-\label{suppfig:type_I_III_correlation}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=\textwidth]{figures/single_cell_figures/p_tsne.pdf}
-\caption{
-Unsupervised t-SNE clustering shows that cell-to-cell variation in expression of influenza, IFN, and ISG transcripts substantially contributes to the structure of the data.
-To generate an unbiased representation of the factors that distinguished the transcriptomes of the cells in our experiments, we used unsupervised t-SNE clustering as implemented in \texttt{Monocle} to generate a two-dimensional representation of the data.
-In the t-SNE plot, each point is a different cell, and cells with similar transcriptomes are closer together.
-Each panel shows the same t-SNE plot, but the cells are colored differently in each panel based on the amount of viral, IFN, or ISG mRNA, shown on a log (top) or linear (bottom) scale.
-As is clear from this plot, expression of influenza, IFN, and ISG genes contributes substantially to the structure of the data, since cells with high expression of these genes clearly group together.
-}
-\label{suppfig:tSNE}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}
-\caption{
-The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
-These sequences were obtained using several PacBio runs, most of which were intentionally loaded with different amounts of the various viral genes to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells (see \SUPPFILE{pacbio_analysis}).
-Because of this unequal loading and the inherently different PCR amplification efficiencies of different viral genes, unlike the transcriptomic data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
-Especially for the polymerase genes (PB2, PB1, and PA), many CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were preferentially amplified during PCR.
-Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
-Note that the disproportionate sequencing of the shorter internally deleted genes does not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse sequences derived from the same cDNA, and cell barcodes were used to collapse sequences from the same cell.
-The bars in the plot are colored by whether the sequence is derived from the wild-type viral variant, the synonymously barcoded viral variant, or represents a mixed-barcode molecule (see panel B).
-From the frequencies of these different forms, we estimate that 5.7\% of molecules are chimeric due to PCR strand exchange.
-About half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
-For some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
-A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcodes.
-}
-\label{suppfig:CCSs}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.8\textwidth]{figures/StrandExchangeSchematic/StrandExchangeSchematic.pdf}
-\caption{
-Strategy for detecting strand exchange during sequencing of full-length viral genes.
-The library preparation for PacBio sequencing of the cDNA for the full-length viral genes required many cycles of PCR.
-A major concern is that strand exchange during this PCR could scramble mutations and 10X cell barcodes / UMIs from different molecules.
-We detect PCR strand exchange by leveraging the fact that our cells were infected with a mix of wild-type virus and virus carrying synonymous barcodes near both termini of each gene.
-If there is no strand exchange, all molecules should either be wild-type or have the synonymous barcoding mutations at \emph{both} termini.
-Strand exchange will create some molecules that have wild-type nucleotides at one termini and synonymous barcoding mutations at the other termini.
-\SUPPFIG{CCSs} shows the frequencies with which these different types of molecules were observed during the PacBio sequencing.
-Note that since the rate of homologous recombination in influenza virus in negligible, such mixed-barcode molecules are \emph{not} expected to be generated naturally during co-infection.
-}
-\label{suppfig:StrandExchange}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}
-\caption{
-Number of cells for which we could determine the full sequences of all genes expressed by the infecting virion(s).
-(A) We could call the complete genotypes of the infecting virion(s) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
-(B) The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNA than cells for which we could not call complete genotypes.
-Both facts make sense.
-Cells with more viral mRNA are more likely to have their viral cDNA captured in the PacBio sequencing, which is only captures a small fraction of the total transcripts identified by the 3'-end sequencing transcriptomic sequencing.
-The lower calling rate for dual-barcode co-infections is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing.
-An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
-Note also that this plot is limited to the cells that were called as infected (\FIG{transcriptomics}C) and could clearly be classified as IFN- or IFN+ (\FIG{transcriptomics}G).
-}
-\label{suppfig:ncells}
-\end{suppfig}
-
-\begin{suppfig}
-{\bf \Large A}\includegraphics[width=0.48\textwidth, valign=t]{figures/Validation_Figure/SNP_flow_plot.pdf}
-\hspace{0.02\textwidth}
-{\bf \Large B} \includegraphics[width=0.39\textwidth, valign=t]{figures/Validation_Figure/del_flow_plot.pdf}
-\caption{
-Flow cytometry data for \FIG{validation}.
-(A)
-Data for \FIG{validation}A.
-A549 cells with the \textit{IFNL1} reporter driving LNGFR$\Delta$C-ZsGreen were infected with stocks of the indicated mutant.
-After 13 hours, cells were stained for HA protein and analyzed by FACS.
-Contour plots show density of all cells, and IFN+ cells are also indicated by orange dots.
-Cells were classified as HA+ or IFN+ based on gates set to put 0.05\% of uninfected cells in these populations.
-For infected cells, the percentage IFN+ was calculated among the HA+ cells (since these are the ones that are infected).
-For uninfected cells, the percentage IFN+ was calculated among all cells, since uninfected cells do not express HA.
-For each viral mutant, two independent stocks were assayed in duplicate (i.e., \#1a and \#1b are one viral stock, and \#2a and \#2b are the other).
-The infections with replicate \#1 of the wild-type virus were performed at an MOI of 0.1 as determined by TCID50, and all other viruses were infected at an equivalent particle number as determined by HI assay.
-(B)
-Data for \FIG{validation}B.
-The virus with the deletion in PB1 cannot be normalized by HA expression since it expresses less HA due to the lack of secondary transcription.
-Therefore, all cells were infected at an equivalent MOI of 0.3 as determined by TCID50 on MDCK-SIAT1 cells for wild type and NS1stop, and on MDCK-SIAT1 cells expressing PB1 for PB1del385to2163.
-\SUPPFIG{delqPCR} shows that at these equivalent TCID50s, all variants had similar amounts of transcriptionally active virus in the absence of secondary transcription.
-The percent IFN+ was calculated for \emph{all} cells (HA+ and HA-) since that is a more fair comparison for PB1del385to2163.
-}
-\label{suppfig:validationflow}
-\end{suppfig}
-
-\begin{suppfig}
-\includegraphics[width=0.6\textwidth]{figures/Validation_Figure/qpcr_plot.pdf}
-\caption{
-Validation that the infections in \FIG{validation}B and \SUPPFIG{validationflow}B were performed at similar doses of virions capable of initiating primary transcription.
-In this experiment, A549 cells were infected at MOI of 0.4 (based on TCID50 as described in \SUPPFIG{validationflow}B), and then after 8 hours mRNA was harvested for qPCR on oligo-dT primed reverse transcription products.
-The y-axis shows the ratio of viral HA mRNA to the housekeeping gene L32.
-These infections were performed in the presence of absence of 50 $\mu$g/ml cycloheximide, which blocks protein synthesis and hence secondary transcription by newly synthesized viral proteins.
-In the absence of cycloheximide, the viruses with deletions in PB1 produced less viral mRNA presumably because they could not produce PB1 protein for secondary transcription.
-But in the presence of cycloheximide, all viruses produced similar amounts of viral mRNA, indicating that the dose of particles active for primary transcription is roughly equivalent across variants.
-Each measurement was performed in quadruplicate.
-}
-\label{suppfig:delqPCR}
-\end{suppfig}
-
-\begin{suppfig}
-\centerline{
-\includegraphics[width=0.6\textwidth, valign=t]{figures/MutantComparison/p_ifn_vs_ha.pdf}
-}
-\caption{
-A more detailed version of the data summarized in \FIG{IFN_vs_HA}.
-For each virus and replicate, we binned the infected cells in \SUPPFIG{validationflow} into HA expression quartiles based on the flow cytometry measurements.
-We then calculated the percent of cells that were IFN+ in each quartile.
-The plots show the mean HA expression of the quartile versus the precent of cells that are IFN+.
-The results clearly show that for the NS1stop and to a lesser extent the NS1-A122V variants, more viral protein (higher HA signal) correlates with IFN induction.
-\FIG{IFN_vs_HA} summarizes these same data by simply showing the ratio of percent IFN+ between the highest and lowest quartile.
-}
-\label{suppfig:IFN_vs_HA}
-\end{suppfig}
-
 \clearpage
 
-\begin{suppfile}
+%%% start IFNrare figure %%%
+\begin{figure}
 \caption{
-Sequences of the IFN reporters in \FIG{IFNrare}A are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/IFN_reporter/plasmids}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
+Reporter cells to identify and enrich infections that activate IFN expression.
+(A) The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
+We created reporters with type I and type III IFN promoters (File S1).
+In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (Fig. S1A).
+(B)
+Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see Fig.
+S2 for full flow cytometry data).
+The plot also shows uninfected cells, and cells infected with Sendai virus.
+The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
 }
-\label{suppfile:reporter_sequences}
-\end{suppfile}
+\end{figure}
+%%% end IFNrare figure %%%
 
-\begin{suppfile}
+%%% start workflow figure
+\begin{figure}
 \caption{
-Genbank files giving sequences of the wild-type and synonymously barcoded viruses are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
+Approach for combined transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
+(A)
+IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
+IFN+ cells are enriched by MACS, and pooled with non-enriched cells and uninfected canine cells that serve as a control for multiplets and mRNA leakage.
+(B)
+The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
+(C)
+Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
+(D)
+viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are sequenced by PacBio).
+(E)
+The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
 }
-\label{suppfile:virus_seqs}
-\end{suppfile}
+\end{figure}
+%%% end workflow figure
 
-\begin{suppfile}
+%%% start transcriptomics figure
+\begin{figure}
 \caption{
-A text file giving the primers used to amplify the influenza cDNAs for PacBio sequencing is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/WorkflowSchematic/PacBio_primer_list.txt}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}. 
+Single-cell transcriptomics of IFN-enriched influenza-infected cells.
+(A) 
+Number of cells for which transcriptomes were obtained.
+From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are derived from multiple cells. 
+(B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
+Green lines show the distribution of cellular mRNAs per cell.
+Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
+(C) Distribution across cells of the fraction of all mRNA derived from influenza.
+Cells called as infected are in blue, while other cells are in green.
+The inset shows the amount of viral mRNA in the human cells that are called as infected.
+(D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
+Fig. S4 shows the frequency that each viral gene is detected.
+(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell derived from each gene.
+(F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
+From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
+(G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
+Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line as IFN+.
+A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
 }
-\label{suppfile:primer_sequences}
-\end{suppfile}
+\end{figure}
+%%% end transcriptomics figure
 
-\begin{suppfile}
+%%% start genotypes figure %%%
+\begin{figure}
 \caption{
-A CSV file giving the genotypes in \FIG{genotypes} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/genotypes.csv}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
+Viral genotypes and infection outcomes in single cells.
+Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
+The second box is framed in orange for cells classified as IFN+ in Fig. 3G.
+Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
+Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
+The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
+For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
+For instance, \textit{cell 5} was co-infected by a virion with one unmutated and one internally deleted copy of PB1.
 }
-\label{suppfile:genotypes}
-\end{suppfile}
+\end{figure}
+%%% end genotypes figure %%%
 
-\begin{suppfile}
+%%% begin mutations figure %%%
+\begin{figure}
 \caption{
-A CSV file giving the viral mutations and related information in \FIG{mutations} is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/paper/figures/single_cell_figures/mutations.csv}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
+Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.
+(A) 
+Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
+Cells infected by fully unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
+(B)
+IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
+(C)
+Specific viral defects associated with IFN induction.
+The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
+The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
+Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
+(D)
+There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
+Throughout this figure, we only consider substitutions that are non-synonymous.
 }
-\label{suppfile:mutations}
-\end{suppfile}
+\end{figure}
+%%% end mutations figure %%%
 
-\begin{suppfile}
+%%% begin validation figure %%%
+\begin{figure}
 \caption{
-Genbank plasmid maps for the mutant genes cloned into the pHW* bi-directional reverse genetics plasmid are at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/FluVariantPlasmidMaps}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
+Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.
+(A)
+Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
+The numbers indicate the median of four measurements for each viral mutant.
+The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for the wild type viral stock is indicated with a dashed blue line.
+Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
+(B)
+Similar to the first panel, but validates increased IFN induction for a large internal deletion in the PB1 gene, and normalizes infecting virion dose rather than calling IFN+ percentage only among infected cells.
+See Fig. S12 and Fig. S13 for details.
+The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
 }
-\label{suppfile:validation_sequences}
-\end{suppfile}
+\end{figure}
+%%% end validation figure %%%
 
-\begin{suppfile}
-\caption{\label{suppfile:pacbio_analysis}
-The Jupyter notebook that analyzes the PacBio data is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/pacbio_analysis.ipynb}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
+%%%%% begin IFN_vs_HA figure %%%%%%%%
+\begin{figure}
+\caption{
+Infected cells that express higher levels of HA protein are much more likely to induce IFN expression only if they are infected by virus with defects in NS1.
+The y-axis shows the ratio of the percent of IFN+ cells in the highest HA-expression quartile relative to the lowest HA-expression quartile.
+Points indicate replicates, and lines indicate the mean.
+This figure is based on joint analysis of the IFN reporter and HA staining for all infected cells in the flow cytometry data in Fig. S12; see Fig. S14 for more details.
 }
-\end{suppfile}
+\end{figure}
+%%%%% end IFN_vs_HA figure %%%%%%%%
 
-\begin{suppfile}
-\caption{\label{suppfile:monocle_analysis}
-The Jupyter notebook that analyzes the annotated cell-gene matrix is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/monocle_analysis.ipynb}.
-This file is also available on Data Dryad Digital Repository at \url{https://doi.org/10.5061/dryad.nh053c6}.
+%%% begin PB1 figure %%%
+\begin{figure}
+\caption{
+IFN-inducing mutations D27N and T677A in the PB1 protein affect polymerase activity. 
+(A) Model of bat influenza A virus polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the influenza B virus polymerase (PDB 5MSG)~\cite{reich2017vitro}.
+The locations of PB1 D27 and T677 (both red) relative to the 5' (blue) and 3' (orange) termini of the RNA template and the PB1 active site (grey; PB1 act) are indicated.
+The PA endonuclease (green; PA endo) and PB2 cap binding domain (pink; PB2 cap) are also indicated. 
+Part of the fingers subdomain of PB1 is hidden to reveal the template in the entry channel. 
+(B) IFN-beta promoter activity measured using a dual luciferase reporter assay in 293T cells transfected with plasmids expressing the indicated PB1 protein, the other polymerase complex proteins (PB2, PA, and NP), and a full-length NA vRNA template.
+PB1a is a catalytically inactive PB1 active site control.
+In this panel and the next two panels, points show three biological replicates; ``n.d.'' indicates not detectable, and orange indicates a variant was significantly different than wild type by a two-sided t-test.
+(C) Polymerase activity on full-length vRNA template in 293T cells transfected as in panel (B).
+Steady-state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging. 
+PB1a was used as negative control and background correction. 
+The 5S rRNA signal was used as loading control. 
+Other panels show Western blot analysis of PB1, NP and GAPDH protein expression, and the graph at the bottom shows quantification by phosphoroimaging. 
+(D) Polymerase activity on a short 246-nucleotide vRNA template. 
+The top panel shows the steady state levels of vRNA template as determined by primer extension and denaturing PAGE. 
+The other two panels show the PB1 and tubulin expression levels analyzed by Western blot, and the graph shows quantification.
 }
-\end{suppfile}
+\end{figure}
+%%% end PB1 figure %%%
 

### 2019-04-27 paper/paper.tex
added Alistair edits

@@
 But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
-As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
+As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription using incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
 The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.

### 2019-04-27 paper/paper_final_version_JVI.tex
added early access PDF
added Alistair edits

--- 2019-04-26 paper/paper.tex
@@
-\documentclass[lineno]{asm-article}
+\documentclass[]{asm-article}
 
@@
 
-\clearpage
+\newpage
+\setcounter{page}{1}
 \section{Supplemental Material}

### 2019-04-29 paper/paper.tex
cite File S4, S5 in text as requested by JVI

@@
 
-We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes in single infected cells.
+We next used PacBio sequencing (\FIG{workflow}D, \SUPPFILE{primer_sequences}) to determine the full sequences of the viral genes in single infected cells.
 We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
@@
 
-The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
+The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes} (see also \SUPPFILE{genotypes}).
 Visual inspection of this figure reveals a wealth of information.
@@
 To systematically assess viral features associated with infection outcome, we divided the 150 cells in \FIG{genotypes} into those that expressed unmutated copies of all eight genes (disregarding synonymous mutations) and those that did not.
-\FIG{mutations}A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
+\FIG{mutations}A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement} (see also \SUPPFILE{mutations}).
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.
@@
 The plot in \FIG{genotypes} summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (\url{https://github.com/wilkox/gggenes}).
+The raw data are in \SUPPFILE{genotypes} and \SUPPFILE{mutations}.
 

### 2019-04-29 paper/paper_final_version_JVI_no-figs-supp.tex
version for JVI final submission
cite File S4, S5 in text as requested by JVI

@@
 \input{asm-biblatex.tex}
+
 \bibliography{references}
@@
 
-We next used PacBio sequencing (Fig. 2D) to determine the full sequences of the viral genes in single infected cells.
+We next used PacBio sequencing (Fig. 2D, File S3) to determine the full sequences of the viral genes in single infected cells.
 We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (Fig. S9).
@@
 
-The 150 cells for which we called the full viral genotypes are shown in Fig. 4.
+The 150 cells for which we called the full viral genotypes are shown in Fig. 4 (see also File S4).
 Visual inspection of this figure reveals a wealth of information.
@@
 Fig.
-5A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement}.
+5A shows that the 49 cells infected by full unmutated virions had a significantly tighter distribution of the amount of viral mRNA per cell than the other 101 cells as quantified by the Gini index~\cite{gini1921measurement} (see also File S5).
 Therefore, viral defects are a major contributor to the heterogeneity in viral transcriptional burden.
@@
 The plot in Fig. 4 summarizes all of the genotypic information, and was created in substantial part using \texttt{gggenes} (https://github.com/wilkox/gggenes).
+The raw data are in File S4 and File S5.
 
@@
 The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
+
+\printbibliography
 
